---
title: "Getting started with admixr2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with admixr2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 5
)
source(Filter(file.exists, c("_common.R", "../_common.R", "vignettes/_common.R"))[1])
```

## What is aggregate-data modelling?

`admixr2` fits pharmacometric PK/PD models to **summary-level data** instead of
individual records — a **meta-analysis** framework for population PK/PD. The
input is **digitised aggregate data** (means, error bars and covariances),
**previously published models**, or both, and the result is one population model
with interpretable fixed, random and covariate effects.

For each study you supply:

- **E** — observed mean vector (one entry per observation time)
- **V** — observed covariance matrix (or variance vector)
- **n** — sample size
- **times** — observation time points
- **ev** — dosing event table

The estimators match E and V against their model-predicted counterparts and
return a standard nlmixr2 fit object. Turning a published figure into `E`, `V`
and `n` is covered in `vignette("aggregate-data")`; letting a study carry its own
published model instead, in `vignette("datagen")`.

Four estimators are available:

| Estimator | `est =` | Control function | Approach |
|-----------|---------|-----------------|---------|
| First-Order | `"adfo"` | `adfoControl()` | First-order Taylor expansion at η = 0; one rxSolve per NLL eval; fastest |
| Monte Carlo | `"admc"` | `admControl()` | Sample average over η; asymptotically exact |
| Gauss-Hermite | `"adgh"` | `adghControl()` | Deterministic quadrature over η; noise-free, unbiased at any IIV |
| Iterative Reweighting MC | `"adirmc"` | `adirmcControl()` | Proposals fixed per phase; inner loop needs no new rxSolve calls |

Start with `adfo` for screening and initial estimates. `admc` is the workhorse
for standard PK models; `adgh` is its noise-free alternative, most efficient when
there are few random effects; `adirmc` suits expensive ODE solves,
high-dimensional IIV or poor starting values. See
`vignette("estimator-comparison")`.

## The examplomycin dataset

`examplomycin` ships with admixr2: 500 simulated subjects from a two-compartment
PK model with first-order absorption (100 mg oral dose, sampled at 0.1, 0.25,
0.5, 1, 2, 3, 5, 8, and 12 h). True parameters: CL = 5 L/h, V1 = 10 L,
V2 = 30 L, Q = 10 L/h, ka = 1 h⁻¹; IIV = 0.3 (SD on log scale) for all
parameters; proportional error SD = 0.2.

```{r data}
library(admixr2)
library(rxode2)
library(nlmixr2)

data("examplomycin")
head(examplomycin[examplomycin$EVID == 0, c("ID", "TIME", "DV")], 9)
```

## Computing aggregate statistics

Reshape individual records into a subjects × times matrix, then compute E and V:

```{r aggregate}
dv_mat <- admVignetteDvMatrix()          # 500 subjects x 9 times

E     <- colMeans(dv_mat)
V     <- cov.wt(dv_mat, method = "ML")$cov
n     <- nrow(dv_mat)                    # 500
times <- as.numeric(colnames(dv_mat))

round(E, 2)
```

`admVignetteDvMatrix()` is a helper defined in this vignette's setup file; it
does nothing but reshape `examplomycin`'s individual records into one row per
subject and one column per time.

`V`'s off-diagonal entries capture within-subject correlation across time. Using
the full matrix (`method = "cov"`) typically tightens estimates against the
diagonal-only approximation (`method = "var"`); admixr2 picks the method from the
structure of `V`.

## Model definition

Models use standard nlmixr2 syntax with mu-referenced log-scale parameters:

```{r model}
pk_model <- function() {
  ini({
    tcl     <- log(5)  ; label("Log clearance (L/hr)")
    tv1     <- log(10) ; label("Log central volume (L)")
    tv2     <- log(30) ; label("Log peripheral volume (L)")
    tq      <- log(10) ; label("Log inter-compartmental CL (L/hr)")
    tka     <- log(1)  ; label("Log absorption rate constant (1/hr)")
    prop.sd <- c(0, 0.2); label("Proportional residual error SD")
    eta.cl ~ 0.09
    eta.v1 ~ 0.09
    eta.v2 ~ 0.09
    eta.q  ~ 0.09
    eta.ka ~ 0.09
  })
  model({
    cl <- exp(tcl + eta.cl)
    v1 <- exp(tv1 + eta.v1)
    v2 <- exp(tv2 + eta.v2)
    q  <- exp(tq  + eta.q)
    ka <- exp(tka + eta.ka)
    d/dt(depot)      <- -ka * depot
    d/dt(central)    <- ka * depot - (cl/v1 + q/v1) * central + (q/v2) * peripheral
    d/dt(peripheral) <- (q/v1) * central - (q/v2) * peripheral
    cp <- central / v1
    cp ~ prop(prop.sd)
  })
}
```

Writing each parameter as `exp(tcl + eta.cl)` is **mu-referencing**: the fixed
effect and its random effect enter additively on the log scale. admixr2 uses that
pairing to get analytical gradients from sensitivity equations — see
[Advanced usage](advanced.html#mu-referencing-and-sensitivity-equations),
which also covers parameters with no random effect.

## Assembling the study specification

Bundle each study's statistics into a named list:

```{r study}
study <- list(
  E     = E,
  V     = V,                       # full 9x9 covariance matrix
  n     = n,
  times = times,
  ev    = rxode2::et(amt = 100)    # single 100 mg oral dose
)
```

## Fitting

Pass one or more named studies to `admControl()`:

```{r fit}
fit <- nlmixr2(
  pk_model, admData(), est = "admc",
  control = admControl(
    studies   = list(examplomycin = study),
    n_sim     = 5000L,
    cov_n_sim = 10000L,
    maxeval   = 300L,
    seed      = 1L
  )
)
```

## Inspecting the fit

```{r print, comment = NA}
print(fit)
```

Key entries in `fit$env$admExtra`:

```{r internals}
fit$objective                    # -2 log-likelihood
fit$env$admExtra$struct          # structural parameters (log scale)
fit$env$admExtra$sigma_var       # residual variance(s)

logLik(fit)
AIC(fit)
```

The estimated between-subject covariance matrix `Omega`:

```{r internals-omega}
knitr::kable(fit$env$admExtra$omega, digits = 4,
             caption = "Estimated Omega (between-subject covariance).")
```

## Diagnostic plots

`plot()` draws up to four panel types and returns them as a named list of ggplot2
objects:

```{r plot, fig.cap="Mean diagnostics (observed vs predicted, with residuals), then the NLL convergence trace."}
plots <- plot(fit, which = c("mean", "nll"))
```

All four panel types and their customisation are covered in
`vignette("diagnostic-plots")`.

## Where to next

- [From a published figure to E, V and n](aggregate-data.html) — turn published summaries into `E`, `V` and `n`
- [Simulating data & using published models](datagen.html) — the model-as-input path
- [Multiple studies](multiple-studies.html) — meta-analysis across several studies
- [Estimator comparison](estimator-comparison.html) — choosing a backend
- [Choosing a residual error model](error-models.html) — what `cp ~ prop(...)` does to `E` and `V`
