---
title: "Advanced usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 5
)
source(Filter(file.exists, c("_common.R", "../_common.R", "vignettes/_common.R"))[1])
```

```{r data-hidden, include=FALSE}
library(admixr2)
library(rxode2)
library(nlmixr2)
library(ggplot2)

# The examplomycin study and model from `vignette("admixr2")`, assembled by the
# shared helpers in this vignette's setup file.
.s       <- admVignetteStats()
E        <- .s$E
V        <- .s$V
n        <- .s$n
times    <- .s$times
pk_model <- admVignetteModel
study    <- c(.s, list(ev = rxode2::et(amt = 100)))

# IIV on CL and V1 only -- the reduced model the AIC/BIC section compares
# against, and the one thing here that is not the running example.
pk_reduced <- function() {
  ini({
    tcl     <- log(5)
    tv1     <- log(10)
    tv2     <- log(30)
    tq      <- log(10)
    tka     <- log(1)
    prop.sd <- c(0, 0.2)
    eta.cl  ~ 0.09
    eta.v1  ~ 0.09
  })
  model({
    cl <- exp(tcl + eta.cl)
    v1 <- exp(tv1 + eta.v1)
    v2 <- exp(tv2)
    q  <- exp(tq)
    ka <- exp(tka)
    d/dt(depot)      <- -ka * depot
    d/dt(central)    <- ka * depot - (cl/v1 + q/v1) * central + (q/v2) * peripheral
    d/dt(peripheral) <- (q/v1) * central - (q/v2) * peripheral
    cp <- central / v1
    cp ~ prop(prop.sd)
  })
}
```

## Full covariance vs variance NLL

Each study's `V` selects a branch of the NLL, auto-detected:

| V structure | Auto-detected `method` | NLL cost |
|-------------|------------------------|----------|
| Non-diagonal matrix | `"cov"` | O(n_t³) Cholesky solve |
| Diagonal matrix | `"var"` | O(n_t) element-wise |
| Plain vector (variances only) | `"var"` | O(n_t) element-wise |

Override by setting `method` explicitly in the study list:

```{r var-vs-cov, eval=FALSE}
# Force full covariance NLL even for a diagonal V
study_cov <- c(study, list(method = "cov"))

# Diagonal approximation when you only have marginal variances
study_var <- list(
  E      = E,
  V      = diag(diag(V)),   # drop off-diagonal entries
  n      = n,
  times  = times,
  ev     = rxode2::et(amt = 100),
  method = "var"
)
```

Use `"cov"` when you have a full covariance matrix and expect the temporal
correlation to inform the estimates; `"var"` when only marginal variances exist —
a published table of means and SDs — or when runtime matters more.

## Gradient modes

Every control takes a `grad` argument. `admControl()` names its analytical mode
`"sens"`; `adfoControl()`, `adghControl()` and `adirmcControl()` call theirs
`"analytical"`. All four also take `"fd"` and `"none"`.

`"fd"` is a **central** difference. Forward differencing was removed in 0.4.1: it
measured 10^2^ to 10^4^ times less accurate than the analytic gradient, and the
one solve per parameter it saved did not pay for a gradient the optimizer
struggles to descend. Nor is the step a constant — it is measured per parameter
by the Shi (2021) procedure, with `grad_h` the fallback when it cannot be.

**`admControl()` gradient modes:**

| `grad =` | Method | Notes |
|----------|--------|-------|
| `"sens"` | Sensitivity equations (default) | Analytical; requires ODE or `linCmt()` model |
| `"fd"` | Central finite differences | Falls back to this if sens unavailable; step measured per parameter |
| `"none"` | BOBYQA (derivative-free) | No gradient; useful for debugging or simple models |

**`adfoControl()` gradient modes:**

| `grad =` | Method | Notes |
|----------|--------|-------|
| `"analytical"` (default) | Chain rule through V\_pred | Omega/sigma analytical; struct thetas from the order-2 sensitivity model, FD only if it cannot be built |
| `"none"` | BOBYQA | Derivative-free; was the default up to 0.4.0 |
| `"fd"` | Central FD of full NLL | All parameters; `2 n_p` NLL evals per step |

**`adghControl()` gradient modes:**

| `grad =` | Method | Notes |
|----------|--------|-------|
| `"analytical"` (default) | Closed-form contractions through sensitivity equations | Exact and cheapest; one batched sensitivity solve per study over the node grid |
| `"fd"` | Central FD of full NLL | All parameters; `2 n_p` NLL evals per step |
| `"none"` | BOBYQA (derivative-free) | No gradient |

The GH objective is **noise-free** — deterministic quadrature, no MC draws — so
its analytical gradient is exact and the Hessian well-conditioned. Keep the
default.

For `adfo`, every mode still uses the sensitivity model inside each NLL
evaluation to get J in one rxSolve; `grad` only controls how the *optimizer*
gradient is formed.

```{r grad-mode, eval=FALSE}
# Analytical gradient (default; fastest for ODE models)
fit_sens <- nlmixr2(pk_model, admData(), est = "admc",
  control = admControl(studies = list(s = study), grad = "sens",
                       n_sim = 5000L, seed = 1L))

# Derivative-free — ignores maxeval; use nloptr stopping criteria instead
fit_bobyqa <- nlmixr2(pk_model, admData(), est = "admc",
  control = admControl(studies = list(s = study), grad = "none",
                       n_sim = 5000L, seed = 1L))
```

Use `"sens"` for any standard ODE or `linCmt()` model. Where the sensitivity
model cannot be built — some rare ODE features — admixr2 falls back to `"fd"`
with a warning.

## Mu-referencing and sensitivity equations

**Mu-referencing** is the nlmixr2 convention of writing each structural parameter
as a fixed effect plus a random effect under a known back-transformation:

```r
cl <- exp(tcl + eta.cl)   # mu-referenced: tcl paired with eta.cl
```

`admixr2` uses this pairing to classify parameters into three cases, each
handled differently by `grad = "sens"`:

**Paired (mu-referenced) parameters** — rxode2 augments the ODE system with
sensitivity equations `d(pred)/d(eta_i)`. Since `tcl` and `eta.cl` enter
additively on the log scale, the NLL gradient with respect to `tcl` equals the
one with respect to `eta.cl`, and a single solve recovers both.

**Non-mu-referenced parameters** — written separately (`cl <- exp(tcl) *
exp(eta.cl)`), the eta-keyed sensitivities do not carry `d(pred)/d(tcl)`. admixr2
therefore emits its own first-order sensitivity model over an explicit *direction
set* — one direction per random effect, one per unpaired theta — compiled with
`eventSens = "jump"` so dosing-modifier (`f`/`lag`/`rate`/`dur`) sensitivities
come too. The unpaired thetas then get an **analytical** gradient from that same
solve, no finite differences, with a message saying so. This is the direction-set
scheme `nlmixr2est`'s fast-focei uses, at first order, cross-validated against
its inner model to ~1e-13. A targeted CRN finite-difference solve is the fallback
when the augmented model cannot be built.

**Parameters without a random effect** — one with no corresponding eta
(`v2 <- exp(tv2)`, IIV dropped from V2) is the same unpaired case: its own
direction in that sensitivity model, and an exact analytical gradient. (`adfo`
needs one thing more: `V_pred = J Ω Jᵀ + Σ` depends on the second derivative
`dJ/dθ`, which comes from the order-2 sensitivity model. Finite differences are
the fallback when that cannot be built, not the rule.)

## IRMC kappa correction

IRMC shifts its importance weights by moving the proposal mean when a structural
parameter changes. For a mu-referenced parameter that shift is a closed-form
function of the back-transformed ratio.

A theta with **no paired random effect** (`ka <- exp(tka)`, no `eta.ka`) has no
such weight path: changing `tka` in the inner loop moves `f(theta, 0)` and
nothing else follows. The kappa correction adds
`kappa = f(theta_cand, 0) - f(theta_outer, 0)` to the predicted mean, anchoring
the inner NLL to the true prediction at the candidate. Kappa is zero when every
theta is mu-referenced.

How `f(theta_cand, 0)` gets evaluated is the choice:

- **`"exact"`** (default) re-solves it at each inner NLL evaluation — one extra
  rxSolve per inner step.
- **`"linearized"`** precomputes `J = df/d(theta)` once per outer iteration from
  a small FD batch and takes
  `kappa_fn(theta_cand) ≈ f0 + J %*% (theta_cand - theta0)` — arithmetic only,
  no rxSolve at all.

The approximation holds while inner steps stay small against the outer box
constraint, which is typical once phases converge, so prefer `"linearized"`
wherever a solve is expensive.

```{r kappa, eval=FALSE}
adirmcControl(..., kappa_method = "exact")       # default
adirmcControl(..., kappa_method = "linearized")  # faster for complex ODE models
```

## Parallel restarts

Multi-restart fitting guards against local optima. `workers > 1` runs the
restarts in parallel, over a pool admixr2 manages itself:

```{r parallel, eval=FALSE}
fit_par <- nlmixr2(
  pk_model, admData(), est = "admc",
  control = admControl(
    studies    = list(examplomycin = study),
    n_sim      = 5000L,
    n_restarts = 4L,
    workers    = 4L,      # background worker processes (mirai daemons)
    cores      = 8L,      # total rxSolve OpenMP threads across workers
    restart_sd = 0.3,     # SD of log-scale perturbation from the starting point
    seed       = 1L
  )
)
```

`cores` is split automatically — `floor(total_cores / n_workers)` each, remainder
cyclically, so 13 cores over 4 workers gives 4, 3, 3, 3. Workers stop after the
restart phase, freeing every core for the covariance Hessian. Call
`admStopWorkers()` by hand if a fit is interrupted before that cleanup.

## Quasi-random sampling

The `sampling` argument controls how eta samples are drawn:

| `sampling =` | Method |
|-------------|--------|
| `"sobol"` | Sobol sequence (default) |
| `"halton"` | Halton sequence |
| `"torus"` | Kronecker torus |
| `"lhs"` | Latin hypercube |
| `"rnorm"` | Plain normal |

Sobol typically needs 2–5× fewer samples than plain normal draws for the same NLL
variance; at small `n_sim`, `"lhs"` covers better.

## Parameter uncertainty

`covMethod = "r"` takes a numerical Hessian after optimisation and reports
standard errors in `print(fit)`. The default is `covMethod = "r,s"` (below),
the sandwich correction of that same Hessian; ask for `"r"` to get the
uncorrected ones. SEs cover the structural, residual-error **and** omega
parameters, omega on its natural scale, with the Hessian rotated off the
Cholesky factor the optimiser works in. A larger `cov_n_sim` cuts MC noise in
the Hessian:

```{r cov-method, eval=FALSE}
fit_cov <- nlmixr2(
  pk_model, admData(), est = "admc",
  control = admControl(
    studies     = list(examplomycin = study),
    n_sim       = 5000L,
    cov_n_sim   = 10000L,   # more samples → lower Hessian noise
    cov_h_outer = 2.5e-3,   # outer FD step scale (default: eps^(1/5) ≈ 7.4e-4)
    covMethod   = "r",
    seed        = 1L
  )
)
```

Skip uncertainty altogether during early development, when only the point
estimates matter:

```{r no-cov, eval=FALSE}
admControl(..., covMethod = "none")
```

If the Hessian is non-positive-definite (SEs printed as `NA`), increase
`cov_h_outer` or `cov_n_sim`.

### `covMethod = "r,s"`: scoring the summary against its own sampling law

The aggregate objective is the exact log-likelihood of `n` iid draws from
`N(yt, Vt)`, which assumes each subject's observation vector is multivariate
normal. It is not: `y_i = f(theta, b_i) + eps_i` is nonlinear in the random
effect, so the marginal is a mixture. The point estimates do **not** pay for
this — the score has expectation zero at the true parameters under any weight,
so the fit stays consistent. The reported uncertainty does, in two ways that
differ in kind: `Cov(V_ij, V_kl)` is mis-sized by the excess kurtosis, and
`Cov(ybar, vech V)` is assumed zero where a real correlation of 0.3–0.6 sits.
Sample mean and sample covariance are exactly independent for a multivariate
normal and for nothing else.

`covMethod = "r,s"` scores `(ybar, vech V)` against its own asymptotic law and
reports the sandwich `H^-1 J H^-1`, where `H` is the same Hessian `"r"`
inverts and `J` is the variance of the score built from that law (the
model-derived weight above, not an empirical fourth moment):

```{r cov-rs, eval=FALSE}
fit_rs <- nlmixr2(
  pk_model, admData(), est = "adgh",
  control = adghControl(studies = list(examplomycin = study), covMethod = "r,s")
)
fit_rs$covMethod   # "r,s" if the correction was applied, "r" if it degraded
```

Point estimates and objective are identical to the `"r"` fit; only the SEs move.
Under correct specification `J = 2H`, so `"r,s"` returns exactly what `"r"`
would: the two agreeing is the expected outcome for a well-specified model, not
a sign that nothing happened.

It covers every residual family whose conditional law is independent across
timepoints — all of them except `ar()`: the conditionally-normal set (`add`,
`prop`, `pow`, `combined1`, `combined2`), the closed-form distributional ones
(`lnorm`, `pois`, `binom`, `nbinomMu`, `beta`, `t(nu > 4)`) and the
transform-both-sides ones (`boxCox`, `yeoJohnson`, `logitNorm`, `probitNorm`).
What it cannot build — `ar()`, `t()` with `nu <= 4`, `ordinal()`, a same-subject
joint study, a singular ingredient — degrades to `"r"` with a warning, and
`fit$covMethod` then reads `"r"`: it records what the covariance **is**, not
what was asked for.

Under `"r,s"` the denominator of a study's reported `V` stops being cosmetic —
see `v_denom` in the [aggregate data](aggregate-data.html) vignette.

## Model comparison: AIC and BIC

Standard information criteria work directly on `admFit` objects. **AIC and BIC
compare only across estimators evaluating the same likelihood.** `admc`, `adgh`
and `adirmc` all target the exact aggregate MVN likelihood — MC, quadrature and
importance-weighted estimates of one integral — so they are mutually comparable.
`adfo`'s is *linearised*, and must not be set against them.

Full model (IIV on all five parameters) against a reduced one (IIV on CL and V1):

```{r model-compare}
ctl <- admControl(
  studies   = list(examplomycin = study),
  n_sim     = 5000L,
  cov_n_sim   = 10000L,
  maxeval   = 300L,
  seed      = 1L
)

fit_full    <- nlmixr2(pk_model,   admData(), est = "admc", control = ctl)
fit_reduced <- nlmixr2(pk_reduced, admData(), est = "admc", control = ctl)

AIC(fit_full, fit_reduced)
BIC(fit_full, fit_reduced)
```

The model with the lower AIC/BIC is preferred; a difference > 10 is generally
considered strong evidence. BIC penalises added parameters more heavily than
AIC, so the two can disagree on a marginal eta.

### `anova()`: a likelihood-ratio test on nested fits

The two are **nested** -- `fit_reduced` is `fit_full` without `eta.v2`, `eta.q`
and `eta.ka` -- so a chi-squared test compares their objectives directly:

```{r model-compare-anova}
anova(fit_full, fit_reduced)
```

`dOFV` is the objective difference and `Df` the number of parameters the
larger model adds (three etas here); `p` is `pchisq(dOFV, Df,
lower.tail = FALSE)`, the ordinary likelihood-ratio test.

`anova()` refuses fits that are not on the same footing. Different estimators
score different approximations (FO-linearised, quadrature, Monte Carlo), and
changing `n_nodes` or `n_sim` moves the objective with the grid or sample size
rather than with the model. All three are errors rather than a number returned
quietly. See `?anova.admFit` for the boundary-variance case — testing whether an
eta's variance is zero — and why a negative `dOFV` is reported, not clamped.

## See also

- [Estimator comparison](estimator-comparison.html) — the mathematical foundations of each backend
- [Diagnostic plots](diagnostic-plots.html)
