---
title: "From a published figure to E, V and n"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From a published figure to E, V and n}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.2, dpi = 120
)
# nlmixr2's fit print method uses cli colour; when a build forces colour on (e.g.
# pkgdown/CI) the raw ANSI escapes leak into the HTML. Force plain-text output
# for this vignette only, saving the previous global state...
.adm_opts    <- options(cli.num_colors = 1, crayon.enabled = FALSE)
.adm_nocolor <- Sys.getenv("NO_COLOR", unset = NA)
Sys.setenv(NO_COLOR = "1")

# ...and restoring it once the document has finished knitting, so the setting
# does not leak into other articles when pkgdown builds them in one process.
knitr::knit_hooks$set(document = function(x) {
  options(.adm_opts)
  if (is.na(.adm_nocolor)) Sys.unsetenv("NO_COLOR") else Sys.setenv(NO_COLOR = .adm_nocolor)
  x
})

# The estimators keep rxode2's progress bar off during fitting (the control's
# nDisplayProgress argument), but a one-off bar still prints while each model is
# *compiled*. Strip those animation frames from the rendered output.
knitr::knit_hooks$set(output = local({
  .default <- knitr::knit_hooks$get("output")
  function(x, options) {
    x <- gsub("[^\n]*\\[[=| ]+\\][^\n]*\n?", "", x, perl = TRUE)
    .default(x, options)
  }
}))
```

## The problem

Every study passed to `admControl()` needs a mean vector `E`, a covariance `V`, a
sample size `n`, the observation `times` and a dosing event table `ev`. A
published figure gives you a mean and an error bar.

Reading the mean off is easy. Getting the variance right is where aggregate
analyses go wrong: a standard error used as a standard deviation is off by a
factor of `sqrt(n)`, and the fit does not tell you — the structural parameters
come back correct and only the between-subject variability collapses. This
vignette converts a published figure into `E`, `V` and `n`, and shows what that
mistake costs.

```{r libs}
library(admixr2)
library(rxode2)
library(nlmixr2)
library(ggplot2)
```

## What V must be

`V` is the spread across *subjects*, not the precision of the mean. admixr2's
likelihood is

$$-2LL = n \left( \log|V_{pred}| + \mathrm{tr}(V_{pred}^{-1} V_{obs}) + r^{\top} V_{pred}^{-1} r \right)$$

where `r` is the mismatch between observed and predicted means, and
`V_pred = J Omega J' + Sigma` — one subject's covariance, built from the
between-subject variability and the residual error. `V_obs` must be the same
object, and the sample size `n` sits outside it. Hand the likelihood a standard
error squared and you have told it about `n` twice.

So `V = SD^2`, never `SEM^2`.

## What is the error bar?

Read the caption. If it does not say, treat the error bar as unknown rather than
assuming:

| Figure reports | Convert to SD | Note |
|---|---|---|
| Standard deviation (SD) | `SD` | Use directly |
| Standard error (SEM) | `SD = SEM * sqrt(n)` | Off by `sqrt(n)` if confused |
| 95% CI of the mean | `SD = (upper - lower) * sqrt(n) / 3.92` | 3.92 = 2 × 1.96; for small `n` use `2 * qt(0.975, n - 1)` |
| Interquartile range | `SD ~ IQR / 1.35` | Assumes normality |
| LS-mean SE (from an MMRM) | `SD ~ SE * sqrt(n)` | Biased small — see below |

```{r convert}
sd_from_sem <- function(sem, n)        sem * sqrt(n)
sd_from_ci  <- function(lower, upper, n) (upper - lower) * sqrt(n) / (2 * qt(0.975, n - 1))
sd_from_iqr <- function(q1, q3)        (q3 - q1) / 1.35
```

An LS-mean SE deserves care because it is the one row that can silently
reproduce the error this vignette is about. It comes out of a model, usually an
MMRM: baseline and covariate adjustment remove variance from the residual, and
the covariance structure borrows across visits, so `SE * sqrt(n)` is usually
**smaller** than the true between-subject SD — the same direction as mistaking a
SEM for an SD. It also describes whatever the MMRM modelled, often a change from
baseline rather than an absolute value. Prefer a descriptive SD from the paper's
own baseline table, or from a comparable study.

## A figure reporting standard errors

A 50 mg arm, 60 subjects, digitised from a concentration–time figure whose
caption reads "mean ± SEM" — the same arm used in
[PD and PK/PD data](pkpd.html), before it was fitted:

```{r data}
times <- c(0.5, 1, 2, 4, 8, 12, 24)
E     <- c(0.969, 0.920, 0.831, 0.679, 0.460, 0.317, 0.111)     # mg/L
SEM   <- c(0.0267, 0.0243, 0.0205, 0.0161, 0.0137, 0.0125, 0.0080)
n     <- 60L

SD <- sd_from_sem(SEM, n)
round(SD, 3)
```

The bars the figure plots are `sqrt(60)` — nearly eight times — smaller than the
standard deviations the fit needs. Plotting both makes it concrete: the SEM band
is far too tight to be a between-subject spread.

```{r band-plot, fig.height = 3.6}
band <- rbind(
  data.frame(t = times, m = E, lo = E - SD,  hi = E + SD,  what = "mean ± SD (what V needs)"),
  data.frame(t = times, m = E, lo = E - SEM, hi = E + SEM, what = "mean ± SEM (what the figure plots)"))

ggplot(band, aes(t, m)) +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = what), alpha = 0.25) +
  geom_line(linewidth = 0.9, colour = "grey20") +
  geom_point(size = 1.8, colour = "grey20") +
  facet_wrap(~ what) +
  scale_fill_manual(values = c("mean ± SD (what V needs)"          = "#0072B2",
                               "mean ± SEM (what the figure plots)" = "#D55E00"),
                    guide = "none") +
  labs(x = "Time (h)", y = "Concentration (mg/L)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        strip.text = element_text(face = "bold", size = 10))
```

Passing `V` as a plain vector of variances is enough; admixr2 expands it to a
diagonal matrix and sets `method = "var"`:

```{r study}
study <- list(E = E, V = SD^2, n = n, times = times,
              ev = rxode2::et(amt = 50, cmt = "central"))
```

```{r model}
pk_model <- function() {
  ini({
    tcl <- log(4)  ; label("Log clearance (L/h)")
    tv  <- log(40) ; label("Log volume (L)")
    prop.cp <- 0.1 ; label("Proportional residual error")
    eta.cl ~ 0.09
    eta.v  ~ 0.04
  })
  model({
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv  + eta.v)
    d/dt(central) <- -(cl/v) * central
    cp <- central / v
    cp ~ prop(prop.cp)
  })
}
```

```{r fit, comment = NA}
fit <- nlmixr2(pk_model, admData(), est = "adgh",
               control = adghControl(studies = list(mg50 = study)))
fit
```

## What reading SEM as SD costs

Now make the mistake: use the plotted `SEM` as if it were an `SD`, so `V` is
`n`-fold too small.

```{r sem-demo, comment = NA}
fit_wrong <- nlmixr2(
  pk_model, admData(), est = "adgh",
  control = adghControl(studies = list(
    mg50 = list(E = E, V = SEM^2, n = n, times = times,   # WRONG: SEM^2 as V
                ev = rxode2::et(amt = 50, cmt = "central")))))
fit_wrong
```

Clearance and volume are unchanged to three figures. The between-subject
variability is not:

```{r sem-omega}
round(diag(fit$omega) / diag(fit_wrong$omega), 1)
```

`Omega` shrinks by more than an order of magnitude on both random effects,
because the model is asked to reproduce a between-subject spread `n` times
tighter than the real one. `V_pred = J Omega J' + Sigma` is linear in `Omega` and
in the residual variance, so both shrink together and the structural parameters
are free to stay where they were. The shrinkage stops short of the full factor of
`n` only because the mean-mismatch term `r' V_pred^-1 r` does not rescale.

Nothing in the *point estimates* warns you. The *precision* does:

```{r sem-rse}
round(c(RSE_CL_correct = fit$parFixedDf["tcl", "%RSE"],
        RSE_CL_wrong   = fit_wrong$parFixedDf["tcl", "%RSE"]), 3)
```

Clearance comes back not just right but implausibly certain — its standard error
tightens about
`r round(fit$parFixedDf["tcl","%RSE"] / fit_wrong$parFixedDf["tcl","%RSE"], 1)`-fold,
of order `sqrt(n)`. An RSE that looks too good for digitised literature data, or
an IIV that comes back near zero, is the tell.

Two things make this worse than the demo suggests:

* **The structural parameters only survive because this model fits.** With a
  misspecified model, `r` is not zero, and a `V_pred` that is `n` times too small
  weights that mismatch `n` times too heavily — dragging the structural estimates
  to close a gap they cannot close.
* **In a multi-study fit, one bad `V` captures the whole fit.** Getting `V` wrong
  in a single study inflates that study's weight in the joint likelihood roughly
  `n`-fold. It then dominates every other arm and biases the shared parameters.
  This is the usual way admixr2 is used, and it is where the mistake is most
  expensive.

## Why V is usually diagonal

A figure gives one error bar per time point and nothing about how the times
covary, so the off-diagonal entries of `V` are not available from published
summaries. A diagonal `V` selects `method = "var"`, which skips the Cholesky
solve the full-covariance path needs. This is the honest default for literature
data. The full-covariance path is available when you have the subject-level
matrix and can compute `cov.wt(dv_mat, method = "ML")$cov` — see
[Getting started](admixr2.html).

Note the denominator. admixr2's likelihood wants the ML (`n`) variance, while a
published SD is the unbiased (`n - 1`) sample SD, so strictly
`V = SD^2 * (n - 1) / n`. At `n = 60` that is a 1.7% change and is usually
ignored; below `n` of about 15 it is worth applying. The `method = "ML"` rule
matters most when you compute `V` yourself from subject-level data.

## Sample size

`n` is the number of subjects contributing to the summary, per arm — not the
total across arms, and not the number of observations.

* **Dropout.** `n` often falls over time, so a figure's late points may rest on
  fewer subjects than its early ones. Convert each point's error bar with the `n`
  that applies to it, and pass the number contributing to the observations you are
  fitting — the number at risk, not the number randomised.
* **Per-endpoint `n`.** PK and PD are not always measured in the same people. A
  study's `observations` entries may each carry their own `n`.

## Absolute values or change from baseline?

Many PD papers report a least-squares-mean change from baseline rather than an
absolute value. Either can be fitted, but the model must predict the same
quantity as the data:

* Fitting absolute values needs a baseline parameter in the model; see
  [PD and PK/PD data](pkpd.html).
* Fitting a change means the model output must itself be a change, and the SD of
  the change — not of the absolute value — is the one `V` needs.

Studies that report different quantities must be converted to a common one before
fitting, not after.

## When there is no variability at all

A paper often gives a mean with no SD, SEM or CI — a placebo arm reported only in
a footnote, for example.

You then have to assume a `V`, and it is worth knowing what that assumption does.
Refit the arm above with the SD deliberately wrong in each direction:

```{r assume-sweep}
assume <- function(mult) {
  f <- nlmixr2(pk_model, admData(), est = "adgh",
               control = adghControl(studies = list(
                 mg50 = list(E = E, V = (SD * mult)^2, n = n, times = times,
                             ev = rxode2::et(amt = 50, cmt = "central")))))
  c(exp(f$theta[["tcl"]]), exp(f$theta[["tv"]]),
    diag(f$omega)[["eta.cl"]], diag(f$omega)[["eta.v"]])
}

mults <- c(0.5, 1, 2, 4)
res   <- t(vapply(mults, assume, numeric(4)))

tbl <- data.frame(
  `Assumed SD` = c("0.5x  (too small)", "1x  (correct)",
                   "2x  (too large)",   "4x  (too large)"),
  CL           = round(res[, 1], 2),
  V            = round(res[, 2], 2),
  `var(eta.cl)` = round(res[, 3], 3),
  `var(eta.v)`  = round(res[, 4], 3),
  check.names  = FALSE
)

knitr::kable(tbl, row.names = FALSE,
             caption = "Fit against a deliberately wrong V. Truth: CL = 5, V = 50, var(eta.cl) = 0.09, var(eta.v) = 0.04.")
```

`Omega` follows the assumption in whichever direction it is wrong: too small and
the between-subject variability collapses, too large and it inflates several-fold.
`V` is data the model must reproduce, not a weight — so an assumed `V` is an
assumption about the IIV you are trying to estimate, and it propagates to every
study through the shared `Omega`. Push it far enough and the fit stops being
sensible at all: at `4x` the structural parameters leave the building.

There is no safe direction to err in. In rough order of preference:

1. Take the variability from a comparable arm, or a comparable study of the same
   endpoint.
2. Use a published typical SD for that endpoint and population.
3. Exclude the study.

Whichever you choose, record it, and refit across the range you consider
plausible. If the estimates move, the assumption is doing the work and the result
belongs in a sensitivity table rather than a headline.

## Notes

* **The error bar is the dominant error source.** Digitisation software is
  accurate to a few percent; mistaking a SEM for an SD is an error of `sqrt(n)`.
  Reading the caption matters more than the pixels.
* **Geometric means.** A paper reporting a geometric mean and CV% is describing a
  log-normal distribution, while `E` and `V` are arithmetic moments. Convert
  before fitting.
* **Digitising.** WebPlotDigitizer is the usual tool. Extracting a figure twice
  and comparing is a cheap check.
* **`%RSE` is on the estimation scale.** These parameters are log-scale, so an
  RSE on `tcl` is not an RSE on clearance.
* **PD specifics** — baselines, placebo arms, two endpoints — are covered in
  [PD and PK/PD data](pkpd.html).
