---
title: "Covariates in aggregate data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covariates in aggregate data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 5
)
source(Filter(file.exists, c("_common.R", "../_common.R", "vignettes/_common.R"))[1])
```

```{r libs}
library(admixr2)
library(rxode2)
library(nlmixr2est)
```

## The problem covariates pose for aggregate data

With individual data, a covariate is a column: each subject has a weight, and
the model reads it. With **aggregate** data there is no such column. A
publication reports one pooled mean vector and one covariance matrix for a
cohort whose weights varied — and it reports, at best, the *distribution* those
weights came from ("median 71 kg, range 48–104").

So a covariate enters admixr2 one of two ways, and the distinction runs through
everything below:

* **Marginalised** — the study declares a `cov_dist`, and the estimator
  integrates the prediction over that distribution. This is what a pooled
  summary requires.
* **Conditioned** — the study declares a `cov` value, and the model is solved
  at that value. This is what a stratified summary allows ("results for
  patients under 70 kg").

A single study can do both: marginalise over weight while sitting in a reported
age stratum.

## Declaring a distribution

`covDist()` takes one named argument per covariate, named as the model reads
it.

```{r covdist}
cd <- covDist(WT = c(mean = 72, sd = 16), dist = "lnorm")
cd$WT
```

The `mean`/`sd` shorthand is moment-matched onto the requested margin, so the
lognormal above really does have mean 72 and SD 16. You can also name a margin
directly (`meanlog`/`sdlog`, or a `quantile` function), which is what you want
when the paper reports a geometric mean.

`covDraw()` shows you what you declared, which is worth doing before fitting:

```{r covdraw}
set.seed(1)
draws <- covDraw(cd, n = 4000L)
round(c(mean = mean(draws[, "WT"]), sd = stats::sd(draws[, "WT"]),
        min = min(draws[, "WT"]), max = max(draws[, "WT"])), 2)
```

A **normal** margin is unbounded below, and the quadrature reaches about 3.75
standard deviations. Any covariate with a CV above roughly 0.27 therefore puts
a node at or below zero, and an allometric term evaluated there is `NaN` —
`covDist()` warns when it sees this. Weight at 72 ± 16 is CV
`r round(16 / 72, 2)`, which is why `dist = "lnorm"` is the right default for
most PK covariates.

## Why several studies matter

Here is the part that has no individual-data analogue. Simulate three trials
that differ in the weight distribution they recruited:

```{r simulate}
set.seed(42)
TIMES <- c(0.5, 1, 2, 4, 8, 16)
DOSE  <- 100
TRUE_CL <- 1; TRUE_V <- 10; OM_CL <- 0.09; SD_ADD <- 0.3
B_WT <- 0.75                       # allometric exponent on volume
CV_WT <- 0.35
sdlog <- sqrt(log(1 + CV_WT^2))

make_study <- function(median_wt, n_sub) {
  meanlog <- log(median_wt)
  WT <- stats::rlnorm(n_sub, meanlog, sdlog)
  cl <- TRUE_CL * exp(stats::rnorm(n_sub, 0, sqrt(OM_CL)))
  v  <- TRUE_V * (WT / 70)^B_WT
  cp <- t(vapply(seq_len(n_sub), function(i)
    DOSE / v[i] * exp(-cl[i] / v[i] * TIMES), numeric(length(TIMES))))
  cp <- cp + matrix(stats::rnorm(length(cp), 0, SD_ADD), nrow(cp), ncol(cp))
  list(E = colMeans(cp),
       V = stats::cov.wt(cp, method = "ML")$cov,
       n = n_sub, times = TIMES, ev = rxode2::et(amt = DOSE),
       cov_dist = covDist(WT = list(meanlog = meanlog, sdlog = sdlog)))
}
studies <- list(
  light = make_study(58, 300L),
  usual = make_study(71, 300L),
  heavy = make_study(93, 300L)
)
knitr::kable(
  data.frame(study = names(studies),
             n     = vapply(studies, `[[`, numeric(1), "n"),
             median_WT = c(58, 71, 93),
             `E at 1h` = vapply(studies, function(s) s$E[2L], numeric(1)),
             check.names = FALSE),
  digits = 2,
  caption = "Three cohorts differing only in the weight distribution recruited."
)
```

**`V` must use the ML denominator `n`, not `n - 1`** — `cov.wt(x, method =
"ML")$cov`, never `cov(x)`. That follows directly from the MVN likelihood for
`n` iid subjects, and it is the single most common way to get an aggregate fit
subtly wrong.

Now fit. The model reads `WT` exactly as an individual-data model would:

```{r fit, comment = NA}
mod <- function() {
  ini({
    tcl <- log(1.3)
    tv  <- log(8)
    bwt <- 0.4
    eta.cl ~ 0.05
    add.err <- 0.4
  })
  model({
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv) * (WT / 70)^bwt
    cp <- linCmt()
    cp ~ add(add.err)
  })
}
fit <- nlmixr2(mod, admData(), est = "adgh",
               control = adghControl(studies = studies, print = 0L,
                                     covMethod = "none"))
print(fit)
```

```{r recovery}
est <- stats::setNames(as.numeric(fit$parFixedDf$Estimate),
                       rownames(fit$parFixedDf))
knitr::kable(
  data.frame(parameter = c("CL", "V", "allometric exponent", "additive SD"),
             truth     = c(TRUE_CL, TRUE_V, B_WT, SD_ADD),
             estimate  = c(exp(est[["tcl"]]), exp(est[["tv"]]),
                           est[["bwt"]], est[["add.err"]])),
  digits = 3, caption = "Recovery from three aggregate summaries."
)
```

### The covariate coefficient needs between-study contrast

The allometric exponent above is identified because the three cohorts have
*different* weight distributions. Fit one of them alone and it is much weaker:

```{r single}
fit1 <- nlmixr2(mod, admData(), est = "adgh",
                control = adghControl(studies = studies["usual"], print = 0L,
                                      covMethod = "none"))
e1 <- stats::setNames(as.numeric(fit1$parFixedDf$Estimate),
                      rownames(fit1$parFixedDf))
knitr::kable(
  data.frame(
    parameter    = c("allometric exponent", "omega (eta.cl)"),
    truth        = c(B_WT, OM_CL),
    three_studies = c(est[["bwt"]], as.numeric(fit$omega[1L, 1L])),
    one_study     = c(e1[["bwt"]], as.numeric(fit1$omega[1L, 1L]))),
  digits = 3,
  caption = "The covariate effect needs contrast across sources; omega absorbs what it cannot explain."
)
```

A single pooled summary sees the covariate only through what it does to that
one cohort's mean and spread — and weight enters `V` in the same place the
between-subject variance does, so `bwt` and `om.eta.cl` trade off. Both drift
upward together above, which is the signature: the exponent is doing work that
`omega` should, and vice versa. One cohort is not *nothing* — a single
covariate still bends the mean profile in a way `omega` cannot mimic exactly —
but it is weak, and it gets rapidly weaker as covariates are added.
**Covariate effects in aggregate data are identified by contrast across
sources**, exactly as in any meta-regression.

Two covariates that share a distribution have the same problem between
*themselves*. With `W1`, `W2`, `W3` all lognormal and independent, the
aggregate depends on the exponents only through `Σb` and `Σb²` — two
functionals for three parameters — so they are not separately identified
however many studies you add. Vary the covariates differently across studies,
or fix all but one.

## Variance or covariance

`V` may be a full covariance matrix or a plain vector of variances:

```{r vvar, comment = NA}
s_var <- studies$usual
s_var$V <- diag(s_var$V)             # keep only the diagonal
fit_var <- nlmixr2(mod, admData(), est = "adgh",
                   control = adghControl(studies = list(usual = s_var),
                                         print = 0L, covMethod = "none"))
c(cov_method = fit$objective, var_method = fit_var$objective)
```

A vector (or a diagonal matrix) is auto-detected and sets `method = "var"`,
which skips the `O(n_t³)` Cholesky solve. Use it when the paper reports SDs per
timepoint and no correlation structure — which is most of the time. Note the
objectives are **not comparable** across the two: they score different
likelihoods.

One thing to be clear about: the covariate marginalisation moves `V`, not just
`E`. Between-subject weight variation is a source of between-subject
variability in the prediction, so a model that marginalises correctly needs
*less* `omega` to explain the same observed spread. Solving at the mean weight
instead pushes that variance into `omega`, which is why an unwired covariate
inflates it.

## Correlated covariates

Real covariates are dependent — weight with height, creatinine with age.
`covDist(cor = )` builds a **Gaussian copula**: each covariate keeps the margin
you declared, and the correlation is imposed on the latent scale.

```{r copula}
R <- matrix(c(1,   0.6, 0.3,
              0.6, 1,   0.4,
              0.3, 0.4, 1  ), 3, 3,
            dimnames = list(c("WT", "HT", "CRCL"), c("WT", "HT", "CRCL")))
cdc <- covDist(WT   = c(mean = 72,  sd = 16),
               HT   = c(mean = 170, sd = 10),
               CRCL = c(mean = 90,  sd = 25),
               cor  = R, dist = "lnorm")
set.seed(2)
d <- covDraw(cdc, n = 8000L)
knitr::kable(round(stats::cor(d), 2),
             caption = "Realised Spearman-like correlation of the drawn covariates.")
```

The realised correlations sit slightly below the requested ones — that is the
copula, not an error: a Gaussian dependence of `r R["WT", "HT"]` on the latent
scale maps to a slightly weaker correlation once each margin is pushed through
a nonlinear quantile function.

For dependence a Gaussian copula cannot express, pass your own sampler as
`joint` — anything that takes an `n × d` matrix of uniforms and returns the
`n × d` covariate values, which is exactly the shape a vine copula produces:

```{r vine, eval = requireNamespace("rvinecopulib", quietly = TRUE)}
vine <- rvinecopulib::vinecop_dist(
  pair_copulas = list(list(rvinecopulib::bicop_dist("clayton", 0, 2),
                           rvinecopulib::bicop_dist("gumbel",  0, 1.5)),
                      list(rvinecopulib::bicop_dist("gaussian", 0, 0.2))),
  structure = rvinecopulib::dvine_structure(1:3))
cdv <- covDist(
  WT = c(mean = 72, sd = 16), HT = c(mean = 170, sd = 10),
  CRCL = c(mean = 90, sd = 25), dist = "lnorm",
  joint = function(u) {
    z <- rvinecopulib::inverse_rosenblatt(u, vine)
    cbind(WT   = stats::qlnorm(z[, 1], log(72),  0.22),
          HT   = stats::qlnorm(z[, 2], log(170), 0.06),
          CRCL = stats::qlnorm(z[, 3], log(90),  0.27))
  })
set.seed(3)
round(stats::cor(covDraw(cdv, n = 4000L)), 2)
```

A user-supplied `joint` is treated as opaque: admixr2 samples from it but
cannot see its latent structure, so the design shortcuts described below are
not available and some studies will fall back to a product grid.

## Discrete covariates

Levels and probabilities, rather than a margin:

```{r discrete}
cdd <- covDist(WT  = c(mean = 72, sd = 16),
               SEX = list(values = c(0, 1), probs = c(0.45, 0.55)),
               dist = "lnorm")
set.seed(4)
table(covDraw(cdd, n = 2000L)[, "SEX"])
```

A discrete covariate is a **stratum**, not a direction: it is enumerated
exactly at its levels, weighted by its probabilities, and crossed with the
continuous design. Nothing is approximated about it.

If a study reports results *separately* by sex, that is conditioning rather
than marginalising — split it into one study per stratum with `cov = list(SEX =
0)` and the stratum's own `n`. `covStrata()` does that split for you.

## What it costs

Every design point is a row in the ODE solve, so a naive product grid over
`p` covariates and `n_eta` random effects costs `n^(n_eta + p)` — which grows
badly. admixr2 avoids that where the model allows it.

```{r cost, comment = NA}
mod3 <- function() {
  ini({
    tcl <- log(1); tv <- log(10)
    b1 <- 0.6; b2 <- 0.4; b3 <- 0.3
    eta.cl ~ 0.09
    add.err <- 0.3
  })
  model({
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv) * (W1/70)^b1 * (W2/70)^b2 * (W3/70)^b3
    cp <- linCmt()
    cp ~ add(add.err)
  })
}
lg <- function() list(meanlog = log(70), sdlog = sdlog)
st3 <- list(s1 = list(E = studies$usual$E, V = studies$usual$V, n = 300L,
                      times = TIMES, ev = rxode2::et(amt = DOSE),
                      cov_dist = covDist(W1 = lg(), W2 = lg(), W3 = lg())))
fit3 <- nlmixr2(mod3, admData(), est = "adgh",
                control = adghControl(studies = st3, print = 0L,
                                      covMethod = "none", maxeval = 5L))
```

admixr2 says what it did when it set that study up (the message is suppressed
in this vignette's chunk options). You can see the same thing in the design it
built:

```{r collapse}
ui3  <- suppressMessages(rxode2::rxode2(mod3))
pin3 <- admixr2:::.admDriverPinfo(
  ui3, adghControl(studies = st3, covMethod = "none"))
st3n <- suppressMessages(admixr2:::.admCheckCovariates(
  ui3, pin3, admixr2:::.admDriverUnits(
    st3, ui3, admixr2:::.admOutputVar(ui3))$studies))
co3  <- st3n[[1L]][[".adm_cov_collapse"]]
c(covariates      = co3$p,
  directions      = co3$r,
  design_points   = nrow(co3$X),
  product_grid    = pin3$cov_nodes^co3$p)
```

The three covariates reach the model through a single scalar — the product
`(W1/70)^b1 (W2/70)^b2 (W3/70)^b3` — so the integral is one-dimensional however
many covariates feed it.

**This is a change of variables, not an approximation.** The prediction depends
on the covariates only through that scalar; integrating over the scalar and
integrating over the product give the same answer, and admixr2 verifies that
numerically before using the reduced design. Where the model does *not* have
that structure — a covariate interacting with a random effect, for instance —
it falls back to the full grid rather than guessing.

`cov_nodes` (default `r pin3$cov_nodes`) is the budget. It is a nodes-per-*direction*
budget, not a literal node count: a direction that absorbs several covariate
axes carries their combined spread and is given proportionally more resolution.
Raise it if a covariate distribution is very wide or its effect strongly
nonlinear.

## Estimator support

| Estimator | Covariates |
|---|---|
| `admc` | yes — per-subject draws |
| `adgh` | yes — quadrature, and the design reductions above |
| `adfo` | **refused** |
| `adirmc` | **refused** |

`adfo` and `adirmc` error rather than silently solving at the covariate mean.
That is deliberate: every study carries a covariate *value* as well as a
distribution, so an unwired estimator would not fail — it would quietly return
a fit whose `omega` had absorbed the between-subject covariate spread. Measured
on the test model, that is `omega` 0.30 → 0.44.

## See also

- [Multiple studies](multiple-studies.html) — the contrast that identifies a covariate effect
- [From a published figure to E, V and n](aggregate-data.html) — preparing each study's summary
- [datagen()](datagen.html) — simulating aggregate data, including covariate distributions
- [Estimator comparison](estimator-comparison.html)
