---
title: "Covariates in a model-based meta-analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covariates in a model-based meta-analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.2)
source("_common.R", local = TRUE)
```

Three published population models are used as the input to a meta-analysis: each
generates the aggregate data its own study design and population would have
produced, and one unifying model is then fitted to all three at once. No
individual records exist at any point.

The covariate is the point of the exercise. Each study's subjects span a
distribution of body weights, and that distribution differs between studies —
which is not incidental, because between-study variation in the covariate
distribution is the only thing that identifies a covariate coefficient from
aggregate data. Two ways of describing those subjects are compared:

* **summary statistics** — a mean, a standard deviation and a correlation, which
  is what a baseline-characteristics table reports;
* **an R-vine copula per study** — a full joint distribution with its own
  margins and dependence structure, which is what you would fit if you had access
  to the individual covariate records even though the outcomes remain aggregate.

Each is run through both estimators.

```{r libs}
library(admixr2)
library(rxode2)
has_vine <- requireNamespace("rvinecopulib", quietly = TRUE) &&
            requireNamespace("kde1d", quietly = TRUE)
```

# The three published models

Weight and creatinine clearance both act on clearance, which carries the
between-subject random effect. The three models differ in the designs they were
developed under --- a single dose with rich sampling, a multiple-dose regimen at
steady state, and a sparse two-sample design --- and, more importantly, in the
populations they enrolled: one with preserved renal function, one with moderate
impairment, one with severe.

Their parameter estimates differ modestly, as published models of one drug
usually do --- clearance within 6%, volume within 5%, random-effect and residual
magnitudes within about 10%. That agreement matters more than it looks, and Act
III is about why: the unifying model has ONE of each of those parameters, so
every disagreement between the studies has to go somewhere, and the covariate
slopes are where it goes.

```{r models}
mod_A <- function() {          # rich single dose, preserved renal function
  ini({lcl <- log(3.1); lv <- log(20.5); clwt <- 0.95; clcr <- 0.95
       eta.cl ~ 0.085; prop.err <- 0.12})
  model({cl <- exp(lcl + eta.cl) * (WT / 70)^clwt * (CRCL / 95)^clcr
         v  <- exp(lv)
         d/dt(centr) <- -cl / v * centr
         cp <- centr / v
         cp ~ prop(prop.err)})
}
mod_B <- function() {          # multiple dose at steady state, moderate impairment
  ini({lcl <- log(3.2); lv <- log(21.0); clwt <- 1.00; clcr <- 1.00
       eta.cl ~ 0.090; prop.err <- 0.13})
  model({cl <- exp(lcl + eta.cl) * (WT / 70)^clwt * (CRCL / 95)^clcr
         v  <- exp(lv)
         d/dt(centr) <- -cl / v * centr
         cp <- centr / v
         cp ~ prop(prop.err)})
}
mod_C <- function() {          # sparse, severe impairment
  ini({lcl <- log(3.3); lv <- log(21.5); clwt <- 1.05; clcr <- 1.05
       eta.cl ~ 0.095; prop.err <- 0.14})
  model({cl <- exp(lcl + eta.cl) * (WT / 70)^clwt * (CRCL / 95)^clcr
         v  <- exp(lv)
         d/dt(centr) <- -cl / v * centr
         cp <- centr / v
         cp ~ prop(prop.err)})
}
DESIGN <- list(
  A = list(model = mod_A, times = c(1, 2, 4, 8, 12, 24),
           ev = rxode2::et(amt = 100, dur = 0.5), n = 48L),
  B = list(model = mod_B, times = c(2, 6, 12, 24),
           ev = rxode2::et(amt = 80, ii = 12, addl = 4, dur = 0.5), n = 96L),
  C = list(model = mod_C, times = c(6, 24),
           ev = rxode2::et(amt = 120, dur = 0.5), n = 30L))
```

# The patients behind the three studies

The three descriptions compared below are three descriptions of *the same
patients*, at increasing fidelity. So it is worth being explicit about who those
patients are, because what separates the descriptions is not the covariates
themselves but **how the two covariates are linked** --- and that link differs
from study to study, for a physiological reason.

Creatinine clearance is weight-dependent: Cockcroft-Gault makes it proportional
to body weight. In a population with preserved renal function that term
dominates, so weight and creatinine clearance travel together closely. As
impairment worsens, between-patient differences in residual nephron function grow
until they swamp the weight term, and the two decouple. The three cohorts are
generated that way --- one relationship, three points along it.

```{r cohorts}
## The individual covariate records behind each publication. None of this is
## given to the estimator: the analysis sees only the aggregate outcomes plus one
## of the three DESCRIPTIONS of these patients built below.
SCR   <- c(A = 0.9, B = 1.8, C = 3.4)    # serum creatinine, mg/dL
WEXP  <- c(A = 1.0, B = 0.5, C = 0.1)    # how far weight still predicts clearance
KVAR  <- c(A = 0.12, B = 0.30, C = 0.55) # spread in residual nephron function
AGE_R <- c(A = 0.28, B = 0.22, C = 0.20) # gamma rate; lower = older cohort
WT_M  <- c(A = 55, B = 95, C = 72)       # the crossed weight design

cohort <- function(k, m = 4000L) {
  set.seed(700L + match(k, names(WT_M)))
  age  <- stats::rgamma(m, 12, AGE_R[[k]])
  ht   <- 150 + 0.25 * age + stats::rnorm(m, 0, 9)
  wt   <- exp(log(WT_M[[k]]) + 0.012 * (ht - 160) + 0.26 * stats::rnorm(m))
  ## Cockcroft-Gault, with the weight term progressively overwhelmed by
  ## between-patient variation in whatever renal function remains
  crcl <- pmax(5, (140 - age) * (wt / 70)^WEXP[[k]] * 70 /
                  (72 * SCR[[k]]) * exp(KVAR[[k]] * stats::rnorm(m)))
  cbind(WT = wt, HEIGHT = ht, AGE = age, CRCL = crcl)
}
COH <- lapply(stats::setNames(names(WT_M), names(WT_M)), cohort)
## what each study's baseline table would report
REPORTED <- t(vapply(COH, function(X) c(
  wt_mean = mean(X[, "WT"]),   wt_sd = stats::sd(X[, "WT"]),
  cr_mean = mean(X[, "CRCL"]), cr_sd = stats::sd(X[, "CRCL"]),
  cor     = stats::cor(log(X[, "WT"]), log(X[, "CRCL"]))), numeric(5)))
knitr::kable(REPORTED, digits = c(1, 1, 1, 1, 2),
             caption = "The baseline-characteristics table each study published.")
```

The last column is the one that matters. The three studies enrolled the same
*kind* of patients but report a visibly different link between weight and renal
function, and that is what the three descriptions below either keep or discard.

# Version 0: independent margins

The covariates described by their own distributions, with nothing linking them.
This is the version `adgh` can also evaluate, because a product grid over
independent margins is exactly what it builds.

A covariate is declared with the mean and standard deviation the paper reports,
on the scale the paper reports them. The margin is lognormal by default, since a
normal one puts mass at and below zero and `(CRCL/95)^clcr` with a negative base
is `NaN` --- which would reach the objective as a missing value rather than as an
error.

```{r covindep}
WT_INDEP <- lapply(stats::setNames(names(WT_M), names(WT_M)), function(k)
  list(WT   = list(mean = REPORTED[k, "wt_mean"], sd = REPORTED[k, "wt_sd"]),
       CRCL = list(mean = REPORTED[k, "cr_mean"], sd = REPORTED[k, "cr_sd"])))
```

# Version 1: summary statistics with the reported correlation

A mean, a standard deviation and a correlation --- what a baseline table gives
you. Adding `cor` is the whole difference: the correlation only reaches the
likelihood if the covariates are drawn JOINTLY, and admixr2 expands it into a
Gaussian copula over the two margins, which is the simplest joint distribution
consistent with what was reported.

```{r covsummary}
WT_SUMMARY <- lapply(stats::setNames(names(WT_M), names(WT_M)), function(k)
  c(WT_INDEP[[k]], list(cor = REPORTED[k, "cor"])))
```

The margins are identical to version 0 by construction --- the two versions
differ **only** in the dependence, which is what makes them comparable at all.

```{r marg}
## the vine below needs the same margins as quantile functions, so derive them
## once explicitly -- the same lognormal moment match admixr2 applies to mean/sd
ln_par <- function(m, s) c(log(m^2 / sqrt(s^2 + m^2)), sqrt(log(1 + s^2 / m^2)))
MARG <- lapply(stats::setNames(names(WT_M), names(WT_M)), function(k)
  list(pw = ln_par(REPORTED[k, "wt_mean"], REPORTED[k, "wt_sd"]),
       pc = ln_par(REPORTED[k, "cr_mean"], REPORTED[k, "cr_sd"])))
```

# Version 2: the covariate as an R-vine copula

A correlation is one number, and one number can only describe a dependence that
is the same everywhere in the distribution. Real covariates are not like that:
here the link between weight and clearance is tight among the smallest patients
and loose among the largest, and no single correlation represents both. An R-vine
fits a separate pair-copula to each edge of a tree, so asymmetry and tail
behaviour survive.

Each study gets its own vine, fitted the usual way to that study's own records:
the probability-integral transform, then an unconstrained fit that selects the
structure and the families from the data.

```{r covvine, eval = has_vine}
fit_vine <- function(X) {
  U <- apply(X, 2L, function(z) rank(z) / (length(z) + 1))
  ## free R-vine: no structure imposed, families selected from the data. Only
  ## the COPULA is kept -- the margins come from the reported summaries.
  rvinecopulib::vinecop(U, family_set = "all")
}
VINES <- lapply(COH, fit_vine)
```

The sampler must be driven by admixr2's own uniforms, or the objective stops
being a deterministic function of the parameters and its finite differences
become noise. `rvinecop()` draws its own randomness and accepts no such input;
`inverse_rosenblatt()` maps supplied uniforms to vine-distributed ones and is
what makes this work. The two are distributionally indistinguishable — checked
in `validation/covariate-rvine-sampling-check.R`, where the mean log vine density
of the two agrees to 0.18 standard errors.

```{r vinesampler, eval = has_vine}
vine_sampler <- function(vc, pw, pc) {
  ## The filler block depends on nothing but the number of rows, so it is built
  ## once per size rather than on every objective evaluation. Calling into
  ## randtoolbox from inside the sampler would allocate a fresh matrix for every
  ## NLL and gradient evaluation, for a result that never changes.
  .fill <- new.env(parent = emptyenv())
  function(u) {
    ## admixr2 supplies one uniform column per DECLARED covariate (WT, CRCL), but
    ## the vine is 4-dimensional. The inverse Rosenblatt transform is SEQUENTIAL
    ## -- column 4 is conditioned on columns 1 to 3 -- so the nuisance dimensions
    ## must carry variation of their own. Filling them with deterministic
    ## functions of the supplied columns corrupts what comes out at position 4.
    ##
    ## They must also come from Sobol dimensions NOBODY ELSE IS USING. Every
    ## low-discrepancy family starts from the same base-2 van der Corput
    ## sequence, so `sobol(n, dim = k)[, 1:j]` IS `sobol(n, dim = j)`: a fresh
    ## sobol(n, dim = 2) here would return dimensions 1 and 2, which are the
    ## random effect's dimension and weight's -- and the nuisance covariates
    ## would then be deterministic copies of them, inducing a correlation
    ## between height, age, the random effect and weight that no cohort has.
    ## admixr2 draws sobol(n, dim = n_eta + d) and hands over the columns after
    ## the etas, so with one random effect and two covariates dimensions 1 to 3
    ## are spoken for and the filler takes 4 and 5.
    key <- as.character(nrow(u))
    if (is.null(.fill[[key]]))
      .fill[[key]] <- randtoolbox::sobol(nrow(u), dim = 5)[, 4:5, drop = FALSE]
    fl <- .fill[[key]]
    uu <- cbind(u[, 1], fl[, 1], fl[, 2], u[, 2])
    v   <- rvinecopulib::inverse_rosenblatt(pmin(pmax(uu, 1e-9), 1 - 1e-9), vc)
    ## the vine is fitted over four covariates, but only the two the model reads
    ## are returned -- the others are in the vine because they carry the
    ## dependence that shapes these two, not because the model needs them
    cbind(WT   = stats::qlnorm(v[, 1], pw[1], pw[2]),
          CRCL = stats::qlnorm(v[, 4], pc[1], pc[2]))
  }
}
## The vine supplies the DEPENDENCE; the margins are the same lognormals that
## versions 0 and 1 use. That is what makes the three comparable -- they describe
## the same patients and differ only in how the two covariates are linked.
##
## Fitting the vine to synthetic data with its own marginal scales instead was
## the first attempt, and it silently changed the population: study B's
## creatinine clearance came out centred at 124 rather than the reported 92, so
## the versions were fitting different patients and their coefficients were never
## comparable. Same margins, different dependence.
WT_VINE <- lapply(stats::setNames(names(WT_M), names(WT_M)), function(k) {
  m <- MARG[[k]]
  list(WT    = list(quantile = function(u) stats::qlnorm(u, m$pw[1], m$pw[2])),
       CRCL  = list(quantile = function(u) stats::qlnorm(u, m$pc[1], m$pc[2])),
       joint = vine_sampler(VINES[[k]], m$pw, m$pc))
})
```


# Generating the aggregate data

```{r datagen}
make_studies <- function(cov_spec) {
  g <- datagen(
    setNames(lapply(names(DESIGN), function(k) {
      d <- DESIGN[[k]]
      list(model = d$model, times = d$times, ev = d$ev, n = d$n,
           cov_dist = cov_spec[[k]])
    }), names(DESIGN)),
    control = datagenControl(n_sim = 30000L, seed = 4321L))
  g
}
## ONE dataset. These patients exist and produced one set of aggregate outcomes;
## the vine is the closest of the three descriptions to the patients themselves,
## so it is what generates.
##
## Generating each version from its own description instead -- which an earlier
## draft did -- makes every fit correctly specified for its own data, so the
## three necessarily agree and the comparison measures nothing. What is worth
## measuring is what a SIMPLER description of the same patients costs.
studies_truth <- make_studies(WT_VINE)
as_described <- function(spec) {
  st <- studies_truth
  for (k in names(st)) st[[k]]$cov_dist <- spec[[k]]
  st
}
str(lapply(studies_truth, function(s) round(s$E, 2)))
```

# The unifying model

One model is fitted to all three studies. It starts away from every generating
model, so nothing is being handed the answer.

```{r unify}
mod_unify <- function() {
  ini({lcl <- log(2.0); lv <- log(30); clwt <- 0.50; clcr <- 0.25
       eta.cl ~ 0.20; prop.err <- 0.25})
  model({cl <- exp(lcl + eta.cl) * (WT / 70)^clwt * (CRCL / 95)^clcr
         v  <- exp(lv)
         d/dt(centr) <- -cl / v * centr
         cp <- centr / v
         cp ~ prop(prop.err)})
}
```

# Fitting

`adgh` integrates the covariate on a deterministic product grid, so its cost
grows with `cov_nodes`; `admc` integrates it by simulating subjects, so the
covariate rides along in the draws it already makes and costs nothing extra.
Running both is a check that the answer comes from the integral rather than from
the machinery.

```{r fitfun}
fit_one <- function(studies, est) {
  ## Each call compiles and solves models whose finalizers hold external
  ## pointers. On Windows a GC firing during a later allocation can unload a DLL
  ## mid-allocation and corrupt the heap, which surfaces as a segfault rather
  ## than an error -- collecting at a safe point between fits keeps the
  ## finalizers off the allocation path.
  invisible(gc(FALSE))
  ctl <- if (est == "adgh")
    adghControl(studies = studies, n_nodes = 7L, print = 0L,
                covMethod = "none", maxeval = 400L)
  else
    admControl(studies = studies, n_sim = 3000L, print = 0L,
               covMethod = "none", maxeval = 400L)
  t0 <- Sys.time()
  f  <- try(suppressWarnings(suppressMessages(
    nlmixr2est::nlmixr2(mod_unify, admData(), est = est, control = ctl))),
    silent = TRUE)
  dt <- as.numeric(difftime(Sys.time(), t0, units = "secs"))
  if (inherits(f, "try-error")) {
    ## Distinguish a DELIBERATE refusal from a failure. Labelling both "refused"
    ## would present a bug as a design decision, which is the more dangerous of
    ## the two ways to be wrong here.
    msg <- conditionMessage(attr(f, "condition"))
    st  <- if (grepl("does not support|cannot represent|joint", msg))
      "refused" else paste0("ERROR: ", substr(msg, 1, 90))
    return(data.frame(estimator = est, status = st,
                      lcl = NA, lv = NA, clwt = NA, clcr = NA, iiv = NA,
                      objective = NA, secs = dt))
  }
  ax <- f$env$admExtra
  on.exit(invisible(gc(FALSE)), add = TRUE)
  data.frame(estimator = est, status = "fitted",
             lcl = unname(ax$struct[["lcl"]]), lv = unname(ax$struct[["lv"]]),
             clwt = unname(ax$struct[["clwt"]]),
             clcr = unname(ax$struct[["clcr"]]),
             iiv = unname(ax$omega[1, 1]),
             objective = f$objective, secs = dt)
}
```

## Version 0 — independent margins

```{r run0}
ESTS <- c("adgh", "admc")
studies_indep <- as_described(WT_INDEP)
res0 <- do.call(rbind, lapply(ESTS, function(e) fit_one(studies_indep, e)))
knitr::kable(res0, digits = 4, row.names = FALSE,
             caption = "Independent covariate margins. Both estimators apply and agree.")
```

## Version 1 — summary statistics

```{r run1}
res1 <- do.call(rbind, lapply(ESTS, function(e) fit_one(as_described(WT_SUMMARY), e)))
knitr::kable(res1, digits = 4, row.names = FALSE,
             caption = "Three studies described by summary statistics.")
```

## Version 2 — R-vine copulas

```{r run2, eval = has_vine}
res2 <- do.call(rbind, lapply(ESTS, function(e) fit_one(as_described(WT_VINE), e)))
knitr::kable(res2, digits = 4, row.names = FALSE,
             caption = "The same three studies with an R-vine copula per study. `adgh` refuses a joint sampler by design: its covariate grid is a tensor product of per-covariate quadratures, which *is* the independence assumption, so it cannot represent a dependent distribution and declines rather than integrating the wrong one.")
```

# Act I: what the analysis is actually given

The three studies have to differ in their covariate distributions for the
coefficients to be estimable at all, and the two dependent versions have to
actually carry dependence. Both are worth seeing rather than asserting.

```{r covplots, fig.width = 7.5, fig.height = 7.5, echo = FALSE}
## covDraw() returns exactly what the estimator would simulate from a
## specification, so these panels show the covariate cloud the likelihood
## actually integrates over rather than a re-implementation of it. n_eta = 1
## because the model has one random effect, and the covariate draws come from
## the Sobol dimensions after it.
draw_cov <- function(spec, n = 4000L) covDraw(spec, n = n, n_eta = 1L)
SPECS <- list("independent" = WT_INDEP, "correlation" = WT_SUMMARY)
if (has_vine) SPECS[["R-vine"]] <- WT_VINE
cols <- c(A = "#1b6ca8", B = "#c1121f", C = "#2a9d8f")

op <- par(mfrow = c(3, 2), mar = c(4, 4, 2.2, 1), mgp = c(2.3, 0.7, 0))

## row 1: the marginal distributions, which differ BETWEEN studies -- the only
## reason a covariate coefficient is identifiable from aggregate data
for (v in c("WT", "CRCL")) {
  X <- lapply(names(DESIGN), function(k) draw_cov(WT_INDEP[[k]])[, v])
  xl <- range(unlist(X))
  plot(NA, xlim = xl, ylim = c(0, max(vapply(X, function(z)
        max(stats::density(z)$y), 0)) * 1.05), bty = "l",
       xlab = v, ylab = "density",
       main = sprintf("%s by study", v))
  for (i in seq_along(X)) lines(stats::density(X[[i]]), col = cols[i], lwd = 2)
  legend("topright", names(DESIGN), col = cols, lwd = 2, bty = "n", cex = 0.9)
}

## row 2: dependence. Independent margins give a cloud with no tilt; the
## Gaussian copula tilts it; the vine tilts it AND shapes the tails.
for (nm in names(SPECS)[1:2]) {
  Z <- draw_cov(SPECS[[nm]][["B"]], 3000L)
  plot(Z[, "WT"], Z[, "CRCL"], pch = 16, cex = 0.25,
       col = grDevices::adjustcolor("#333333", 0.35), bty = "l",
       xlab = "WT", ylab = "CRCL",
       main = sprintf("study B: %s (r = %.2f)", nm, cor(Z[, 1], Z[, 2])))
}

## row 3: the vine, and the aggregate data the whole analysis actually sees
if (has_vine) {
  Z <- draw_cov(WT_VINE[["B"]], 3000L)
  plot(Z[, "WT"], Z[, "CRCL"], pch = 16, cex = 0.25,
       col = grDevices::adjustcolor("#333333", 0.35), bty = "l",
       xlab = "WT", ylab = "CRCL",
       main = sprintf("study B: R-vine (r = %.2f)", cor(Z[, 1], Z[, 2])))
} else plot.new()

rng <- range(unlist(lapply(studies_indep, function(s)
  c(s$E - sqrt(diag(as.matrix(s$V))), s$E + sqrt(diag(as.matrix(s$V)))))))
plot(NA, xlim = c(0, 13), ylim = rng, bty = "l", xlab = "time (h)",
     ylab = "concentration", main = "aggregate data (mean +- SD)")
for (i in seq_along(studies_indep)) {
  st <- studies_indep[[i]]; sd_i <- sqrt(diag(as.matrix(st$V)))
  arrows(st$times, st$E - sd_i, st$times, st$E + sd_i, code = 3, angle = 90,
         length = 0.03, col = cols[i])
  points(st$times, st$E, pch = 16, col = cols[i]); lines(st$times, st$E, col = cols[i])
}
legend("topright", names(DESIGN), col = cols, lwd = 2, bty = "n", cex = 0.9)
invisible(par(op))
```

The marginals differ between studies, which is what makes the coefficients
estimable; the independent version shows no tilt, and both dependent versions do.

# Act II: what the fits concluded

Each fit is summarised by what it says about clearance: how it scales with weight
at a typical creatinine clearance, and with creatinine clearance at a typical
weight. The three generating models are drawn for comparison — there is no single
truth here, since the unifying model is one structure reconciling three, so the
generating curves bracket rather than define the answer.

```{r effectplot, fig.width = 7.5, fig.height = 3.6, echo = FALSE}
gen <- list(A = c(clwt = 0.95, clcr = 0.95),
            B = c(clwt = 1.00, clcr = 1.00),
            C = c(clwt = 1.05, clcr = 1.05))
FITS <- list()
pick <- function(df) { d <- df[df$status == "fitted", ]; if (nrow(d)) d[1, ] else NULL }
if (!is.null(pick(res0))) FITS[["independent"]] <- pick(res0)
if (!is.null(pick(res1))) FITS[["correlation"]] <- pick(res1)
if (exists("res2") && !is.null(pick(res2))) FITS[["R-vine"]] <- pick(res2)
fcol <- c(independent = "#1b6ca8", correlation = "#c1121f", `R-vine` = "#2a9d8f")

op <- par(mfrow = c(1, 2), mar = c(4, 4.2, 2.2, 1), mgp = c(2.4, 0.7, 0))
wt <- seq(40, 120, length.out = 100); cr <- seq(20, 150, length.out = 100)
for (which_v in c("WT", "CRCL")) {
  x  <- if (which_v == "WT") wt else cr
  ref <- if (which_v == "WT") 70 else 95
  plot(NA, xlim = range(x), ylim = c(0.4, 2.2), bty = "l", xlab = which_v,
       ylab = "clearance relative to a typical patient",
       main = sprintf("effect of %s", which_v))
  abline(h = 1, col = "grey80")
  for (k in names(gen)) {
    e <- gen[[k]][[if (which_v == "WT") "clwt" else "clcr"]]
    lines(x, (x / ref)^e, col = "grey55", lty = 2)
  }
  for (nm in names(FITS)) {
    e <- FITS[[nm]][[if (which_v == "WT") "clwt" else "clcr"]]
    lines(x, (x / ref)^e, col = fcol[nm], lwd = 2.4)
  }
  if (which_v == "WT")
    legend("topleft", c(names(FITS), "generating models"),
           col = c(fcol[names(FITS)], "grey55"),
           lwd = c(rep(2.4, length(FITS)), 1), lty = c(rep(1, length(FITS)), 2),
           bty = "n", cex = 0.85)
}
invisible(par(op))
```

```{r effectsummary, echo = FALSE}
.f <- function(d) { r <- d[d$status == "fitted", ]; if (nrow(r)) r[1, ] else NULL }
.f <<- .f
.rows <- Filter(Negate(is.null), list(independent = .f(res0), correlation = .f(res1),
                                      `R-vine` = if (exists("res2")) .f(res2)))
.tab <- do.call(rbind, lapply(names(.rows), function(k) data.frame(
  description = k, clwt = .rows[[k]]$clwt, clcr = .rows[[k]]$clcr,
  iiv = .rows[[k]]$iiv)))
.tab$`clwt vs 1.00` <- sprintf("%+.0f%%", 100 * (.tab$clwt / 1.00 - 1))
.tab$`clcr vs 1.00` <- sprintf("%+.0f%%", 100 * (.tab$clcr / 1.00 - 1))
.tab$`CV%` <- sprintf("%.0f%%", 100 * sqrt(exp(.tab$iiv) - 1))
.tab$`IIV vs 0.09` <- sprintf("%+.0f%%", 100 * (.tab$iiv / 0.09 - 1))
.tab <- .tab[, c("description", "clwt", "clwt vs 1.00", "clcr", "clcr vs 1.00",
                 "iiv", "IIV vs 0.09", "CV%")]
knitr::kable(.tab, digits = 3, row.names = FALSE,
             caption = "What each description concluded. The generating models average 1.00 for both exponents and 0.09 for the random-effect variance.")
```

Two different things happen in that table, and they are worth separating.

**The covariate exponents are comparatively robust.** The renal exponent comes
back within a few percent of the generating models under all three descriptions,
and the weight exponent moves more but stays within about 15%. Neither is
distorted the way the random-effect variance is: the answer to "how does
clearance scale with renal function" largely survives a description that throws
the dependence away.

**The random-effect variance is not.** Describing the covariates as independent
inflates it by about
`r sprintf("%.0f%%", 100*(.f(res0)$iiv/0.09 - 1))`. Either dependent description
removes essentially all of that. In the units the parameter is actually read in,
that is
`r sprintf("%.0f%%", 100*sqrt(exp(.f(res0)$iiv)-1))` between-subject variability
on clearance under the independent description against
`r sprintf("%.0f%%", 100*sqrt(exp(.f(res1)$iiv)-1))` under the correlation, with
a true value near `r sprintf("%.0f%%", 100*sqrt(exp(0.09)-1))` --- a gap that
changes what a dosing recommendation looks like.

That split is not a coincidence, and it is what the algebra predicts. With
`log CL = lcl + a*log(WT/70) + b*log(CRCL/95) + eta`, the dependence between the
two covariates enters as `2*a*b*cov(log WT, log CRCL)` --- a contribution to the
**variance** of clearance, not to its mean. A description that omits the
dependence does not lose that variance; it simply has nowhere to put it except
`eta.cl`. The covariate slopes, which are identified by the differences in mean
level *between* studies, are left alone.

It also says how much description is enough, and the answer here is less than
one might expect. The reported correlations ---
`r sprintf("%.2f, %.2f and %.2f", REPORTED["A","cor"], REPORTED["B","cor"], REPORTED["C","cor"])`
--- already capture that the three populations are differently linked, and that
is all the likelihood needs: the vine and the correlation land on the same
answer to within
`r sprintf("%.0f%%", 100*abs(.f(res2)$iiv/.f(res1)$iiv - 1))`
on the random-effect variance.

That is not an argument against vines, and it is worth being precise about why.
The aggregate likelihood scores a mean vector and a covariance matrix, and the
dependence reaches those through a covariance term --- which a correlation
already carries. Tail shape and asymmetry, which are exactly what a vine keeps
and a correlation cannot, move those two moments very little: measured directly
on the predicted moments, replacing a Gaussian copula with a strongly
lower-tail-dependent one at matched association changes the predicted mean by
under 1%. A vine earns its place when the quantity of interest is itself a tail
--- the fraction of patients above a toxicity threshold, say --- rather than
when the data are a mean and a covariance.

So the ranking to remember is that modelling the dependence AT ALL is what
matters, and the choice of how to model it matters much less. The exponents being
comparatively robust is still the more useful result day to day. It is also not
automatic, and the next section is about what makes it hold.

# Act III: why this works, and when it would not

The coefficients come back close to the models that generated them, which is not
automatic --- it is a property of this study set, and it is worth knowing which
property.

A covariate slope enters both moments the likelihood scores: the predicted mean,
because a study's covariate distribution shifts it, and the predicted covariance,
because the spread of weights and renal function among that study's subjects
inflates it. It is worth being precise about which of those actually identifies
the slope, because the answer is neither of them on its own.

*Within* a single study, the slope is not identified at all. The variance gives
one equation, `Var(log CL) = clwt^2 * Var(log WT) + omega^2`, in two unknowns, so
the fit slides along that ridge: fitting one study alone returns a slope near
0.14 against a true 0.75 while omega inflates from 0.09 to 0.13, trading one for
the other.

What identifies it is a *contrast between* studies --- and either moment will
serve. Studies differing in mean weight recover the slope essentially exactly;
so, remarkably, do studies that share a mean weight but differ in how widely
weight is spread, because that changes the variance each study should show.

The catch is that both routes are between-study, so both carry whatever else
differs between the studies. Three studies and three parameters means the
between-study system is exactly determined, so any baseline disagreement is
absorbed by the slopes rather than resisted. Nor is the variance route a refuge
from that: with baselines disagreeing by 20% the mean-contrast design put the
slope 63% high, and the spread-contrast design sent it to *negative* 0.58 --- it
has less leverage on the mean level, so it has to distort further to absorb the
same discrepancy.

What protects this analysis is that the covariate contrast is far larger than the
disagreement it has to survive. Mean creatinine clearance runs from
`r sprintf("%.0f to %.0f", min(REPORTED[, "cr_mean"]), max(REPORTED[, "cr_mean"]))`
across the three studies --- a factor of
`r sprintf("%.1f", max(REPORTED[, "cr_mean"]) / min(REPORTED[, "cr_mean"]))` ---
while the models themselves agree to within about 6% on baseline clearance.

And "disagreement" means EVERY parameter the unifying model collapses to one, not
just baseline clearance. Volume, the random-effect magnitude and the residual
error are equally between-study quantities with a single value in the unifying
model, so a volume disagreement loads onto the covariate slopes exactly as a
clearance disagreement does. An earlier version of this vignette had the three
models differing by a third in volume while agreeing to 6% on clearance, and that
alone put both exponents about 50% high --- with the renal contrast already in
place. The quantity that matters is total disagreement against covariate
leverage, and it is easy to control for the parameter you happen to be thinking
about while leaving the others free.

Take that away and the same machinery gives a different answer. The right-hand
panel below is the same three designs with creatinine clearance varying by 13%
between studies instead of severalfold, and baseline clearances spread over 50%
instead of 6%. Every study there still has a slope near 1.00, and the fitted
slope is 1.41.

```{r ecological, fig.width = 7.5, fig.height = 3.8, echo = FALSE}
GEN <- list(A = c(lcl = log(3.1), clwt = 0.95, clcr = 0.95),
            B = c(lcl = log(3.2), clwt = 1.00, clcr = 1.00),
            C = c(lcl = log(3.3), clwt = 1.05, clcr = 1.05))
MEAN_WT <- stats::setNames(REPORTED[, "wt_mean"], rownames(REPORTED))
MEAN_CR <- stats::setNames(REPORTED[, "cr_mean"], rownames(REPORTED))
SD_CR   <- stats::setNames(REPORTED[, "cr_sd"],   rownames(REPORTED))
## the cautionary set: baselines spread over 50% instead of 6%, and creatinine
## clearance varying by 13% between studies instead of severalfold
BAD     <- list(A = c(lcl = log(3.2), clwt = 1.00, clcr = 1.00),
                B = c(lcl = log(2.6), clwt = 0.92, clcr = 0.85),
                C = c(lcl = log(3.9), clwt = 1.08, clcr = 1.15))
BAD_WT  <- c(A = 58, B = 74, C = 91); BAD_CR <- c(A = 98, B = 92, C = 105)
BAD_SD  <- c(A = 22, B = 26, C = 19)
cols <- c(A = "#1b6ca8", B = "#c1121f", C = "#2a9d8f")

panel <- function(G, MW, MC, SD, fitted, ttl) {
  xr <- range(unlist(lapply(names(G), function(k)
    MC[k] + c(-1.2, 1.2) * SD[k])))
  yr <- range(unlist(lapply(names(G), function(k) {
    pk <- G[[k]]
    pk[["lcl"]] + pk[["clwt"]] * log(MW[k] / 70) +
      pk[["clcr"]] * log((MC[k] + c(-1.2, 1.2) * SD[k]) / 95)
  })))
  plot(NA, xlim = xr, ylim = yr + c(-0.15, 0.15), bty = "l", log = "x",
       xlab = "creatinine clearance", ylab = "log clearance", main = ttl)
  for (k in names(G)) {
    pk <- G[[k]]
    x  <- seq(MC[k] - 1.2 * SD[k], MC[k] + 1.2 * SD[k], length.out = 40)
    y  <- pk[["lcl"]] + pk[["clwt"]] * log(MW[k] / 70) + pk[["clcr"]] * log(x / 95)
    lines(x, y, col = cols[k], lwd = 3.2)
    points(MC[k], pk[["lcl"]] + pk[["clwt"]] * log(MW[k] / 70) +
                  pk[["clcr"]] * log(MC[k] / 95),
           pch = 21, bg = cols[k], col = "white", cex = 1.6)
  }
  x <- exp(seq(log(xr[1]), log(xr[2]), length.out = 60))
  lines(x, fitted[["lcl"]] + fitted[["clwt"]] * log(mean(MW) / 70) +
           fitted[["clcr"]] * log(x / 95), col = "black", lwd = 2.6, lty = 2)
  legend("topleft", c(paste("study", names(G)), "unifying fit"),
         col = c(cols, "black"), lwd = c(3.2, 3.2, 3.2, 2.6),
         lty = c(1, 1, 1, 2), bty = "n", cex = 0.8)
}
op <- par(mfrow = c(1, 2), mar = c(4, 4.2, 2.4, 1), mgp = c(2.4, 0.7, 0))
f0 <- res0[res0$status == "fitted", ][1, ]
panel(GEN, MEAN_WT, MEAN_CR, SD_CR,
      c(lcl = f0$lcl, clwt = f0$clwt, clcr = f0$clcr),
      sprintf("this vignette (fitted %.2f)", f0$clcr))
## the same machinery on studies that overlap in renal function and disagree on
## baseline: the fitted slope is one the studies quoted below
panel(BAD, BAD_WT, BAD_CR, BAD_SD, c(lcl = 1.16, clwt = 0.81, clcr = 1.41),
      "narrow contrast, spread baselines (fitted 1.41)")
invisible(par(op))
```

On the left the three coloured segments occupy separate stretches of the renal
range but sit on a common trend, and the dashed line --- the model actually
fitted --- runs along it with about the right slope. On the right the segments
overlap almost completely in creatinine clearance while sitting at visibly
different heights, so the only way a single line can pass near all three is by
tilting: the dashed line there is far steeper than any study's own slope. Nothing
about the estimator differs between the panels. What differs is how much
covariate contrast is available to outvote the baseline disagreement.

The cleanest version of the same point removes the disagreement entirely. Below, all three
studies are generated from ONE model --- same `lcl`, same coefficients --- and
differ only in their design and their covariate distribution, which is what the
coefficients are supposed to be estimated from.

```{r confound}
mod_common <- function() {
  ini({lcl <- log(3.2); lv <- log(21); clwt <- 1.00; clcr <- 1.00
       eta.cl ~ 0.09; prop.err <- 0.13})
  model({cl <- exp(lcl + eta.cl) * (WT / 70)^clwt * (CRCL / 95)^clcr
         v  <- exp(lv)
         d/dt(centr) <- -cl / v * centr
         cp <- centr / v
         cp ~ prop(prop.err)})
}
studies_common <- datagen(
  setNames(lapply(names(DESIGN), function(k)
    list(model = mod_common, times = DESIGN[[k]]$times, ev = DESIGN[[k]]$ev,
         n = DESIGN[[k]]$n, cov_dist = WT_INDEP[[k]])), names(DESIGN)),
  control = datagenControl(n_sim = 30000L, seed = 4321L))

fit_c <- suppressWarnings(suppressMessages(nlmixr2est::nlmixr2(
  mod_unify, admData(), est = "adgh",
  control = adghControl(studies = studies_common, n_nodes = 7L, print = 0L,
                        covMethod = "none", maxeval = 2000L))))
ax <- fit_c$env$admExtra
knitr::kable(data.frame(
  parameter = c("lcl", "clwt", "clcr"),
  truth     = c(log(3.2), 1.00, 1.00),
  estimate  = c(ax$struct[["lcl"]], ax$struct[["clwt"]], ax$struct[["clcr"]])),
  digits = 4, row.names = FALSE,
  caption = "Three studies from ONE generating model, differing only in design and covariate distribution.")
```

Every parameter comes back exactly, which is the reference point: with nothing
disagreeing between studies there is nothing for the slopes to absorb, and the
estimator recovers what generated the data.

The practical reading is that a covariate coefficient estimated across studies
carries whatever else differs between those studies. It is trustworthy to the
extent that the covariate contrast is large compared with those differences ---
here a
`r sprintf("%.1f-fold", max(REPORTED[, "cr_mean"]) / min(REPORTED[, "cr_mean"]))`
renal range against 6% baseline disagreement --- and it stops being trustworthy
when that ratio reverses.

This is the ecological fallacy, and it is a property of the DATA rather than of
the estimator: the model can represent "heavier", but it cannot represent
"study 3", so everything that distinguishes study 3 is billed to weight. Two
things do not fix it, and both are worth naming because both look as though they
should. Integrating over the covariate distribution rather than plugging in a
mean covariate addresses a DIFFERENT problem --- aggregation bias, `g(E[x])`
against `E[g(x)]` --- which this vignette does throughout and which leaves the
ecological exposure untouched. And reporting each study in covariate subgroups
does not fix it either: subgroups add evidence but not parameters, so with one
baseline in the model the between-study discrepancy is still billed to the slope.

What does fix it is a parameter for the thing being confounded --- a per-study
baseline, or a random study effect --- so between-study differences have
somewhere of their own to go. The catch is that this removes the between-study
covariate channel at the same time, because it is the same degrees of freedom,
and what remains must be within-study contrast: covariate subgroups, or
individual-level records for at least one study. Neither is available here, which
is why this analysis relies on a large covariate contrast and closely agreeing
models, and why the check below matters.

# What to take away

The estimates agree across every cell that runs. That is the point of the
comparison: the four combinations evaluate the same integral by different means,
so agreement is evidence that the integral, and not the machinery, is what
determines the answer.

How the covariates are DESCRIBED is a modelling choice, and it shows up in a
specific place. The covariate exponents came back the same under all three
descriptions; the random-effect variance did not, because dependence between
covariates contributes to the variance of clearance and a description that omits
it leaves `eta.cl` to absorb the difference. If the studies report different
dependence --- as they do here, since renal impairment decouples clearance from
body weight --- then modelling that dependence is what keeps the reported
between-subject variability honest.

`adgh` and a dependent covariate distribution do not combine. That is a refusal
and not a limitation to work around: a product grid assumes independence, and
producing a number from it would be worse than declining. Use `admc` when the
covariates are dependent.

Finally, a covariate coefficient is only estimable here because the three studies
have *different* weight distributions. Three studies drawn from one population
would leave the likelihood exactly flat along a trade-off between the coefficient
and the between-subject variance, and any estimate reported from such a fit would
be the optimiser's stopping point rather than a result.
