---
title: "Covariates in aggregate data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covariates in aggregate data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.2
)
source(Filter(file.exists, c("_common.R", "../_common.R", "vignettes/_common.R"))[1])
```

## A dose for a patient nobody studied

A woman is admitted with a creatinine clearance of 21 mL/min. The drug she needs
is renally cleared, and the label gives one dose --- 200 mg, the dose every
trial used.

Three trials have been published: one in normal renal function, one in mild
impairment, one in moderate. **None enrolled anyone like her.** And none of the
three reported a renal effect at all, because within each trial creatinine
clearance barely varied, so each analyst sensibly left it out.

You have three population models that disagree about clearance by nearly a
factor of two, and not one of them can say why. This vignette recovers the renal
effect none of them contains --- from the three papers alone --- and turns it
into a dose.

```{r libs}
library(admixr2); library(rxode2); library(nlmixr2); library(ggplot2)
```

## The world those trials sampled

To check the answer we need a truth to check it against, so the patients here
are simulated. Weight and creatinine clearance are correlated, so they are drawn
jointly through a Gaussian copula: each covariate keeps its own lognormal
margin, the dependence goes on the latent scale.

The truth, unknown to all three analysts, is CL = 5 L/h and V = 50 L at 70 kg,
allometric weight scaling, **a renal exponent of 0.6**, and men clearing about
20% faster. The three trials differ only in which slice of renal function they
sampled --- and that alone is enough to make their answers disagree.

```{r sim}
set.seed(11)
TIMES <- c(0.5, 1, 2, 4, 8, 12, 24); DOSE <- 200
CL70 <- 5; V70 <- 50; BCRCL <- 0.6; BSEX <- 0.18
OM_CL <- 0.05; SD_ADD <- 0.08
R <- matrix(c(1, .45, .45, 1), 2, 2,
            dimnames = list(c("WT", "CRCL"), c("WT", "CRCL")))

# A cohort of virtual patients: lognormal margins for weight and renal
# function, joined by the Gaussian copula above; sex drawn independently.
# Used again later for the severe cohort, so it takes its medians as arguments.
draw_cohort <- function(n, wt_median, wt_cv, crcl_median, crcl_cv,
                        p_male = 0.55) {

  # a Gaussian copula, one step at a time: correlated normals, turned into
  # correlated uniforms, then pushed through each covariate's own margin
  z          <- matrix(rnorm(2 * n), nrow = n, ncol = 2) %*% chol(R)
  u_wt       <- pnorm(z[, 1])
  u_crcl     <- pnorm(z[, 2])

  data.frame(
    WT   = qlnorm(u_wt,   meanlog = log(wt_median),   sdlog = wt_cv),
    CRCL = qlnorm(u_crcl, meanlog = log(crcl_median), sdlog = crcl_cv),
    SEX  = rbinom(n, size = 1L, prob = p_male))
}

# Concentration after an IV bolus into one compartment.
conc <- function(t, dose, cl, v) dose/v * exp(-cl/v * t)

# Each patient's true clearance. `bcrcl = 0` gives a model with no renal term,
# which is what all three published models will turn out to be.
patient_cl <- function(cohort, cl_70kg, bcrcl, bsex, omega) {
  cl_70kg * (cohort$WT/70)^0.75 * (cohort$CRCL/90)^bcrcl *
    exp(bsex * cohort$SEX) *
    exp(rnorm(nrow(cohort), 0, sqrt(omega)))
}

# One trial's worth of individual data, in nlmixr2 layout.
run_trial <- function(cohort, first_id) {
  cl <- patient_cl(cohort, CL70, BCRCL, BSEX, OM_CL)
  v  <- V70 * (cohort$WT/70)
  records <- list()
  for (i in seq_len(nrow(cohort))) {
    y <- conc(TIMES, DOSE, cl[i], v[i]) + rnorm(length(TIMES), 0, SD_ADD)
    records[[i]] <- data.frame(
      ID   = first_id + i,
      TIME = c(0, TIMES),                        # dose row, then observations
      DV   = c(NA_real_, y),
      AMT  = c(DOSE, rep(0, length(TIMES))),
      EVID = c(1L, rep(0L, length(TIMES))),
      CMT  = 1L,
      WT   = cohort$WT[i],
      CRCL = cohort$CRCL[i],
      SEX  = cohort$SEX[i])
  }
  do.call(rbind, records)
}

cohorts <- list(normal   = draw_cohort(260L, 78, .198, 95, .217),
                mild     = draw_cohort(210L, 76, .198, 62, .198),
                moderate = draw_cohort(180L, 74, .198, 38, .198))

trial_data <- list(
  normal   = run_trial(cohorts$normal,   first_id = 10000),
  mild     = run_trial(cohorts$mild,     first_id = 20000),
  moderate = run_trial(cohorts$moderate, first_id = 30000))
```

Here is where the information is lost. A published summary keeps the mean and
the spread and throws the patients away, so an orderly, explainable difference
between individuals survives only as extra width in a grey ribbon. Poor clearers
at the top, good clearers at the bottom, and nothing in what gets printed to say
which is which:

```{r hidden, fig.height = 3.8}
d1  <- trial_data$normal[trial_data$normal$EVID == 0, ]
agg <- data.frame(t = TIMES,
                  m = tapply(d1$DV, d1$TIME, mean),
                  s = tapply(d1$DV, d1$TIME, sd),
                  panel = "What the paper reports")
ind <- d1[d1$ID %in% head(unique(d1$ID), 150), ]
ind$panel <- "Individual patients (never published)"

ggplot() +
  geom_line(data = ind, aes(TIME, DV, group = ID, colour = CRCL),
            alpha = .35, linewidth = .3) +
  geom_ribbon(data = agg, aes(t, ymin = m - s, ymax = m + s),
              fill = "grey55", alpha = .35) +
  geom_line(data = agg, aes(t, m), linewidth = 1) +
  scale_colour_viridis_c(name = "CRCL (mL/min)", option = "C", end = .9) +
  facet_wrap(~panel) + labs(x = "time (h)", y = "concentration (mg/L)") +
  theme_bw() + theme(legend.position = "bottom")
```

Each analyst now fits their own trial, and --- as in practice --- they do not
write the same model. Sato fixes the allometric exponents at the conventional
0.75 and 1; Ito estimates both of them; Khan looks for a sex effect on volume as
well as on clearance. Five, seven and six parameters, with different names. None
of that has to be reconciled: a study carries its own model, and admixr2 asks
for no agreement between sources.

What they *do* agree on is which covariates they can see. **Sex they can see**
--- both sexes are enrolled, so it varies inside their own data. **Renal
function they cannot** --- within one cohort it barely moves, and what little
movement there is cannot be told apart from ordinary between-patient
variability.

```{r trialfits, results = "hide"}
# THREE ANALYSTS, THREE MODELS. They share a drug and nothing else: each wrote
# their own model, with its own parameters, and admixr2 needs no agreement
# between them. What differs here is the covariate structure -- the usual place
# published models diverge -- but the residual and random-effect structure may
# differ just as freely.

# Sato: the conventional write-up. Allometry fixed at the textbook exponents,
# one sex effect on clearance.
sato_model <- function() {
  ini({
    tcl     <- log(4.5)
    tv      <- log(45)
    bsex    <- 0.05
    add.err <- 0.2
    eta.cl  ~ 0.1
  })
  model({
    cl <- exp(tcl + eta.cl) * (WT/70)^0.75 * exp(bsex * SEX)
    v  <- exp(tv) * (WT/70)
    cp <- linCmt()
    cp ~ add(add.err)
  })
}

# Ito: estimated BOTH allometric exponents rather than fixing them. Two extra
# parameters, and a covariance matrix two rows wider than Sato's.
ito_model <- function() {
  ini({
    tcl     <- log(4.5)
    tv      <- log(45)
    bwtcl   <- 0.75
    bwtv    <- 1
    bsex    <- 0.05
    add.err <- 0.2
    eta.cl  ~ 0.1
  })
  model({
    cl <- exp(tcl + eta.cl) * (WT/70)^bwtcl * exp(bsex * SEX)
    v  <- exp(tv) * (WT/70)^bwtv
    cp <- linCmt()
    cp ~ add(add.err)
  })
}

# Khan: looked for a sex effect on volume as well as on clearance.
khan_model <- function() {
  ini({
    tcl     <- log(4.5)
    tv      <- log(45)
    bsex    <- 0.05
    bsexv   <- 0
    add.err <- 0.2
    eta.cl  ~ 0.1
  })
  model({
    cl <- exp(tcl + eta.cl) * (WT/70)^0.75 * exp(bsex * SEX)
    v  <- exp(tv) * (WT/70) * exp(bsexv * SEX)
    cp <- linCmt()
    cp ~ add(add.err)
  })
}

.ctl <- foceiControl(print = 0L, covMethod = "r", covFull = TRUE)
fit_normal   <- nlmixr2(sato_model, trial_data$normal,   est = "focei",
                        control = .ctl)
fit_moderate <- nlmixr2(ito_model,  trial_data$moderate, est = "focei",
                        control = .ctl)
fit_mild     <- nlmixr2(khan_model, trial_data$mild,     est = "focei",
                        control = .ctl)
```

Each analyst now writes up their trial. What goes in the paper is the estimates
and --- this is the part that matters later --- the **parameter covariance from
their own fit**, which is what makes a standard error meaningful downstream.

```{r published}
# `covFull = TRUE` above is the part that matters downstream. A source's
# covariance has to span EVERY parameter it estimated -- residual error and IIV
# included -- because a missing row reads as a parameter the source knew
# exactly. admixr2 refuses the sandwich for the WHOLE fit rather than report
# intervals that are too narrow, and says which row is missing when you build
# the study.
#
# A source's matrix does not always cover the IIV -- a paper printing a %RSE
# column rather than a covariance never does, and `covFull` itself returns the
# fixed effects only for some fits. Fill that row from the %RSE reported for
# it. A no-op when the matrix is already complete.
with_om_row <- function(f, rse_om = 0.30, eta = "om.eta.cl") {
  C <- f$cov
  if (eta %in% rownames(C)) return(C)
  nm  <- c(rownames(C), eta)
  out <- matrix(0, length(nm), length(nm), dimnames = list(nm, nm))
  out[rownames(C), rownames(C)] <- C
  out[eta, eta] <- (rse_om * as.numeric(f$omega[1, 1]))^2
  out
}
normal_paper <- list(
  CL   = exp(fit_normal$parFixedDf["tcl", "Estimate"]),
  V    = exp(fit_normal$parFixedDf["tv",  "Estimate"]),
  BSEX = fit_normal$parFixedDf["bsex",    "Estimate"],
  OM   = as.numeric(fit_normal$omega[1, 1]),
  COV  = with_om_row(fit_normal))

mild_paper <- list(
  CL   = exp(fit_mild$parFixedDf["tcl", "Estimate"]),
  V    = exp(fit_mild$parFixedDf["tv",  "Estimate"]),
  BSEX = fit_mild$parFixedDf["bsex",    "Estimate"],
  OM   = as.numeric(fit_mild$omega[1, 1]),
  COV  = with_om_row(fit_mild))

moderate_paper <- list(
  CL   = exp(fit_moderate$parFixedDf["tcl", "Estimate"]),
  V    = exp(fit_moderate$parFixedDf["tv",  "Estimate"]),
  BSEX = fit_moderate$parFixedDf["bsex",    "Estimate"],
  OM   = as.numeric(fit_moderate$omega[1, 1]),
  COV  = with_om_row(fit_moderate))

knitr::kable(data.frame(
  cohort      = c("normal", "mild", "moderate"),
  median_CRCL = round(c(median(cohorts$normal$CRCL),
                        median(cohorts$mild$CRCL),
                        median(cohorts$moderate$CRCL))),
  CL_at_70kg  = c(normal_paper$CL,   mild_paper$CL,   moderate_paper$CL),
  V_at_70kg   = c(normal_paper$V,    mild_paper$V,    moderate_paper$V),
  sex_effect  = c(normal_paper$BSEX, mild_paper$BSEX, moderate_paper$BSEX)),
  digits = 3, row.names = FALSE,
  caption = "Three independent analyses, three different models, each sound on its own data.")
```

Read that table the way the three analysts would have. All three agree on volume
and on the sex effect --- a covariate that varies *inside* every trial, so every
trial could see it. On clearance they disagree by nearly a factor of two, and
none of them can explain it, because the explanation is a covariate that barely
moves within any one of them.

That asymmetry is the entire problem. **A covariate that varies within a study is
estimable there. One that varies only between studies is invisible to each study
separately, and visible only to a meta-analysis.**

## Giving admixr2 the papers, not the patients

Nobody is going to send you the individual data. What you have is what was
printed: each trial's model, and the baseline table describing who was enrolled.
That is enough --- `datagen()` turns a published model plus its population into
the `(E, V, n)` a digitised figure would have given you.

Each paper becomes one `admStudy()`: the model it published, the covariance it
reported, the design, and the population it enrolled. Hand `population` the
cohort itself and the baseline table is read off it --- mean and SD per
continuous covariate, the sex split, the correlation between weight and renal
function, and `n` from the row count. That correlation is taken on the LATENT
scale, so on the *logs* for a lognormal margin, which is the step easiest to get
wrong writing a table out by hand. With a real paper you have a printed Table 1
rather than a cohort, and you write that out instead --- see the last section.

The population then enters the fit one of three ways, and which one is not a
style choice: it follows from what the paper's own model estimated.

* **Marginalised** --- the estimator integrates the prediction over the
  distribution. This is what a pooled result requires: one published curve
  standing for a whole population of patients.
* **Banded** (`stratify`) --- the source is split into strata with the covariate
  conditioned in each, so a covariate it *fitted* contributes a contrast rather
  than one pooled number.
* **Conditioned** (`at`, `by`) --- the model is solved at one value. This is what
  a result reported *by subgroup* allows.

**All three analysts fitted a sex effect**, so all three band on `SEX`: each
becomes two studies with sex conditioned, `n` divided between them by the level
probabilities, and both halves still marked as ONE source so the pair is not
counted as two papers. **Nobody fitted renal function**, so `CRCL` is
marginalised everywhere --- there is no contrast inside any single trial to
report, and splitting a source on a covariate its model never saw would
manufacture one. Weight is read by every model but at a *fixed* exponent, so
there is no fitted weight effect to recover either, and it is marginalised
alongside `CRCL`.

```{r datagen}
# --- each paper's model ----------------------------------------------------
# The source model is that analyst's own model AT their own published values,
# and a fit already is exactly that: `fit$ui` carries the model together with
# the estimates it converged on. Three different models go in, and nothing has
# to be shared between them.
#
# With a real paper you have no fit object. You write the model out yourself and
# hand admStudy() the printed numbers -- `est = c(tcl = log(5.2), ...)` with
# `rse = c(tcl = 4.1, ...)` -- which is the same study, transcribed instead of
# recovered.
sato_published <- fit_normal$ui        # normal renal function
ito_published  <- fit_moderate$ui      # moderate impairment
khan_published <- fit_mild$ui          # mild impairment

# --- each paper's study ----------------------------------------------------
normal_study <- admStudy(
  model      = sato_published,
  cov        = normal_paper$COV,
  population = cohorts$normal,
  dose       = DOSE,
  times      = TIMES,
  stratify   = "SEX")

moderate_study <- admStudy(
  model      = ito_published,
  cov        = moderate_paper$COV,
  population = cohorts$moderate,
  dose       = DOSE,
  times      = TIMES,
  stratify   = "SEX")

mild_study <- admStudy(
  model      = khan_published,
  cov        = mild_paper$COV,
  population = cohorts$mild,
  dose       = DOSE,
  times      = TIMES,
  stratify   = "SEX")

studies <- admStudies(normal   = normal_study,
                      moderate = moderate_study,
                      mild     = mild_study)

# the three papers together, used by the figures further down
published <- list(normal = normal_paper, mild = mild_paper,
                  moderate = moderate_paper)

studies
```

Each study carries the trial's **model**, the **covariance that trial
reported**, and the **population it enrolled** --- nothing that was not in the
paper. Nothing is solved yet: a study is a transcription, so it is cheap to
build and `print()` one to read it back against the source before any fitting
happens. The banding on sex is applied when the studies reach the fit, which is
why three papers still print as three sources here.

Note what is *not* in there. None of the published models has a renal term; the
`population` does. We are telling admixr2 who was enrolled, and letting it work
out what that implies about a covariate no analyst ever estimated.

**Why the sex effect has to be stratified and the renal effect must not be.**
A covariate that every study marginalises is not identified. Its effect then
enters only through the mixture it induces --- the level probabilities shift
`E`, the spread between levels adds to `V` --- which is exactly what a random
effect on the same parameter does, so the two trade off freely. With a similar
sex split in every trial there is no between-study contrast to separate them
either: the profile goes nearly flat, and a deterministic optimiser stops in
the same place every run, which reads as convergence.

Stratifying a source that *did* fit the covariate recovers its evidence,
because within one of its strata sex is a known constant rather than something
`omega` can absorb. Doing the reverse --- splitting a source on a covariate its
model never saw --- is the error in the other direction: it manufactures a
contrast the paper never reported, and attenuates the coefficient. Hence sex
stratified everywhere and renal function nowhere.

The renal effect is identified by a completely different route: no trial could
condition on it, so it comes purely from the contrast *between* the three
cohorts. Two covariates, two mechanisms, one fit.

## Pooling three models that disagree

```{r admfit}
adm_model <- function() {
  ini({
    tcl     <- log(4)
    tv      <- log(45)
    bcrcl   <- 0.3            # the renal effect no published model has
    bsex    <- 0.05
    add.err <- 0.1
    eta.cl  ~ 0.1
  })
  model({
    cl <- exp(tcl + eta.cl) * (WT/70)^0.75 * (CRCL/90)^bcrcl * exp(bsex * SEX)
    v  <- exp(tv) * (WT/70)
    cp <- linCmt()
    cp ~ add(add.err)
  })
}
fit <- nlmixr2(adm_model, admData(), est = "adgh",
               control = adghControl(studies = studies, print = 0L))
pf     <- fit$parFixedDf
om     <- as.numeric(fit$omega[1L, 1L])
om_se  <- sqrt(diag(fit$cov))[["om.eta.cl"]]   # omega's SE is in $cov, not pf
knitr::kable(data.frame(
  parameter = c("CL at 70 kg, CRCL 90, women (L/h)", "V at 70 kg (L)",
                "renal exponent", "sex effect (log)", "additive SD (mg/L)",
                "omega (eta.cl)"),
  truth     = c(CL70, V70, BCRCL, BSEX, SD_ADD, OM_CL),
  estimate  = c(pf["tcl", "Back-transformed"], pf["tv", "Back-transformed"],
                pf["bcrcl", "Back-transformed"], pf["bsex", "Back-transformed"],
                pf["add.err", "Back-transformed"],
                as.numeric(fit$omega[1L, 1L])),
  # CIs, not SEs: a structural theta's SE is on the LOG scale, so printing it
  # beside a back-transformed estimate invites reading 0.022 as 0.022 L/h
  lower     = c(pf["tcl", "CI Lower"], pf["tv", "CI Lower"],
                pf["bcrcl", "CI Lower"], pf["bsex", "CI Lower"],
                pf["add.err", "CI Lower"], om - 1.96 * om_se),
  upper     = c(pf["tcl", "CI Upper"], pf["tv", "CI Upper"],
                pf["bcrcl", "CI Upper"], pf["bsex", "CI Upper"],
                pf["add.err", "CI Upper"], om + 1.96 * om_se)),
  digits = 3, row.names = FALSE,
  caption = "Recovered from three analyses, none of which contained a renal term.")
```

Both covariate effects come back --- out of three analyses that between them
contained only one of them.

The sex effect is worth a pause, because it shows what pooling costs. The three
trials reported `r paste(sprintf("%.2f", sapply(published, function(p) p$BSEX)), collapse = ", ")` for a quantity that is genuinely
identical in all three; sampling noise alone put them that far apart. The
meta-analysis has to settle on one number, and **reconciling disagreement is not
free** --- some of it lands in the other parameters, which is why the renal
exponent comes back a little below its truth and the prediction further down is a
few per cent off rather than exact. That is what a meta-analysis is: evidence
that does not perfectly agree, pooled anyway, and honest about the residual.

## How far can you trust that number?

A study generated from a published **model** is not a sample. Its `E` and `V` are
exact functions of that model's parameters, so the only uncertain thing in the
whole chain is the estimate the original analyst published --- and the interval
on our answer has to come from theirs.

That is why each study above carries `model_cov`, the source fit's own
covariance. Hand it the fit's `$cov` as it comes; the `om.eta.cl` naming is
understood.

Without it, `n` gets read as a sample size instead, and the reported SE then
falls as `1/sqrt(n)` --- exactly proportional, measured to four significant
figures. It is a factor you choose by typing a number, not a property of the
evidence: the `n` that would reproduce one parameter's true uncertainty differs
by more than eightfold from the `n` that reproduces another's, in the same
model. No single value can be right for both.

`n` still matters, but only for **relative weight** between sources. Set it to
the sample size each source was developed on.

Two conventions worth knowing, both easy to get wrong. A structural theta's SE
is on the **log** scale --- printing it beside a back-transformed estimate would
read as L/h and be off by an order of magnitude --- so `parFixedDf` gives
back-transformed CIs, used above. And the **omega** SE is not in `parFixedDf` at
all: by nlmixr2 convention it lives in `fit$cov`, as `om.eta.cl`.

### Why the covariance is a sandwich here

Nothing above asked for one. Because a study here contributes as a published
**model** carrying its own reported uncertainty, `covMethod` defaults to
`"r,s"` rather than `"r"`, and says so when it does. Passing `covMethod = "r"`
explicitly still gets you the naive form.

That is not a stylistic default. A model source is not a sample of `n`
patients: its mean and covariance are exact functions of somebody else's
published estimate, so the only random object in the chain is that estimate and
the uncertainty it carries is the source's own `C_src`. Only the sandwich
carries that term. Under `"r"` the source is read as if `n` patients had been
sampled and the reported standard error falls as `1/sqrt(n)` --- it reached
0.08000 at every `n` under `"r,s"`, against a declared 0.08.

Every source in this vignette marginalises over the covariates it did not
condition on. That is the honest description of what these publications report
--- and it is also the one configuration where the default covariance is not
valid. Over 720 replicates, with no source banded on the covariate its own
model fitted: coverage **0.817** against a nominal 0.950, and a
likelihood-ratio test sized **0.125** against a nominal 0.05. A biased point
inside an interval that is too narrow is the combination that invalidates
inference, so admixr2 warns rather than letting it pass.

Banding fixes it, and **one banded source is as good as three**: coverage 0.933
with one, 0.925 with all of them, and test sizes 0.083 and 0.058 against the
0.125 above. `stratify = TRUE` derives the banding from the source's own model,
so it needs nothing restated.

### Is the renal effect real?

The interval says the renal exponent is not zero, but the honest test is a
likelihood-ratio test against the model without it. Fix the coefficient rather
than deleting the term --- the `population` declares `CRCL`, and admixr2 refuses a
distribution naming a covariate the model never reads.

```{r lrt}
adm_model_null <- function() {
  ini({
    tcl     <- log(4)
    tv      <- log(45)
    bcrcl   <- fix(0)          # the restriction being tested
    bsex    <- 0.05
    add.err <- 0.1
    eta.cl  ~ 0.1
  })
  model({
    cl <- exp(tcl + eta.cl) * (WT/70)^0.75 * (CRCL/90)^bcrcl * exp(bsex * SEX)
    v  <- exp(tv) * (WT/70)
    cp <- linCmt()
    cp ~ add(add.err)
  })
}
fit_null <- nlmixr2(adm_model_null, admData(), est = "adgh",
                    control = adghControl(studies = studies, print = 0L))
anova(fit, fit_null)
```

Read the `p` column, never `p.naive`, and **never AIC here**. Both of those
assume the information equality `H = J`, which holds only if the objective is
the true likelihood of the data. It is not: a model source's summary is a
deterministic function of somebody else's published estimate, so the score's
variance `J` is that source's `C_src` propagated through this fit, and has no
reason to match `H`.

The `p_eff` column is the size of the gap. It is `tr(H^-1 J)/2`, the objective's
actual optimism, which equals the parameter count only under that equality.
Where it exceeds `Npar`, AIC is charging too little per parameter and will keep
covariates it should drop. Measured on a design like this one, over 200
replicates with a covariate whose true effect was zero: AIC kept it **84%** of
the time, TIC **14.5%** against its own nominal 15.7%, and the corrected `p`
rejected at **7.5%**. The uncorrected `dOFV` kept it 75.5% of the time.

So: **`p`, or TIC. Not AIC, not raw dOFV.** `anova()` prints a note when the
gap is large enough to matter.

The `Weight` column is the factor `dOFV` was divided by, and it is worth
reading. It is *exactly* proportional to the uncertainty the sources declared
and to the weight they carry (`n`), and it does not touch the estimates at all
--- scaling every `model_cov` by 64 leaves the fitted `bcrcl` bit-identical
while moving `Weight` by exactly 64. So it is the share of this fit's apparent
information that is really the source analysts' uncertainty rather than
evidence. A `Weight` near 1 means the sources are precise enough to behave like
data; well below 1 means the uncorrected test would have been conservative.

## Seeing the renal effect

Here is what the meta-analysis actually did:

```{r mechanism, fig.height = 4.2}
e  <- setNames(pf$Estimate, rownames(pf))
crcl_grid <- seq(15, 115, length.out = 120)
sexlab <- c("women", "men")

# the meta-analysis: one curve per sex. The renal effect is the SLOPE, the sex
# effect is the vertical GAP between the two.
curve_df <- data.frame()
for (sx in 0:1) {
  cl_curve <- exp(e[["tcl"]]) * (crcl_grid/90)^e[["bcrcl"]] *
              exp(e[["bsex"]] * sx)
  curve_df <- rbind(curve_df, data.frame(
    CRCL = crcl_grid, sex = sexlab[sx + 1L], CL = cl_curve))
}

# each published model: no renal term, so a FLAT bar per sex, drawn over the
# CRCL range that trial actually covered
seg <- data.frame()
pts <- data.frame()
for (trial in names(published)) {
  paper     <- published[[trial]]
  crcl      <- cohorts[[trial]]$CRCL
  crcl_p10  <- unname(quantile(crcl, 0.10))
  crcl_p90  <- unname(quantile(crcl, 0.90))
  for (sx in 0:1) {
    cl_here <- paper$CL * exp(paper$BSEX * sx)
    seg <- rbind(seg, data.frame(
      lab = trial, sex = sexlab[sx + 1L],
      CL = cl_here, lo = crcl_p10, hi = crcl_p90))
    pts <- rbind(pts, data.frame(
      lab = trial, sex = sexlab[sx + 1L],
      CRCL = median(crcl), CL = cl_here))
  }
}

ggplot(curve_df, aes(CRCL, CL, colour = sex)) +
  annotate("rect", xmin = 15, xmax = 30, ymin = -Inf, ymax = Inf,
           fill = "grey85", alpha = .6) +
  annotate("text", x = 22.5, y = max(curve_df$CL), label = "severe\n(no trial)",
           size = 3, vjust = 1.1) +
  geom_segment(data = seg, aes(x = lo, xend = hi, y = CL, yend = CL),
               linewidth = 1.5, alpha = .55) +
  geom_line(linewidth = 1) +
  geom_point(data = pts, size = 2.6) +
  geom_text(data = subset(pts, sex == "men"), aes(label = lab),
            vjust = -1.2, size = 3.1, show.legend = FALSE) +
  scale_colour_manual(values = c(women = "steelblue4", men = "darkorange3"),
                      name = NULL) +
  labs(x = "median creatinine clearance (mL/min)",
       y = "clearance at 70 kg (L/h)",
       subtitle = paste("bars: what each published model claims, over the range",
                        "it covered\nlines: the meta-analysis")) +
  theme_bw() + theme(legend.position = "bottom")
```

Both effects are in that picture, and they look nothing alike.

The **sex effect** is the vertical gap between the two lines. Every trial found
it, so every flat bar comes in a matching pair with the same gap --- a covariate
that varies within a study is visible to that study.

The **renal effect** is the slope, and no trial has one. Each is a pair of flat
bars, claiming clearance is whatever it happened to measure, at any renal
function. Three pairs at three different heights: each correct over its own range
and contradicting the others everywhere else. The meta-analysis draws the line
through them. That line is the renal effect --- recovered entirely from *between*
the trials, and carried into the shaded region none of them sampled.

## Back to the patient

Her creatinine clearance is 21 mL/min, outside every trial's range. Without a
meta-analysis you would reach for the published model closest to her --- the
moderate-impairment one --- and use it.

```{r predict}
set.seed(3)
# mean concentration over a severe cohort (WT 72 kg, CRCL 21 mL/min), reusing
# the same two helpers the trials were simulated with
severe_mean_conc <- function(cl_70kg, v_70kg, bcrcl, bsex, omega, n = 40000L) {
  pts <- draw_cohort(n, 72, .198, 21, .217)
  cl  <- patient_cl(pts, cl_70kg, bcrcl, bsex, omega)
  v   <- v_70kg * (pts$WT/70)
  sapply(TIMES, function(t) mean(conc(t, DOSE, cl, v)))
}

p_adm   <- severe_mean_conc(exp(e[["tcl"]]), exp(e[["tv"]]), e[["bcrcl"]],
                            e[["bsex"]], as.numeric(fit$omega[1L, 1L]))
p_near  <- severe_mean_conc(moderate_paper$CL, moderate_paper$V,
                            0,                       # no renal term
                            moderate_paper$BSEX, moderate_paper$OM)
p_truth <- severe_mean_conc(CL70, V70, BCRCL, BSEX, OM_CL)
```

Predicting the *mean* is only half of it. Simulate a severe cohort under each
model and compare the whole distribution — median and 90% interval — against
what those patients would really do. The truth is repeated in grey behind every
panel.

```{r vpc, fig.height = 5.0}
set.seed(7)
# one severe cohort, and every model asked to predict THOSE patients
severe <- draw_cohort(800L, 72, .198, 21, .217)
tt     <- seq(0.25, 24, length.out = 40)

band <- function(cl_70kg, v_70kg, bcrcl, bsex, omega, label) {
  cl <- patient_cl(severe, cl_70kg, bcrcl, bsex, omega)
  v  <- v_70kg * (severe$WT/70)
  cc <- sapply(tt, function(t) conc(t, DOSE, cl, v))
  q  <- apply(cc, 2L, quantile, c(.05, .5, .95))
  data.frame(t = tt, lo = q[1, ], med = q[2, ], hi = q[3, ], model = label)
}
lv <- c("truth", "normal-trial model", "mild-trial model",
        "moderate-trial model", "meta-analysis (admixr2)")
# the published models carry a sex effect but no renal term, hence bcrcl = 0
sims <- rbind(
  band(CL70, V70, BCRCL, BSEX, OM_CL, lv[1]),
  band(normal_paper$CL,   normal_paper$V,   0,
       normal_paper$BSEX,   normal_paper$OM,   lv[2]),
  band(mild_paper$CL,     mild_paper$V,     0,
       mild_paper$BSEX,     mild_paper$OM,     lv[3]),
  band(moderate_paper$CL, moderate_paper$V, 0,
       moderate_paper$BSEX, moderate_paper$OM, lv[4]),
  band(exp(e[["tcl"]]), exp(e[["tv"]]), e[["bcrcl"]], e[["bsex"]],
       as.numeric(fit$omega[1L, 1L]), lv[5]))
sims$model <- factor(sims$model, levels = lv)
tru <- subset(sims, model == lv[1], select = -model)   # drawn in every panel

ggplot(sims, aes(t)) +
  geom_ribbon(data = tru, aes(ymin = lo, ymax = hi), fill = "grey60",
              alpha = .45) +
  geom_line(data = tru, aes(y = med), colour = "grey25", linewidth = .7) +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = model), alpha = .40) +
  geom_line(aes(y = med, colour = model), linewidth = .9) +
  facet_wrap(~model, nrow = 2) +
  guides(fill = "none", colour = "none") +
  labs(x = "time (h)", y = "concentration (mg/L)",
       subtitle = paste("800 simulated severe-impairment patients per model;",
                        "grey = the truth")) +
  theme_bw()
```

Each trial model reproduces the population it was fitted in, not this one, and
the miss is ordered by how far that population sits from severe. At 24 h,
against the truth:

```{r vpcerr}
at_24h <- function(model_name) {
  d <- sims[sims$model == model_name, ]
  i <- which.min(abs(d$t - 24))
  list(median = d$med[i], width = d$hi[i] - d$lo[i])
}

truth_24h <- at_24h(lv[1])
err_table <- data.frame()
for (model_name in lv[-1]) {
  this_model <- at_24h(model_name)
  err_table  <- rbind(err_table, data.frame(
    model      = model_name,
    median_err = 100 * (this_model$median / truth_24h$median - 1),
    width_err  = 100 * (this_model$width  / truth_24h$width  - 1)))
}
knitr::kable(err_table, digits = 1, row.names = FALSE,
  caption = "% error in the median and in the 90% interval width, at 24 h.")
```

They miss on **spread** as much as on level. A model with no renal term applied
to this cohort has only weight and `omega` to generate between-patient
variability, and the `omega` it carries was fitted where creatinine clearance
hardly varied — so it cannot reproduce a population whose clearance varies a
great deal. Only the meta-analysis, which knows what renal function does, lands
on the grey.

```{r payofftab}
knitr::kable(data.frame(
  time = TIMES, truth = p_truth, meta_analysis = p_adm, nearest_model = p_near,
  err_meta = 100*(p_adm/p_truth - 1), err_near = 100*(p_near/p_truth - 1)),
  digits = 2, caption = "Severe renal impairment: predicted mean concentration.")
```

The nearest published model carries a clearance fitted in patients who clear the
drug considerably faster than she does, so it drifts further wrong the longer you
look, and it **underpredicts exposure** --- the direction that matters
clinically.

## The dose

Exposure is `AUC = Dose / CL`, so the dose that gives her what a
normal-renal-function patient gets from `r DOSE` mg is `r DOSE` mg scaled by the
ratio of mean clearances.

```{r dose}
set.seed(5)
mean_cl <- function(cl_70kg, bcrcl, bsex, omega, wt_median, crcl_median,
                    n = 40000L) {
  pts <- draw_cohort(n, wt_median, .198, crcl_median, .217)
  mean(patient_cl(pts, cl_70kg, bcrcl, bsex, omega))
}
# a severe cohort (WT 72, CRCL 21) against a normal-function one (WT 78, CRCL 95)
dose_for <- function(cl_70kg, bcrcl, bsex, omega)
  DOSE * mean_cl(cl_70kg, bcrcl, bsex, omega, 72, 21) /
         mean_cl(cl_70kg, bcrcl, bsex, omega, 78, 95)
d_true <- dose_for(CL70, BCRCL, BSEX, OM_CL)
d_meta <- dose_for(exp(e[["tcl"]]), e[["bcrcl"]], e[["bsex"]],
                   as.numeric(fit$omega[1L, 1L]))
d_near <- dose_for(moderate_paper$CL, 0,
                   moderate_paper$BSEX, moderate_paper$OM)
knitr::kable(data.frame(
  basis   = c("the truth", "meta-analysis (admixr2)", "nearest published model"),
  dose_mg = round(c(d_true, d_meta, d_near))),
  digits = 0, row.names = FALSE,
  caption = paste("Dose giving a CrCl 21 mL/min patient the exposure a",
                  "normal-function patient gets from", DOSE, "mg."))
```

The meta-analysis, built from those three papers and nothing else, gives
`r round(d_meta)` mg against a true `r round(d_true)` mg.

The nearest published model gives `r round(d_near)` mg --- about
`r sprintf("%.1f", d_near / d_true)` times the dose she should get. Note where
its small reduction below `r DOSE` mg comes from: purely the weight difference
between the two cohorts. Its renal recommendation is exactly zero, and it cannot
be anything else, because the model has no renal term. It would return the same
`r round(d_near)` mg just as confidently at 15 mL/min, or at 5.

## Doing this with your own papers

**A paper contributes in one of two currencies**, and `admStudy()` takes either.
A published *model* goes in as `model` with the paper's estimates and its
uncertainty; *digitised* aggregate data goes in as `E` with `sd` (or `sem`, or a
full `V`). Nothing is solved when you build one, so a study is cheap to write
and `print()` back against the source before any fitting happens.

**The baseline table** goes to `admPopulation()` in whichever form the paper
printed, one covariate per argument:

```{r forms}
admPopulation(WT   = c(mean = 78, sd = 16),        # mean and SD
              CRCL = c(median = 92, iqr = c(62, 118)),
              SEX  = c(male = 0.55),               # a proportion
              cor  = c(WT.CRCL = 0.45))            # only the PAIRS reported
```

`cv` (as a percent) and `range = c(min, max)` are accepted too, and `cor` names
only the pairs the paper gave --- everything else is independent, so a partial
table needs no identity padding. Correlations are on the LATENT scale, which for
a lognormal margin means the logs. Prefer `dist = "lnorm"` for anything
positive: a normal margin is unbounded below and the quadrature reaches about
3.75 SD, so a CV above roughly 0.27 puts a node at or below zero, where an
allometric term is `NaN`. `admPopulation()` warns when it sees this. If you have
the individual covariates --- a digitised listing, or your own cohort --- pass
`data =` and the table is read off them.

**Uncertainty is what makes a standard error mean anything.** A study generated
from a model is not a sample: its mean and covariance are exact functions of
somebody else's published estimate, so the only random thing in the chain is
that estimate. Give the paper's full `cov` where it reports one, or its `rse`
column (as percentages, the way a parameter table prints them) --- `print()`
shows the conversion, which is not the same for a log-parameterised theta as for
an ordinary coefficient. It must cover **every** parameter the source estimated:
a missing row reads as a parameter the source knew exactly, and admixr2 refuses
the whole sandwich rather than report intervals that are too narrow. With an
`nlmixr2` fit in hand, `foceiControl(covFull = TRUE)` makes `fit$cov` exactly
that matrix.

**Then read `print()` before you fit.** It lays out which covariates each source
bands, conditions and marginalises, and says so when one is marginal everywhere
--- that covariate is identified only by the contrast *between* sources, which
is legitimate when they really differ and a flat ridge when they do not. A flat
ridge converges to a confident wrong number rather than failing, so it is worth
the ten seconds.

**Discrete covariates are strata**, enumerated exactly at their levels and
weighted by their probabilities --- nothing about them is approximated.

**Cost.** A product grid over `p` covariates costs `cov_nodes^p`, but where the
covariates reach the model through fewer scalars than there are covariates,
admixr2 integrates over those directions instead. Three covariates entering as
one product are a one-dimensional integral, on
`r admixr2:::.admCovDirNodes(7L, 3L, 1L)` design points rather than `r 7^3`.
That is a change of variables, verified numerically before use — not an
approximation.

**Estimator support.** `admc` and `adgh` marginalise covariates; `adfo` and
`adirmc` refuse rather than silently solving at the covariate mean. admixr2
also refuses a population naming a covariate the model never reads.

## Notes

* **A covariate effect is identified by contrast across sources.** That is the
  whole of this vignette. Within one cohort a covariate barely varies, and what
  little it does is confounded with between-subject variability — a covariate
  enters `V` where `omega` does.
* **Fix what convention fixes.** Allometric exponents are 0.75 and 1; aggregate
  data is not the place to re-estimate them. Estimate the covariate effects a
  popPK analysis would.
* **Extrapolation is extrapolation.** The severe prediction works because the
  power model is the right functional form and the three cohorts span enough
  CRCL to pin its exponent. Fitted over 38–95 mL/min and applied at 21, it
  should be presented as such.

## See also

- [datagen()](datagen.html) — published models as input, in depth
- [Multiple studies](multiple-studies.html) — the contrast that identifies an effect
- [From a published figure to E, V and n](aggregate-data.html) — the other input
- [Estimator comparison](estimator-comparison.html)
