---
title: "Reuniting a split study"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reuniting a split study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 3.0, dpi = 110, warning = FALSE,
                      message = FALSE)
source("_common.R", local = TRUE)
```

```{r libs}
library(admixr2)
library(rxode2)
library(nlmixr2)
library(ggplot2)
```

```{r theme, include = FALSE}
NAVY <- "#001158"; ORANGE <- "#DD6B2A"
theme_set(theme_bw(base_size = 10))
```

The [covariates vignette](covariates.html) shows how to describe a study
population and how admixr2 integrates over it. This one asks the question that
matters: **when you pool published models, how much of the truth comes back?**

The only way to answer that is to know the truth. So we manufacture it. One
data-generating model, one set of patients, split into two studies. Each study
is analysed on its own by FOCEI, exactly as its authors would have. Then the two
*fitted models* --- not the patients --- are pooled. Because we made the truth,
we can check what survived.

```{r truth}
LCL <- log(3.0); LV <- log(20)
BWT <- 0.70      # clearance scales with weight
BCR <- 0.55      # ... and with renal function
OM  <- sqrt(0.060); PE <- 0.12
TT  <- c(0.5, 1, 2, 4, 8, 12); DOSE <- 100
```

# One truth, two studies

The two studies enrolled different people. That is not incidental: it is what
makes the pooled analysis able to say anything at all. Study A is heavier with
ordinary renal function, study B lighter with impairment --- so between them they
span a covariate range neither covers alone.

```{r populations}
popA <- covDist(WT = c(mean = 88, sd = 16), CRCL = c(mean = 95, sd = 22),
                cor = 0.2, dist = "lnorm")
popB <- covDist(WT = c(mean = 70, sd = 13), CRCL = c(mean = 55, sd = 16),
                cor = 0.2, dist = "lnorm")
```

```{r simulate}
sim_ipd <- function(n, cd, seed, id0) {
  set.seed(seed)
  X   <- covDraw(cd, n = n)
  cl  <- exp(LCL + OM * rnorm(n)) * (X[, "WT"]/70)^BWT * (X[, "CRCL"]/90)^BCR
  f   <- DOSE / exp(LV) * exp(outer(-cl / exp(LV), TT))
  y   <- f * (1 + PE * matrix(rnorm(length(f)), nrow(f)))
  do.call(rbind, lapply(seq_len(n), function(i) rbind(
    data.frame(ID = id0+i, TIME = 0,  DV = NA_real_, AMT = DOSE, EVID = 1,
               WT = X[i,"WT"], CRCL = X[i,"CRCL"]),
    data.frame(ID = id0+i, TIME = TT, DV = y[i, ],   AMT = 0,    EVID = 0,
               WT = X[i,"WT"], CRCL = X[i,"CRCL"]))))
}
ipdA <- sim_ipd(150L, popA, 11L, 0L)
ipdB <- sim_ipd(150L, popB, 12L, 1000L)
```

```{r clouds, fig.height = 3.0, echo = FALSE}
dd <- rbind(
  data.frame(study = "A: heavier, normal renal", unique(ipdA[, c("WT","CRCL")])),
  data.frame(study = "B: lighter, impaired",     unique(ipdB[, c("WT","CRCL")])))
ggplot(dd, aes(WT, CRCL, colour = study)) +
  geom_point(alpha = 0.7, size = 1.2) +
  scale_colour_manual(values = c(NAVY, ORANGE)) +
  labs(x = "weight (kg)", y = "creatinine clearance (mL/min)", colour = NULL) +
  theme(legend.position = "top")
```

# Each group analyses its own patients

This is the step that makes the exercise realistic. Neither group has the
other's data. Each runs FOCEI on its own 150 subjects and publishes the result.

```{r focei, results = "hide"}
mod_full <- function() {
  ini({lcl <- log(2.5); lv <- log(18); bwt <- 0.5; bcr <- 0.5
       eta.cl ~ 0.05; prop.err <- 0.2})
  model({cl <- exp(lcl + eta.cl) * (WT/70)^bwt * (CRCL/90)^bcr
         v  <- exp(lv); d/dt(centr) <- -cl/v*centr; cp <- centr/v
         cp ~ prop(prop.err)})
}
mod_wtonly <- function() {                   # renal function never considered
  ini({lcl <- log(2.5); lv <- log(18); bwt <- 0.5
       eta.cl ~ 0.05; prop.err <- 0.2})
  model({cl <- exp(lcl + eta.cl) * (WT/70)^bwt
         v  <- exp(lv); d/dt(centr) <- -cl/v*centr; cp <- centr/v
         cp ~ prop(prop.err)})
}
foceiFit <- function(m, d)
  nlmixr2(m, d, est = "focei", control = foceiControl(print = 0L, covMethod = ""))

fitA  <- foceiFit(mod_full,   ipdA)              # A analyses both covariates
fitB  <- foceiFit(mod_full,   ipdB)              # so does B
fitA0 <- foceiFit(mod_wtonly, ipdA)              # ... and an A that did not
fitG  <- foceiFit(mod_full,   rbind(ipdA, ipdB)) # the gold standard: all 300
```

The last fit is the benchmark. It is what you would get **if both groups shared
their patients** --- the best any synthesis could hope to match. It is not the
truth either: FOCEI on 300 subjects has its own estimation error.

```{r focei-tab, echo = FALSE}
grab <- function(f, lbl) {
  e <- setNames(f$parFixedDf$Estimate, rownames(f$parFixedDf))
  data.frame(published = lbl,
             bwt = round(e[["bwt"]], 3),
             bcr = if (is.na(e["bcr"])) NA else round(e[["bcr"]], 3),
             omega = round(sqrt(f$omega[1,1]), 3))
}
knitr::kable(rbind(grab(fitA, "A, both covariates"),
                   grab(fitB, "B, both covariates"),
                   grab(fitA0, "A, weight only"),
                   grab(fitG, "GOLD: all 300 patients"),
                   data.frame(published = "TRUTH", bwt = BWT, bcr = BCR,
                              omega = round(OM, 3))),
             row.names = FALSE,
             caption = "What each group would publish, and the pooled-IPD benchmark.")
```

Two things in that table are the whole point of what follows.

**The individual fits are not the truth either.** Each is 150 patients; A's
renal coefficient came out high, B's weight coefficient low. Any pooled analysis
inherits that noise --- it cannot do better than its inputs.

**The study that omitted renal function did not report a null.** Its weight
coefficient came out *higher* than A's full fit, because weight and renal
function are correlated and the weight term absorbed part of an effect that was
really renal. And its random effect came out *larger*, because the rest of that
effect had nowhere else to go. Both are exactly what theory says:
$\hat\beta_{wt} \to \beta_{wt} + \beta_{cr}\,\mathrm{Cov}/\mathrm{Var}$ and
$\omega'^2 = \omega^2 + \beta_{cr}^2\mathrm{Var}(a)(1-\rho^2)$.

# Pooling the published models

Now the models --- not the patients --- go into admixr2. Each study contributes
its own fitted model applied to its own population, stratified on the covariates
that model actually conditions on.

```{r as-model, echo = FALSE}
as_model <- function(fit) {
  e <- setNames(fit$parFixedDf$Estimate, rownames(fit$parFixedDf))
  om <- fit$omega[1, 1]
  if (!is.na(e["bcr"])) {
    f <- function() {
      ini({lcl <- .LCL; lv <- .LV; bwt <- .BWT; bcr <- .BCR
           eta.cl ~ .OM; prop.err <- .PE})
      model({cl <- exp(lcl + eta.cl) * (WT/70)^bwt * (CRCL/90)^bcr
             v <- exp(lv); d/dt(centr) <- -cl/v*centr; cp <- centr/v
             cp ~ prop(prop.err)})
    }
    body(f) <- do.call(substitute, list(body(f), list(
      .LCL = e[["lcl"]], .LV = e[["lv"]], .BWT = e[["bwt"]],
      .BCR = e[["bcr"]], .OM = om, .PE = e[["prop.err"]])))
  } else {
    f <- function() {
      ini({lcl <- .LCL; lv <- .LV; bwt <- .BWT; eta.cl ~ .OM; prop.err <- .PE})
      model({cl <- exp(lcl + eta.cl) * (WT/70)^bwt
             v <- exp(lv); d/dt(centr) <- -cl/v*centr; cp <- centr/v
             cp ~ prop(prop.err)})
    }
    body(f) <- do.call(substitute, list(body(f), list(
      .LCL = e[["lcl"]], .LV = e[["lv"]], .BWT = e[["bwt"]],
      .OM = om, .PE = e[["prop.err"]])))
  }
  f
}
```

```{r generate}
make_studies <- function(mA, mB)
  datagen(list(
    A = list(model = mA, n = 150L, times = TT, ev = et(amt = DOSE),
             cov_dist = popA, stratify = TRUE),
    B = list(model = mB, n = 150L, times = TT, ev = et(amt = DOSE),
             cov_dist = popB, stratify = TRUE)),
    control = datagenControl(method = "mc", n_sim = 3000L))

unified <- function() {
  ini({lcl <- log(2.0); lv <- log(15); bwt <- 0.20; bcr <- 0.20
       eta.cl ~ 0.03; prop.err <- 0.30})
  model({cl <- exp(lcl + eta.cl) * (WT/70)^bwt * (CRCL/90)^bcr
         v  <- exp(lv); d/dt(centr) <- -cl/v*centr; cp <- centr/v
         cp ~ prop(prop.err)})
}
pool <- function(st) nlmixr2(unified, admData(), est = "adgh",
  control = adghControl(studies = st, print = 0L, covMethod = "none"))

pooled_both <- pool(make_studies(as_model(fitA),  as_model(fitB)))
pooled_omit <- pool(make_studies(as_model(fitA0), as_model(fitB)))
```

`stratify = TRUE` reads each model to decide the split. Study A's full model
conditions on both covariates, so it is stratified on both. The weight-only
version of A is stratified on weight alone, with renal function marginalised
inside each stratum --- because A has no renal contrast to give, and asking for
one would manufacture a null.

```{r results, echo = FALSE}
res <- function(f, lbl) {
  e <- setNames(f$parFixedDf$Estimate, rownames(f$parFixedDf))
  data.frame(analysis = lbl, bwt = round(e[["bwt"]], 3),
             bcr = round(e[["bcr"]], 3), omega = round(sqrt(f$omega[1,1]), 3))
}
knitr::kable(rbind(
  res(pooled_both, "MBMA: both studies fitted both"),
  res(pooled_omit, "MBMA: A omitted renal function"),
  res(fitG,        "GOLD: FOCEI on all 300 patients"),
  data.frame(analysis = "TRUTH", bwt = BWT, bcr = BCR, omega = round(OM, 3))),
  row.names = FALSE,
  caption = "What survives the round trip through aggregate data.")
```

```{r recovery-fig, echo = FALSE, fig.height = 2.6}
e1 <- setNames(pooled_both$parFixedDf$Estimate, rownames(pooled_both$parFixedDf))
e2 <- setNames(pooled_omit$parFixedDf$Estimate, rownames(pooled_omit$parFixedDf))
df <- data.frame(
  coef = rep(c("weight", "renal"), each = 4),
  what = rep(c("start", "pooled (both)", "pooled (A omits)", "truth"), 2),
  value = c(0.20, e1[["bwt"]], e2[["bwt"]], BWT,
            0.20, e1[["bcr"]], e2[["bcr"]], BCR))
df$what <- factor(df$what,
  levels = c("start", "pooled (both)", "pooled (A omits)", "truth"))
ggplot(df, aes(value, coef, colour = what, shape = what)) +
  geom_point(size = 4, alpha = 0.85) +
  scale_colour_manual(values = c("grey65", NAVY, ORANGE, "black")) +
  labs(x = "coefficient", y = NULL, colour = NULL, shape = NULL) +
  theme(legend.position = "top")
```

# What this shows

**Read the pooled rows against the GOLD row, not against TRUTH.** A synthesis
of published models cannot be better than a pooled analysis of the patients
those models were fitted to --- and that pooled analysis is not the truth
either. FOCEI on 150 subjects scatters; FOCEI on 300 still misses.

**A covariate no single study could measure well comes back.** Both
coefficients move from a starting value of 0.20 to the neighbourhood of the
gold-standard fit, from aggregate summaries alone --- no patient-level record
ever left either group.

**A study that never measured renal function still contributes.** Compare the
two pooled rows: dropping renal function from A's model does not destroy the
renal estimate. A's information now arrives through its *inflated random
effect* rather than a mean contrast --- the
$\omega'^2 = \omega^2 + eta^2\mathrm{Var}(a)$ channel --- and it works only
because A's prediction marginalises over renal function instead of stratifying
on one it never fitted.

## Where the remaining gap comes from

Two different things are at work, and only one of them is about aggregate data.

**The pooling is faithful to each study.** Hand it a single study and it returns
that study's own model: a study published with $eta_{wt}=0.669$ and
$eta_{cr}=0.730$ comes back as **0.673** and **0.731**; one published with
0.606 and 0.546 comes back as **0.606** and **0.546**. The marginalisation and
the within-stratum conditioning are exact --- what goes in is what comes out.

**When two studies disagree, it interpolates.** Two studies differing only in
their covariate coefficients (0.669 vs 0.606 for weight, 0.730 vs 0.546 for
renal) pool to **0.621** and **0.607** --- each between its own two inputs, as a
precision-weighted compromise should be. And this is not simulation noise: the
same numbers come back at 3 000, 10 000 and 40 000 generation draws.

**So the distance from TRUTH is inherited, not introduced.** Each published
model is a single draw from a sampling distribution whose standard error, at
150 subjects, is around 0.04 on each coefficient. admixr2 reproduces those
published models faithfully and compromises between them sensibly --- but two
noisy estimates cannot average into a precise one.

The contrast with the [covariates vignette](covariates.html) is the point. There,
two studies that never measured weight at all recovered its coefficient at
**0.7007** against a true 0.700 --- because their *published models were exact*,
constructed from the omega-inflation identity rather than fitted to a finite
sample. Give the method exact inputs and it returns the truth; give it noisy
inputs and it returns an honest compromise between them.

# The honest limits

- **Point estimates are targeted; the reported uncertainty is approximate.** The
  aggregate likelihood treats the observed covariance as Wishart, which stops
  being exactly right once a covariate has been integrated out.
- **This is one replicate.** The numbers above will move if you change the
  seed; they illustrate the mechanism and are not a bias estimate. A bias study
  needs many replicates.
- **The individual fits are noisy, and that is the whole gap.** Over six
  independent replicates at n = 400, FOCEI recovered the weight coefficient at
  0.675 with a standard error of 0.023 --- unbiased, but scattering from 0.584
  to 0.721 replicate to replicate. At the 150 subjects each study has here the
  scatter is wider still. A synthesis of two such estimates inherits that
  scatter; it cannot average away noise that only two studies contain.
- **Studies that disagree on more than the covariates** --- different baseline
  clearance, different random-effect magnitude --- pull the compromise in ways
  no single coefficient explains. That is real heterogeneity, and worth looking
  at before pooling.
- **The design must not be confounded.** The two studies here differ in weight
  *and* renal function in different directions. Had they differed along a single
  axis, the pooled analysis could identify only a combination of the two
  coefficients, however much data you fed it.

# See also

- [Covariates](covariates.html) --- describing a population, and the three ways it goes wrong
- [Joint fitting across studies](multiple-studies.html)
- [Comparing the estimators](estimator-comparison.html)
- [Getting started](admixr2.html)
