---
title: "Covariates in aggregate data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covariates in aggregate data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.2
)
source(Filter(file.exists, c("_common.R", "../_common.R", "vignettes/_common.R"))[1])
```

## The problem

You need a dose for **severe renal impairment**. Three trials have been
published — normal renal function, mild impairment, moderate — and none
enrolled a severe patient. None of them reported a renal effect either: within
each trial creatinine clearance barely varied, so each analyst sensibly left it
out.

You are left with three population models that disagree about clearance and
none that says why.

```{r libs}
library(admixr2); library(rxode2); library(nlmixr2); library(ggplot2)
```

## Three trials, three answers

Weight and creatinine clearance are correlated, so the virtual patients are
drawn jointly through a Gaussian copula — each covariate keeps its own
lognormal margin, the dependence goes on the latent scale. The truth, unknown
to all three analysts, is CL = 5 L/h and V = 50 L at 70 kg, allometric weight
scaling, and a renal exponent of 0.6.

```{r sim}
set.seed(11)
TIMES <- c(0.5, 1, 2, 4, 8, 12, 24); DOSE <- 200
CL70 <- 5; V70 <- 50; BCRCL <- 0.6; OM_CL <- 0.06; SD_ADD <- 0.08
R <- matrix(c(1, .45, .45, 1), 2, 2,
            dimnames = list(c("WT", "CRCL"), c("WT", "CRCL")))

draw_pop <- function(n, mw, sw, mc, sc) {
  Z <- matrix(stats::rnorm(2 * n), n, 2) %*% chol(R)
  data.frame(WT   = stats::qlnorm(stats::pnorm(Z[, 1]), log(mw), sw),
             CRCL = stats::qlnorm(stats::pnorm(Z[, 2]), log(mc), sc))
}
sim_trial <- function(cv, id0) {
  n  <- nrow(cv)
  cl <- CL70 * (cv$WT/70)^0.75 * (cv$CRCL/90)^BCRCL *
        exp(stats::rnorm(n, 0, sqrt(OM_CL)))
  v  <- V70 * (cv$WT/70)
  do.call(rbind, lapply(seq_len(n), function(i) rbind(
    data.frame(ID = id0+i, TIME = 0, DV = NA_real_, AMT = DOSE, EVID = 1,
               CMT = 1, WT = cv$WT[i], CRCL = cv$CRCL[i]),
    data.frame(ID = id0+i, TIME = TIMES,
               DV = DOSE/v[i] * exp(-cl[i]/v[i]*TIMES) +
                    stats::rnorm(length(TIMES), 0, SD_ADD),
               AMT = 0, EVID = 0, CMT = 1, WT = cv$WT[i], CRCL = cv$CRCL[i]))))
}
cohorts <- list(normal   = draw_pop(120L, 78, .198, 95, .217),
                mild     = draw_pop( 90L, 76, .198, 62, .198),
                moderate = draw_pop( 70L, 74, .198, 38, .198))
trial_data <- Map(function(cv, k) sim_trial(cv, k * 10000L),
                  cohorts, seq_along(cohorts))
```

A published summary keeps only the mean and the spread, so the covariate
disappears into the width of the ribbon — poor clearers at the top, good
clearers at the bottom:

```{r hidden, fig.height = 3.8}
d1  <- trial_data$normal[trial_data$normal$EVID == 0, ]
agg <- data.frame(t = TIMES,
                  m = tapply(d1$DV, d1$TIME, mean),
                  s = tapply(d1$DV, d1$TIME, stats::sd),
                  panel = "What the paper reports")
ind <- d1[d1$ID %in% head(unique(d1$ID), 150), ]
ind$panel <- "Individual patients (never published)"

ggplot() +
  geom_line(data = ind, aes(TIME, DV, group = ID, colour = CRCL),
            alpha = .35, linewidth = .3) +
  geom_ribbon(data = agg, aes(t, ymin = m - s, ymax = m + s),
              fill = "grey55", alpha = .35) +
  geom_line(data = agg, aes(t, m), linewidth = 1) +
  scale_colour_viridis_c(name = "CRCL (mL/min)", option = "C", end = .9) +
  facet_wrap(~panel) + labs(x = "time (h)", y = "concentration (mg/L)") +
  theme_bw() + theme(legend.position = "bottom")
```

Each analyst now fits their own trial: weight scaled allometrically with the
exponents fixed at 0.75 and 1, and no renal term.

```{r trialfits}
trial_model <- function() {
  ini({ tcl <- log(4.5); tv <- log(45); eta.cl ~ 0.1; add.err <- 0.2 })
  model({ cl <- exp(tcl + eta.cl) * (WT/70)^0.75
          v  <- exp(tv) * (WT/70)
          cp <- linCmt(); cp ~ add(add.err) })
}
trial_fits <- lapply(trial_data, function(d)
  suppressWarnings(nlmixr2(trial_model, d, est = "focei",
                           control = foceiControl(print = 0L))))
published <- lapply(trial_fits, function(f) list(
  CL = exp(f$parFixedDf["tcl", "Estimate"]),
  V  = exp(f$parFixedDf["tv",  "Estimate"]),
  OM = as.numeric(f$omega[1L, 1L]),
  ADD = f$parFixedDf["add.err", "Estimate"]))
knitr::kable(data.frame(
  cohort      = names(published),
  median_CRCL = round(vapply(cohorts, function(c) stats::median(c$CRCL), 0)),
  CL_at_70kg  = vapply(published, `[[`, 0, "CL"),
  V_at_70kg   = vapply(published, `[[`, 0, "V")),
  digits = 2, row.names = FALSE,
  caption = "Three independent analyses, each sound on its own data.")
```

Clearance differs by nearly a factor of two; volume does not move. That
difference *is* the renal effect — unattributed, because no single trial could
attribute it.

## Their models as input

admixr2 takes two kinds of input: a digitised summary, or a **published model
supplied directly**. This is the second. `datagen()` turns each published model,
plus the population it was fitted in, into the `(E, V, n)` a figure would have
given us.

```{r datagen}
CLp <- Vp <- OMp <- ADDp <- NULL             # read by ini() below
published_model <- function() {
  ini({ tcl <- log(CLp); tv <- log(Vp); eta.cl ~ OMp; add.err <- ADDp })
  model({ cl <- exp(tcl + eta.cl) * (WT/70)^0.75
          v  <- exp(tv) * (WT/70)
          cp <- linCmt(); cp ~ add(add.err) })
}
cov_dists <- lapply(cohorts, function(cv) covDist(
  WT   = list(meanlog = mean(log(cv$WT)),   sdlog = stats::sd(log(cv$WT))),
  CRCL = list(meanlog = mean(log(cv$CRCL)), sdlog = stats::sd(log(cv$CRCL))),
  cor  = R))
studies <- list()
for (nm in names(published)) {
  CLp <- published[[nm]]$CL; Vp   <- published[[nm]]$V
  OMp <- published[[nm]]$OM; ADDp <- published[[nm]]$ADD
  studies[[nm]] <- datagen(
    list(s = list(times = TIMES, ev = rxode2::et(amt = DOSE),
                  n = nrow(cohorts[[nm]]), cov_dist = cov_dists[[nm]])),
    model = published_model,
    control = datagenControl(method = "gh", seed = 1L))$s
}
```

Two things go in per study: the trial's **model**, and the **`cov_dist`** it was
fitted in — which the paper describes in its baseline table. The models have no
renal term; the `cov_dist` does. We are telling admixr2 who was enrolled and
letting it work out what that implies.

## The meta-analysis

```{r admfit}
adm_model <- function() {
  ini({ tcl <- log(4); tv <- log(45); bcrcl <- 0.3
        eta.cl ~ 0.1; add.err <- 0.1 })
  model({ cl <- exp(tcl + eta.cl) * (WT/70)^0.75 * (CRCL/90)^bcrcl
          v  <- exp(tv) * (WT/70)
          cp <- linCmt(); cp ~ add(add.err) })
}
fit <- nlmixr2(adm_model, admData(), est = "adgh",
               control = adghControl(studies = studies, print = 0L,
                                     covMethod = "r"))
pf     <- fit$parFixedDf
om     <- as.numeric(fit$omega[1L, 1L])
om_se  <- sqrt(diag(fit$cov))[["om.eta.cl"]]   # omega's SE is in $cov, not pf
knitr::kable(data.frame(
  parameter = c("CL at 70 kg, CRCL 90 (L/h)", "V at 70 kg (L)",
                "renal exponent", "additive SD (mg/L)", "omega (eta.cl)"),
  truth     = c(CL70, V70, BCRCL, SD_ADD, OM_CL),
  estimate  = c(pf["tcl", "Back-transformed"], pf["tv", "Back-transformed"],
                pf["bcrcl", "Back-transformed"], pf["add.err", "Back-transformed"],
                as.numeric(fit$omega[1L, 1L])),
  # CIs, not SEs: a structural theta's SE is on the LOG scale, so printing it
  # beside a back-transformed estimate invites reading 0.022 as 0.022 L/h
  lower     = c(pf["tcl", "CI Lower"], pf["tv", "CI Lower"],
                pf["bcrcl", "CI Lower"], pf["add.err", "CI Lower"],
                om - 1.96 * om_se),
  upper     = c(pf["tcl", "CI Upper"], pf["tv", "CI Upper"],
                pf["bcrcl", "CI Upper"], pf["add.err", "CI Upper"],
                om + 1.96 * om_se)),
  digits = 3, row.names = FALSE,
  caption = "Recovered from three analyses, none of which contained a renal term.")
```

Two things about standard errors, both easy to get wrong. A structural theta's
SE is on the **log** scale — printing it beside a back-transformed estimate
would read as L/h and be off by an order of magnitude — so `parFixedDf` gives
back-transformed CIs, used above. And the **omega** SE is not in `parFixedDf`
at all: by nlmixr2 convention it lives in `fit$cov`, as `om.eta.cl`.

Here is what the meta-analysis actually did:

```{r mechanism, fig.height = 3.8}
e  <- stats::setNames(pf$Estimate, rownames(pf))
crcl_grid <- seq(15, 110, length.out = 120)
curve_df <- data.frame(
  CRCL = crcl_grid,
  CL   = exp(e[["tcl"]]) * (crcl_grid / 90)^e[["bcrcl"]])
pts <- data.frame(
  CRCL = vapply(cohorts, function(c) stats::median(c$CRCL), 0),
  CL   = vapply(published, `[[`, 0, "CL"),
  lab  = names(published))
# each published model has NO renal term, so what it says about clearance is a
# FLAT line -- drawn over the CRCL range that trial actually covered
seg <- data.frame(
  lab = names(published),
  CL  = vapply(published, `[[`, 0, "CL"),
  lo  = vapply(cohorts, function(c) unname(stats::quantile(c$CRCL, .10)), 0),
  hi  = vapply(cohorts, function(c) unname(stats::quantile(c$CRCL, .90)), 0))

ggplot(curve_df, aes(CRCL, CL)) +
  annotate("rect", xmin = 15, xmax = 30, ymin = -Inf, ymax = Inf,
           fill = "grey85", alpha = .6) +
  annotate("text", x = 22.5, y = max(curve_df$CL), label = "severe\n(no trial)",
           size = 3, vjust = 1.1) +
  # what the nearest published model would claim down there
  geom_segment(data = subset(seg, lab == "moderate"),
               aes(x = 15, xend = lo, y = CL, yend = CL),
               linetype = "dotted", linewidth = .7, colour = "grey45") +
  geom_segment(data = seg, aes(x = lo, xend = hi, y = CL, yend = CL,
                               colour = lab), linewidth = 1.6) +
  geom_line(linewidth = 1, colour = "steelblue4") +
  geom_point(data = pts, size = 3.2) +
  geom_text(data = pts, aes(label = lab), vjust = -1.2, size = 3.2) +
  guides(colour = "none") +
  labs(x = "median creatinine clearance (mL/min)",
       y = "clearance at 70 kg (L/h)",
       subtitle = paste("bars: what each published model says, over the range",
                        "it covered\nblue: the meta-analysis")) +
  theme_bw()
```

Each trial's model is a **flat bar**: having no renal term, it says clearance
is whatever it measured, at any renal function. Three flat bars at three
different heights, each correct over its own range and mutually contradictory.

The blue curve is the meta-analysis, and it passes through all three. (Just
above, in fact, by 1-4%: a trial with no renal term fits roughly the *average*
of `(CRCL/90)^0.6` over its cohort, and that average sits below the value at
the cohort's median because the power function is concave.) **That curve is the
renal effect** — recovered from the contrast between trials, and
carried into a region none of them sampled. The dotted extension shows what the
nearest published model would claim there, which is the comparison the next
section quantifies.

## The cohort nobody ran

Without admixr2 you would take the published model closest to your patients —
the moderate-impairment one — and use it.

```{r predict}
set.seed(3)
marg_mean <- function(cl_ref, v_ref, bcrcl, om, n = 40000L) {
  Z    <- matrix(stats::rnorm(2 * n), n, 2) %*% chol(R)
  WT   <- stats::qlnorm(stats::pnorm(Z[, 1]), log(72), .198)
  CRCL <- stats::qlnorm(stats::pnorm(Z[, 2]), log(21), .217)
  cl <- cl_ref * (WT/70)^0.75 *
        (if (is.na(bcrcl)) 1 else (CRCL/90)^bcrcl) *
        exp(stats::rnorm(n, 0, sqrt(om)))
  vapply(TIMES, function(t)
    mean(DOSE / (v_ref * WT/70) * exp(-cl / (v_ref * WT/70) * t)), numeric(1))
}
p_adm   <- marg_mean(exp(e[["tcl"]]), exp(e[["tv"]]), e[["bcrcl"]],
                     as.numeric(fit$omega[1L, 1L]))
p_near  <- marg_mean(published$moderate$CL, published$moderate$V, NA,
                     published$moderate$OM)
p_truth <- marg_mean(CL70, V70, BCRCL, OM_CL)
```

Predicting the *mean* is only half of it. Simulate a severe cohort under each
model and compare the whole distribution — median and 90% interval — against
what those patients would really do. The truth is repeated in grey behind every
panel.

```{r vpc, fig.height = 5.0}
set.seed(7)
n_pt <- 800L
Zs   <- matrix(stats::rnorm(2 * n_pt), n_pt, 2) %*% chol(R)
WTs  <- stats::qlnorm(stats::pnorm(Zs[, 1]), log(72), .198)
CRs  <- stats::qlnorm(stats::pnorm(Zs[, 2]), log(21), .217)
tt   <- seq(0.25, 24, length.out = 40)

band <- function(cl_ref, v_ref, bcrcl, om, label) {
  cl <- cl_ref * (WTs/70)^0.75 *
        (if (is.na(bcrcl)) 1 else (CRs/90)^bcrcl) *
        exp(stats::rnorm(n_pt, 0, sqrt(om)))
  v  <- v_ref * (WTs/70)
  M  <- vapply(tt, function(t) DOSE / v * exp(-cl / v * t), numeric(n_pt))
  q  <- apply(M, 2L, stats::quantile, c(.05, .5, .95))
  data.frame(t = tt, lo = q[1, ], med = q[2, ], hi = q[3, ], model = label)
}
lv  <- c("truth", "normal-trial model", "mild-trial model",
         "moderate-trial model", "meta-analysis (admixr2)")
sims <- rbind(
  band(CL70, V70, BCRCL, OM_CL, lv[1]),
  band(published$normal$CL,   published$normal$V,   NA, published$normal$OM,   lv[2]),
  band(published$mild$CL,     published$mild$V,     NA, published$mild$OM,     lv[3]),
  band(published$moderate$CL, published$moderate$V, NA, published$moderate$OM, lv[4]),
  band(exp(e[["tcl"]]), exp(e[["tv"]]), e[["bcrcl"]],
       as.numeric(fit$omega[1L, 1L]), lv[5]))
sims$model <- factor(sims$model, levels = lv)
tru <- subset(sims, model == lv[1], select = -model)   # drawn in every panel

ggplot(sims, aes(t)) +
  geom_ribbon(data = tru, aes(ymin = lo, ymax = hi), fill = "grey60",
              alpha = .45) +
  geom_line(data = tru, aes(y = med), colour = "grey25", linewidth = .7) +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = model), alpha = .40) +
  geom_line(aes(y = med, colour = model), linewidth = .9) +
  facet_wrap(~model, nrow = 2) +
  guides(fill = "none", colour = "none") +
  labs(x = "time (h)", y = "concentration (mg/L)",
       subtitle = paste("800 simulated severe-impairment patients per model;",
                        "grey = the truth")) +
  theme_bw()
```

Each trial model reproduces the population it was fitted in, not this one, and
the miss is ordered by how far that population sits from severe. At 24 h,
against the truth:

```{r vpcerr}
at24 <- function(m) {
  d <- sims[sims$model == m, ]; i <- which.min(abs(d$t - 24))
  c(med = d$med[i], width = d$hi[i] - d$lo[i])
}
tr <- at24(lv[1])
knitr::kable(data.frame(
  model      = lv[-1],
  median_err = vapply(lv[-1], function(m) 100*(at24(m)[["med"]]/tr[["med"]]-1), 0),
  width_err  = vapply(lv[-1], function(m)
                      100*(at24(m)[["width"]]/tr[["width"]]-1), 0)),
  digits = 1, row.names = FALSE,
  caption = "% error in the median and in the 90% interval width, at 24 h.")
```

They miss on **spread** as much as on level. A model with no renal term applied
to this cohort has only weight and `omega` to generate between-patient
variability, and the `omega` it carries was fitted where creatinine clearance
hardly varied — so it cannot reproduce a population whose clearance varies a
great deal. Only the meta-analysis, which knows what renal function does, lands
on the grey.

```{r payofftab}
knitr::kable(data.frame(
  time = TIMES, truth = p_truth, meta_analysis = p_adm, nearest_model = p_near,
  err_meta = 100*(p_adm/p_truth - 1), err_near = 100*(p_near/p_truth - 1)),
  digits = 2, caption = "Severe renal impairment: predicted mean concentration.")
```

The nearest published model is carrying a clearance fitted in patients who
clear the drug faster than these ones do, so it drifts further wrong the longer
you look. Underpredicting exposure in renal impairment is the direction that
matters clinically.

## Practicalities

**Declaring the population.** `cov_dist` is data, not a parameter — admixr2
never estimates it. Take it from the paper's baseline table:

```{r forms}
covDist(WT = c(mean = 78, sd = 16), dist = "lnorm")$WT   # mean and SD
covDist(SEX = c(male = .55, female = .45))$SEX           # a categorical
```

| the paper reports | use |
|---|---|
| mean and SD | `c(mean = , sd = )` with `dist = "lnorm"` |
| geometric mean `G`, CV | `meanlog = log(G)`, `sdlog = sqrt(log(1 + CV^2))` |
| median `M`, IQR `Q1`–`Q3` | `meanlog = log(M)`, `sdlog = (log(Q3) - log(Q1))/1.349` |

A whole baseline table goes in at once as a data frame with `covariate`,
`mean`, `sd` and `dist` columns. Prefer `"lnorm"` for anything positive: a
normal margin is unbounded below and the quadrature reaches about 3.75 SD, so a
CV above roughly 0.27 puts a node at or below zero where an allometric term is
`NaN`. `covDist()` warns when it sees this.

**Marginalised or conditioned.** A `cov_dist` is integrated over; a `cov` value
is solved at. A pooled summary needs the first, a stratified one the second,
and a study can do both. `covStrata()` splits a study reported by subgroup.

**Discrete covariates are strata**, enumerated exactly at their levels and
weighted by their probabilities — nothing about them is approximated.

**Cost.** A product grid over `p` covariates costs `cov_nodes^p`, but where the
covariates reach the model through fewer scalars than there are covariates,
admixr2 integrates over those directions instead. Three covariates entering as
one product are a one-dimensional integral, on
`r admixr2:::.admCovDirNodes(7L, 3L, 1L)` design points rather than `r 7^3`.
That is a change of variables, verified numerically before use — not an
approximation.

**Estimator support.** `admc` and `adgh` marginalise covariates; `adfo` and
`adirmc` refuse rather than silently solving at the covariate mean. admixr2
also refuses a `cov_dist` naming a covariate the model never reads.

## Notes

* **A covariate effect is identified by contrast across sources.** That is the
  whole of this vignette. Within one cohort a covariate barely varies, and what
  little it does is confounded with between-subject variability — a covariate
  enters `V` where `omega` does.
* **Fix what convention fixes.** Allometric exponents are 0.75 and 1; aggregate
  data is not the place to re-estimate them. Estimate the covariate effects a
  popPK analysis would.
* **Extrapolation is extrapolation.** The severe prediction works because the
  power model is the right functional form and the three cohorts span enough
  CRCL to pin its exponent. Fitted over 38–95 mL/min and applied at 21, it
  should be presented as such.

## See also

- [datagen()](datagen.html) — published models as input, in depth
- [Multiple studies](multiple-studies.html) — the contrast that identifies an effect
- [From a published figure to E, V and n](aggregate-data.html) — the other input
- [Estimator comparison](estimator-comparison.html)
