---
title: "Covariate marginalisation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covariate marginalisation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 5
)
```

## The problem

Aggregate data are frequently **stratified by a covariate**: each study
arm reports a mean vector and covariance matrix for a sub-population with
a known covariate value (e.g. a body-weight band). To recover the
population model we must marginalise the prediction over the covariate
distribution.

`admixr2` treats each stratum as a *study* carrying its covariate value
and an integration `weight`, then combines the per-stratum $-2\log L$
contributions. Three rules are available:

| Method     | Combine                                              | Use when                          |
|------------|------------------------------------------------------|-----------------------------------|
| `"gl"`     | weighted sum (Gauss–Legendre)                        | general purpose, truncated normal |
| `"gh"`     | weighted sum (Gauss–Hermite)                         | exact normal, few nodes           |
| `"taylor"` | Laplace (Hessian combine); `order=2`: 3 nodes, `order=4`: 5 nodes | fastest, mild–moderate non-linearity |

All three methods expand the covariate study into **per-stratum studies**, each
evaluated at a fixed covariate value. For `"gl"` and `"gh"` the per-study
$-2\log L$ values are combined as a weighted sum. For `"taylor"` exactly three
per-stratum studies are built (covariate $-h$, centre, $+h$) and combined via
the 2nd-order Laplace approximation:

$$\mathrm{NLL}_\mathrm{Taylor}
  = \mathrm{NLL}_0
    + \tfrac{1}{2}\,\sigma^2_\mathrm{cov}\,
      \frac{\mathrm{NLL}_+ - 2\,\mathrm{NLL}_0 + \mathrm{NLL}_-}{h^2}$$

where $\sigma^2_\mathrm{cov}$ is the covariate variance. For correlated
multi-covariate distributions, additional cross-node evaluations supply the
off-diagonal Hessian terms.

The whole setup — building the quadrature grid, simulating aggregate data, and
assembling the studies — is done by a single [`datagen()`](datagen.html) call.

## Two models: data-generating and fitting

Clearance scales allometrically with body weight,
$\mathrm{CL} = \theta_{CL}\,e^{\eta_{CL}}\,(WT/70)^{\theta_{CLWT}}$, with
$WT \sim N(70, 10^2)$. `datagen()` reads true parameters from a model's
`ini()` block, so we use one model at the truth to *generate* data and a
second with off-truth starting values to *fit*.

```{r model}
library(admixr2)
library(rxode2)
library(nlmixr2)

# data-generating model: ini() = truth
pk_true <- function() {
  ini({
    tcl <- log(5); tv <- log(30); tka <- log(1); clwt <- 1.25
    prop.err <- 0.2
    eta.cl ~ 0.09; eta.v ~ 0.04
  })
  model({
    cl <- exp(tcl + eta.cl) * (wt / 70)^clwt
    v  <- exp(tv  + eta.v); ka <- exp(tka)
    d/dt(depot)   <- -ka * depot
    d/dt(central) <- ka * depot - (cl/v) * central
    cp <- central / v
    cp ~ prop(prop.err)
  })
}

# fitting model: same structure, off-truth ini() = starting guess
pk_fit <- function() {
  ini({
    tcl <- log(4); tv <- log(25); tka <- log(0.8); clwt <- 1.0
    prop.err <- 0.3
    eta.cl ~ 0.05; eta.v ~ 0.05
  })
  model({
    cl <- exp(tcl + eta.cl) * (wt / 70)^clwt
    v  <- exp(tv  + eta.v); ka <- exp(tka)
    d/dt(depot)   <- -ka * depot
    d/dt(central) <- ka * depot - (cl/v) * central
    cp <- central / v
    cp ~ prop(prop.err)
  })
}
```

## Generate the covariate-stratified studies in one call

Add a `covariate` field to the study spec. `datagen()` builds the
quadrature grid, simulates the aggregate `E`/`V` at each weight node, and
returns a weighted `studies` list with the quadrature attached:

```{r datagen}
times <- c(0.5, 1, 2, 4, 8, 12)

cov_data <- datagen(
  studies = list(pop = list(
    times     = times,
    ev        = rxode2::et(amt = 100, cmt = "depot"),
    n         = 300L,
    covariate = list(wt      = list(mu = 70, sd = 10),
                     method  = "gl",
                     n_nodes = 11L))),
  model   = pk_true,
  control = datagenControl(n_sim = 10000L))

length(cov_data)                       # 7 weight strata
attr(cov_data, "quadrature")$method    # "gl"
```

## Fit — exactly like a multiple-studies model

Pass the studies straight to `admControl()` and fit with `nlmixr2()`; `admData()`
is the usual placeholder (no covariate column needed — the estimator supplies
the covariate from each stratum automatically). The quadrature rides along on
the studies as an attribute:

```{r fit}
fit_gl <- nlmixr2(
  pk_fit, admData(), est = "admc",
  control = admControl(studies = cov_data, n_sim = 1000L,
                       grad = "sens", maxeval = 150L, print = 1L, cores = 10))
```

The result is a genuine `admFit` object — `print()`, `AIC()`, standard
errors and `plot()` all work:

```{r results}
ax <- fit_gl$env$admExtra
data.frame(
  param    = c("CL", "V", "KA", "clwt", "om.cl", "sig2"),
  true     = c(5, 30, 1, 1.25, 0.09, 0.04),
  estimate = round(c(exp(ax$struct["tcl"]), exp(ax$struct["tv"]),
                     exp(ax$struct["tka"]), ax$struct["clwt"],
                     ax$omega[1, 1], ax$sigma_var[1]), 3),
  row.names = NULL)
```

## Covariate diagnostics

The `"covariate"` panel (added automatically to `plot()` for covariate
fits) shows observed vs predicted mean across the weight strata plus the
quadrature grid:

```{r plot, fig.height=6}
plot(fit_gl, which = "covariate")
```

## Switching methods

All three methods are one-word (or one-argument) swaps in the `datagen()`
study spec — everything downstream is identical:

```{r switch, eval=FALSE}
# Gauss–Hermite
gh_data <- datagen(
  studies = list(pop = list(
    times = times, ev = rxode2::et(amt = 100, cmt = "depot"), n = 300L,
    covariate = list(wt = list(mu = 70, sd = 10), method = "gh", n_nodes = 5L))),
  model = pk_true, control = datagenControl(n_sim = 2000L))

# 2nd-order Taylor (3 nodes per covariate)
ty_data <- datagen(
  studies = list(pop = list(
    times = times, ev = rxode2::et(amt = 100, cmt = "depot"), n = 300L,
    covariate = list(wt = list(mu = 70, sd = 10), method = "taylor", h = 2))),
  model = pk_true, control = datagenControl(n_sim = 10000L))

# 4th-order Taylor (5 nodes per covariate — adds sigma^4 * NLL'''' / 8 term)
ty4_data <- datagen(
  studies = list(pop = list(
    times = times, ev = rxode2::et(amt = 100, cmt = "depot"), n = 300L,
    covariate = list(wt = list(mu = 70, sd = 10), method = "taylor", h = 2, order = 4L))),
  model = pk_true, control = datagenControl(n_sim = 10000L))
fit_gh  <- nlmixr2(pk_fit, admData(), est = "admc",
                   control = admControl(studies = gh_data,  n_sim = 600L,
                                        grad = "sens", maxeval = 150L))
fit_ty  <- nlmixr2(pk_fit, admData(), est = "admc",
                   control = admControl(studies = ty_data,  n_sim = 1000L,
                                        grad = "sens", maxeval = 150L, print = 1, cores = 10))
fit_ty4 <- nlmixr2(pk_fit, admData(), est = "admc",
                   control = admControl(studies = ty4_data, n_sim = 1000L,
                                        grad = "sens", maxeval = 150L, print = 1, cores = 10))
```

The default `order = 2` Taylor uses 3 per-stratum studies and combines them
via $\mathrm{NLL}_0 + \tfrac{1}{2}\sigma^2\,\mathrm{NLL}''$. Setting
`order = 4` adds two more axis nodes (at covariate $\pm 2h$) and includes the
next correction term $\tfrac{\sigma^4}{8}\,\mathrm{NLL}''''$, which matters
when the model response is moderately curved over the covariate range.
Gauss–Legendre/Hermite are more accurate still for strongly nonlinear
covariate effects.
