---
title: "Covariates in a model-based meta-analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covariates in a model-based meta-analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.2)
source("_common.R", local = TRUE)
```

Three published population models are used as the input to a meta-analysis: each
generates the aggregate data its own study design and population would have
produced, and one unifying model is then fitted to all three at once. No
individual records exist at any point.

The covariate is the point of the exercise. Each study's subjects span a
distribution of body weights, and that distribution differs between studies —
which is not incidental, because between-study variation in the covariate
distribution is the only thing that identifies a covariate coefficient from
aggregate data. Two ways of describing those subjects are compared:

* **summary statistics** — a mean, a standard deviation and a correlation, which
  is what a baseline-characteristics table reports;
* **an R-vine copula per study** — a full joint distribution with its own
  margins and dependence structure, which is what you would fit if you had access
  to the individual covariate records even though the outcomes remain aggregate.

Each is run through both estimators, and each estimator is run in its ordinary
form and with covariate reweighting enabled.

```{r libs}
library(admixr2)
library(rxode2)
has_vine <- requireNamespace("rvinecopulib", quietly = TRUE) &&
            requireNamespace("kde1d", quietly = TRUE)
```

# The three published models

Weight acts on clearance, which carries the between-subject random effect. The
three models differ in structure, in their parameter values, and in the designs
they were developed under — a single dose with rich sampling, a multiple-dose
regimen sampled at steady state, and a sparse two-sample design.

```{r models}
mod_A <- function() {          # rich single dose
  ini({lcl <- log(3.2); lv <- log(21); clwt <- 0.75; clcr <- 0.45
       eta.cl ~ 0.09; prop.err <- 0.13})
  model({cl <- exp(lcl + eta.cl) * (WT / 70)^clwt * (CRCL / 95)^clcr
         v  <- exp(lv)
         d/dt(centr) <- -cl / v * centr
         cp <- centr / v
         cp ~ prop(prop.err)})
}
mod_B <- function() {          # multiple dose, steady state
  ini({lcl <- log(2.6); lv <- log(25); clwt <- 0.68; clcr <- 0.38
       eta.cl ~ 0.055; prop.err <- 0.16})
  model({cl <- exp(lcl + eta.cl) * (WT / 70)^clwt * (CRCL / 95)^clcr
         v  <- exp(lv)
         d/dt(centr) <- -cl / v * centr
         cp <- centr / v
         cp ~ prop(prop.err)})
}
mod_C <- function() {          # sparse
  ini({lcl <- log(3.9); lv <- log(18); clwt <- 0.82; clcr <- 0.52
       eta.cl ~ 0.12; prop.err <- 0.11})
  model({cl <- exp(lcl + eta.cl) * (WT / 70)^clwt * (CRCL / 95)^clcr
         v  <- exp(lv)
         d/dt(centr) <- -cl / v * centr
         cp <- centr / v
         cp ~ prop(prop.err)})
}
DESIGN <- list(
  A = list(model = mod_A, times = c(0.5, 1, 2, 4, 6, 8, 12),
           ev = rxode2::et(amt = 100, dur = 0.5), n = 48L),
  B = list(model = mod_B, times = c(1, 3, 6, 12),
           ev = rxode2::et(amt = 80, ii = 12, addl = 4, dur = 0.5), n = 96L),
  C = list(model = mod_C, times = c(2, 8),
           ev = rxode2::et(amt = 120, dur = 0.5), n = 30L))
```

# Version 1: the covariate as summary statistics

Each study reports a mean and standard deviation of weight. This is the ordinary
case, and it is all a baseline-characteristics table gives you.

```{r covsummary}
## A baseline table gives a mean and SD per covariate and a correlation between
## them. That correlation only reaches the likelihood if the covariates are drawn
## JOINTLY, so it is expressed as a Gaussian copula over the two margins -- the
## simplest joint distribution consistent with what was reported.
## Both margins are LOGNORMAL, matched to the reported mean and SD. A normal
## margin would be the obvious reading of "mean and standard deviation", but it
## puts mass at and below zero -- and (CRCL/95)^clcr with a negative base and a
## fractional exponent is NaN, which reaches the objective as a missing value
## rather than as an error. A positive covariate needs a positive margin.
ln_par <- function(m, s) c(meanlog = log(m^2 / sqrt(s^2 + m^2)),
                           sdlog   = sqrt(log(1 + s^2 / m^2)))
gauss_pair <- function(pw, pc, rho) function(u) {
  z  <- stats::qnorm(u)
  z2 <- rho * z[, 1] + sqrt(1 - rho^2) * z[, 2]
  cbind(WT   = stats::qlnorm(stats::pnorm(z[, 1]), pw[1], pw[2]),
        CRCL = stats::qlnorm(stats::pnorm(z2),     pc[1], pc[2]))
}
mk_summary <- function(m_wt, s_wt, m_cr, s_cr, rho) {
  pw <- ln_par(m_wt, s_wt); pc <- ln_par(m_cr, s_cr)
  list(WT    = list(quantile = function(u) stats::qlnorm(u, pw[1], pw[2])),
       CRCL  = list(quantile = function(u) stats::qlnorm(u, pc[1], pc[2])),
       joint = gauss_pair(pw, pc, rho))
}
WT_SUMMARY <- list(
  A = mk_summary(58, 13,  98, 22, 0.45),
  B = mk_summary(74, 23,  92, 26, 0.38),
  C = mk_summary(91, 16, 105, 19, 0.52))
```

# Version 0: independent margins

Before the two dependent versions, the case where the covariates are described
only by their own distributions and nothing links them. This is the one `adgh`
can also evaluate, because a product grid over independent margins is exactly
what it builds.

```{r covindep}
WT_INDEP <- lapply(list(c(58, 13, 98, 22), c(74, 23, 92, 26), c(91, 16, 105, 19)),
  function(p) {
    pw <- ln_par(p[1], p[2]); pc <- ln_par(p[3], p[4])
    list(WT   = list(meanlog = pw[[1]], sdlog = pw[[2]]),
         CRCL = list(meanlog = pc[[1]], sdlog = pc[[2]]))
  })
names(WT_INDEP) <- names(DESIGN)
```

# Version 2: the covariate as an R-vine copula

Real patient statistics are dependent: weight travels with height, and renal
function with age. An R-vine describes that joint distribution directly, with its
own margins and its own pair-copula for each edge of the tree — so the tails and
the asymmetries survive, which a correlation matrix cannot represent.

Each study gets its own vine, fitted the usual way: non-parametric marginals, the
probability-integral transform, then an unconstrained fit that selects the
structure from the data.

```{r covvine, eval = has_vine}
fit_vine <- function(seed, wt_med, wt_cv) {
  set.seed(seed); m <- 4000
  age  <- stats::rgamma(m, 9, 0.25)
  ht   <- 150 + 0.25 * age + stats::rnorm(m, 0, 9)
  wt   <- exp(log(wt_med) + 0.012 * (ht - 160) + wt_cv * stats::rnorm(m))
  crcl <- pmax(20, 130 - 0.9 * age + 0.35 * wt + stats::rnorm(m, 0, 12))
  X    <- cbind(WT = wt, HEIGHT = ht, AGE = age, CRCL = crcl)
  mg   <- lapply(seq_len(4), function(j) kde1d::kde1d(X[, j]))
  U    <- vapply(seq_len(4), function(j) kde1d::pkde1d(X[, j], mg[[j]]),
                 numeric(m))
  ## free R-vine: no structure imposed, families selected from the data
  list(vc = rvinecopulib::vinecop(U, family_set = "all"), mg = mg)
}
```

The sampler must be driven by admixr2's own uniforms, or the objective stops
being a deterministic function of the parameters and its finite differences
become noise. `rvinecop()` draws its own randomness and accepts no such input;
`inverse_rosenblatt()` maps supplied uniforms to vine-distributed ones and is
what makes this work. The two are distributionally indistinguishable — checked
in `validation/covariate-rvine-sampling-check.R`, where the mean log vine density
of the two agrees to 0.18 standard errors.

```{r vinesampler, eval = has_vine}
vine_sampler <- function(fitted) {
  vc <- fitted$vc; mg <- fitted$mg
  function(u) {
    ## admixr2 supplies one uniform column per DECLARED covariate (WT, CRCL);
    ## the vine is 4-dimensional, so the two nuisance dimensions are filled from
    ## the same stream by a fixed deterministic rotation -- still a function of
    ## `u` alone, which is what keeps the objective deterministic.
    uu <- cbind(u[, 1], (u[, 1] + u[, 2]) %% 1, (u[, 1] + 2 * u[, 2]) %% 1, u[, 2])
    v  <- rvinecopulib::inverse_rosenblatt(
      pmin(pmax(uu, 1e-9), 1 - 1e-9), vc)
    ## the vine is fitted over four covariates, but only the two the model
    ## reads are returned -- the others are in the vine because they carry the
    ## dependence that shapes these two, not because the model needs them
    cbind(WT   = kde1d::qkde1d(v[, 1], mg[[1]]),
          CRCL = kde1d::qkde1d(v[, 4], mg[[4]]))
  }
}
VINES <- list(A = fit_vine(11, 58, 0.22),
              B = fit_vine(12, 74, 0.31),
              C = fit_vine(13, 91, 0.18))
WT_VINE <- lapply(names(VINES), function(k) {
  mg <- VINES[[k]]$mg
  list(WT    = list(quantile = function(u) kde1d::qkde1d(u, mg[[1]])),
       CRCL  = list(quantile = function(u) kde1d::qkde1d(u, mg[[4]])),
       joint = vine_sampler(VINES[[k]]))
})
names(WT_VINE) <- names(VINES)
```


# Generating the aggregate data

```{r datagen}
make_studies <- function(cov_spec) {
  g <- datagen(
    setNames(lapply(names(DESIGN), function(k) {
      d <- DESIGN[[k]]
      list(model = d$model, times = d$times, ev = d$ev, n = d$n,
           cov_dist = cov_spec[[k]])
    }), names(DESIGN)),
    control = datagenControl(n_sim = 30000L, seed = 4321L))
  g
}
studies_summary <- make_studies(WT_SUMMARY)
str(lapply(studies_summary, function(s) round(s$E, 2)))
```

# The unifying model

One model is fitted to all three studies. It starts away from every generating
model, so nothing is being handed the answer.

```{r unify}
mod_unify <- function() {
  ini({lcl <- log(2.0); lv <- log(30); clwt <- 0.50; clcr <- 0.25
       eta.cl ~ 0.20; prop.err <- 0.25})
  model({cl <- exp(lcl + eta.cl) * (WT / 70)^clwt * (CRCL / 95)^clcr
         v  <- exp(lv)
         d/dt(centr) <- -cl / v * centr
         cp <- centr / v
         cp ~ prop(prop.err)})
}
```

# Fitting: two estimators, ordinary and reweighted

`adgh` integrates the covariate on a deterministic product grid; `admc`
integrates it by simulating subjects. `cov_reweight = TRUE` changes how that
integral is evaluated in both: the covariate is carried in an importance weight
rather than in the grid or the sample, so `adgh` drops the covariate dimension
from its grid and `admc` shares one simulation ensemble across studies that
differ only in their covariate distribution.

It is opt-in and self-declining. The substitution it relies on — that the
prediction depends on the covariate and the random effect only through their
sum — is a property of the model, and it is checked numerically once per fit
rather than inferred from how the model is written. Anything that fails the check
keeps the ordinary path, so the flag can only make a qualifying fit cheaper.

```{r fitfun}
fit_one <- function(studies, est, reweight) {
  ctl <- if (est == "adgh")
    adghControl(studies = studies, n_nodes = 7L, print = 0L,
                covMethod = "none", maxeval = 400L, cov_reweight = reweight)
  else
    admControl(studies = studies, n_sim = 3000L, print = 0L,
               covMethod = "none", maxeval = 400L, cov_reweight = reweight)
  t0 <- Sys.time()
  f  <- try(suppressWarnings(suppressMessages(
    nlmixr2est::nlmixr2(mod_unify, admData(), est = est, control = ctl))),
    silent = TRUE)
  dt <- as.numeric(difftime(Sys.time(), t0, units = "secs"))
  if (inherits(f, "try-error")) {
    ## Distinguish a DELIBERATE refusal from a failure. Labelling both "refused"
    ## would present a bug as a design decision, which is the more dangerous of
    ## the two ways to be wrong here.
    msg <- conditionMessage(attr(f, "condition"))
    st  <- if (grepl("does not support|cannot represent|joint", msg))
      "refused" else paste0("ERROR: ", substr(msg, 1, 90))
    return(data.frame(estimator = est, reweight = reweight, status = st,
                      lcl = NA, lv = NA, clwt = NA, clcr = NA,
                      objective = NA, secs = dt))
  }
  ax <- f$env$admExtra
  data.frame(estimator = est, reweight = reweight, status = "fitted",
             lcl = unname(ax$struct[["lcl"]]), lv = unname(ax$struct[["lv"]]),
             clwt = unname(ax$struct[["clwt"]]),
             clcr = unname(ax$struct[["clcr"]]),
             objective = f$objective, secs = dt)
}
```

## Version 0 — independent margins

```{r run0}
studies_indep <- make_studies(WT_INDEP)
grid1 <- expand.grid(est = c("adgh", "admc"), rw = c(FALSE, TRUE),
                     stringsAsFactors = FALSE)
res0 <- do.call(rbind, Map(function(e, r)
  fit_one(studies_indep, e, r), grid1$est, grid1$rw))
knitr::kable(res0, digits = 4, row.names = FALSE,
             caption = "Independent covariate margins. Both estimators apply, and reweighting reproduces the unweighted answer.")
```

## Version 1 — summary statistics

```{r run1}
res1 <- do.call(rbind, Map(function(e, r)
  fit_one(studies_summary, e, r), grid1$est, grid1$rw))
knitr::kable(res1, digits = 4, row.names = FALSE,
             caption = "Three studies described by summary statistics.")
```

## Version 2 — R-vine copulas

```{r run2, eval = has_vine}
studies_vine <- make_studies(WT_VINE)
res2 <- do.call(rbind, Map(function(e, r)
  fit_one(studies_vine, e, r), grid1$est, grid1$rw))
knitr::kable(res2, digits = 4, row.names = FALSE,
             caption = "The same three studies with an R-vine copula per study. `adgh` refuses a joint sampler by design: its covariate grid is a tensor product of per-covariate quadratures, which *is* the independence assumption, so it cannot represent a dependent distribution and declines rather than integrating the wrong one.")
```

# What to take from this

The estimates agree across every cell that runs. That is the point of the
comparison: the four combinations evaluate the same integral by different means,
so agreement is evidence that the integral, and not the machinery, is what
determines the answer.

Reweighting is a speed setting, not a modelling choice. It changes how the
covariate integral is evaluated and nothing about what is being estimated, which
is why the estimates match rather than merely resemble each other.

`adgh` and a dependent covariate distribution do not combine. That is a refusal
and not a limitation to work around: a product grid assumes independence, and
producing a number from it would be worse than declining. Use `admc` when the
covariates are dependent.

Finally, a covariate coefficient is only estimable here because the three studies
have *different* weight distributions. Three studies drawn from one population
would leave the likelihood exactly flat along a trade-off between the coefficient
and the between-subject variance, and any estimate reported from such a fit would
be the optimiser's stopping point rather than a result.
