---
title: "Covariates in a model-based meta-analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covariates in a model-based meta-analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.2)
source("_common.R", local = TRUE)
```

Three published population models are used as the input to a meta-analysis: each
generates the aggregate data its own study design and population would have
produced, and one unifying model is then fitted to all three at once. No
individual records exist at any point.

The covariate is the point of the exercise. Each study's subjects span a
distribution of body weights, and that distribution differs between studies —
which is not incidental, because between-study variation in the covariate
distribution is the only thing that identifies a covariate coefficient from
aggregate data. Two ways of describing those subjects are compared:

* **summary statistics** — a mean, a standard deviation and a correlation, which
  is what a baseline-characteristics table reports;
* **an R-vine copula per study** — a full joint distribution with its own
  margins and dependence structure, which is what you would fit if you had access
  to the individual covariate records even though the outcomes remain aggregate.

Each is run through both estimators, and each estimator is run in its ordinary
form and with covariate reweighting enabled.

```{r libs}
library(admixr2)
library(rxode2)
has_vine <- requireNamespace("rvinecopulib", quietly = TRUE)
```

# The three published models

Weight acts on clearance, which carries the between-subject random effect. The
three models differ in structure, in their parameter values, and in the designs
they were developed under — a single dose with rich sampling, a multiple-dose
regimen sampled at steady state, and a sparse two-sample design.

```{r models}
mod_A <- function() {          # rich single dose
  ini({lcl <- log(3.2); lv <- log(21); clwt <- 0.75
       eta.cl ~ 0.09; prop.err <- 0.13})
  model({cl <- exp(lcl + eta.cl) * (WT / 70)^clwt
         v  <- exp(lv)
         d/dt(centr) <- -cl / v * centr
         cp <- centr / v
         cp ~ prop(prop.err)})
}
mod_B <- function() {          # multiple dose, steady state
  ini({lcl <- log(2.6); lv <- log(25); clwt <- 0.68
       eta.cl ~ 0.055; prop.err <- 0.16})
  model({cl <- exp(lcl + eta.cl) * (WT / 70)^clwt
         v  <- exp(lv)
         d/dt(centr) <- -cl / v * centr
         cp <- centr / v
         cp ~ prop(prop.err)})
}
mod_C <- function() {          # sparse
  ini({lcl <- log(3.9); lv <- log(18); clwt <- 0.82
       eta.cl ~ 0.12; prop.err <- 0.11})
  model({cl <- exp(lcl + eta.cl) * (WT / 70)^clwt
         v  <- exp(lv)
         d/dt(centr) <- -cl / v * centr
         cp <- centr / v
         cp ~ prop(prop.err)})
}
DESIGN <- list(
  A = list(model = mod_A, times = c(0.5, 1, 2, 4, 6, 8, 12),
           ev = rxode2::et(amt = 100, dur = 0.5), n = 48L),
  B = list(model = mod_B, times = c(1, 3, 6, 12),
           ev = rxode2::et(amt = 80, ii = 12, addl = 4, dur = 0.5), n = 96L),
  C = list(model = mod_C, times = c(2, 8),
           ev = rxode2::et(amt = 120, dur = 0.5), n = 30L))
```

# Version 1: the covariate as summary statistics

Each study reports a mean and standard deviation of weight. This is the ordinary
case, and it is all a baseline-characteristics table gives you.

```{r covsummary}
WT_SUMMARY <- list(
  A = list(WT = list(meanlog = log(58), sdlog = 0.22)),
  B = list(WT = list(meanlog = log(74), sdlog = 0.31)),
  C = list(WT = list(meanlog = log(91), sdlog = 0.18)))
```

# Version 2: the covariate as an R-vine copula

Real patient statistics are dependent: weight travels with height and with age.
An R-vine describes that joint distribution directly. `cov_dist$joint` takes the
uniforms admixr2 draws and returns covariate values, which is exactly the shape a
vine produces — sample the vine on the uniform scale, then push each margin
through its own quantile function.

The uniforms have to come from admixr2's own stream (`U = u` below) rather than
from the sampler's internal randomness. That stream is what makes the objective a
deterministic function of the parameters; a sampler that reseeds itself turns the
objective into noise and the gradients with it.

```{r covvine, eval = has_vine}
vine_sampler <- function(meanlog, sdlog, rho) {
  ## a 3-dimensional vine over (weight, height, age); only weight enters the
  ## model, but it is drawn jointly with the others rather than in isolation
  str <- rvinecopulib::rvine_structure(order = 1:3)
  pcs <- list(
    list(rvinecopulib::bicop_dist("gaussian", parameters = rho),
         rvinecopulib::bicop_dist("clayton",  parameters = 1.2)),
    list(rvinecopulib::bicop_dist("gaussian", parameters = 0.3)))
  vc  <- rvinecopulib::vinecop_dist(pcs, str)
  function(u) {
    v <- rvinecopulib::rvinecop(nrow(u), vc, U = u)
    cbind(WT     = stats::qlnorm(v[, 1], meanlog, sdlog),
          HEIGHT = stats::qnorm( v[, 2], 150, 12),
          AGE    = stats::qgamma(v[, 3], 9, 0.25))
  }
}
WT_VINE <- list(
  A = list(WT = list(quantile = function(u) stats::qlnorm(u, log(58), 0.22)),
           joint = vine_sampler(log(58), 0.22, 0.70)),
  B = list(WT = list(quantile = function(u) stats::qlnorm(u, log(74), 0.31)),
           joint = vine_sampler(log(74), 0.31, 0.55)),
  C = list(WT = list(quantile = function(u) stats::qlnorm(u, log(91), 0.18)),
           joint = vine_sampler(log(91), 0.18, 0.65)))
```

The model reads only `WT`, so the vine's other margins do not enter the
prediction — they are there because the weight distribution is a margin of a
joint distribution, and drawing it from that joint is what makes the marginal
draws right when the dependence matters.

# Generating the aggregate data

```{r datagen}
make_studies <- function(cov_spec) {
  g <- datagen(
    setNames(lapply(names(DESIGN), function(k) {
      d <- DESIGN[[k]]
      list(model = d$model, times = d$times, ev = d$ev, n = d$n,
           cov_dist = cov_spec[[k]])
    }), names(DESIGN)),
    control = datagenControl(n_sim = 30000L, seed = 4321L))
  g
}
studies_summary <- make_studies(WT_SUMMARY)
str(lapply(studies_summary, function(s) round(s$E, 2)))
```

# The unifying model

One model is fitted to all three studies. It starts away from every generating
model, so nothing is being handed the answer.

```{r unify}
mod_unify <- function() {
  ini({lcl <- log(2.0); lv <- log(30); clwt <- 0.50
       eta.cl ~ 0.20; prop.err <- 0.25})
  model({cl <- exp(lcl + eta.cl) * (WT / 70)^clwt
         v  <- exp(lv)
         d/dt(centr) <- -cl / v * centr
         cp <- centr / v
         cp ~ prop(prop.err)})
}
```

# Fitting: two estimators, ordinary and reweighted

`adgh` integrates the covariate on a deterministic product grid; `admc`
integrates it by simulating subjects. `cov_reweight = TRUE` changes how that
integral is evaluated in both: the covariate is carried in an importance weight
rather than in the grid or the sample, so `adgh` drops the covariate dimension
from its grid and `admc` shares one simulation ensemble across studies that
differ only in their covariate distribution.

It is opt-in and self-declining. The substitution it relies on — that the
prediction depends on the covariate and the random effect only through their
sum — is a property of the model, and it is checked numerically once per fit
rather than inferred from how the model is written. Anything that fails the check
keeps the ordinary path, so the flag can only make a qualifying fit cheaper.

```{r fitfun}
fit_one <- function(studies, est, reweight) {
  ctl <- if (est == "adgh")
    adghControl(studies = studies, n_nodes = 7L, print = 0L,
                covMethod = "none", maxeval = 400L, cov_reweight = reweight)
  else
    admControl(studies = studies, n_sim = 3000L, print = 0L,
               covMethod = "none", maxeval = 400L, cov_reweight = reweight)
  t0 <- Sys.time()
  f  <- try(suppressWarnings(suppressMessages(
    nlmixr2est::nlmixr2(mod_unify, admData(), est = est, control = ctl))),
    silent = TRUE)
  dt <- as.numeric(difftime(Sys.time(), t0, units = "secs"))
  if (inherits(f, "try-error"))
    return(data.frame(estimator = est, reweight = reweight, status = "refused",
                      lcl = NA, lv = NA, clwt = NA, objective = NA, secs = dt))
  ax <- f$env$admExtra
  data.frame(estimator = est, reweight = reweight, status = "fitted",
             lcl = unname(ax$struct[["lcl"]]), lv = unname(ax$struct[["lv"]]),
             clwt = unname(ax$struct[["clwt"]]),
             objective = f$objective, secs = dt)
}
```

## Version 1 — summary statistics

```{r run1}
grid1 <- expand.grid(est = c("adgh", "admc"), rw = c(FALSE, TRUE),
                     stringsAsFactors = FALSE)
res1 <- do.call(rbind, Map(function(e, r)
  fit_one(studies_summary, e, r), grid1$est, grid1$rw))
knitr::kable(res1, digits = 4, row.names = FALSE,
             caption = "Three studies described by summary statistics.")
```

## Version 2 — R-vine copulas

```{r run2, eval = has_vine}
studies_vine <- make_studies(WT_VINE)
res2 <- do.call(rbind, Map(function(e, r)
  fit_one(studies_vine, e, r), grid1$est, grid1$rw))
knitr::kable(res2, digits = 4, row.names = FALSE,
             caption = "The same three studies with an R-vine copula per study. `adgh` refuses a joint sampler by design: its covariate grid is a tensor product of per-covariate quadratures, which *is* the independence assumption, so it cannot represent a dependent distribution and declines rather than integrating the wrong one.")
```

# What to take from this

The estimates agree across every cell that runs. That is the point of the
comparison: the four combinations evaluate the same integral by different means,
so agreement is evidence that the integral, and not the machinery, is what
determines the answer.

Reweighting is a speed setting, not a modelling choice. It changes how the
covariate integral is evaluated and nothing about what is being estimated, which
is why the estimates match rather than merely resemble each other.

`adgh` and a dependent covariate distribution do not combine. That is a refusal
and not a limitation to work around: a product grid assumes independence, and
producing a number from it would be worse than declining. Use `admc` when the
covariates are dependent.

Finally, a covariate coefficient is only estimable here because the three studies
have *different* weight distributions. Three studies drawn from one population
would leave the likelihood exactly flat along a trade-off between the coefficient
and the between-subject variance, and any estimate reported from such a fit would
be the optimiser's stopping point rather than a result.
