---
title: "Covariates in a model-based meta-analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covariates in a model-based meta-analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.2)
source("_common.R", local = TRUE)
```

Three published population models are used as the input to a meta-analysis: each
generates the aggregate data its own study design and population would have
produced, and one unifying model is then fitted to all three at once. No
individual records exist at any point.

The covariate is the point of the exercise. Each study's subjects span a
distribution of body weights, and that distribution differs between studies —
which is not incidental, because between-study variation in the covariate
distribution is the only thing that identifies a covariate coefficient from
aggregate data. Two ways of describing those subjects are compared:

* **summary statistics** — a mean, a standard deviation and a correlation, which
  is what a baseline-characteristics table reports;
* **an R-vine copula per study** — a full joint distribution with its own
  margins and dependence structure, which is what you would fit if you had access
  to the individual covariate records even though the outcomes remain aggregate.

Each is run through both estimators, and each estimator is run in its ordinary
form and with covariate reweighting enabled.

```{r libs}
library(admixr2)
library(rxode2)
has_vine <- requireNamespace("rvinecopulib", quietly = TRUE) &&
            requireNamespace("kde1d", quietly = TRUE)
```

# The three published models

Weight and creatinine clearance both act on clearance, which carries the
between-subject random effect. The three models differ in the designs they were
developed under --- a single dose with rich sampling, a multiple-dose regimen at
steady state, and a sparse two-sample design --- and, more importantly, in the
populations they enrolled: one with preserved renal function, one with moderate
impairment, one with severe.

Their parameter estimates differ modestly, as published models of one drug
usually do --- clearance within 6%, volume within 5%, random-effect and residual
magnitudes within about 10%. That agreement matters more than it looks, and Act
III is about why: the unifying model has ONE of each of those parameters, so
every disagreement between the studies has to go somewhere, and the covariate
slopes are where it goes.

```{r models}
mod_A <- function() {          # rich single dose, preserved renal function
  ini({lcl <- log(3.1); lv <- log(20.5); clwt <- 0.72; clcr <- 0.40
       eta.cl ~ 0.085; prop.err <- 0.12})
  model({cl <- exp(lcl + eta.cl) * (WT / 70)^clwt * (CRCL / 95)^clcr
         v  <- exp(lv)
         d/dt(centr) <- -cl / v * centr
         cp <- centr / v
         cp ~ prop(prop.err)})
}
mod_B <- function() {          # multiple dose at steady state, moderate impairment
  ini({lcl <- log(3.2); lv <- log(21.0); clwt <- 0.75; clcr <- 0.45
       eta.cl ~ 0.090; prop.err <- 0.13})
  model({cl <- exp(lcl + eta.cl) * (WT / 70)^clwt * (CRCL / 95)^clcr
         v  <- exp(lv)
         d/dt(centr) <- -cl / v * centr
         cp <- centr / v
         cp ~ prop(prop.err)})
}
mod_C <- function() {          # sparse, severe impairment
  ini({lcl <- log(3.3); lv <- log(21.5); clwt <- 0.78; clcr <- 0.50
       eta.cl ~ 0.095; prop.err <- 0.14})
  model({cl <- exp(lcl + eta.cl) * (WT / 70)^clwt * (CRCL / 95)^clcr
         v  <- exp(lv)
         d/dt(centr) <- -cl / v * centr
         cp <- centr / v
         cp ~ prop(prop.err)})
}
DESIGN <- list(
  A = list(model = mod_A, times = c(0.5, 1, 2, 4, 6, 8, 12),
           ev = rxode2::et(amt = 100, dur = 0.5), n = 48L),
  B = list(model = mod_B, times = c(1, 3, 6, 12),
           ev = rxode2::et(amt = 80, ii = 12, addl = 4, dur = 0.5), n = 96L),
  C = list(model = mod_C, times = c(2, 8),
           ev = rxode2::et(amt = 120, dur = 0.5), n = 30L))
```

# Version 1: the covariate as summary statistics

Each study reports a mean and standard deviation of weight. This is the ordinary
case, and it is all a baseline-characteristics table gives you.

```{r covsummary}
## A baseline table gives a mean and SD per covariate and a correlation between
## them. That correlation only reaches the likelihood if the covariates are drawn
## JOINTLY, so it is expressed as a Gaussian copula over the two margins -- the
## simplest joint distribution consistent with what was reported.
## Both margins are LOGNORMAL, matched to the reported mean and SD. A normal
## margin would be the obvious reading of "mean and standard deviation", but it
## puts mass at and below zero -- and (CRCL/95)^clcr with a negative base and a
## fractional exponent is NaN, which reaches the objective as a missing value
## rather than as an error. A positive covariate needs a positive margin.
ln_par <- function(m, s) c(meanlog = log(m^2 / sqrt(s^2 + m^2)),
                           sdlog   = sqrt(log(1 + s^2 / m^2)))
gauss_pair <- function(pw, pc, rho) function(u) {
  z  <- stats::qnorm(u)
  z2 <- rho * z[, 1] + sqrt(1 - rho^2) * z[, 2]
  cbind(WT   = stats::qlnorm(stats::pnorm(z[, 1]), pw[1], pw[2]),
        CRCL = stats::qlnorm(stats::pnorm(z2),     pc[1], pc[2]))
}
mk_summary <- function(m_wt, s_wt, m_cr, s_cr, rho) {
  pw <- ln_par(m_wt, s_wt); pc <- ln_par(m_cr, s_cr)
  list(WT    = list(quantile = function(u) stats::qlnorm(u, pw[1], pw[2])),
       CRCL  = list(quantile = function(u) stats::qlnorm(u, pc[1], pc[2])),
       joint = gauss_pair(pw, pc, rho))
}
WT_SUMMARY <- list(
  A = mk_summary(55, 12, 120, 26, 0.45),
  B = mk_summary(95, 20,  55, 14, 0.38),
  C = mk_summary(72, 16,  28,  8, 0.52))
```

# Version 0: independent margins

Before the two dependent versions, the case where the covariates are described
only by their own distributions and nothing links them. This is the one `adgh`
can also evaluate, because a product grid over independent margins is exactly
what it builds.

```{r covindep}
WT_INDEP <- lapply(list(c(55, 12, 120, 26), c(95, 20, 55, 14), c(72, 16, 28, 8)),
  function(p) {
    pw <- ln_par(p[1], p[2]); pc <- ln_par(p[3], p[4])
    list(WT   = list(meanlog = pw[[1]], sdlog = pw[[2]]),
         CRCL = list(meanlog = pc[[1]], sdlog = pc[[2]]))
  })
names(WT_INDEP) <- names(DESIGN)
```

# Version 2: the covariate as an R-vine copula

Real patient statistics are dependent: weight travels with height, and renal
function with age. An R-vine describes that joint distribution directly, with its
own margins and its own pair-copula for each edge of the tree — so the tails and
the asymmetries survive, which a correlation matrix cannot represent.

Each study gets its own vine, fitted the usual way: non-parametric marginals, the
probability-integral transform, then an unconstrained fit that selects the
structure from the data.

```{r covvine, eval = has_vine}
fit_vine <- function(seed) {
  set.seed(seed); m <- 4000
  ## an individual covariate record, of the kind a vine would be fitted to
  age  <- stats::rgamma(m, 9, 0.25)
  ht   <- 150 + 0.25 * age + stats::rnorm(m, 0, 9)
  wt   <- exp(log(70) + 0.012 * (ht - 160) + 0.25 * stats::rnorm(m))
  crcl <- pmax(20, 130 - 0.9 * age + 0.35 * wt + stats::rnorm(m, 0, 12))
  U    <- apply(cbind(wt, ht, age, crcl), 2L,
                function(z) rank(z) / (length(z) + 1))
  ## free R-vine: no structure imposed, families selected from the data. Only
  ## the COPULA is kept -- the margins come from the reported summaries.
  list(vc = rvinecopulib::vinecop(U, family_set = "all"))
}
```

The sampler must be driven by admixr2's own uniforms, or the objective stops
being a deterministic function of the parameters and its finite differences
become noise. `rvinecop()` draws its own randomness and accepts no such input;
`inverse_rosenblatt()` maps supplied uniforms to vine-distributed ones and is
what makes this work. The two are distributionally indistinguishable — checked
in `validation/covariate-rvine-sampling-check.R`, where the mean log vine density
of the two agrees to 0.18 standard errors.

```{r vinesampler, eval = has_vine}
vine_sampler <- function(vc, pw, pc) {
  function(u) {
    ## admixr2 supplies one uniform column per DECLARED covariate (WT, CRCL);
    ## the vine is 4-dimensional, so the two nuisance dimensions are filled from
    ## the same stream by a fixed deterministic rotation -- still a function of
    ## `u` alone, which is what keeps the objective deterministic.
    uu <- cbind(u[, 1], (u[, 1] + u[, 2]) %% 1, (u[, 1] + 2 * u[, 2]) %% 1, u[, 2])
    v  <- rvinecopulib::inverse_rosenblatt(
      pmin(pmax(uu, 1e-9), 1 - 1e-9), vc)
    ## the vine is fitted over four covariates, but only the two the model
    ## reads are returned -- the others are in the vine because they carry the
    ## dependence that shapes these two, not because the model needs them
    cbind(WT   = stats::qlnorm(v[, 1], pw[1], pw[2]),
          CRCL = stats::qlnorm(v[, 4], pc[1], pc[2]))
  }
}
## The vine supplies the DEPENDENCE; the margins are the same lognormals version
## 1 reports. That is what makes the two versions comparable -- they describe the
## same patients and differ only in how the two covariates are linked.
##
## Fitting the vine to synthetic data with its own marginal scales instead was
## the first attempt, and it silently changed the population: study B's
## creatinine clearance came out centred at 124 rather than the reported 92, so
## the two versions were fitting different patients and their coefficients were
## never comparable. Same margins, different dependence.
VINES <- list(A = fit_vine(11), B = fit_vine(12), C = fit_vine(13))
PARS  <- list(A = c(55, 12, 120, 26), B = c(95, 20, 55, 14),
              C = c(72, 16,  28,  8))
WT_VINE <- lapply(names(VINES), function(k) {
  vc <- VINES[[k]]$vc
  pw <- ln_par(PARS[[k]][1], PARS[[k]][2])
  pc <- ln_par(PARS[[k]][3], PARS[[k]][4])
  list(WT    = list(quantile = function(u) stats::qlnorm(u, pw[1], pw[2])),
       CRCL  = list(quantile = function(u) stats::qlnorm(u, pc[1], pc[2])),
       joint = vine_sampler(vc, pw, pc))
})
names(WT_VINE) <- names(VINES)
```


# Generating the aggregate data

```{r datagen}
make_studies <- function(cov_spec) {
  g <- datagen(
    setNames(lapply(names(DESIGN), function(k) {
      d <- DESIGN[[k]]
      list(model = d$model, times = d$times, ev = d$ev, n = d$n,
           cov_dist = cov_spec[[k]])
    }), names(DESIGN)),
    control = datagenControl(n_sim = 30000L, seed = 4321L))
  g
}
studies_summary <- make_studies(WT_SUMMARY)
str(lapply(studies_summary, function(s) round(s$E, 2)))
```

# The unifying model

One model is fitted to all three studies. It starts away from every generating
model, so nothing is being handed the answer.

```{r unify}
mod_unify <- function() {
  ini({lcl <- log(2.0); lv <- log(30); clwt <- 0.50; clcr <- 0.25
       eta.cl ~ 0.20; prop.err <- 0.25})
  model({cl <- exp(lcl + eta.cl) * (WT / 70)^clwt * (CRCL / 95)^clcr
         v  <- exp(lv)
         d/dt(centr) <- -cl / v * centr
         cp <- centr / v
         cp ~ prop(prop.err)})
}
```

# Fitting: two estimators, ordinary and reweighted

`adgh` integrates the covariate on a deterministic product grid; `admc`
integrates it by simulating subjects. `cov_reweight = TRUE` changes how that
integral is evaluated in both: the covariate is carried in an importance weight
rather than in the grid or the sample, so `adgh` drops the covariate dimension
from its grid and `admc` shares one simulation ensemble across studies that
differ only in their covariate distribution.

It is opt-in and self-declining. The substitution it relies on — that the
prediction depends on the covariate and the random effect only through their
sum — is a property of the model, and it is checked numerically once per fit
rather than inferred from how the model is written. Anything that fails the check
keeps the ordinary path, so the flag can only make a qualifying fit cheaper.

```{r fitfun}
fit_one <- function(studies, est, reweight) {
  ctl <- if (est == "adgh")
    adghControl(studies = studies, n_nodes = 7L, print = 0L,
                covMethod = "none", maxeval = 400L, cov_reweight = reweight)
  else
    admControl(studies = studies, n_sim = 3000L, print = 0L,
               covMethod = "none", maxeval = 400L, cov_reweight = reweight)
  t0 <- Sys.time()
  f  <- try(suppressWarnings(suppressMessages(
    nlmixr2est::nlmixr2(mod_unify, admData(), est = est, control = ctl))),
    silent = TRUE)
  dt <- as.numeric(difftime(Sys.time(), t0, units = "secs"))
  if (inherits(f, "try-error")) {
    ## Distinguish a DELIBERATE refusal from a failure. Labelling both "refused"
    ## would present a bug as a design decision, which is the more dangerous of
    ## the two ways to be wrong here.
    msg <- conditionMessage(attr(f, "condition"))
    st  <- if (grepl("does not support|cannot represent|joint", msg))
      "refused" else paste0("ERROR: ", substr(msg, 1, 90))
    return(data.frame(estimator = est, reweight = reweight, status = st,
                      lcl = NA, lv = NA, clwt = NA, clcr = NA,
                      objective = NA, secs = dt))
  }
  ax <- f$env$admExtra
  data.frame(estimator = est, reweight = reweight, status = "fitted",
             lcl = unname(ax$struct[["lcl"]]), lv = unname(ax$struct[["lv"]]),
             clwt = unname(ax$struct[["clwt"]]),
             clcr = unname(ax$struct[["clcr"]]),
             objective = f$objective, secs = dt)
}
```

## Version 0 — independent margins

```{r run0}
studies_indep <- make_studies(WT_INDEP)
grid1 <- expand.grid(est = c("adgh", "admc"), rw = c(FALSE, TRUE),
                     stringsAsFactors = FALSE)
res0 <- do.call(rbind, Map(function(e, r)
  fit_one(studies_indep, e, r), grid1$est, grid1$rw))
knitr::kable(res0, digits = 4, row.names = FALSE,
             caption = "Independent covariate margins. Both estimators apply, and reweighting reproduces the unweighted answer.")
```

## Version 1 — summary statistics

```{r run1}
res1 <- do.call(rbind, Map(function(e, r)
  fit_one(studies_summary, e, r), grid1$est, grid1$rw))
knitr::kable(res1, digits = 4, row.names = FALSE,
             caption = "Three studies described by summary statistics.")
```

## Version 2 — R-vine copulas

```{r run2, eval = has_vine}
studies_vine <- make_studies(WT_VINE)
res2 <- do.call(rbind, Map(function(e, r)
  fit_one(studies_vine, e, r), grid1$est, grid1$rw))
knitr::kable(res2, digits = 4, row.names = FALSE,
             caption = "The same three studies with an R-vine copula per study. `adgh` refuses a joint sampler by design: its covariate grid is a tensor product of per-covariate quadratures, which *is* the independence assumption, so it cannot represent a dependent distribution and declines rather than integrating the wrong one.")
```

# Act I: what the analysis is actually given

The three studies have to differ in their covariate distributions for the
coefficients to be estimable at all, and the two dependent versions have to
actually carry dependence. Both are worth seeing rather than asserting.

```{r covplots, fig.width = 7.5, fig.height = 7.5, echo = FALSE}
draw_cov <- function(spec, n = 4000L) {
  u <- randtoolbox::sobol(n, dim = 2)
  u <- pmin(pmax(u, 1e-9), 1 - 1e-9)
  if (is.function(spec$joint)) return(spec$joint(u))
  cbind(WT   = stats::qlnorm(u[, 1], spec$WT$meanlog,   spec$WT$sdlog),
        CRCL = stats::qlnorm(u[, 2], spec$CRCL$meanlog, spec$CRCL$sdlog))
}
SPECS <- list("independent" = WT_INDEP, "correlation" = WT_SUMMARY)
if (has_vine) SPECS[["R-vine"]] <- WT_VINE
cols <- c(A = "#1b6ca8", B = "#c1121f", C = "#2a9d8f")

op <- par(mfrow = c(3, 2), mar = c(4, 4, 2.2, 1), mgp = c(2.3, 0.7, 0))

## row 1: the marginal distributions, which differ BETWEEN studies -- the only
## reason a covariate coefficient is identifiable from aggregate data
for (v in c("WT", "CRCL")) {
  X <- lapply(names(DESIGN), function(k) draw_cov(WT_INDEP[[k]])[, v])
  xl <- range(unlist(X))
  plot(NA, xlim = xl, ylim = c(0, max(vapply(X, function(z)
        max(stats::density(z)$y), 0)) * 1.05), bty = "l",
       xlab = v, ylab = "density",
       main = sprintf("%s by study", v))
  for (i in seq_along(X)) lines(stats::density(X[[i]]), col = cols[i], lwd = 2)
  legend("topright", names(DESIGN), col = cols, lwd = 2, bty = "n", cex = 0.9)
}

## row 2: dependence. Independent margins give a cloud with no tilt; the
## Gaussian copula tilts it; the vine tilts it AND shapes the tails.
for (nm in names(SPECS)[1:2]) {
  Z <- draw_cov(SPECS[[nm]][["B"]], 3000L)
  plot(Z[, "WT"], Z[, "CRCL"], pch = 16, cex = 0.25,
       col = grDevices::adjustcolor("#333333", 0.35), bty = "l",
       xlab = "WT", ylab = "CRCL",
       main = sprintf("study B: %s (r = %.2f)", nm, cor(Z[, 1], Z[, 2])))
}

## row 3: the vine, and the aggregate data the whole analysis actually sees
if (has_vine) {
  Z <- draw_cov(WT_VINE[["B"]], 3000L)
  plot(Z[, "WT"], Z[, "CRCL"], pch = 16, cex = 0.25,
       col = grDevices::adjustcolor("#333333", 0.35), bty = "l",
       xlab = "WT", ylab = "CRCL",
       main = sprintf("study B: R-vine (r = %.2f)", cor(Z[, 1], Z[, 2])))
} else plot.new()

rng <- range(unlist(lapply(studies_indep, function(s)
  c(s$E - sqrt(diag(as.matrix(s$V))), s$E + sqrt(diag(as.matrix(s$V)))))))
plot(NA, xlim = c(0, 13), ylim = rng, bty = "l", xlab = "time (h)",
     ylab = "concentration", main = "aggregate data (mean +- SD)")
for (i in seq_along(studies_indep)) {
  st <- studies_indep[[i]]; sd_i <- sqrt(diag(as.matrix(st$V)))
  arrows(st$times, st$E - sd_i, st$times, st$E + sd_i, code = 3, angle = 90,
         length = 0.03, col = cols[i])
  points(st$times, st$E, pch = 16, col = cols[i]); lines(st$times, st$E, col = cols[i])
}
legend("topright", names(DESIGN), col = cols, lwd = 2, bty = "n", cex = 0.9)
invisible(par(op))
```

The marginals differ between studies, which is what makes the coefficients
estimable; the independent version shows no tilt, and both dependent versions do.

# Act II: what the fits concluded

Each fit is summarised by what it says about clearance: how it scales with weight
at a typical creatinine clearance, and with creatinine clearance at a typical
weight. The three generating models are drawn for comparison — there is no single
truth here, since the unifying model is one structure reconciling three, so the
generating curves bracket rather than define the answer.

```{r effectplot, fig.width = 7.5, fig.height = 3.6, echo = FALSE}
gen <- list(A = c(clwt = 0.72, clcr = 0.40),
            B = c(clwt = 0.75, clcr = 0.45),
            C = c(clwt = 0.78, clcr = 0.50))
FITS <- list()
pick <- function(df) { d <- df[df$status == "fitted", ]; if (nrow(d)) d[1, ] else NULL }
if (!is.null(pick(res0))) FITS[["independent"]] <- pick(res0)
if (!is.null(pick(res1))) FITS[["correlation"]] <- pick(res1)
if (exists("res2") && !is.null(pick(res2))) FITS[["R-vine"]] <- pick(res2)
fcol <- c(independent = "#1b6ca8", correlation = "#c1121f", `R-vine` = "#2a9d8f")

op <- par(mfrow = c(1, 2), mar = c(4, 4.2, 2.2, 1), mgp = c(2.4, 0.7, 0))
wt <- seq(40, 120, length.out = 100); cr <- seq(20, 150, length.out = 100)
for (which_v in c("WT", "CRCL")) {
  x  <- if (which_v == "WT") wt else cr
  ref <- if (which_v == "WT") 70 else 95
  plot(NA, xlim = range(x), ylim = c(0.4, 2.2), bty = "l", xlab = which_v,
       ylab = "clearance relative to a typical patient",
       main = sprintf("effect of %s", which_v))
  abline(h = 1, col = "grey80")
  for (k in names(gen)) {
    e <- gen[[k]][[if (which_v == "WT") "clwt" else "clcr"]]
    lines(x, (x / ref)^e, col = "grey55", lty = 2)
  }
  for (nm in names(FITS)) {
    e <- FITS[[nm]][[if (which_v == "WT") "clwt" else "clcr"]]
    lines(x, (x / ref)^e, col = fcol[nm], lwd = 2.4)
  }
  if (which_v == "WT")
    legend("topleft", c(names(FITS), "generating models"),
           col = c(fcol[names(FITS)], "grey55"),
           lwd = c(rep(2.4, length(FITS)), 1), lty = c(rep(1, length(FITS)), 2),
           bty = "n", cex = 0.85)
}
invisible(par(op))
```

```{r effectsummary, echo = FALSE}
.f <- function(d) { r <- d[d$status == "fitted", ]; if (nrow(r)) r[1, ] else NULL }
.rows <- Filter(Negate(is.null), list(independent = .f(res0), correlation = .f(res1),
                                      `R-vine` = if (exists("res2")) .f(res2)))
.tab <- do.call(rbind, lapply(names(.rows), function(k) data.frame(
  description = k, clwt = .rows[[k]]$clwt, clcr = .rows[[k]]$clcr)))
.tab$`clwt vs 0.75` <- sprintf("%+.0f%%", 100 * (.tab$clwt / 0.75 - 1))
.tab$`clcr vs 0.45` <- sprintf("%+.0f%%", 100 * (.tab$clcr / 0.45 - 1))
knitr::kable(.tab, digits = 3, row.names = FALSE,
             caption = "Estimated covariate exponents against the average of the three generating models.")
```

Both coefficients come back close to the models that generated them, and --- more
usefully --- they barely move between the three covariate descriptions. How the
covariates are described, and whether their dependence is modelled at all, shifts
the estimates by less than the generating models differ from each other.

That is the robustness result. It is also not automatic, and the next section is
about what makes it hold.

# Act III: why this works, and when it would not

The coefficients come back close to the models that generated them, which is not
automatic --- it is a property of this study set, and it is worth knowing which
property.

A covariate slope is learned two ways, and they are very unequal. *Within* a
study it enters through the variance, because the spread of weights and renal
function among that study's subjects inflates the observed covariance. That is
weak information. *Between* studies it enters through the mean levels, which is
strong --- but it is also exactly where a difference in baseline clearance would
land. Three studies and three parameters means the between-study system is
exactly determined, so any baseline disagreement is absorbed by the slopes
rather than resisted.

What protects this analysis is that the covariate contrast is far larger than the
disagreement it has to survive. Creatinine clearance runs from 28 to 120 across
the three studies --- a factor of four --- while the models themselves agree to
within about 6%.

And "disagreement" means EVERY parameter the unifying model collapses to one, not
just baseline clearance. Volume, the random-effect magnitude and the residual
error are equally between-study quantities with a single value in the unifying
model, so a volume disagreement loads onto the covariate slopes exactly as a
clearance disagreement does. An earlier version of this vignette had the three
models differing by a third in volume while agreeing to 6% on clearance, and that
alone put both exponents about 50% high --- with the renal contrast already in
place. The quantity that matters is total disagreement against covariate
leverage, and it is easy to control for the parameter you happen to be thinking
about while leaving the others free.

Take that away and the same machinery gives a different answer. The right-hand
panel below is the same three designs with creatinine clearance varying by 13%
between studies instead of fourfold, and baseline clearances spread over 50%
instead of 6%. Every study still has a slope near 0.45, and the fitted slope is
1.41.

```{r ecological, fig.width = 7.5, fig.height = 3.8, echo = FALSE}
GEN <- list(A = c(lcl = log(3.1), clwt = 0.72, clcr = 0.40),
            B = c(lcl = log(3.2), clwt = 0.75, clcr = 0.45),
            C = c(lcl = log(3.3), clwt = 0.78, clcr = 0.50))
MEAN_WT <- c(A = 55, B = 95, C = 72); MEAN_CR <- c(A = 120, B = 55, C = 28)
SD_CR   <- c(A = 26, B = 14, C = 8)
## the cautionary set: baselines spread over 50% instead of 6%, and creatinine
## clearance varying by 13% between studies instead of fourfold
BAD     <- list(A = c(lcl = log(3.2), clwt = 0.75, clcr = 0.45),
                B = c(lcl = log(2.6), clwt = 0.68, clcr = 0.38),
                C = c(lcl = log(3.9), clwt = 0.82, clcr = 0.52))
BAD_WT  <- c(A = 58, B = 74, C = 91); BAD_CR <- c(A = 98, B = 92, C = 105)
BAD_SD  <- c(A = 22, B = 26, C = 19)
cols <- c(A = "#1b6ca8", B = "#c1121f", C = "#2a9d8f")

panel <- function(G, MW, MC, SD, fitted, ttl) {
  xr <- range(unlist(lapply(names(G), function(k)
    MC[k] + c(-1.2, 1.2) * SD[k])))
  yr <- range(unlist(lapply(names(G), function(k) {
    pk <- G[[k]]
    pk[["lcl"]] + pk[["clwt"]] * log(MW[k] / 70) +
      pk[["clcr"]] * log((MC[k] + c(-1.2, 1.2) * SD[k]) / 95)
  })))
  plot(NA, xlim = xr, ylim = yr + c(-0.15, 0.15), bty = "l", log = "x",
       xlab = "creatinine clearance", ylab = "log clearance", main = ttl)
  for (k in names(G)) {
    pk <- G[[k]]
    x  <- seq(MC[k] - 1.2 * SD[k], MC[k] + 1.2 * SD[k], length.out = 40)
    y  <- pk[["lcl"]] + pk[["clwt"]] * log(MW[k] / 70) + pk[["clcr"]] * log(x / 95)
    lines(x, y, col = cols[k], lwd = 3.2)
    points(MC[k], pk[["lcl"]] + pk[["clwt"]] * log(MW[k] / 70) +
                  pk[["clcr"]] * log(MC[k] / 95),
           pch = 21, bg = cols[k], col = "white", cex = 1.6)
  }
  x <- exp(seq(log(xr[1]), log(xr[2]), length.out = 60))
  lines(x, fitted[["lcl"]] + fitted[["clwt"]] * log(mean(MW) / 70) +
           fitted[["clcr"]] * log(x / 95), col = "black", lwd = 2.6, lty = 2)
  legend("topleft", c(paste("study", names(G)), "unifying fit"),
         col = c(cols, "black"), lwd = c(3.2, 3.2, 3.2, 2.6),
         lty = c(1, 1, 1, 2), bty = "n", cex = 0.8)
}
op <- par(mfrow = c(1, 2), mar = c(4, 4.2, 2.4, 1), mgp = c(2.4, 0.7, 0))
f0 <- res0[res0$status == "fitted", ][1, ]
panel(GEN, MEAN_WT, MEAN_CR, SD_CR,
      c(lcl = f0$lcl, clwt = f0$clwt, clcr = f0$clcr),
      sprintf("this vignette (fitted %.2f)", f0$clcr))
## the same machinery on studies that overlap in renal function and disagree on
## baseline: the fitted slope is one the studies quoted below
panel(BAD, BAD_WT, BAD_CR, BAD_SD, c(lcl = 1.16, clwt = 0.81, clcr = 1.41),
      "narrow contrast, spread baselines (fitted 1.41)")
invisible(par(op))
```

On the left the three coloured segments sit at visibly different heights, and
the dashed line --- the model actually fitted --- is far steeper than any of
them. On the right, where the studies share a generating model, the segments line
up and a single line reproduces them with the right slope. Nothing about the
estimator differs between the two panels; only whether the studies were the same
underneath.

The cleanest version of the same point removes the disagreement entirely. Below, all three
studies are generated from ONE model --- same `lcl`, same coefficients --- and
differ only in their design and their covariate distribution, which is what the
coefficients are supposed to be estimated from.

```{r confound}
mod_common <- function() {
  ini({lcl <- log(3.2); lv <- log(21); clwt <- 0.75; clcr <- 0.45
       eta.cl ~ 0.09; prop.err <- 0.13})
  model({cl <- exp(lcl + eta.cl) * (WT / 70)^clwt * (CRCL / 95)^clcr
         v  <- exp(lv)
         d/dt(centr) <- -cl / v * centr
         cp <- centr / v
         cp ~ prop(prop.err)})
}
studies_common <- datagen(
  setNames(lapply(names(DESIGN), function(k)
    list(model = mod_common, times = DESIGN[[k]]$times, ev = DESIGN[[k]]$ev,
         n = DESIGN[[k]]$n, cov_dist = WT_INDEP[[k]])), names(DESIGN)),
  control = datagenControl(n_sim = 30000L, seed = 4321L))

fit_c <- suppressWarnings(suppressMessages(nlmixr2est::nlmixr2(
  mod_unify, admData(), est = "adgh",
  control = adghControl(studies = studies_common, n_nodes = 7L, print = 0L,
                        covMethod = "none", maxeval = 2000L))))
ax <- fit_c$env$admExtra
knitr::kable(data.frame(
  parameter = c("lcl", "clwt", "clcr"),
  truth     = c(log(3.2), 0.75, 0.45),
  estimate  = c(ax$struct[["lcl"]], ax$struct[["clwt"]], ax$struct[["clcr"]])),
  digits = 4, row.names = FALSE,
  caption = "Three studies from ONE generating model, differing only in design and covariate distribution.")
```

Every parameter comes back exactly, which is the reference point: with nothing
disagreeing between studies there is nothing for the slopes to absorb, and the
estimator recovers what generated the data.

The practical reading is that a covariate coefficient estimated across studies
carries whatever else differs between those studies. It is trustworthy to the
extent that the covariate contrast is large compared with those differences ---
here a fourfold renal range against 6% baseline disagreement --- and it stops
being trustworthy when that ratio reverses, whether the analysis is of aggregate
data or of individual records.

# What to take away

The estimates agree across every cell that runs. That is the point of the
comparison: the four combinations evaluate the same integral by different means,
so agreement is evidence that the integral, and not the machinery, is what
determines the answer.

Reweighting is a speed setting, not a modelling choice. It changes how the
covariate integral is evaluated and nothing about what is being estimated, which
is why the estimates match rather than merely resemble each other.

`adgh` and a dependent covariate distribution do not combine. That is a refusal
and not a limitation to work around: a product grid assumes independence, and
producing a number from it would be worse than declining. Use `admc` when the
covariates are dependent.

Finally, a covariate coefficient is only estimable here because the three studies
have *different* weight distributions. Three studies drawn from one population
would leave the likelihood exactly flat along a trade-off between the coefficient
and the between-subject variance, and any estimate reported from such a fit would
be the optimiser's stopping point rather than a result.
