---
title: "Covariates in aggregate data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covariates in aggregate data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.2
)
source(Filter(file.exists, c("_common.R", "../_common.R", "vignettes/_common.R"))[1])
```

## The problem

You need a dose for **severe renal impairment**. Three trials have been
published — normal renal function, mild impairment, moderate — and none
enrolled a severe patient. None of them reported a renal effect either: within
each trial creatinine clearance barely varied, so each analyst sensibly left it
out.

You are left with three population models that disagree about clearance and
none that says why.

```{r libs}
library(admixr2); library(rxode2); library(nlmixr2); library(ggplot2)
```

## Three trials, three answers

Weight and creatinine clearance are correlated, so the virtual patients are
drawn jointly through a Gaussian copula — each covariate keeps its own
lognormal margin, the dependence goes on the latent scale. The truth, unknown
to all three analysts, is CL = 5 L/h and V = 50 L at 70 kg, allometric weight
scaling, a renal exponent of 0.6, and men clearing about 20% faster.

```{r sim}
set.seed(11)
TIMES <- c(0.5, 1, 2, 4, 8, 12, 24); DOSE <- 200
CL70 <- 5; V70 <- 50; BCRCL <- 0.6; BSEX <- 0.18
OM_CL <- 0.05; SD_ADD <- 0.08
R <- matrix(c(1, .45, .45, 1), 2, 2,
            dimnames = list(c("WT", "CRCL"), c("WT", "CRCL")))

draw_pop <- function(n, mw, sw, mc, sc, p_male = .55) {
  Z <- matrix(stats::rnorm(2 * n), n, 2) %*% chol(R)
  data.frame(WT   = stats::qlnorm(stats::pnorm(Z[, 1]), log(mw), sw),
             CRCL = stats::qlnorm(stats::pnorm(Z[, 2]), log(mc), sc),
             SEX  = stats::rbinom(n, 1L, p_male))
}
sim_trial <- function(cv, id0) {
  n  <- nrow(cv)
  cl <- CL70 * (cv$WT/70)^0.75 * (cv$CRCL/90)^BCRCL * exp(BSEX * cv$SEX) *
        exp(stats::rnorm(n, 0, sqrt(OM_CL)))
  v  <- V70 * (cv$WT/70)
  do.call(rbind, lapply(seq_len(n), function(i) rbind(
    data.frame(ID = id0+i, TIME = 0, DV = NA_real_, AMT = DOSE, EVID = 1,
               CMT = 1, WT = cv$WT[i], CRCL = cv$CRCL[i], SEX = cv$SEX[i]),
    data.frame(ID = id0+i, TIME = TIMES,
               DV = DOSE/v[i] * exp(-cl[i]/v[i]*TIMES) +
                    stats::rnorm(length(TIMES), 0, SD_ADD),
               AMT = 0, EVID = 0, CMT = 1, WT = cv$WT[i], CRCL = cv$CRCL[i],
               SEX = cv$SEX[i]))))
}
cohorts <- list(normal   = draw_pop(260L, 78, .198, 95, .217),
                mild     = draw_pop(210L, 76, .198, 62, .198),
                moderate = draw_pop(180L, 74, .198, 38, .198))
trial_data <- Map(function(cv, k) sim_trial(cv, k * 10000L),
                  cohorts, seq_along(cohorts))
```

A published summary keeps only the mean and the spread, so the covariate
disappears into the width of the ribbon — poor clearers at the top, good
clearers at the bottom:

```{r hidden, fig.height = 3.8}
d1  <- trial_data$normal[trial_data$normal$EVID == 0, ]
agg <- data.frame(t = TIMES,
                  m = tapply(d1$DV, d1$TIME, mean),
                  s = tapply(d1$DV, d1$TIME, stats::sd),
                  panel = "What the paper reports")
ind <- d1[d1$ID %in% head(unique(d1$ID), 150), ]
ind$panel <- "Individual patients (never published)"

ggplot() +
  geom_line(data = ind, aes(TIME, DV, group = ID, colour = CRCL),
            alpha = .35, linewidth = .3) +
  geom_ribbon(data = agg, aes(t, ymin = m - s, ymax = m + s),
              fill = "grey55", alpha = .35) +
  geom_line(data = agg, aes(t, m), linewidth = 1) +
  scale_colour_viridis_c(name = "CRCL (mL/min)", option = "C", end = .9) +
  facet_wrap(~panel) + labs(x = "time (h)", y = "concentration (mg/L)") +
  theme_bw() + theme(legend.position = "bottom")
```

Each analyst now fits their own trial. Weight is scaled allometrically with
the exponents fixed at 0.75 and 1. **Sex they can estimate** — both sexes are
enrolled, so it varies where they can see it. **Renal function they cannot** —
within one cohort it barely moves.

```{r trialfits}
trial_model <- function() {
  ini({ tcl <- log(4.5); tv <- log(45); bsex <- 0.05
        eta.cl ~ 0.1; add.err <- 0.2 })
  model({ cl <- exp(tcl + eta.cl) * (WT/70)^0.75 * exp(bsex * SEX)
          v  <- exp(tv) * (WT/70)
          cp <- linCmt(); cp ~ add(add.err) })
}
trial_fits <- lapply(trial_data, function(d)
  suppressWarnings(nlmixr2(trial_model, d, est = "focei",
                           control = foceiControl(print = 0L))))
published <- lapply(trial_fits, function(f) list(
  CL   = exp(f$parFixedDf["tcl", "Estimate"]),
  V    = exp(f$parFixedDf["tv",  "Estimate"]),
  BSEX = f$parFixedDf["bsex", "Estimate"],
  OM   = as.numeric(f$omega[1L, 1L]),
  ADD  = f$parFixedDf["add.err", "Estimate"]))
knitr::kable(data.frame(
  cohort      = names(published),
  median_CRCL = round(vapply(cohorts, function(c) stats::median(c$CRCL), 0)),
  CL_at_70kg  = vapply(published, `[[`, 0, "CL"),
  V_at_70kg   = vapply(published, `[[`, 0, "V"),
  sex_effect  = vapply(published, `[[`, 0, "BSEX")),
  digits = 3, row.names = FALSE,
  caption = "Three independent analyses, each sound on its own data.")
```

All three agree on volume and on the sex effect — a covariate that varies
*inside* each trial, so each could estimate it. They disagree wildly on
clearance, by nearly a factor of two, and none of them can say why: the
explanation is a covariate that barely varies within any one of them.

That is the asymmetry to hold onto. **A covariate that varies within a study is
estimable there. One that varies only between studies is not — and needs a
meta-analysis.**

## Their models as input

admixr2 takes two kinds of input: a digitised summary, or a **published model
supplied directly**. This is the second. `datagen()` turns each published model,
plus the population it was fitted in, into the `(E, V, n)` a figure would have
given us.

A covariate reaches admixr2 one of two ways, and both appear here:

* **Marginalised** — the study declares a `cov_dist` and the estimator
  integrates over it. That is what a pooled summary requires.
* **Conditioned** — the study declares a `cov` value and the model is solved
  there. That is what a summary reported *by subgroup* allows.

The normal and moderate trials reported pooled results, so everything is
marginalised. The mild trial reported **separately for men and women**, so it
becomes two studies with sex *conditioned* and weight and renal function still
marginalised.

```{r datagen}
CLp <- Vp <- BSEXp <- OMp <- ADDp <- NULL      # read by ini() below
published_model <- function() {
  ini({ tcl <- log(CLp); tv <- log(Vp); bsex <- BSEXp
        eta.cl ~ OMp; add.err <- ADDp })
  model({ cl <- exp(tcl + eta.cl) * (WT/70)^0.75 * exp(bsex * SEX)
          v  <- exp(tv) * (WT/70)
          cp <- linCmt(); cp ~ add(add.err) })
}
use_published <- function(nm) {
  p <- published[[nm]]
  CLp <<- p$CL; Vp <<- p$V; BSEXp <<- p$BSEX; OMp <<- p$OM; ADDp <<- p$ADD
}
# `cor` must span EVERY declared covariate, so sex joins it as independent --
# a discrete covariate correlated with a continuous one would make each level a
# truncation of the latent normal rather than a point, which admixr2 refuses.
R3 <- rbind(cbind(R, SEX = c(0, 0)), SEX = c(0, 0, 1))
dimnames(R3) <- list(c("WT", "CRCL", "SEX"), c("WT", "CRCL", "SEX"))
pop_dist <- function(cv, with_sex) {
  a <- list(WT   = list(meanlog = mean(log(cv$WT)),   sdlog = stats::sd(log(cv$WT))),
            CRCL = list(meanlog = mean(log(cv$CRCL)), sdlog = stats::sd(log(cv$CRCL))),
            cor  = R)
  if (with_sex) {
    a$SEX <- list(values = c(0, 1), probs = c(1 - mean(cv$SEX), mean(cv$SEX)))
    a$cor <- R3
  }
  do.call(covDist, a)
}
gen1 <- function(n, cov_dist, cov = NULL) {
  sp <- list(times = TIMES, ev = rxode2::et(amt = DOSE), n = n,
             cov_dist = cov_dist)
  if (!is.null(cov)) sp$cov <- cov
  datagen(list(s = sp), model = published_model,
          control = datagenControl(method = "gh", seed = 1L))$s
}
studies <- list()
for (nm in c("normal", "moderate")) {             # POOLED -> all marginalised
  use_published(nm)
  studies[[nm]] <- gen1(nrow(cohorts[[nm]]), pop_dist(cohorts[[nm]], TRUE))
}
use_published("mild")                             # REPORTED BY SEX
for (sx in c(0, 1))                               # -> sex conditioned
  studies[[paste0("mild_sex", sx)]] <-
    gen1(sum(cohorts$mild$SEX == sx), pop_dist(cohorts$mild, FALSE),
         cov = list(SEX = sx))
names(studies)
```

Each study carries the trial's **model** and the **population** it was fitted
in, which the paper describes in its baseline table. The `mild_sex*` studies
carry a `cov` as well: sex is pinned, weight and renal function are still
integrated over. None of the published models has a renal term — the
`cov_dist` does. We are telling admixr2 who was enrolled and letting it work
out what that implies.

## The meta-analysis

```{r admfit}
adm_model <- function() {
  ini({ tcl <- log(4); tv <- log(45); bcrcl <- 0.3; bsex <- 0.05
        eta.cl ~ 0.1; add.err <- 0.1 })
  model({ cl <- exp(tcl + eta.cl) * (WT/70)^0.75 * (CRCL/90)^bcrcl *
                exp(bsex * SEX)
          v  <- exp(tv) * (WT/70)
          cp <- linCmt(); cp ~ add(add.err) })
}
fit <- nlmixr2(adm_model, admData(), est = "adgh",
               control = adghControl(studies = studies, print = 0L,
                                     covMethod = "r"))
pf     <- fit$parFixedDf
om     <- as.numeric(fit$omega[1L, 1L])
om_se  <- sqrt(diag(fit$cov))[["om.eta.cl"]]   # omega's SE is in $cov, not pf
knitr::kable(data.frame(
  parameter = c("CL at 70 kg, CRCL 90, women (L/h)", "V at 70 kg (L)",
                "renal exponent", "sex effect (log)", "additive SD (mg/L)",
                "omega (eta.cl)"),
  truth     = c(CL70, V70, BCRCL, BSEX, SD_ADD, OM_CL),
  estimate  = c(pf["tcl", "Back-transformed"], pf["tv", "Back-transformed"],
                pf["bcrcl", "Back-transformed"], pf["bsex", "Back-transformed"],
                pf["add.err", "Back-transformed"],
                as.numeric(fit$omega[1L, 1L])),
  # CIs, not SEs: a structural theta's SE is on the LOG scale, so printing it
  # beside a back-transformed estimate invites reading 0.022 as 0.022 L/h
  lower     = c(pf["tcl", "CI Lower"], pf["tv", "CI Lower"],
                pf["bcrcl", "CI Lower"], pf["bsex", "CI Lower"],
                pf["add.err", "CI Lower"], om - 1.96 * om_se),
  upper     = c(pf["tcl", "CI Upper"], pf["tv", "CI Upper"],
                pf["bcrcl", "CI Upper"], pf["bsex", "CI Upper"],
                pf["add.err", "CI Upper"], om + 1.96 * om_se)),
  digits = 3, row.names = FALSE,
  caption = "Recovered from three analyses, none of which contained a renal term.")
```

Both covariate effects come back, from three analyses that between them
contained only one of them. But look at the sex effect: the three trials
reported `r paste(sprintf("%.2f", vapply(published, `[[`, 0, "BSEX")), collapse = ", ")`
for a quantity that is genuinely the same in all of them, and the meta-analysis has to settle on one number. **Published models
disagree, and reconciling them is not free** — some of that disagreement lands
in the other parameters, which is why the renal exponent sits a little below
its truth and the prediction below is a few per cent off rather than exact.
That is what a meta-analysis is: evidence that does not perfectly agree, pooled
anyway, and honest about the residual.

Two things about standard errors, both easy to get wrong. A structural theta's
SE is on the **log** scale — printing it beside a back-transformed estimate
would read as L/h and be off by an order of magnitude — so `parFixedDf` gives
back-transformed CIs, used above. And the **omega** SE is not in `parFixedDf`
at all: by nlmixr2 convention it lives in `fit$cov`, as `om.eta.cl`.

Here is what the meta-analysis actually did:

```{r mechanism, fig.height = 4.2}
e  <- stats::setNames(pf$Estimate, rownames(pf))
crcl_grid <- seq(15, 115, length.out = 120)
sexlab <- c("women", "men")

# the meta-analysis: one curve per sex. The renal effect is the SLOPE, the sex
# effect is the vertical GAP between the two.
curve_df <- do.call(rbind, lapply(0:1, function(sx) data.frame(
  CRCL = crcl_grid, sex = sexlab[sx + 1L],
  CL = exp(e[["tcl"]]) * (crcl_grid/90)^e[["bcrcl"]] * exp(e[["bsex"]] * sx))))

# each published model: no renal term, so a FLAT bar per sex, drawn over the
# CRCL range that trial actually covered
seg <- do.call(rbind, lapply(names(published), function(nm) do.call(rbind,
  lapply(0:1, function(sx) data.frame(
    lab = nm, sex = sexlab[sx + 1L],
    CL  = published[[nm]]$CL * exp(published[[nm]]$BSEX * sx),
    lo  = unname(stats::quantile(cohorts[[nm]]$CRCL, .10)),
    hi  = unname(stats::quantile(cohorts[[nm]]$CRCL, .90)))))))
pts <- do.call(rbind, lapply(names(published), function(nm) do.call(rbind,
  lapply(0:1, function(sx) data.frame(
    lab = nm, sex = sexlab[sx + 1L],
    CRCL = stats::median(cohorts[[nm]]$CRCL),
    CL   = published[[nm]]$CL * exp(published[[nm]]$BSEX * sx))))))

ggplot(curve_df, aes(CRCL, CL, colour = sex)) +
  annotate("rect", xmin = 15, xmax = 30, ymin = -Inf, ymax = Inf,
           fill = "grey85", alpha = .6) +
  annotate("text", x = 22.5, y = max(curve_df$CL), label = "severe\n(no trial)",
           size = 3, vjust = 1.1) +
  geom_segment(data = seg, aes(x = lo, xend = hi, y = CL, yend = CL),
               linewidth = 1.5, alpha = .55) +
  geom_line(linewidth = 1) +
  geom_point(data = pts, size = 2.6) +
  geom_text(data = subset(pts, sex == "men"), aes(label = lab),
            vjust = -1.2, size = 3.1, show.legend = FALSE) +
  scale_colour_manual(values = c(women = "steelblue4", men = "darkorange3"),
                      name = NULL) +
  labs(x = "median creatinine clearance (mL/min)",
       y = "clearance at 70 kg (L/h)",
       subtitle = paste("bars: what each published model claims, over the range",
                        "it covered\nlines: the meta-analysis")) +
  theme_bw() + theme(legend.position = "bottom")
```

Both effects are in that picture, and they look completely different.

The **sex effect** is the vertical gap between the two lines. Every trial found
it, so every flat bar comes in a matching pair with the same gap — a covariate
that varies *within* a study is visible to that study.

The **renal effect** is the slope. No trial has one: each is a pair of flat
bars, claiming clearance is whatever it measured at any renal function. Three
pairs at three different heights, each right over its own range and mutually
contradictory everywhere else. The meta-analysis draws the line through them,
and that line is the renal effect — recovered from between trials, and carried
into the shaded region none of them sampled.

## The cohort nobody ran

Without admixr2 you would take the published model closest to your patients —
the moderate-impairment one — and use it.

```{r predict}
set.seed(3)
marg_mean <- function(cl_ref, v_ref, bcrcl, bsex, om, n = 40000L) {
  Z    <- matrix(stats::rnorm(2 * n), n, 2) %*% chol(R)
  WT   <- stats::qlnorm(stats::pnorm(Z[, 1]), log(72), .198)
  CRCL <- stats::qlnorm(stats::pnorm(Z[, 2]), log(21), .217)
  SX   <- stats::rbinom(n, 1L, .55)
  cl <- cl_ref * (WT/70)^0.75 *
        (if (is.na(bcrcl)) 1 else (CRCL/90)^bcrcl) * exp(bsex * SX) *
        exp(stats::rnorm(n, 0, sqrt(om)))
  vapply(TIMES, function(t)
    mean(DOSE / (v_ref * WT/70) * exp(-cl / (v_ref * WT/70) * t)), numeric(1))
}
p_adm   <- marg_mean(exp(e[["tcl"]]), exp(e[["tv"]]), e[["bcrcl"]], e[["bsex"]],
                     as.numeric(fit$omega[1L, 1L]))
p_near  <- marg_mean(published$moderate$CL, published$moderate$V, NA,
                     published$moderate$BSEX, published$moderate$OM)
p_truth <- marg_mean(CL70, V70, BCRCL, BSEX, OM_CL)
```

Predicting the *mean* is only half of it. Simulate a severe cohort under each
model and compare the whole distribution — median and 90% interval — against
what those patients would really do. The truth is repeated in grey behind every
panel.

```{r vpc, fig.height = 5.0}
set.seed(7)
n_pt <- 800L
Zs   <- matrix(stats::rnorm(2 * n_pt), n_pt, 2) %*% chol(R)
WTs  <- stats::qlnorm(stats::pnorm(Zs[, 1]), log(72), .198)
CRs  <- stats::qlnorm(stats::pnorm(Zs[, 2]), log(21), .217)
SXs  <- stats::rbinom(n_pt, 1L, .55)
tt   <- seq(0.25, 24, length.out = 40)

band <- function(cl_ref, v_ref, bcrcl, bsex, om, label) {
  cl <- cl_ref * (WTs/70)^0.75 *
        (if (is.na(bcrcl)) 1 else (CRs/90)^bcrcl) * exp(bsex * SXs) *
        exp(stats::rnorm(n_pt, 0, sqrt(om)))
  v  <- v_ref * (WTs/70)
  M  <- vapply(tt, function(t) DOSE / v * exp(-cl / v * t), numeric(n_pt))
  q  <- apply(M, 2L, stats::quantile, c(.05, .5, .95))
  data.frame(t = tt, lo = q[1, ], med = q[2, ], hi = q[3, ], model = label)
}
lv  <- c("truth", "normal-trial model", "mild-trial model",
         "moderate-trial model", "meta-analysis (admixr2)")
pb   <- function(nm, lab) with(published[[nm]],
  band(CL, V, NA, BSEX, OM, lab))     # published models: sex yes, renal no
sims <- rbind(
  band(CL70, V70, BCRCL, BSEX, OM_CL, lv[1]),
  pb("normal", lv[2]), pb("mild", lv[3]), pb("moderate", lv[4]),
  band(exp(e[["tcl"]]), exp(e[["tv"]]), e[["bcrcl"]], e[["bsex"]],
       as.numeric(fit$omega[1L, 1L]), lv[5]))
sims$model <- factor(sims$model, levels = lv)
tru <- subset(sims, model == lv[1], select = -model)   # drawn in every panel

ggplot(sims, aes(t)) +
  geom_ribbon(data = tru, aes(ymin = lo, ymax = hi), fill = "grey60",
              alpha = .45) +
  geom_line(data = tru, aes(y = med), colour = "grey25", linewidth = .7) +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = model), alpha = .40) +
  geom_line(aes(y = med, colour = model), linewidth = .9) +
  facet_wrap(~model, nrow = 2) +
  guides(fill = "none", colour = "none") +
  labs(x = "time (h)", y = "concentration (mg/L)",
       subtitle = paste("800 simulated severe-impairment patients per model;",
                        "grey = the truth")) +
  theme_bw()
```

Each trial model reproduces the population it was fitted in, not this one, and
the miss is ordered by how far that population sits from severe. At 24 h,
against the truth:

```{r vpcerr}
at24 <- function(m) {
  d <- sims[sims$model == m, ]; i <- which.min(abs(d$t - 24))
  c(med = d$med[i], width = d$hi[i] - d$lo[i])
}
tr <- at24(lv[1])
knitr::kable(data.frame(
  model      = lv[-1],
  median_err = vapply(lv[-1], function(m) 100*(at24(m)[["med"]]/tr[["med"]]-1), 0),
  width_err  = vapply(lv[-1], function(m)
                      100*(at24(m)[["width"]]/tr[["width"]]-1), 0)),
  digits = 1, row.names = FALSE,
  caption = "% error in the median and in the 90% interval width, at 24 h.")
```

They miss on **spread** as much as on level. A model with no renal term applied
to this cohort has only weight and `omega` to generate between-patient
variability, and the `omega` it carries was fitted where creatinine clearance
hardly varied — so it cannot reproduce a population whose clearance varies a
great deal. Only the meta-analysis, which knows what renal function does, lands
on the grey.

```{r payofftab}
knitr::kable(data.frame(
  time = TIMES, truth = p_truth, meta_analysis = p_adm, nearest_model = p_near,
  err_meta = 100*(p_adm/p_truth - 1), err_near = 100*(p_near/p_truth - 1)),
  digits = 2, caption = "Severe renal impairment: predicted mean concentration.")
```

The nearest published model is carrying a clearance fitted in patients who
clear the drug faster than these ones do, so it drifts further wrong the longer
you look. Underpredicting exposure in renal impairment is the direction that
matters clinically.

## Practicalities

**Declaring the population.** `cov_dist` is data, not a parameter — admixr2
never estimates it. Take it from the paper's baseline table:

```{r forms}
covDist(WT = c(mean = 78, sd = 16), dist = "lnorm")$WT   # mean and SD
covDist(SEX = c(male = .55, female = .45))$SEX           # a categorical
```

| the paper reports | use |
|---|---|
| mean and SD | `c(mean = , sd = )` with `dist = "lnorm"` |
| geometric mean `G`, CV | `meanlog = log(G)`, `sdlog = sqrt(log(1 + CV^2))` |
| median `M`, IQR `Q1`–`Q3` | `meanlog = log(M)`, `sdlog = (log(Q3) - log(Q1))/1.349` |

A whole baseline table goes in at once as a data frame with `covariate`,
`mean`, `sd` and `dist` columns. Prefer `"lnorm"` for anything positive: a
normal margin is unbounded below and the quadrature reaches about 3.75 SD, so a
CV above roughly 0.27 puts a node at or below zero where an allometric term is
`NaN`. `covDist()` warns when it sees this.

**Marginalised or conditioned.** A `cov_dist` is integrated over; a `cov`
value is solved at. Both appear above: the pooled trials marginalise all three
covariates, the sex-stratified one pins `SEX` and marginalises the rest.
`covStrata()` will do that split for you when a study reports subgroups.

**Discrete covariates are strata**, enumerated exactly at their levels and
weighted by their probabilities — nothing about them is approximated.

**Cost.** A product grid over `p` covariates costs `cov_nodes^p`, but where the
covariates reach the model through fewer scalars than there are covariates,
admixr2 integrates over those directions instead. Three covariates entering as
one product are a one-dimensional integral, on
`r admixr2:::.admCovDirNodes(7L, 3L, 1L)` design points rather than `r 7^3`.
That is a change of variables, verified numerically before use — not an
approximation.

**Estimator support.** `admc` and `adgh` marginalise covariates; `adfo` and
`adirmc` refuse rather than silently solving at the covariate mean. admixr2
also refuses a `cov_dist` naming a covariate the model never reads.

## Notes

* **A covariate effect is identified by contrast across sources.** That is the
  whole of this vignette. Within one cohort a covariate barely varies, and what
  little it does is confounded with between-subject variability — a covariate
  enters `V` where `omega` does.
* **Fix what convention fixes.** Allometric exponents are 0.75 and 1; aggregate
  data is not the place to re-estimate them. Estimate the covariate effects a
  popPK analysis would.
* **Extrapolation is extrapolation.** The severe prediction works because the
  power model is the right functional form and the three cohorts span enough
  CRCL to pin its exponent. Fitted over 38–95 mL/min and applied at 21, it
  should be presented as such.

## See also

- [datagen()](datagen.html) — published models as input, in depth
- [Multiple studies](multiple-studies.html) — the contrast that identifies an effect
- [From a published figure to E, V and n](aggregate-data.html) — the other input
- [Estimator comparison](estimator-comparison.html)
