---
title: "Covariates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covariates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 3.0, dpi = 110, warning = FALSE,
                      message = FALSE)
source("_common.R", local = TRUE)
```

```{r libs}
library(admixr2)
library(rxode2)
library(nlmixr2)
library(ggplot2)
```

```{r theme, include = FALSE}
NAVY <- "#001158"; ORANGE <- "#DD6B2A"
theme_set(theme_bw(base_size = 10))
```

A published model tells you what a drug did in the patients that study enrolled.
Pool several such models and you can, in principle, describe patients none of
them enrolled --- but only if you are careful about who was in each study and
what each study actually measured.

This vignette covers that: how to describe a study's population, how admixr2
integrates over it, and the three ways it can go wrong.

# The covariate distribution

The aggregate `E` and `V` a paper reports are **marginal** over its patients'
covariates. So the prediction has to be marginal too. Evaluating the model at
the *average* patient and comparing that to the *average* concentration is the
ecological fallacy, because $\mathbb{E}[g(a)] \neq g(\mathbb{E}[a])$ for any
nonlinear model.

`covDist()` takes what a baseline-characteristics table reports:

```{r covdist}
pop <- covDist(WT   = c(mean = 78, sd = 14),
               CRCL = c(mean = 112, sd = 24),
               cor  = 0.55, dist = "lnorm")
pop
```

It prints the distribution it *realised*, not the numbers you typed, so a
transcription slip surfaces here instead of inside a fit.

## Choosing the margin

`dist = "normal"` is the default --- `mean` and `sd` read as a normal
distribution, which is what most people expect. **But weight, renal function,
age and most other PK covariates are positive and right-skewed, and belong in
`dist = "lnorm"`.**

That is not a stylistic preference. A normal margin is unbounded below, and the
quadrature reaches 3.75 standard deviations at the default `cov_nodes = 7`:

```{r reach, echo = FALSE}
z7 <- max(abs(admixr2:::.adghNodes1(7L)$x))
knitr::kable(data.frame(
  covariate = c("weight", "creatinine clearance", "age"),
  mean = c(78, 112, 54), sd = c(14, 24, 11),
  `lowest node if normal` = round(c(78, 112, 54) - z7 * c(14, 24, 11), 1),
  check.names = FALSE),
  caption = "A normal margin reaches below zero for ordinary PK covariates.")
```

A negative weight in `(WT/70)^theta` is `NaN`. A lognormal margin is positive by
construction at any node count, so this cannot happen. `covDist()` warns when a
normal margin would reach zero:

```{r warn}
tryCatch(covDist(WT = c(mean = 78, sd = 14)),
         warning = function(w) cat(strwrap(conditionMessage(w), 76), sep = "\n"))
```

Ignore it for a genuinely centred covariate --- a log-ratio, a change from
baseline, a z-score.

## The rest of the grammar

```{r grammar, eval = FALSE}
covDist(WT  = c(mean = 78, sd = 14), dist = "lnorm")   # natural scale
covDist(WT  = c(mu = 0, sd = 1))                       # normal, own scale
covDist(WT  = c(meanlog = log(70), sdlog = 0.2))       # lognormal, own scale
covDist(SEX = c(female = 0.55, male = 0.45))           # categorical
covDist(WT  = list(quantile = function(u) qgamma(u, 9, 0.125)))   # anything
covDist(baseline_table)                                # covariate/mean/sd cols
```

Dependence is `cor` (a scalar or a matrix), `rho`, `Sigma`, or `joint` for an
arbitrary copula --- an R-vine included. A `joint` sampler receives the uniforms
admixr2 supplies and must consume them deterministically; that is what keeps the
objective a deterministic function of the parameters.
`rvinecopulib::inverse_rosenblatt()` qualifies, `rvinecop()` does not.

```{r cloud, fig.height = 3.0}
X <- covDraw(pop, n = 5000L)
ggplot(as.data.frame(X), aes(WT, CRCL)) +
  geom_point(alpha = 0.12, size = 0.5, colour = NAVY) +
  labs(x = "weight (kg)", y = "creatinine clearance (mL/min)")
```

# Asking each source only what it measured

A study that never fitted renal function has no answer about renal function.
Evaluate its model across a range of CrCl values and it returns the same
prediction every time --- and if you feed that flat answer to the likelihood as
evidence, it argues against the effect a study that *did* measure it found. In a
noiseless simulation this pulled a true coefficient of 0.450 down to 0.211.

The rule is per covariate, not per study:

> **Stratify** on the covariates a source actually fitted.
> **Marginalise** the ones it did not.

`stratify = TRUE` reads which is which off the source's own model, so the split
cannot disagree with the model that produced the data:

```{r generate}
mod_wt <- function() {                      # a source that fitted WEIGHT only
  ini({lcl <- log(3.0); lv <- log(20); bwt <- 0.70
       eta.cl ~ 0.0736; prop.err <- 0.12})
  model({cl <- exp(lcl + eta.cl) * (WT / 70)^bwt; v <- exp(lv)
         d/dt(centr) <- -cl / v * centr; cp <- centr / v
         cp ~ prop(prop.err)})
}
tms <- c(0.5, 1, 2, 4, 8, 12)
dat <- datagen(list(
  S = list(model = mod_wt, n = 180L, times = tms, ev = et(amt = 100),
           cov_dist = pop, stratify = TRUE)),
  control = datagenControl(method = "mc", n_sim = 3000L))
names(dat)
```

One study became five --- one per weight stratum, because weight is what this
source's model conditions on. Renal function is left to be marginalised inside
each.

## What a stratum contains

```{r strata-fig, fig.height = 3.2}
st  <- covStrata(pop, stratify = "WT", n_nodes = 4L, n = 180)
st  <- st[order(vapply(st, function(s) s$cov$WT, 0))]
lab <- c("lightest", "light", "heavy", "heaviest")
cloud <- do.call(rbind, lapply(seq_along(st), function(k)
  data.frame(band = lab[k], covDraw(st[[k]]$cov_dist, n = 2500L))))
cloud$band <- factor(cloud$band, levels = lab)

p1 <- ggplot(cloud, aes(WT, CRCL, colour = band)) +
  geom_point(alpha = 0.10, size = 0.5) +
  labs(x = "weight (kg)", y = "creatinine clearance") +
  theme(legend.position = "none")
p2 <- ggplot(cloud, aes(CRCL, colour = band)) +
  geom_density(linewidth = 0.8) +
  geom_vline(xintercept = mean(covDraw(pop, n = 5e4L)[, "CRCL"]),
             linetype = 2, colour = "grey40") +
  labs(x = "creatinine clearance within each stratum", y = NULL, colour = NULL)
if (requireNamespace("patchwork", quietly = TRUE)) p1 + p2 else {print(p1); p2}
```

**Left:** the strata cut the study's patients by weight. **Right:** within each
stratum renal function has its *own* distribution, shifted away from the
study-wide mean (dashed) because the two covariates are correlated.

That right panel is the part that is easy to get wrong. When this source's
prediction marginalises over renal function it must use **those** curves. Using
the study-wide distribution instead is wrong by a *different amount in each
stratum*, which is precisely the shape that biases the weight coefficient the
stratification was meant to expose.

```{r strata-tab}
data.frame(stratum   = lab,
           mean_WT   = round(vapply(st, function(s) s$cov$WT, 0), 1),
           n_k       = round(vapply(st, function(s) s$n, 0), 1),
           mean_CRCL = round(vapply(st, function(s)
                         mean(covDraw(s$cov_dist, n = 4000L)[, "CRCL"]), 0), 1))
```

`n_k` sums to 180: the strata **partition** the study's patients rather than
multiplying them. More strata resolve the covariate range more finely but do not
buy accuracy --- on a matched fit the coefficient came back at 0.7000 / 0.7002 /
0.7005 for 3 / 4 / 10 strata against a true 0.700 --- and each one costs a solve,
which is why the default is 5.

### A published subgroup is an interval, not a value

The strata above are pinned at a covariate *value*, which is the right
representation for a source you can evaluate there --- a fitted model, or a
simulation you are generating. A subgroup read off a publication is not that: it
is a **band**, and its patients are spread across it.

Scoring a band at a single representative value leaves the band's own spread out
of the prediction. Writing `tau_k` for the covariate spread within band *k*, the
predicted covariance is short by `tau_k^2 * D D'` with `D` the derivative of the
profile in the covariate --- rank one, and first order in the *slope*, so it does
not vanish where the model is locally linear and the mean is unaffected. It is
the same plug-in error that stratifying was meant to avoid, committed inside a
stratum instead of across the study.

The fix needs no new machinery: give that study the band's **distribution**
rather than its midpoint, and `admixr2` integrates over it exactly as it
integrates a study-wide distribution.

```{r band-spec, eval = FALSE}
# a published subgroup "weight 60-75 kg", n = 84
study_band <- list(E = E_band, V = V_band, n = 84,
                   cov_dist = list(WT = list(dist = "unif", min = 60, max = 75)))
```

Two practical notes. Use the band's reported **mean**, not the midpoint of its
limits --- they differ most for the open-ended outer bands (`>= 65 years`) that
publications favour, and that is where the coefficient moves most. And a band
narrow enough that `tau_k` is negligible can still be given as a value: the
correction is then empty, and `cov = list(WT = 67)` is exact.

# A covariate nobody measured

Here is the result that makes aggregate synthesis worth doing.

Take two studies that **never fitted weight at all**. Each folded the weight
effect into its intercept and its random effect, exactly as theory says a
covariate-free fit must:

$$	heta' = 	heta + eta\,\mathbb{E}[a], \qquad
  \omega'^2 = \omega^2 + eta^2\mathrm{Var}(a)$$

Neither study can tell you the weight coefficient on its own. Together they can,
because their populations differ in **both mean and spread** --- which gives two
equations for the two unknowns.

```{r nobody}
LCLt <- log(3.0); BWt <- 0.70; OM2t <- 0.060
absorb <- function(m, s) {                    # what such a study would publish
  sl <- sqrt(log(1 + s^2/m^2)); ml <- log(m) - sl^2/2
  list(lcl = LCLt + BWt * (ml - log(70)), om2 = OM2t + BWt^2 * sl^2)
}
no_wt <- function(a) {
  g <- function() {
    ini({lcl <- .L; lv <- log(20); eta.cl ~ .O; prop.err <- 0.12})
    model({cl <- exp(lcl + eta.cl); v <- exp(lv)
           d/dt(centr) <- -cl / v * centr; cp <- centr / v
           cp ~ prop(prop.err)})
  }
  body(g) <- do.call(substitute, list(body(g), list(.L = a$lcl, .O = a$om2)))
  g
}
A <- absorb(95, 24); B <- absorb(65, 10)      # heavier/spread vs lighter/tight
dat2 <- datagen(list(
  A = list(model = no_wt(A), n = 200L, times = tms, ev = et(amt = 100),
           cov_dist = covDist(WT = c(mean = 95, sd = 24), dist = "lnorm")),
  B = list(model = no_wt(B), n = 200L, times = tms, ev = et(amt = 100),
           cov_dist = covDist(WT = c(mean = 65, sd = 10), dist = "lnorm"))),
  control = datagenControl(method = "mc", n_sim = 6000L))
```

Neither study is stratified --- neither has a weight contrast to give. Both are
purely **marginal** over weight, which is exactly what their models support.

```{r nobody-fit}
joint_wt <- function() {
  ini({lcl <- log(2.0); lv <- log(15); bwt <- 0.20
       eta.cl ~ 0.03; prop.err <- 0.30})
  model({cl <- exp(lcl + eta.cl) * (WT / 70)^bwt; v <- exp(lv)
         d/dt(centr) <- -cl / v * centr; cp <- centr / v
         cp ~ prop(prop.err)})
}
f2 <- nlmixr2(joint_wt, admData(), est = "adgh",
              control = adghControl(studies = dat2, print = 0L,
                                    covMethod = "none"))
round(c(bwt = setNames(f2$parFixedDf$Estimate, rownames(f2$parFixedDf))[["bwt"]],
        truth = BWt,
        omega = sqrt(f2$omega[1, 1]), omega_truth = sqrt(OM2t)), 4)
```

**The coefficient comes back, from two studies neither of which measured it.**
The random effect comes back to its true value too, because the variability each
study had to absorb is now explained by the covariate.

This is the $\omega'^2 = \omega^2 + eta^2\mathrm{Var}(a)$ channel doing real
work --- not a curiosity. It is why a covariate-free publication is still
evidence, and why marginalising over what a source did not measure is the right
thing to do with it rather than a concession.

# How the integral is done

```{r integration, eval = FALSE}
adghControl(studies = st, cov_integration = "quadrature", cov_nodes = 7L)
adghControl(studies = st, cov_integration = "taylor",     cov_taylor_h = 0.5)
```

**`"quadrature"`** (the default) evaluates the model on a product Gauss--Hermite
grid over the *sampler's uniforms* --- which stays exact whatever the dependence,
because a copula maps independent uniforms to dependent values. At
`cov_nodes = 7` the marginal moments are accurate to 2e-16; the default is set
where accuracy saturates but the nodes have not yet run far into the tails.

**`"taylor"`** expands the marginal moments to second order instead: $1+2p$ model
evaluations rather than `cov_nodes`$^p$, at any correlation, by differencing
along the eigenvectors of the covariate covariance. The expansion runs in the
*latent normal*, so every design point maps back through the covariate's own
quantile function and stays inside its support --- a lognormal covariate can
never be handed a negative value. It is an approximation; check an important fit
against `"quadrature"`.

```{r support, echo = FALSE}
knitr::kable(data.frame(
  Estimator = c("`admc`", "`adgh`", "`adfo`", "`adirmc`"),
  Covariates = c("yes", "yes", "refused", "refused"),
  How = c("per-subject draws (Sobol)",
          "product grid, or Taylor expansion", "---", "---"),
  Dependence = c("yes", "yes", "---", "---"), check.names = FALSE),
  caption = "Covariate marginalisation by estimator.")
```

`adfo` and `adirmc` **error** rather than run: solving at the covariate mean and
inflating omega would look like a fit and be the ecological fallacy.

# When a coefficient cannot be trusted

Inside one study the model sees a covariate and a random effect only through
their sum, $u = \beta a + \omega z$. For a log-linear effect on a normal
covariate,

$$\beta a + \omega z \;\sim\; N\!\left(\beta\mu_a,\; \beta^2 s_a^2 + \omega^2\right)$$

so every marginal moment depends on $(\beta, \omega)$ **only** through
$\beta^2 s_a^2 + \omega^2$. Halve the coefficient, inflate the random effect to
compensate, and pooled aggregate data cannot tell the difference: the objective
moves by $1.7\times10^{-5}$ across $\beta = 0.30$ to $1.20$, and a fit will still
report a number.

Three things break the ridge, and you need at least one:

1. **Studies whose covariate distributions differ** --- differing spreads break
   the variance equation, differing means the mean one.
2. **A source reporting subgroups**, or held as a *model* that can be evaluated
   at any covariate value --- which is what the section above does. The same
   source, stratified, moves the objective by 155 with its minimum at the truth.
3. **A covariate on a parameter with no random effect**, identified by shape
   because there is no $\omega$ for its variance to hide in.

admixr2 warns when it can see that none of them holds.

## What a source that omitted a covariate still tells you

A study that never fitted a covariate did not report a null result. It folded
that effect into its random effect, exactly:

$$\omega'^2 = \omega^2 + \beta^2\,\mathrm{Var}(a)$$

for a log-linear effect on a normally distributed covariate. So it still carries
information about that coefficient --- through its *variance* rather than its
mean --- provided its prediction marginalises over the covariate rather than
stratifying on one it never fitted.

Two caveats. A covariate acting on more than one parameter also induces
off-diagonal $\mathrm{Cov}(\eta_i,\eta_j) = \beta_i\beta_j\mathrm{Var}(a)$ that a
published *diagonal* Omega cannot represent. And when the omitted covariate is
correlated with one the source *did* fit, the inflation uses the residual
variance $\mathrm{Var}(a)(1-\rho^2)$, and the published coefficient is itself
confounded.

**How much this actually buys is a separate question**, and
[Reuniting a split study](covariates-split.html) measures it: two studies from
one known truth, one of them omitting a covariate, reunited.

# What is refused, and why

```{r refusals, echo = FALSE}
knitr::kable(data.frame(
  `You will see` = c("does not support covariate marginalisation",
                     "does not describe covariate(s)",
                     "which the model never reads",
                     "no contrast to stratify on",
                     "is discrete (`values`)",
                     "are (near-)perfectly collinear",
                     "too few draws in the smallest stratum"),
  Meaning = c("`adfo`/`adirmc` have no covariate path",
              "a model covariate no study described",
              "a covariate name typo",
              "the source model reads none of these covariates",
              "`taylor` has no second derivative of a step function",
              "the expansion step collapses along a null direction",
              "the strata are finer than the pool can describe"),
  `What to do` = c("use `adgh` or `admc`",
                   "add a `cov_dist` entry, or a fixed `cov`",
                   "check the spelling against the model",
                   "drop `stratify`; it is generated marginal instead",
                   "use `cov_integration = \"quadrature\"`",
                   "drop one covariate",
                   "fewer strata, or raise `n_pool`"),
  check.names = FALSE), caption = "Every refusal, and its cure.")
```

Each is an error rather than a silent approximation, because every one of them
would otherwise produce a finite, plausible, wrong number.

# See also

- [Reuniting a split study](covariates-split.html) --- what covariate synthesis can and cannot recover
- [From a published figure to E, V and n](aggregate-data.html) --- where `E` and `V` come from
- [Joint fitting across studies](multiple-studies.html) --- the machinery strata reuse
- [Comparing the estimators](estimator-comparison.html) --- `admc` vs `adgh`
- [Getting started](admixr2.html)
