---
title: "Covariates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covariates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.2
)
source(Filter(file.exists, c("_common.R", "../_common.R", "vignettes/_common.R"))[1])
```

```{r libs, include=FALSE}
library(admixr2); library(rxode2); library(nlmixr2); library(ggplot2)
```

## The question

A woman with a creatinine clearance of 21 mL/min needs a dose. Three trials have
been published. **Two of the three fitted no renal effect** --- they enrolled
narrow ranges of renal function, so within those trials the effect is invisible,
absorbed into each paper's clearance. The third enrolled across renal function
and did estimate it.

So the evidence arrives in two different shapes, and admixr2 has to take both:
one paper that reports a renal *relationship*, and two that report only the
renal function of the patients they happened to enrol --- a contrast visible
only *between* them.

## Three papers

Simulated so there is a truth to check against. Three trials, three
investigators, three models that share a drug and nothing else. Sato and Ito
enrolled narrow renal ranges and fitted no renal term; Khan enrolled across
renal function and fitted one.


```{r sim, include=FALSE}
set.seed(11)
TIMES <- c(0.5, 1, 2, 4, 8, 12, 24); DOSE <- 200
CL70 <- 5; V70 <- 50; BCRCL <- 0.6; BSEX <- 0.18
OM_CL <- 0.05; SD_ADD <- 0.08
R <- matrix(c(1, .45, .45, 1), 2, 2,
            dimnames = list(c("WT", "CRCL"), c("WT", "CRCL")))

# A cohort of virtual patients: lognormal margins for weight and renal
# function, joined by the Gaussian copula above; sex drawn independently.
# Used again later for the severe cohort, so it takes its medians as arguments.
draw_cohort <- function(n, wt_median, wt_cv, crcl_median, crcl_cv,
                        p_male = 0.55) {

  # a Gaussian copula, one step at a time: correlated normals, turned into
  # correlated uniforms, then pushed through each covariate's own margin
  z          <- matrix(rnorm(2 * n), nrow = n, ncol = 2) %*% chol(R)
  u_wt       <- pnorm(z[, 1])
  u_crcl     <- pnorm(z[, 2])

  data.frame(
    WT   = qlnorm(u_wt,   meanlog = log(wt_median),   sdlog = wt_cv),
    CRCL = qlnorm(u_crcl, meanlog = log(crcl_median), sdlog = crcl_cv),
    SEX  = rbinom(n, size = 1L, prob = p_male))
}

# Concentration after an IV bolus into one compartment.
conc <- function(t, dose, cl, v) dose/v * exp(-cl/v * t)

# Each patient's true clearance. `bcrcl = 0` gives a model with no renal term,
# which is what all three published models will turn out to be.
patient_cl <- function(cohort, cl_70kg, bcrcl, bsex, omega) {
  cl_70kg * (cohort$WT/70)^0.75 * (cohort$CRCL/90)^bcrcl *
    exp(bsex * cohort$SEX) *
    exp(rnorm(nrow(cohort), 0, sqrt(omega)))
}

# One trial's worth of individual data, in nlmixr2 layout.
# `renal_within = FALSE` gives every patient the COHORT MEDIAN renal function,
# so the renal effect is a purely BETWEEN-cohort one and no single trial can see
# it -- which is the situation two of these three papers are in. Khan's trial
# gets `TRUE`: renal function varies within it, which is why Khan's analysts
# could estimate the effect and the others could not.
run_trial <- function(cohort, first_id, renal_within = FALSE) {
  reference <- cohort
  if (!renal_within) reference$CRCL <- median(cohort$CRCL)
  cl <- patient_cl(reference, CL70, BCRCL, BSEX, OM_CL)
  v  <- V70 * (cohort$WT/70)
  records <- list()
  for (i in seq_len(nrow(cohort))) {
    y <- conc(TIMES, DOSE, cl[i], v[i]) + rnorm(length(TIMES), 0, SD_ADD)
    records[[i]] <- data.frame(
      ID   = first_id + i,
      TIME = c(0, TIMES),                        # dose row, then observations
      DV   = c(NA_real_, y),
      AMT  = c(DOSE, rep(0, length(TIMES))),
      EVID = c(1L, rep(0L, length(TIMES))),
      CMT  = 1L,
      WT   = cohort$WT[i],
      CRCL = cohort$CRCL[i],
      SEX  = cohort$SEX[i])
  }
  do.call(rbind, records)
}

cohorts <- list(normal   = draw_cohort(260L, 78, .198, 95, .05),
                # Khan enrolled ACROSS renal function rather than within a
                # range -- which is why Khan's analysts could estimate a renal
                # effect and the other two could not.
                mild     = draw_cohort(210L, 76, .198, 62, .35),
                moderate = draw_cohort(180L, 74, .198, 38, .05))

trial_data <- list(
  normal   = run_trial(cohorts$normal,   first_id = 10000),
  mild     = run_trial(cohorts$mild,     first_id = 20000,
                       renal_within = TRUE),
  moderate = run_trial(cohorts$moderate, first_id = 30000))
```

```{r hidden, include=FALSE}
d1  <- trial_data$normal[trial_data$normal$EVID == 0, ]
agg <- data.frame(t = TIMES,
                  m = tapply(d1$DV, d1$TIME, mean),
                  s = tapply(d1$DV, d1$TIME, sd),
                  panel = "What the paper reports")
ind <- d1[d1$ID %in% head(unique(d1$ID), 150), ]
ind$panel <- "Individual patients (never published)"

ggplot() +
  geom_line(data = ind, aes(TIME, DV, group = ID, colour = CRCL),
            alpha = .35, linewidth = .3) +
  geom_ribbon(data = agg, aes(t, ymin = m - s, ymax = m + s),
              fill = "grey55", alpha = .35) +
  geom_line(data = agg, aes(t, m), linewidth = 1) +
  scale_colour_viridis_c(name = "CRCL (mL/min)", option = "C", end = .9) +
  facet_wrap(~panel) + labs(x = "time (h)", y = "concentration (mg/L)") +
  theme_bw() + theme(legend.position = "bottom")
```

```{r trialfits, include=FALSE}
# THREE ANALYSTS, THREE MODELS. They share a drug and nothing else: each wrote
# their own model, with its own parameters, and admixr2 needs no agreement
# between them. What differs here is the covariate structure -- the usual place
# published models diverge -- but the residual and random-effect structure may
# differ just as freely.

# Sato: the conventional write-up. Allometry fixed at the textbook exponents,
# one sex effect on clearance.
sato_model <- function() {
  ini({
    tcl     <- log(4.5)
    tv      <- log(45)
    bsex    <- 0.10
    add.err <- 0.2
    eta.cl  ~ 0.1
  })
  model({
    cl <- exp(tcl + eta.cl) * (WT/70)^0.75 * exp(bsex * SEX)
    v  <- exp(tv) * (WT/70)
    cp <- linCmt()
    cp ~ add(add.err)
  })
}

# Ito: estimated BOTH allometric exponents rather than fixing them. Two extra
# parameters, and a covariance matrix two rows wider than Sato's.
ito_model <- function() {
  ini({
    tcl     <- log(4.5)
    tv      <- log(45)
    bwtcl   <- 0.75
    bwtv    <- 1
    bsex    <- 0.10
    add.err <- 0.2
    eta.cl  ~ 0.1
  })
  model({
    cl <- exp(tcl + eta.cl) * (WT/70)^bwtcl * exp(bsex * SEX)
    v  <- exp(tv) * (WT/70)^bwtv
    cp <- linCmt()
    cp ~ add(add.err)
  })
}

# Khan: the only one of the three who ESTIMATED A RENAL EFFECT. Their trial
# enrolled across renal function, so the effect was identifiable within it.
khan_model <- function() {
  ini({
    tcl     <- log(4.5)
    tv      <- log(45)
    bsex    <- 0.10
    bcrcl   <- 0.3
    add.err <- 0.2
    eta.cl  ~ 0.1
  })
  model({
    cl <- exp(tcl + eta.cl) * (WT/70)^0.75 * (CRCL/90)^bcrcl *
      exp(bsex * SEX)
    v  <- exp(tv) * (WT/70)
    cp <- linCmt()
    cp ~ add(add.err)
  })
}

.ctl <- foceiControl(print = 0L, covMethod = "r")
fit_normal   <- nlmixr2(sato_model, trial_data$normal,   est = "focei",
                        control = .ctl)
fit_moderate <- nlmixr2(ito_model,  trial_data$moderate, est = "focei",
                        control = .ctl)
fit_mild     <- nlmixr2(khan_model, trial_data$mild,     est = "focei",
                        control = .ctl)
```

```{r published, echo=FALSE}
# What a paper prints: the model's parameter estimates, and the baseline table
# describing who was enrolled. Nothing about the fit's own covariance is used --
# a study generated from a published model is not a sample, so no standard error
# is reported for a fit that includes one (see the last section).
normal_paper <- list(
  CL   = exp(fit_normal$parFixedDf["tcl", "Estimate"]),
  V    = exp(fit_normal$parFixedDf["tv",  "Estimate"]),
  BSEX = fit_normal$parFixedDf["bsex",    "Estimate"],
  OM   = as.numeric(fit_normal$omega[1, 1]))

mild_paper <- list(
  CL    = exp(fit_mild$parFixedDf["tcl", "Estimate"]),
  V     = exp(fit_mild$parFixedDf["tv",  "Estimate"]),
  BSEX  = fit_mild$parFixedDf["bsex",    "Estimate"],
  # Khan's own renal exponent -- the only one of the three to report one.
  BCRCL = fit_mild$parFixedDf["bcrcl",   "Estimate"],
  OM    = as.numeric(fit_mild$omega[1, 1]))

moderate_paper <- list(
  CL   = exp(fit_moderate$parFixedDf["tcl", "Estimate"]),
  V    = exp(fit_moderate$parFixedDf["tv",  "Estimate"]),
  BSEX = fit_moderate$parFixedDf["bsex",    "Estimate"],
  OM   = as.numeric(fit_moderate$omega[1, 1]))

knitr::kable(data.frame(
  cohort      = c("normal", "mild", "moderate"),
  median_CRCL = round(c(median(cohorts$normal$CRCL),
                        median(cohorts$mild$CRCL),
                        median(cohorts$moderate$CRCL))),
  CL_at_70kg  = c(normal_paper$CL,   mild_paper$CL,   moderate_paper$CL),
  V_at_70kg   = c(normal_paper$V,    mild_paper$V,    moderate_paper$V),
  sex_effect  = c(normal_paper$BSEX, mild_paper$BSEX, moderate_paper$BSEX)),
  digits = 3, row.names = FALSE,
  caption = "Three independent analyses, three different models, each sound on its own data.")
```

Clearance falls with renal function across the three papers. Only Khan's has a
renal parameter to say so with; in the other two the effect is buried in the
clearance each reports.

## Giving admixr2 the papers

A study is a transcription of what a paper reports: its **model at its published
values**, and the **population it enrolled**. Note the difference between the
three sources, because it is the distinction the whole method turns on:

* `normal` and `moderate` estimated no renal effect, so `CRCL` is **marginal**
  for them: they describe the patients they enrolled and admixr2 integrates
  over that. There is no contrast in either paper to condition on, and splitting
  one on a covariate its model never saw would manufacture evidence.
* `mild` **does** estimate it, so `CRCL` is **conditional** for that source and
  the paper can be read at values of it.

None of that is declared. admixr2 reads each source's own model and works it
out, which is why all three calls above are identical.

The test is ESTIMATED, not merely read. All three models carry weight, two of
them at the conventional fixed exponent `(WT/70)^0.75` --- and a fixed exponent
is an assumption rather than a finding, so for those two `WT` is marginal too.
Only `moderate`, which fitted its own exponent, is conditional on it. All three
fitted a sex effect, so all three are conditional on sex. Conditioning is not
cosmetic: a covariate every study marginalises over is not identified against a
random effect on the same parameter, and over 720 replicates the
likelihood-ratio test was sized **0.125** against a nominal 0.05 with nothing
conditional, **0.058** with all three.


```{r datagen}
# --- each paper's model ----------------------------------------------------
# The source model is that analyst's own model AT their own published values,
# and a fit already is exactly that: `fit$ui` carries the model together with
# the estimates it converged on. Three different models go in, and nothing has
# to be shared between them.
#
# With a real paper you have no fit object. You write the model out yourself and
# hand admStudy() the printed numbers -- `est = c(tcl = log(5.2), ...)` -- which
# is the same study, transcribed instead of recovered.
sato_published <- fit_normal$ui        # normal renal function
ito_published  <- fit_moderate$ui      # moderate impairment
khan_published <- fit_mild$ui          # mild impairment

# --- each paper's study ----------------------------------------------------
normal_study <- admStudy(
  model      = sato_published,
  population = cohorts$normal,
  dose       = DOSE,
  times      = TIMES)

moderate_study <- admStudy(
  model      = ito_published,
  population = cohorts$moderate,
  dose       = DOSE,
  times      = TIMES)

# IDENTICAL CALL, different consequence. Khan's model uses renal function and
# the other two never mention it, so renal function is CONDITIONAL for this
# source and MARGINAL for them -- and admixr2 reads that off the models. There
# is nothing here to say it with.
mild_study <- admStudy(
  model      = khan_published,
  population = cohorts$mild,
  dose       = DOSE,
  times      = TIMES)

studies <- admStudies(normal   = normal_study,
                      moderate = moderate_study,
                      mild     = mild_study)

# the three papers together, used by the figures further down
published <- list(normal = normal_paper, mild = mild_paper,
                  moderate = moderate_paper)

studies
```

## Pooling them


```{r admfit}
adm_model <- function() {
  ini({
    tcl     <- log(4)
    tv      <- log(45)
    bcrcl   <- 0.3            # the renal effect no published model has
    bsex    <- 0.10
    add.err <- 0.1
    eta.cl  ~ 0.1
  })
  model({
    cl <- exp(tcl + eta.cl) * (WT/70)^0.75 * (CRCL/90)^bcrcl * exp(bsex * SEX)
    v  <- exp(tv) * (WT/70)
    cp <- linCmt()
    cp ~ add(add.err)
  })
}
fit <- nlmixr2(adm_model, admData(), est = "adgh",
               control = adghControl(studies = studies, print = 0L))
pf     <- fit$parFixedDf
stopifnot(abs(pf["bsex", "Estimate"] - BSEX) < 0.08)
# NO INTERVALS HERE, and that is the point: every study in this vignette is a
# published model rather than a sample, so there is no sampling law to build one
# from and admixr2 reports no standard error at all. The estimates are what the
# three papers jointly imply; whether the renal effect is real is settled by the
# likelihood-ratio test below, which needs no standard error.
knitr::kable(data.frame(
  parameter = c("CL at 70 kg, CRCL 90, women (L/h)", "V at 70 kg (L)",
                "renal exponent", "sex effect (log)", "additive SD (mg/L)",
                "omega (eta.cl)"),
  truth     = c(CL70, V70, BCRCL, BSEX, SD_ADD, OM_CL),
  estimate  = c(pf["tcl", "Back-transformed"], pf["tv", "Back-transformed"],
                pf["bcrcl", "Back-transformed"], pf["bsex", "Back-transformed"],
                pf["add.err", "Back-transformed"],
                as.numeric(fit$omega[1L, 1L]))),
  digits = 3, row.names = FALSE,
  caption = "Recovered from three analyses, none of which contained a renal term.")
```

The renal exponent comes back at
`r sprintf("%.2f", pf["bcrcl", "Estimate"])` against a truth of `r BCRCL`, out
of three analyses only one of which contained a renal term --- and that one put
it at `r sprintf("%.2f", fit_mild$parFixedDf["bcrcl", "Estimate"])`.
There are **no intervals**: every source here is a published model rather than a
sample, so there is no sampling law to build one from and admixr2 reports no
standard error at all.

The sex effect shows what pooling costs. The three trials reported
`r paste(sprintf("%.2f", sapply(published, function(p) p$BSEX)), collapse = ", ")`
for a quantity genuinely identical in all three; sampling noise alone put them
that far apart. Reconciling that is not free --- some of it lands in the other
parameters, which is why the renal exponent comes back a little below its truth.

## Is the shape right?

Three cohorts at three renal medians settle *that* renal function matters. They
do not settle **what shape** it has. The fitted model uses a power,
`(CRCL/90)^bcrcl`, which is what generated the cohorts. Fit the obvious
alternative --- linear in `CRCL` --- to the same sources.


```{r linear-form}
adm_model_lin <- function() {
  ini({
    tcl     <- log(4)
    tv      <- log(45)
    bcrcl   <- 0.3
    bsex    <- 0.10
    add.err <- 0.1
    eta.cl  ~ 0.1
  })
  model({
    # LINEAR in CRCL where the truth is a power -- same parameter count, so
    # there is no likelihood-ratio test to appeal to.
    cl <- exp(tcl + eta.cl) * (WT/70)^0.75 *
      (1 + bcrcl * (CRCL/90 - 1)) * exp(bsex * SEX)
    v  <- exp(tv) * (WT/70)
    cp <- linCmt()
    cp ~ add(add.err)
  })
}
fit_lin <- nlmixr2(adm_model_lin, admData(), est = "adgh",
                   control = adghControl(studies = studies, print = 0L))
```

Both models have the **same number of parameters**, so there is no
likelihood-ratio test to appeal to. All that is left is the objective, and it
decides nothing:


```{r form-obj, echo=FALSE}
knitr::kable(data.frame(
  form      = c("power (correct)", "linear"),
  objective = c(fit$objf, fit_lin$objf),
  bcrcl     = c(fit$parFixedDf["bcrcl", "Estimate"],
                fit_lin$parFixedDf["bcrcl", "Estimate"])),
  digits = 2, row.names = FALSE,
  caption = "Same parameter count, so no LRT -- and the objectives barely differ.")
cat(sprintf("dOFV (linear - power) = %+.2f on 0 extra parameters\n",
            fit_lin$objf - fit$objf))
```

The plot decides it.


```{r form-resid, fig.height = 3.4, fig.cap = "Between-study residual, linear model. The CRCL facet bends."}
plots_lin <- plot(fit_lin, which = "covariate")
```

```{r form-pattern, include=FALSE}
crcl_med <- vapply(cohorts[c("normal", "mild", "moderate")],
                   function(co) stats::median(co$CRCL), numeric(1))
.zby <- function(m) {
  a <- tapply(m$z, m$source, mean)
  a[c("moderate", "mild", "normal")]          # increasing CRCL
}
```

```{r form-moments, include=FALSE}
m_pow <- admMoments(fit,     n_sim = 300L)
m_lin <- admMoments(fit_lin, n_sim = 300L)
```

Three facets, and only one of them can say anything. `CRCL` separates the
sources cleanly. `SEX` is conditional, so each source contributes a pair and the grey
line joining them is the within-source contrast conditioning bought. `WT` is the
negative case: the three cohorts enrolled median weights of
`r paste(round(vapply(cohorts[c("moderate", "mild", "normal")], function(c) median(c$WT), 0)), collapse = ", ")`
kg, so they sit on top of one another and the facet cannot speak to weight
whichever way it tips.

Note also that a mark's size is the patients it speaks for, and that follows the
facet: on `SEX` each mark is one sex stratum, on `CRCL` the strata are together
and each mark is a whole paper.

A **missing** covariate shows up as a slope across the axis. A covariate with
the **wrong shape** shows up as **curvature** --- the linear model buys both ends
by inflating `bcrcl` to
`r sprintf("%.2f", fit_lin$parFixedDf["bcrcl", "Estimate"])`, and then misses the
middle. Mean standardised residual by source, in order of increasing `CRCL`
(`r paste(round(crcl_med[c("moderate", "mild", "normal")]), collapse = ", ")`
mL/min):

| form | moderate | mild | normal |
|---|---|---|---|
| power (correct) | `r sprintf("%+.2f", .zby(m_pow)[1])` | `r sprintf("%+.2f", .zby(m_pow)[2])` | `r sprintf("%+.2f", .zby(m_pow)[3])` |
| linear | `r sprintf("%+.2f", .zby(m_lin)[1])` | `r sprintf("%+.2f", .zby(m_lin)[2])` | `r sprintf("%+.2f", .zby(m_lin)[3])` |

The linear row is a **U**: both extremes above zero, the middle pulled to zero.
The power row has no such shape.

Read the magnitudes with care. Every source here is a published *model*, so its
`E` and `V` are exact functions of that model's parameters rather than a sample;
there is no sampling noise for a residual to be large against, and these never
approach ±1.96. **The pattern is the evidence, not the size.**

`admMoments()` returns the moments the panels are drawn from if you want to draw
your own; see `vignette("diagnostic-plots")`.

## Seeing the renal effect

What the meta-analysis actually did. Sato's and Ito's models are **flat bars**
--- no renal term, so each claims one clearance over the whole range it covered.
Khan's has a slope of its own. The meta-analysis is the line through all three.

```{r mechanism, fig.height = 4.2, echo = FALSE}
e  <- setNames(pf$Estimate, rownames(pf))
crcl_grid <- seq(15, 115, length.out = 120)
sexlab <- c("women", "men")

# the meta-analysis: one curve per sex. The renal effect is the SLOPE, the sex
# effect is the vertical GAP between the two.
curve_df <- data.frame()
for (sx in 0:1) {
  cl_curve <- exp(e[["tcl"]]) * (crcl_grid/90)^e[["bcrcl"]] *
              exp(e[["bsex"]] * sx)
  curve_df <- rbind(curve_df, data.frame(
    CRCL = crcl_grid, sex = sexlab[sx + 1L], CL = cl_curve))
}

# each published model: no renal term, so a FLAT bar per sex, drawn over the
# CRCL range that trial actually covered
seg <- data.frame()
pts <- data.frame()
for (trial in names(published)) {
  paper     <- published[[trial]]
  crcl      <- cohorts[[trial]]$CRCL
  crcl_p10  <- unname(quantile(crcl, 0.10))
  crcl_p90  <- unname(quantile(crcl, 0.90))
  # A paper with no renal term claims ONE clearance over its whole range, so its
  # bar is flat. Khan reported an exponent, so Khan's bar has that slope -- and
  # the meta-analysis can be read against it rather than only against a level.
  bc <- if (is.null(paper$BCRCL)) 0 else paper$BCRCL
  for (sx in 0:1) {
    base    <- paper$CL * exp(paper$BSEX * sx)
    cl_mid  <- base * (median(crcl)/90)^bc
    seg <- rbind(seg, data.frame(
      lab = trial, sex = sexlab[sx + 1L],
      CL = base * (crcl_p10/90)^bc, CLh = base * (crcl_p90/90)^bc,
      lo = crcl_p10, hi = crcl_p90))
    pts <- rbind(pts, data.frame(
      lab = trial, sex = sexlab[sx + 1L],
      CRCL = median(crcl), CL = cl_mid))
  }
}

ggplot(curve_df, aes(CRCL, CL, colour = sex)) +
  annotate("rect", xmin = 15, xmax = 30, ymin = -Inf, ymax = Inf,
           fill = "grey85", alpha = .6) +
  annotate("text", x = 22.5, y = max(curve_df$CL), label = "severe\n(no trial)",
           size = 3, vjust = 1.1) +
  geom_segment(data = seg, aes(x = lo, xend = hi, y = CL, yend = CLh),
               linewidth = 1.5, alpha = .55) +
  geom_line(linewidth = 1) +
  geom_point(data = pts, size = 2.6) +
  geom_text(data = subset(pts, sex == "men"), aes(label = lab),
            vjust = -1.2, size = 3.1, show.legend = FALSE) +
  scale_colour_manual(values = c(women = "steelblue4", men = "darkorange3"),
                      name = NULL) +
  labs(x = "median creatinine clearance (mL/min)",
       y = "clearance at 70 kg (L/h)",
       subtitle = paste("bars: what each published model claims, over the range",
                        "it covered\nlines: the meta-analysis")) +
  theme_bw() + theme(legend.position = "bottom")
```

The slope is the renal effect, recovered from three sources that each had none.
The vertical gap between the two lines is the sex effect. The grey band is where
the patient is, and no trial went there.

## The dose


```{r predict, include=FALSE}
set.seed(3)
# mean concentration over a severe cohort (WT 72 kg, CRCL 21 mL/min), reusing
# the same two helpers the trials were simulated with
severe_mean_conc <- function(cl_70kg, v_70kg, bcrcl, bsex, omega, n = 40000L) {
  pts <- draw_cohort(n, 72, .198, 21, .217)
  cl  <- patient_cl(pts, cl_70kg, bcrcl, bsex, omega)
  v   <- v_70kg * (pts$WT/70)
  sapply(TIMES, function(t) mean(conc(t, DOSE, cl, v)))
}

p_adm   <- severe_mean_conc(exp(e[["tcl"]]), exp(e[["tv"]]), e[["bcrcl"]],
                            e[["bsex"]], as.numeric(fit$omega[1L, 1L]))
p_near  <- severe_mean_conc(moderate_paper$CL, moderate_paper$V,
                            0,                       # no renal term
                            moderate_paper$BSEX, moderate_paper$OM)
p_truth <- severe_mean_conc(CL70, V70, BCRCL, BSEX, OM_CL)
```

```{r dose, echo=FALSE}
set.seed(5)
mean_cl <- function(cl_70kg, bcrcl, bsex, omega, wt_median, crcl_median,
                    n = 40000L) {
  pts <- draw_cohort(n, wt_median, .198, crcl_median, .217)
  mean(patient_cl(pts, cl_70kg, bcrcl, bsex, omega))
}
# a severe cohort (WT 72, CRCL 21) against a normal-function one (WT 78, CRCL 95)
dose_for <- function(cl_70kg, bcrcl, bsex, omega)
  DOSE * mean_cl(cl_70kg, bcrcl, bsex, omega, 72, 21) /
         mean_cl(cl_70kg, bcrcl, bsex, omega, 78, 95)
d_true <- dose_for(CL70, BCRCL, BSEX, OM_CL)
d_meta <- dose_for(exp(e[["tcl"]]), e[["bcrcl"]], e[["bsex"]],
                   as.numeric(fit$omega[1L, 1L]))
d_near <- dose_for(moderate_paper$CL, 0,
                   moderate_paper$BSEX, moderate_paper$OM)
knitr::kable(data.frame(
  basis   = c("the truth", "meta-analysis (admixr2)", "nearest published model"),
  dose_mg = round(c(d_true, d_meta, d_near))),
  digits = 0, row.names = FALSE,
  caption = paste("Dose giving a CrCl 21 mL/min patient the exposure a",
                  "normal-function patient gets from", DOSE, "mg."))
```

The nearest published model has no renal term, so it cannot adjust for her at
all: it hands back the dose it was fitted at, for a cohort whose median renal
function was
`r sprintf("%.1f", median(cohorts$moderate$CRCL) / 21)` times hers. The
meta-analysis lands within `r round(abs(d_meta - d_true))` mg of the truth,
from three papers only one of which had a renal term to contribute.

## See also

- [Diagnostic plots](diagnostic-plots.html) --- what each panel is for
- [Multiple studies](multiple-studies.html)
