---
title: "Covariates in aggregate data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covariates in aggregate data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.4
)
source(Filter(file.exists, c("_common.R", "../_common.R", "vignettes/_common.R"))[1])
```

## The problem

You need a dose for patients with **severe renal impairment**. Three trials of
the drug have been published — one in normal renal function, one in mild
impairment, one in moderate — and none enrolled a severe patient, because that
population is routinely excluded.

Worse, none of the three published a renal effect. Each analysed its own trial,
each found a population where creatinine clearance barely varied, and each
sensibly left it out of the model. So you have three population models that
disagree about clearance and none of them says why.

This vignette builds that situation from scratch, then shows that pooling the
three published models recovers the renal effect none of them could estimate —
and predicts the cohort nobody ran.

```{r libs}
library(admixr2)
library(rxode2)
library(nlmixr2)
library(ggplot2)
```

## Three populations

Start with the patients. Weight and creatinine clearance are correlated —
larger patients clear more — so they are drawn jointly through a Gaussian
copula: each keeps its own lognormal margin, and the dependence is imposed on
the latent scale.

```{r populations}
set.seed(11)
R <- matrix(c(1, 0.45, 0.45, 1), 2, 2,
            dimnames = list(c("WT", "CRCL"), c("WT", "CRCL")))
draw_pop <- function(n, med_wt, sd_wt, med_crcl, sd_crcl) {
  Z <- matrix(stats::rnorm(2 * n), n, 2) %*% chol(R)
  data.frame(WT   = stats::qlnorm(stats::pnorm(Z[, 1]), log(med_wt),   sd_wt),
             CRCL = stats::qlnorm(stats::pnorm(Z[, 2]), log(med_crcl), sd_crcl))
}
cohorts <- list(
  normal   = draw_pop(120L, 78, 0.198, 95, 0.217),
  mild     = draw_pop( 90L, 76, 0.198, 62, 0.198),
  moderate = draw_pop( 70L, 74, 0.198, 38, 0.198)
)
```

```{r popfig, fig.height = 3.2}
dd <- do.call(rbind, Map(function(cv, nm)
  data.frame(cohort = nm, CRCL = cv$CRCL), cohorts, names(cohorts)))
dd$cohort <- factor(dd$cohort, levels = names(cohorts))
ggplot(dd, aes(CRCL, fill = cohort)) +
  geom_density(alpha = 0.45, colour = NA) +
  geom_vline(xintercept = 21, linetype = "dashed") +
  annotate("text", x = 21, y = Inf, label = " severe: the cohort we need",
           hjust = 0, vjust = 1.6, size = 3.2) +
  labs(x = "creatinine clearance (mL/min)", y = NULL, fill = NULL) +
  theme_bw() + theme(legend.position = "bottom")
```

Each trial spans a narrow band, and the population we need a dose for sits
below all of them. Note what that means for any single trial: within one
cohort, creatinine clearance barely varies, so no one trial can estimate what
it does.

## What each trial measured

A 200 mg intravenous dose, seven samples. The truth — never known to any of the
three analysts — is CL = 5 L/h and V = 50 L at 70 kg, allometric scaling on
weight, and a renal exponent of 0.6.

```{r simulate}
TIMES <- c(0.5, 1, 2, 4, 8, 12, 24); DOSE <- 200
CL70 <- 5; V70 <- 50; BCRCL <- 0.6; OM_CL <- 0.06; SD_ADD <- 0.08

sim_trial <- function(cv, id0) {
  n  <- nrow(cv)
  cl <- CL70 * (cv$WT / 70)^0.75 * (cv$CRCL / 90)^BCRCL *
        exp(stats::rnorm(n, 0, sqrt(OM_CL)))
  v  <- V70 * (cv$WT / 70)
  do.call(rbind, lapply(seq_len(n), function(i) rbind(
    data.frame(ID = id0 + i, TIME = 0, DV = NA_real_, AMT = DOSE, EVID = 1,
               CMT = 1, WT = cv$WT[i], CRCL = cv$CRCL[i]),
    data.frame(ID = id0 + i, TIME = TIMES,
               DV = DOSE / v[i] * exp(-cl[i] / v[i] * TIMES) +
                    stats::rnorm(length(TIMES), 0, SD_ADD),
               AMT = 0, EVID = 0, CMT = 1,
               WT = cv$WT[i], CRCL = cv$CRCL[i]))))
}
trial_data <- Map(function(cv, k) sim_trial(cv, k * 10000L),
                  cohorts, seq_along(cohorts))
```

Here is the normal-function trial as it happened, and as a paper would report
it. Every patient has a creatinine clearance, and it fans the profiles out —
but the published summary keeps only the mean and the spread.

```{r hidden, fig.height = 4.0}
d1 <- trial_data$normal[trial_data$normal$EVID == 0, ]
ind <- d1[d1$ID %in% head(unique(d1$ID), 150), ]
ind$panel <- "Individual patients (never published)"
agg <- do.call(rbind, lapply(split(d1$DV, d1$TIME), function(x)
  data.frame(m = mean(x), s = stats::sd(x))))
agg <- data.frame(t = as.numeric(names(split(d1$DV, d1$TIME))), agg,
                  panel = "What the paper reports")

ggplot() +
  geom_line(data = ind, aes(TIME, DV, group = ID, colour = CRCL),
            alpha = 0.35, linewidth = 0.3) +
  geom_ribbon(data = agg, aes(t, ymin = m - s, ymax = m + s),
              fill = "grey55", alpha = 0.35) +
  geom_line(data = agg, aes(t, m), linewidth = 1) +
  geom_point(data = agg, aes(t, m), size = 2) +
  scale_colour_viridis_c(name = "CRCL (mL/min)", option = "C", end = 0.9) +
  facet_wrap(~panel) +
  labs(x = "time (h)", y = "concentration (mg/L)") +
  theme_bw() + theme(legend.position = "bottom")
```

The covariate is *invisible* in what gets published. The only trace renal
function leaves is in the **width** of that ribbon — poor clearers at the top,
good clearers at the bottom.

## What each trial published

Each analyst fits their own trial. Weight is scaled allometrically, with the
exponents fixed at 0.75 and 1 as convention requires. None of them includes
creatinine clearance, because within their own cohort it hardly moves.

```{r trialfits}
trial_model <- function() {
  ini({ tcl <- log(4.5); tv <- log(45); eta.cl ~ 0.1; add.err <- 0.2 })
  model({
    cl <- exp(tcl + eta.cl) * (WT / 70)^0.75
    v  <- exp(tv) * (WT / 70)
    cp <- linCmt()
    cp ~ add(add.err)
  })
}
trial_fits <- lapply(trial_data, function(d)
  suppressWarnings(nlmixr2(trial_model, d, est = "focei",
                           control = foceiControl(print = 0L))))
published <- lapply(trial_fits, function(f) list(
  CL  = exp(f$parFixedDf["tcl", "Estimate"]),
  V   = exp(f$parFixedDf["tv",  "Estimate"]),
  OM  = as.numeric(f$omega[1L, 1L]),
  ADD = f$parFixedDf["add.err", "Estimate"]))
knitr::kable(
  data.frame(
    cohort      = names(published),
    median_CRCL = round(vapply(cohorts, function(c) stats::median(c$CRCL),
                               numeric(1))),
    CL          = vapply(published, `[[`, numeric(1), "CL"),
    V           = vapply(published, `[[`, numeric(1), "V"),
    omega       = vapply(published, `[[`, numeric(1), "OM"),
    add_SD      = vapply(published, `[[`, numeric(1), "ADD")),
  digits = 3, row.names = FALSE,
  caption = "Three independent analyses, each of its own trial."
)
```

**This is the whole problem in one table.** Clearance differs by nearly a
factor of two across the three publications; volume is the same in all three.
Each analysis is perfectly sound on its own data, and none of them can say why
the clearances differ — the explanation lies in a covariate that does not vary
within any single trial.

A reader of these three papers is left with three incompatible models and no
way to choose.

## Their models as our input

Aggregate data modelling takes two kinds of input: a digitised summary read off
a published figure, or a **previously published model, supplied directly**.
This is the second kind. We are not simulating profiles from a model of our own
invention — we are using three published analyses as evidence, exactly as their
authors reported them.

`datagen()` turns a published model, plus the design and population it applies
to, into the `(E, V, n)` a figure would have given us:

```{r datagen}
CLp <- Vp <- OMp <- ADDp <- NULL          # filled per trial, read by ini()
published_model <- function() {
  ini({ tcl <- log(CLp); tv <- log(Vp); eta.cl ~ OMp; add.err <- ADDp })
  model({
    cl <- exp(tcl + eta.cl) * (WT / 70)^0.75
    v  <- exp(tv) * (WT / 70)
    cp <- linCmt()
    cp ~ add(add.err)
  })
}
cov_dists <- lapply(cohorts, function(cv) covDist(
  WT   = list(meanlog = mean(log(cv$WT)),   sdlog = stats::sd(log(cv$WT))),
  CRCL = list(meanlog = mean(log(cv$CRCL)), sdlog = stats::sd(log(cv$CRCL))),
  cor  = R))

studies <- list()
for (nm in names(published)) {
  CLp <- published[[nm]]$CL;  Vp   <- published[[nm]]$V
  OMp <- published[[nm]]$OM;  ADDp <- published[[nm]]$ADD
  studies[[nm]] <- datagen(
    list(s = list(times = TIMES, ev = rxode2::et(amt = DOSE),
                  n = nrow(cohorts[[nm]]), cov_dist = cov_dists[[nm]])),
    model   = published_model,
    control = datagenControl(method = "gh", seed = 1L))$s
}
str(studies$normal[c("E", "n")])
```

Two things are being supplied per study. The **model** is the trial's published
one. The **`cov_dist`** is the population it was fitted in — which the paper
describes in its baseline table, and which we now declare so admixr2 can
integrate over it. `method = "gh"` uses quadrature, so no Monte-Carlo noise
enters the meta-analysis.

The published models have no renal term. The `cov_dist` does. That asymmetry is
the point: we are telling admixr2 *who was enrolled*, and letting it work out
what that implies.

### Specifying the marginal

Those `cov_dist` values came from simulated patients, which a real analysis
never has. What a paper gives you is a baseline table, and `covDist()` takes
whichever summary that is:

```{r forms}
# mean and SD -- moment-matched onto the margin you name
covDist(WT = c(mean = 78, sd = 16), dist = "lnorm")$WT

# a categorical covariate as the proportions a baseline table prints
covDist(SEX = c(male = 0.55, female = 0.45))$SEX
```

| the paper reports | use |
|---|---|
| mean and SD | `c(mean = , sd = )` with `dist = "lnorm"` |
| geometric mean `G`, CV as a fraction | `meanlog = log(G)`, `sdlog = sqrt(log(1 + CV^2))` |
| median `M` and IQR `Q1`–`Q3` | `meanlog = log(M)`, `sdlog = (log(Q3) - log(Q1)) / 1.349` |
| median and full range only | `sdlog ≈ (log(max) - log(min)) / 4`, and say you approximated |

```{r iqr}
# "median weight 78 kg (IQR 68-89)"
covDist(WT = c(meanlog = log(78), sdlog = (log(89) - log(68)) / 1.349))$WT
```

A whole baseline table goes in at once as a data frame with `covariate`,
`mean`, `sd` and `dist` columns, which is usually the least error-prone route.
For anything these do not cover, pass a `quantile` function.

Prefer `"lnorm"` for anything positive: a **normal** margin is unbounded below
and the quadrature reaches about 3.75 SD, so a covariate with a CV above
roughly 0.27 puts a node at or below zero, where an allometric term is `NaN`.
`covDist()` warns when it sees this.

**The marginal is data, not a parameter.** admixr2 never estimates it. You are
asserting what the enrolled population looked like, on the paper's authority —
and if the paper is vague, that is a real limitation of your analysis rather
than something to bury in an `sdlog`.

## The meta-analysis

Now fit all three together, with the renal term none of them could estimate.

```{r admfit}
adm_model <- function() {
  ini({ tcl <- log(4); tv <- log(45); bcrcl <- 0.3
        eta.cl ~ 0.1; add.err <- 0.1 })
  model({
    cl <- exp(tcl + eta.cl) * (WT / 70)^0.75 * (CRCL / 90)^bcrcl
    v  <- exp(tv) * (WT / 70)
    cp <- linCmt()
    cp ~ add(add.err)
  })
}
fit <- nlmixr2(adm_model, admData(), est = "adgh",
               control = adghControl(studies = studies, print = 0L,
                                     covMethod = "r"))
```

```{r admtable}
e  <- stats::setNames(fit$parFixedDf$Estimate, rownames(fit$parFixedDf))
se <- stats::setNames(fit$parFixedDf$SE,       rownames(fit$parFixedDf))
knitr::kable(
  data.frame(
    quantity = c("CL at 70 kg, CRCL 90 (L/h)", "V at 70 kg (L)",
                 "renal exponent", "omega (eta.cl)", "additive SD (mg/L)"),
    truth    = c(CL70, V70, BCRCL, OM_CL, SD_ADD),
    estimate = c(exp(e[["tcl"]]), exp(e[["tv"]]), e[["bcrcl"]],
                 as.numeric(fit$omega[1L, 1L]), e[["add.err"]]),
    SE       = c(NA, NA, se[["bcrcl"]], NA, se[["add.err"]])),
  digits = 3,
  caption = "The meta-analysis across three published models."
)
```

The renal exponent is recovered, with a standard error small enough to be
useful — **from three analyses none of which contained it**. It is identified
by the *contrast* between the cohorts: three populations, three clearances, and
a covariate that explains the difference.

```{r diag, fig.height = 4.0}
plot(fit, which = "mean")$mean_moderate
```

## Predicting the cohort nobody ran

What would you do without this? You would take the published model closest to
your patients — the moderate-impairment one — and use it. Compare that against
the meta-analysis, for a severe population with median CRCL 21 mL/min and
median weight 72 kg.

```{r predict}
set.seed(3)
marg_mean <- function(cl_ref, v_ref, bcrcl, om, n = 40000L) {
  Z    <- matrix(stats::rnorm(2 * n), n, 2) %*% chol(R)
  WT   <- stats::qlnorm(stats::pnorm(Z[, 1]), log(72), 0.198)
  CRCL <- stats::qlnorm(stats::pnorm(Z[, 2]), log(21), 0.217)
  cl <- cl_ref * (WT / 70)^0.75 *
        (if (is.na(bcrcl)) 1 else (CRCL / 90)^bcrcl) *
        exp(stats::rnorm(n, 0, sqrt(om)))
  v  <- v_ref * (WT / 70)
  vapply(TIMES, function(t) mean(DOSE / v * exp(-cl / v * t)), numeric(1))
}
p_adm   <- marg_mean(exp(e[["tcl"]]), exp(e[["tv"]]), e[["bcrcl"]],
                     as.numeric(fit$omega[1L, 1L]))
p_near  <- marg_mean(published$moderate$CL, published$moderate$V, NA,
                     published$moderate$OM)
p_truth <- marg_mean(CL70, V70, BCRCL, OM_CL)
```

```{r payoff, fig.height = 4.0}
pp <- rbind(
  data.frame(t = TIMES, cp = p_adm,  what = "meta-analysis, with CRCL"),
  data.frame(t = TIMES, cp = p_near, what = "nearest published model"))
ggplot(pp, aes(t, cp, colour = what)) +
  geom_line(linewidth = 1) +
  geom_point(data = data.frame(t = TIMES, cp = p_truth),
             aes(t, cp), inherit.aes = FALSE, size = 2.6, shape = 21,
             fill = "white", stroke = 1) +
  labs(x = "time (h)", y = "concentration (mg/L)", colour = NULL,
       caption = "points: the truth for severe renal impairment") +
  theme_bw() + theme(legend.position = "bottom")
```

```{r payofftab}
knitr::kable(
  data.frame(time = TIMES, truth = p_truth,
             meta_analysis = p_adm, nearest_model = p_near,
             err_meta = 100 * (p_adm  / p_truth - 1),
             err_near = 100 * (p_near / p_truth - 1)),
  digits = 2,
  caption = "Severe renal impairment: predicted mean concentration and % error."
)
```

The nearest published model is fine early and drifts badly late, because it is
carrying a clearance fitted in patients who clear the drug faster than these
ones do. Underpredicting exposure in renal impairment is the direction that
matters clinically.

## Discrete covariates

Not every covariate is continuous. Levels and probabilities rather than a
margin:

```{r discrete}
cdd <- covDist(WT  = c(mean = 78, sd = 16),
               SEX = list(values = c(0, 1), probs = c(0.45, 0.55)),
               dist = "lnorm")
set.seed(5)
table(covDraw(cdd, 2000L)[, "SEX"])
```

A discrete covariate is a **stratum**, not a distribution to integrate: it is
enumerated exactly at its levels, weighted by its probabilities, and crossed
with the continuous design. Nothing about it is approximated.

If a paper reports results *separately* by sex, that is conditioning rather
than marginalising — split it into one study per stratum, each with
`cov = list(SEX = 0)` and its own `n`. `covStrata()` does that split for you.

## What it costs

Every design point is a row in the ODE solve, so a product grid over `p`
covariates and `n_eta` random effects costs `n^(n_eta + p)`. admixr2 avoids
that where the model's structure allows.

```{r cost}
lg <- function() list(meanlog = log(78), sdlog = 0.198)
st3 <- list(s = list(E = as.numeric(studies$normal$E), V = studies$normal$V,
                     n = 120L, times = TIMES, ev = rxode2::et(amt = DOSE),
                     cov_dist = covDist(W1 = lg(), W2 = lg(), W3 = lg())))
mod3 <- function() {
  ini({ tcl <- log(5); tv <- log(50); b1 <- .6; b2 <- .4; b3 <- .3
        eta.cl ~ 0.06; add.err <- 0.08 })
  model({
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv) * (W1/70)^b1 * (W2/70)^b2 * (W3/70)^b3
    cp <- linCmt()
    cp ~ add(add.err)
  })
}
ui3  <- suppressMessages(rxode2::rxode2(mod3))
pin3 <- admixr2:::.admDriverPinfo(ui3, adghControl(studies = st3))
st3n <- suppressMessages(admixr2:::.admCheckCovariates(
  ui3, pin3, admixr2:::.admDriverUnits(
    st3, ui3, admixr2:::.admOutputVar(ui3))$studies))
co3 <- st3n[[1L]][[".adm_cov_collapse"]]
c(covariates = co3$p, directions = co3$r,
  design_points = nrow(co3$X), product_grid = pin3$cov_nodes^co3$p)
```

Those three covariates reach the model through a single scalar — the product
`(W1/70)^b1 (W2/70)^b2 (W3/70)^b3` — so the integral is one-dimensional however
many covariates feed it.

**This is a change of variables, not an approximation.** The prediction depends
on the covariates only through that scalar, so integrating over the scalar and
integrating over the product give the same answer — and admixr2 verifies that
numerically before using the reduced design. Where the model does not have that
structure it falls back to the full grid rather than guessing.

`cov_nodes` (default `r pin3$cov_nodes`) is the budget, per *direction* rather
than per covariate: a direction absorbing several covariate axes carries their
combined spread and is given proportionally more resolution.

## Notes

* **A covariate effect is identified by contrast across sources.** That is the
  whole of this vignette: three trials that individually could not estimate a
  renal effect, pooled into one that can. Within a single cohort the covariate
  barely varies, and what little it does is confounded with between-subject
  variability — a covariate enters `V` in the same place `omega` does.
* **Fix what convention fixes.** Allometric exponents are 0.75 and 1. Aggregate
  data is not the place to re-estimate them, and trying mostly buys a trade-off
  against `omega`. Estimate the covariate effects a popPK analysis would.
* **Covariates that share a distribution are not separately identified.** With
  `W1`, `W2`, `W3` independent and identically distributed, the aggregate
  depends on their exponents only through `Σb` and `Σb²` — two functionals for
  three parameters. Vary them differently across studies, or fix all but one.
* **Extrapolation is only as good as the covariate model.** The severe
  prediction works because the power model is the right functional form and the
  three cohorts span enough of the CRCL range to identify its exponent. A model
  fitted over 38–95 mL/min and applied at 21 is an extrapolation and should be
  presented as one.
* **Estimator support**: `admc` (per-subject draws) and `adgh` (quadrature)
  marginalise covariates. `adfo` and `adirmc` **refuse** rather than silently
  solving at the covariate mean — every study carries a covariate value as well
  as a distribution, so an unwired estimator would not fail, it would quietly
  return a fit whose `omega` had absorbed the covariate spread.
* admixr2 also refuses a `cov_dist` naming a covariate the model never reads.
  A declared-but-unread covariate is almost always a typo or a half-finished
  edit, and ignoring it would give you a fit that merely looked like it
  accounted for something.

## See also

- [datagen()](datagen.html) — published models as input, in depth
- [Multiple studies](multiple-studies.html) — the contrast a covariate effect is identified by
- [From a published figure to E, V and n](aggregate-data.html) — the other kind of input
- [Estimator comparison](estimator-comparison.html)
