---
title: "Generating aggregate data with datagen"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generating aggregate data with datagen}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 5
)
source(Filter(file.exists, c("_common.R", "../_common.R", "vignettes/_common.R"))[1])
```

## When would you use `datagen`?

Two situations call for generating aggregate data programmatically:

**Simulation studies.** To check that an estimator recovers the truth: define a
data-generating model with known parameters, call `datagen()` for E and V, fit
your analysis model, compare.

**Published models as input.** Each published study was analysed with its own
structural model — a different number of compartments, a different
parameterisation, only some of the IIV terms. `datagen()` turns such a model into
the aggregate data it implies, making it a direct input alongside digitised
summaries. Several of them in `admControl(studies = ...)` is a meta-analysis
across the literature (see `vignette("multiple-studies")`).

Either way the output is a named list of `(E, V, n, times, ev)` objects that
plugs straight into `admControl(studies = ...)`.

## Simulation study: same model across studies

A one-compartment oral model generates two synthetic studies — a low-dose and a
high-dose arm — which the estimator should recover the truth from.

### Data-generating model

```{r true-model}
library(admixr2)
library(rxode2)
library(nlmixr2)
library(ggplot2)

true_model <- function() {
  ini({
    tcl     <- log(5)  ; label("Log clearance (L/h)")
    tv      <- log(10) ; label("Log volume (L)")
    tka     <- log(1)  ; label("Log absorption rate (1/h)")
    prop.sd <- c(0, 0.2) ; label("Proportional error SD")
    eta.cl  ~ 0.09
    eta.v   ~ 0.04
    eta.ka  ~ 0.04
  })
  model({
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv  + eta.v)
    ka <- exp(tka + eta.ka)
    d/dt(depot)   <- -ka * depot
    d/dt(central) <- ka * depot - (cl/v) * central
    cp <- central / v
    cp ~ prop(prop.sd)
  })
}
```

True parameter values: CL = 5 L/h, V = 10 L, ka = 1 h⁻¹; IIV of 0.3 (SD on
log scale) for CL and 0.2 for V and ka; proportional residual error SD = 0.2.

### Generating the data

A top-level `model` is the default every study inherits, so each spec needs only
`times`, `ev` and `n`:

```{r datagen-same}
times <- c(0.5, 1, 2, 4, 8, 12, 24)

study_data <- datagen(
  studies = list(
    low_dose  = list(times = times, ev = rxode2::et(amt =  50), n = 250L),
    high_dose = list(times = times, ev = rxode2::et(amt = 100), n = 250L)
  ),
  model   = true_model,
  control = datagenControl(n_sim = 10000L, seed = 1L)
)

# Each study returns E, V, n, times, ev
names(study_data$low_dose)
round(study_data$low_dose$E, 2)   # population mean at each time
```

`V` carries the IIV-driven between-time correlation and, on the diagonal, the
residual error:

```{r show-v}
knitr::kable(round(study_data$low_dose$V, 3),
             caption = "Generated covariance matrix V for the low-dose study")
```

### Inspecting the generated profiles

```{r plot-generated, fig.cap="Generated population mean ± 1 SD for both dose levels."}
df_gen <- do.call(rbind, lapply(names(study_data), function(nm) {
  s  <- study_data[[nm]]
  sd <- sqrt(diag(s$V))
  data.frame(study = nm, time = s$times,
             mean = s$E, lo = s$E - sd, hi = s$E + sd)
}))

ggplot(df_gen, aes(x = time, y = mean, colour = study, fill = study)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_x_log10(breaks = times, labels = times) +
  scale_colour_manual(values = c(low_dose = "#0072B2", high_dose = "#D55E00")) +
  scale_fill_manual(  values = c(low_dose = "#0072B2", high_dose = "#D55E00")) +
  labs(title  = "Generated aggregate data — mean ± 1 SD",
       x = "Time (h, log scale)", y = "Concentration (mg/L)",
       colour = NULL, fill = NULL) +
  theme_bw()
```

The ribbon is the square root of the diagonal of V — IIV spread plus residual
error. The off-diagonal within-subject correlation is not drawn, but it reaches
the estimator and enters the likelihood.

### Parameter recovery

The generated data plug straight into `admControl(studies = ...)`. Using the same
structural model for the analysis is what makes this a consistency check:

```{r analysis-model}
analysis_model <- function() {
  ini({
    tcl     <- log(4)  ; label("Log clearance (L/h)")
    tv      <- log(12) ; label("Log volume (L)")
    tka     <- log(1.5); label("Log absorption rate (1/h)")
    prop.sd <- c(0, 0.3); label("Proportional error SD")
    eta.cl  ~ 0.12
    eta.v   ~ 0.05
    eta.ka  ~ 0.05
  })
  model({
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv  + eta.v)
    ka <- exp(tka + eta.ka)
    d/dt(depot)   <- -ka * depot
    d/dt(central) <- ka * depot - (cl/v) * central
    cp <- central / v
    cp ~ prop(prop.sd)
  })
}
```

The starting values are deliberately off the truth (CL 4 against 5, V 12 against
10), so finding it is the estimator's doing rather than the initialisation's.

```{r fit-simulation, comment = NA}
fit_sim <- nlmixr2(
  analysis_model, admData(), est = "admc",
  control = admControl(
    studies   = study_data,
      maxeval = 300L
  )
)

print(fit_sim)
```

**No standard errors, and the blank `SE` / `%RSE` / `95%CI` columns are why.** A
study generated from a model is not a sample: `n` is the source study's true
sample size, but the parameter uncertainty a sampling law would need is not
available, so admixr2 sets `covMethod = "none"` and reports no uncertainty at
all. Naming a `covMethod` explicitly is an error rather than an override. The
columns are nlmixr2's table; the emptiness is the answer.

Structural parameter estimates and the truth:

```{r recovery-table}
est   <- fit_sim$env$admExtra$struct
truth <- c(tcl = log(5), tv = log(10), tka = log(1))

knitr::kable(
  data.frame(
    parameter = names(truth),
    truth     = round(exp(truth), 2),
    estimate  = round(exp(est[names(truth)]), 2)
  ),
  row.names = FALSE, caption = "Structural parameter estimates vs true values")
```

## Literature workflow: per-study models

Each publication comes with its own model — a 2016 paper with one compartment, a
2022 follow-up with two. A `model` inside a study spec overrides the top-level
default:

```{r lit-models}
# One-compartment oral model (as published in a simpler earlier study)
model_1cmt <- function() {
  ini({
    tcl     <- log(5)
    tv      <- log(40)    # apparent volume, peripheral compartment lumped in
    tka     <- log(1)
    prop.sd <- c(0, 0.25)
    eta.cl  ~ 0.04
    eta.v   ~ 0.01
    eta.ka  ~ 0.01
  })
  model({
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv  + eta.v)
    ka <- exp(tka + eta.ka)
    d/dt(depot)   <- -ka * depot
    d/dt(central) <- ka * depot - (cl/v) * central
    cp <- central / v
    cp ~ prop(prop.sd)
  })
}

# Two-compartment oral model (as published in a more detailed later study)
model_2cmt <- function() {
  ini({
    tcl     <- log(5)
    tv1     <- log(10)
    tv2     <- log(30)
    tq      <- log(10)
    tka     <- log(1)
    prop.sd <- c(0, 0.2)
    eta.cl  ~ 0.09
    eta.v1  ~ 0.04
  })
  model({
    cl <- exp(tcl + eta.cl)
    v1 <- exp(tv1 + eta.v1)
    v2 <- exp(tv2)
    q  <- exp(tq)
    ka <- exp(tka)
    d/dt(depot)      <- -ka * depot
    d/dt(central)    <- ka * depot - (cl/v1 + q/v1) * central + (q/v2) * peripheral
    d/dt(peripheral) <- (q/v1) * central - (q/v2) * peripheral
    cp <- central / v1
    cp ~ prop(prop.sd)
  })
}
```

Both share a CL of 5 L/h and differ only in how they describe distribution: the
earlier study lumped the periphery into one larger apparent volume (V = 40 L),
the later one resolved it (V₁ = 10 L, V₂ = 30 L, Q = 10 L/h).

```{r lit-datagen}
lit_data <- datagen(
  studies = list(
    early_study = list(
      model = model_1cmt,
      times = c(0.5, 1, 2, 4, 8, 12),
      ev    = rxode2::et(amt = 100),
      n     = 120L
    ),
    later_study = list(
      model = model_2cmt,
      times = c(2, 4, 8, 12, 24),
      ev    = rxode2::et(amt = 200),
      n     = 180L
    )
  ),
  control = datagenControl(n_sim = 10000L, seed = 2L)
)
```

Same drug, same CL, different E profiles — the dosing and the distributional
assumptions differ:

```{r lit-plot, fig.cap="Simulated aggregate profiles from two publications with different structural models."}
df_lit <- do.call(rbind, lapply(names(lit_data), function(nm) {
  s  <- lit_data[[nm]]
  sd <- sqrt(diag(s$V))
  data.frame(study = nm, time = s$times,
             mean = s$E, lo = s$E - sd, hi = s$E + sd)
}))

ggplot(df_lit, aes(x = time, y = mean, colour = study, fill = study)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_colour_manual(values = c(early_study = "#009E73", later_study = "#CC79A7")) +
  scale_fill_manual(  values = c(early_study = "#009E73", later_study = "#CC79A7")) +
  labs(title  = "Literature-derived aggregate profiles",
       x = "Time (h)", y = "Concentration (mg/L)",
       colour = NULL, fill = NULL) +
  theme_bw()
```

### Fitting an analysis model to literature data

The data-generating models encode each publication's assumptions; the analysis
model is your unified structural hypothesis, fitted to both studies at once:

```{r lit-fit, eval=FALSE}
lit_analysis <- function() {
  ini({
    tcl     <- log(5)
    tv1     <- log(10)
    tv2     <- log(30)
    tq      <- log(10)
    tka     <- log(1)
    prop.sd <- c(0, 0.25)
    eta.cl  ~ 0.09
    eta.v1  ~ 0.04
  })
  model({
    cl <- exp(tcl + eta.cl)
    v1 <- exp(tv1 + eta.v1)
    v2 <- exp(tv2)
    q  <- exp(tq)
    ka <- exp(tka)
    d/dt(depot)      <- -ka * depot
    d/dt(central)    <- ka * depot - (cl/v1 + q/v1) * central + (q/v2) * peripheral
    d/dt(peripheral) <- (q/v1) * central - (q/v2) * peripheral
    cp <- central / v1
    cp ~ prop(prop.sd)
  })
}

fit_lit <- nlmixr2(
  lit_analysis, admData(), est = "admc",
  control = admControl(
    studies   = lit_data,
    n_sim     = 5000L,
    cov_n_sim = 10000L,
    seed      = 3L
  )
)
```

Because the two used genuinely different structural models, the analysis model
lands on a weighted compromise between the two parameterisations, minimising the
joint discrepancy rather than matching either exactly. That is the point:
aggregate-data modelling takes each study as published and looks for one set of
parameters consistent with all of them.

## Examining individual simulated samples

`return_samples = TRUE` adds the raw `n_sim × n_times` prediction matrix to each
study's output — useful for inspecting the shape of the simulated distribution,
or for custom summaries:

```{r samples, fig.cap="First 200 individual simulated trajectories (grey) with the population mean (blue)."}
study_with_samples <- datagen(
  studies = list(
    single_study = list(times = times, ev = rxode2::et(amt = 100), n = 250L)
  ),
  model   = true_model,
  control = datagenControl(n_sim = 2000L, seed = 4L, return_samples = TRUE)
)

cp_mat <- study_with_samples$single_study$samples   # 2000 x 7
mu     <- study_with_samples$single_study$E

df_samp <- data.frame(
  time  = rep(times, each = 200),
  conc  = as.vector(cp_mat[1:200, ]),
  id    = rep(seq_len(200), times = length(times))
)

ggplot(df_samp, aes(x = time, y = conc, group = id)) +
  geom_line(alpha = 0.06, colour = "grey30") +
  geom_line(data = data.frame(time = times, conc = mu),
            aes(group = NULL), colour = "#0072B2", linewidth = 1.5) +
  labs(title = "Simulated individual trajectories (n = 200 shown)",
       x = "Time (h)", y = "Concentration (mg/L)") +
  theme_bw()
```

Blue is the population mean E, the grey spaghetti the IIV spread. `samples` holds
concentrations **before** residual error; only the diagonal of V carries that.

## Choosing the moment method

`datagen()` integrates over the IIV by Monte Carlo, the same engine as
`est = "admc"`. Two deterministic alternatives match the other estimators, and
both ignore `n_sim`, `sampling` and `seed`, so their output is exactly
reproducible:

```{r fo-gh-moments}
one <- list(single_study = list(times = times, ev = rxode2::et(amt = 100),
                                n = 250L))
fo_data <- datagen(studies = one, model = true_model,
                   control = datagenControl(method = "fo"))
gh_data <- datagen(studies = one, model = true_model,
                   control = datagenControl(method = "gh", n_nodes = 5L))
round(rbind(fo = fo_data$single_study$E, gh = gh_data$single_study$E), 3)
```

`method = "fo"` is the first-order expansion `est = "adfo"` uses, `E = f(theta,
0)` and `V = J Omega J' + Sigma`; `method = "gh"` is the Gauss-Hermite
quadrature `est = "adgh"` uses, unbiased at any IIV. `n_nodes` (per eta
dimension) trades accuracy against cost: 3 is fast, 5 is near-exact to IIV SD
~0.5, 7 reaches ~0.7.

For design work, **generate and analyse under the same approximation**. The
expected information matrix is the Hessian of the log-likelihood at the
generating parameters, which is only a genuine maximum when the two agree --
generate with Monte Carlo, analyse under FO, and those parameters are not an FO
maximum-likelihood estimate, so the FIM is biased.

## See also

- [Multiple studies](multiple-studies.html) — feed per-study models into a meta-analysis
- [From a published figure to E, V and n](aggregate-data.html) — the other input type
- [Estimator comparison](estimator-comparison.html)
