---
title: "Diagnostic plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnostic plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 5
)
```

```{r fit-hidden, include=FALSE}
library(admixr2)
library(rxode2)
library(nlmixr2)
library(ggplot2)

data("examplomycin")
obs   <- examplomycin[examplomycin$EVID == 0, ]
obs   <- obs[order(obs$ID, obs$TIME), ]
times <- sort(unique(obs$TIME))
ids   <- unique(obs$ID)
n     <- length(ids)
dv_mat <- matrix(NA_real_, nrow = n, ncol = length(times))
for (i in seq_along(ids)) {
  sub         <- obs[obs$ID == ids[i], ]
  dv_mat[i, ] <- sub$DV[order(sub$TIME)]
}
E <- colMeans(dv_mat)
V <- cov.wt(dv_mat, method = "ML")$cov

pk_model <- function() {
  ini({
    tcl     <- log(5)
    tv1     <- log(10)
    tv2     <- log(30)
    tq      <- log(10)
    tka     <- log(1)
    prop.sd <- c(0, 0.2)
    eta.cl ~ 0.09
    eta.v1 ~ 0.09
    eta.v2 ~ 0.09
    eta.q  ~ 0.09
    eta.ka ~ 0.09
  })
  model({
    cl <- exp(tcl + eta.cl)
    v1 <- exp(tv1 + eta.v1)
    v2 <- exp(tv2 + eta.v2)
    q  <- exp(tq  + eta.q)
    ka <- exp(tka + eta.ka)
    d/dt(depot)      <- -ka * depot
    d/dt(central)    <- ka * depot - (cl/v1 + q/v1) * central + (q/v2) * peripheral
    d/dt(peripheral) <- (q/v1) * central - (q/v2) * peripheral
    cp <- central / v1
    cp ~ prop(prop.sd)
  })
}

fit <- nlmixr2(
  pk_model, admData(), est = "admc",
  control = admControl(
    studies   = list(examplomycin = list(E = E, V = V, n = n,
                                         times = times, ev = rxode2::et(amt = 100))),
    n_sim     = 5000L,
    cov_n_sim = 10000L,
    maxeval   = 300L,
    seed      = 1L
  )
)
```

`plot.admFit()` produces up to four diagnostic panel types. Call with `which` to
select a subset; the default is all four. Each panel is returned as a named
ggplot2 object in a list.

```r
plots <- plot(fit, which = c("mean", "cov", "nll", "par"))
```

## Mean diagnostics

The `"mean"` panel is a 2×2 grid per study. Requires `patchwork` for the
composite layout; without it, four separate plots are returned.

```{r mean-panel, fig.height=6, fig.cap="Mean diagnostics for the examplomycin study."}
plots <- plot(fit, which = "mean")
```

**Top-left — Observed:** sample mean with ±1 SD ribbon, where
SD = √diag(**V**_obs).

**Top-right — Predicted:** predicted mean with ±1 SD ribbon, where
SD = √diag(**V**_pred). **V**_pred combines between-subject variability (Omega)
and residual error (sigma). Both top panels share the same y-axis scale so
differences in magnitude are immediately visible.

**Bottom-left — Raw residual:** `E_obs[t] − μ_pred[t]` as a lollipop. The grey
band is ±2 SE(mean), where SE = √(**V**_pred[t,t] / n). Points outside the band
indicate systematic bias at that time point.

**Bottom-right — Standardised residual:**
`z[t] = residual[t] / SE[t]`. Under a well-specified model, z should be
approximately N(0, 1). Stars flag |z| > 1.96 (∗), 2.58 (∗∗), 3.29 (∗∗∗). With
9 time points and no multiplicity correction, one flagged time point is expected
by chance alone.

## Covariance diagnostics

The `"cov"` panel compares observed and predicted (co)variance as heatmaps.
Requires `patchwork` for the composite layout.

```{r cov-panel, fig.height=6, fig.cap="Covariance diagnostics: observed and predicted covariance matrices with residuals."}
plots <- plot(fit, which = "cov")
```

**Top row — Observed | Predicted:** both matrices plotted on a shared
blue-white-red scale. A good fit shows matching patterns of magnitude and
sign — including the off-diagonal temporal correlation structure.

**Bottom-left — Residual (V_obs − V_pred):** diverging scale (purple-white-teal).
Persistent positive residuals on the diagonal mean the model under-predicts
variance; negative residuals indicate over-prediction.

**Bottom-right — Standardised residual:** each entry divided by its asymptotic
SE. Diagonal SE = √(2 V[i,i]² / (n−1)); off-diagonal SE =
√((V[i,i]·V[j,j] + V[i,j]²) / (n−1)). Stars use the same thresholds as the
mean panel.

## NLL trace

The `"nll"` panel shows how the objective function evolved across optimizer
iterations, coloured by restart:

```{r nll-panel, fig.height=4, fig.cap="NLL convergence trace. Each line is one optimizer restart."}
plots <- plot(fit, which = "nll")
```

Restarts that converge to the same final value support a unimodal landscape. A
spread of final NLL values suggests local optima — increase `n_restarts` and
`restart_sd` in `admControl()`.

## Parameter trace

The `"par"` panel facets each parameter's trajectory over optimizer iterations:

```{r par-panel, fig.height=6, fig.cap="Parameter trace on the natural scale. Struct thetas back-transformed; sigma shown as SD; V(eta) = variance."}
plots <- plot(fit, which = "par")
```

Parameters are displayed on the natural scale:

- Structural thetas: back-transformed (e.g. `exp()` for log-scale params)
- Sigma: displayed as SD
- Omega diagonal: displayed as variance, labelled `V(eta.x)`
- Omega off-diagonal: raw Cholesky L[i,j]

Smoothly converging traces that stabilise well before `maxeval` indicate the
optimizer is not being artificially cut off. If traces still drift at the end,
increase `maxeval`.

## Accessing individual panels

`plot()` returns its list invisibly. Assign it to retrieve or modify panels:

```{r access}
plots <- plot(fit, which = c("nll", "par"))
names(plots)
```

Panel names follow the pattern `<type>_<study>` for per-study panels, or
`nll_trace` / `par_trace` for the trace panels. For a study with several observed
outputs the study label gains the output name — e.g. `mean_lit.plasma`,
`cov_lit.brain` — so each compartment gets its own panel:

```{r access2, eval=FALSE}
plots$nll_trace
plots$par_trace
plots$mean_examplomycin
plots$cov_examplomycin
```

## Customising with ggplot2

All returned objects are standard ggplot2 plots:

```{r custom, fig.height=4, fig.cap="NLL trace with a custom theme."}
plots$nll_trace +
  ggplot2::theme_minimal(base_size = 13) +
  ggplot2::labs(title = "Optimiser convergence", subtitle = NULL)
```

## IIV correlation heatmap

The estimated Omega matrix can be visualised as a correlation heatmap to inspect
the inter-individual variability structure. This is especially informative for
models with off-diagonal omega entries.

```{r omega-heatmap, fig.height=4.5, fig.cap="IIV correlation matrix derived from the estimated Omega."}
omega   <- fit$env$admExtra$omega
eta_nms <- fit$env$admExtra$eta_col_names
if (is.null(eta_nms)) eta_nms <- paste0("eta.", seq_len(nrow(omega)))

corr_mat <- cov2cor(omega)
rownames(corr_mat) <- colnames(corr_mat) <- eta_nms

df_corr <- expand.grid(
  eta_i = factor(eta_nms, levels = rev(eta_nms)),
  eta_j = factor(eta_nms, levels = eta_nms),
  stringsAsFactors = FALSE
)
df_corr$r <- as.vector(t(corr_mat))

ggplot2::ggplot(df_corr, ggplot2::aes(x = eta_j, y = eta_i, fill = r)) +
  ggplot2::geom_tile(colour = "white", linewidth = 0.5) +
  ggplot2::geom_text(ggplot2::aes(label = round(r, 2)), size = 3.2) +
  ggplot2::scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#D6604D",
    midpoint = 0, limits = c(-1, 1), name = "r"
  ) +
  ggplot2::labs(
    title    = "IIV correlation (estimated Omega)",
    subtitle = "Off-diagonal entries are zero for a diagonal Omega model",
    x = NULL, y = NULL
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
```

## See also

- [Getting started](admixr2.html)
- [Multiple studies](multiple-studies.html) — meta-analysis across studies
- [Estimator comparison](estimator-comparison.html)
