---
title: "Diagnostic plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnostic plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 5
)
source(Filter(file.exists, c("_common.R", "../_common.R", "vignettes/_common.R"))[1])
```

```{r fit-hidden, include=FALSE}
library(admixr2)
library(rxode2)
library(nlmixr2)
library(ggplot2)

# The examplomycin fit from `vignette("admixr2")`, assembled by the shared
# helpers in this vignette's setup file.
s <- admVignetteStats()

fit <- nlmixr2(
  admVignetteModel, admData(), est = "admc",
  control = admControl(
    studies   = list(examplomycin = c(s, list(ev = rxode2::et(amt = 100)))),
    n_sim     = 5000L,
    cov_n_sim = 10000L,
    maxeval   = 300L,
    seed      = 1L
  )
)
```

`plot.admFit()` draws up to five panel types, all five by default. Each is
returned as a named ggplot2 object in a list.

```r
plots <- plot(fit, which = c("mean", "cov", "covariate", "nll", "par"))
```

`"covariate"` needs studies that declare covariates and is silently absent
otherwise, so the single-study examplomycin fit used below produces the other
four. It has its own section, with its own fit, further down.

## Mean diagnostics

`sqrt(diag(V))` is the total SD of **one observation across subjects**, not
between-subject variability --- it carries BSV, residual error, and, wherever a
study marginalises a covariate, the spread that covariate induces. The panels
name the terms from the fit rather than assuming them, and the predicted ribbon
draws the **pre-sigma part inside** the total, so the gap between the two bands
is the residual error. A predicted spread that misses the observed one then
tells you whether to look at omega and the covariates or at the error model.

A 2×2 grid **per source**. A source is banded into one study per covariate
level or quadrature node, for every covariate its own model fitted --- a covariate
a study marginalises over is not identified against a random effect on the same
parameter, so the objective has to sum over the strata. But `A_s1` is an index
into that expansion, and this panel asks "does the fit reproduce this paper", so
the strata are put back together first.

The mean is the n-weighted mean of the strata. The variance is **not**: it is
the law of total variance, within plus **between**, and the between term is the
covariate effect the banding created. Averaging the strata's variances would
report the within-level spread, which is not what the paper published --- a
correctly specified fit would draw as under-predicting it.

The covariate panels keep the strata, because there the between-level contrast
is the entire signal. `patchwork` composes it; without it you get four separate
plots.

```{r mean-panel, fig.height=6, fig.cap="Mean diagnostics for the examplomycin study."}
plots <- plot(fit, which = "mean")
```

**Top-left — Observed:** sample mean with ±1 SD ribbon, where
SD = √diag(**V**_obs).

**Top-right — Predicted:** predicted mean with ±1 SD ribbon, where
SD = √diag(**V**_pred), combining between-subject variability (Omega) and
residual error (sigma). Both top panels share a y-axis, so a difference in
magnitude is visible at once.

**Bottom-left — Raw residual:** `E_obs[t] − μ_pred[t]` as a lollipop. The grey
band is ±2 SE(mean), SE = √(**V**_pred[t,t] / n); points outside it are
systematic bias at that time.

**Bottom-right — Standardised residual:** `z[t] = residual[t] / SE[t]`, which a
well-specified model leaves approximately N(0, 1). Stars flag |z| > 1.96 (∗),
2.58 (∗∗), 3.29 (∗∗∗). Across 9 uncorrected time points, one flag is expected by
chance.

## Covariance diagnostics

Observed against predicted (co)variance, as heatmaps. Also composed with
`patchwork`.

```{r cov-panel, fig.height=6, fig.cap="Covariance diagnostics: observed and predicted covariance matrices with residuals."}
plots <- plot(fit, which = "cov")
```

**Top row — Observed | Predicted:** one shared blue-white-red scale. A good fit
matches in magnitude and sign, off-diagonal temporal structure included.

**Bottom-left — Residual (V_obs − V_pred):** diverging purple-white-teal.
Positive on the diagonal means the model under-predicts variance; negative,
over-predicts.

**Bottom-right — Standardised residual:** each entry over its asymptotic SE —
√(2 V[i,i]² / (n−1)) on the diagonal, √((V[i,i]·V[j,j] + V[i,j]²) / (n−1)) off
it. Stars as above.

## Covariate diagnostics

The mean and covariance panels are per study, in isolation. With aggregate data
a covariate effect is identified *between* studies --- the renal exponent in
`vignette("covariates")` comes from three cohorts sitting at three different
creatinine-clearance medians, none of which fitted a renal term --- so every
source can sit beautifully on its own panel while the relation tying them
together is wrong. `which = "covariate"` draws that contrast, and returns two
plots.

The examplomycin fit declares no covariates, so these two panels need a fit
that does. Three renal cohorts, generated *with* a renal effect. Two of them
report only who they enrolled, so creatinine clearance is **marginal** for
them; `mild` reports its result *by* renal band, so it is **conditional**. One
figure then carries both readings of the same covariate.

```{r cov-fit, message=FALSE, results="hide"}
set.seed(11)

renal_model <- function() {
  ini({
    tcl   <- log(5) ; tv    <- log(50)
    bcrcl <- 0.6    ; bsex  <- 0.18
    add.err <- 0.08
    eta.cl ~ 0.05
  })
  model({
    cl <- exp(tcl + eta.cl) * (WT / 70)^0.75 * (CRCL / 90)^bcrcl *
      exp(bsex * SEX)
    v  <- exp(tv) * (WT / 70)
    cp <- linCmt()
    cp ~ add(add.err)
  })
}

TIMES <- c(0.5, 2, 8, 24)

# MARGINAL on renal function: the paper reports the population it enrolled and
# admixr2 integrates over it. Banded on sex, which every analyst fitted.
enrolled <- function(n, crclm) data.frame(
  WT   = rlnorm(n, log(76), 0.198),
  CRCL = rlnorm(n, log(crclm), 0.05),
  SEX  = rbinom(n, 1, 0.5))

marginal_source <- function(n, crclm) admStudy(
  model = renal_model, population = enrolled(n, crclm),
  dose = 200, times = TIMES)

# CONDITIONAL on renal function -- and the call says nothing about that either.
# `renal_model` uses CRCL, so CRCL is conditional for any source built on it and
# is banded. What IS worth declaring is `range`: the population here is a
# `mean +/- SD` transcription rather than the patients themselves, so its tails
# reach where nobody was enrolled, and `strata_nodes` keeps the band count to
# the one renal band this paper reported.
banded_source <- function(crclm, band) admStudy(
  model = renal_model,
  population = covDist(WT   = c(meanlog = log(76), sdlog = 0.198),
                       CRCL = c(meanlog = log(crclm), sdlog = 0.05),
                       SEX  = c(female = 0.5, male = 0.5)),
  n = 210L, dose = 200, times = TIMES,
  strata_nodes = 1L, range = list(CRCL = band))

renal_studies <- admStudies(
  normal   = marginal_source(260L, 95),
  mild     = banded_source(62, c(50, 75)),
  moderate = marginal_source(180L, 38))

renal_fit <- nlmixr2(
  renal_model, admData(), est = "adgh",
  control = adghControl(studies = renal_studies, print = 0L,
                        n_restart = 1L, maxeval = 40L))
```

```{r cov-plots, include=FALSE, fig.show="hide"}
cov_plots <- plot(renal_fit, which = "covariate", n_sim = 300L)
```

```{r cov-names}
names(cov_plots)
```

**`covariate_effect`** asks one question: **does the estimated covariate effect
agree with the sources it was fitted from?** One small panel per (parameter,
covariate) pair, carrying three things and nothing else.

* the **estimated effect**, as a dotted black line across the covariate axis ---
  the fitted model's parameter at the fitted thetas, with every other covariate
  at the pooled centre.
* **each source at its own parameter value**, taken from that source's own
  published model at its own published estimates. This is the comparison: a
  source sitting off the dotted line is one the meta-analysis does not
  reproduce.
* **how far along the axis each source speaks for**, which is where conditional
  and marginal differ --- see the table below.

A covariate whose coefficient the model does **not** estimate gets no panel at
all. A fixed allometric exponent is not a finding; drawing a facet for it invites
a reader to check an agreement that was never in question.

The region no source sampled is shaded: that is where the fit extrapolates. A
discrete covariate is **marked at its levels** with open squares --- those are
the places the model was actually asked --- and the dotted line joins them, so
the effect reads the same way on both kinds of axis. The line is a connector
between marked points, not a claim about the space between them.

```{r covariate-effect, fig.height=4.2, fig.cap="Estimated covariate effect against its sources. `cl vs CRCL`: `mild` reported by renal band, so its own regression runs across 50-75; `normal` and `moderate` reported only who they enrolled, so each is a whisker. `cl vs SEX`: every analyst banded on sex, so each source draws its own line between the two levels."}
cov_plots$covariate_effect
```

There is no `WT` facet. Every source reads weight at a *fixed* 0.75 exponent, so
no weight coefficient is estimated and there is no agreement to check.

The fit is black, dotted, open squares. Sources are coloured, solid, filled.
Nothing about the fit borrows a source's vocabulary: the shape key
(circle/diamond) is a statement about a *source*, and the fit is not one.

**`covariate_resid`** plots each study's mean standardised residual against the
covariate value it sits at. How it is read depends on the covariate:

* **continuous** --- a trend. The dashed `lm` across studies; a slope is a
  mis-specified covariate form.
* **discrete** --- a contrast. The axis is ticked at the levels the sources were
  *conditioned* at and nowhere
  else, and no regression is drawn: a line across the levels of a factor
  reports as a slope what is really a difference between groups. Instead the
  strata cut from one source are joined in grey, because that
  pairing is what banding buys --- the same study, one covariate moved. Several
  sources tilting the same way is the mis-specification.

```{r covariate-resid, fig.height=3.4, fig.cap="Between-study residual against each covariate. A slope across sources, or a consistent tilt in the grey within-source pairs, is a mis-specified covariate form."}
cov_plots$covariate_resid
```

This fit has the renal term the data were generated with, so its residuals are
flat --- which is what a correctly specified covariate form looks like. For the
same panel on a model that is *missing* the effect, where the three sources run
from +5 to −4.5 monotone in creatinine clearance, see
`vignette("covariates")`.

On both panels **point area is the study's sample size**, as `multinma`'s
`weight_nodes` does for its network nodes. `scale_size_area()`, so area rather
than radius carries it and zero maps to zero; the legend is suppressed when
every study reports the same `n`, since there is nothing to compare. A banded
source's `n` is divided among its strata, so it contributes half its patients at
each sex level but its whole `n` on any other covariate's axis.

Both panels key colour on the **source**, not the stratum, so a source is one
colour across the figure. Strata of one source are told apart by where they sit
and by the grey line joining them, which is what makes them a pair.

A covariate facet is dropped from `covariate_resid` when the sources have no
real contrast along it --- when the spread of their centres is small against a
typical within-study 10th--90th. Three cohorts drawing weight from the same
distribution differ only in what a finite sample estimated, and a free x scale
would blow that noise up to full panel width and fit a trend through it.

Read it with the coverage bars: three sources spanning the *same* range of a
covariate cannot say anything about it, whatever the fitted line does, and with
few sources two covariates whose study medians move together cannot be told
apart.

Both panels distinguish how a study enters a covariate. This is a property of
**the source's own published model** --- did that analyst estimate the effect
--- and both kinds feed the same fit:

| | drawn as | meaning |
|---|---|---|
| **conditional** | **its own regression line**, over the range it covers | the source's own model ESTIMATED this covariate's coefficient, so it is banded --- or it reported at a value of it (`at`, `by`). It reported a *relationship* here, so its own model is drawn: a slope different from the dotted line is a paper the fit does not reproduce, and parallel-but-offset is a different finding from crossing. A source that reported only one value has no slope of its own and stays a point |
| **marginal** | round point and a **whisker** --- 10th–90th over 2.5th–97.5th | the source reported no contrast here, so admixr2 integrates over the population it enrolled. What it covers is a distribution, and drawing that as a line would claim it covers its tails as evenly as its middle |

Splitting a source on a covariate its model never saw would manufacture a
contrast that is not in the literature --- see `vignette("covariates")`.

Read the panel by looking for sources that sit **off** the dotted line. Black is
the meta-analysis; the coloured marks are the papers, each at the value its own
model publishes. Where they part company over a stretch a source actually
enrolled, the fit is disagreeing with a paper about that paper's own patients.

On a level axis the source's own effect is the **gap between its own marks** ---
one per level it reported --- against the gap between the black points.

A banded source appears once per **level** it reported on a covariate banded at
its levels, and once at its own centre on every other --- including a continuous
covariate it was banded on, where the values are quadrature nodes rather than
anything the paper reported. The effect panel is per source; only the residual
panel keeps the strata apart, and only on a level axis.

A covariate the model **does not read** still gets a `covariate_resid` facet.
That is the case the panel exists for: the null model of a covariate test simply
omits the term, admixr2 takes the covariate off the design because it cannot
change the prediction, and the question you are asking is precisely whether it
belonged.

## Building your own from the moments

`plot()` draws a fixed set of panels. `admMoments()` returns the numbers behind
them --- one row per study and observation time --- so you can draw the rest.

```{r moments}
m <- admMoments(renal_fit, n_sim = 300L)
str(m, give.attr = FALSE)
```

`obs_sd` and `pred_sd` are the SD of one observation across subjects, and
`struct_sd` is the same quantity *before* residual error is composed on. So
`pred_sd - struct_sd` is what sigma contributes, which is what lets a departure
be attributed rather than merely noticed.

Strata are collapsed to their source by default; pass `by = "stratum"` to keep
them apart.

To see the three panels below do something, they are drawn against a
deliberately **mis-specified** fit: the data carry additive residual error and
this model fits proportional. Everything else is correct.

```{r wrong-err, message=FALSE, results="hide"}
wrong_err <- function() {
  ini({
    tcl   <- log(5) ; tv   <- log(50)
    bcrcl <- 0.6    ; bsex <- 0.18
    prop.sd <- c(0, 0.1)
    eta.cl ~ 0.05
  })
  model({
    cl <- exp(tcl + eta.cl) * (WT / 70)^0.75 * (CRCL / 90)^bcrcl *
      exp(bsex * SEX)
    v  <- exp(tv) * (WT / 70)
    cp <- linCmt()
    cp ~ prop(prop.sd)
  })
}

bad_fit <- nlmixr2(
  wrong_err, admData(), est = "adgh",
  control = adghControl(studies = renal_studies, print = 0L,
                        n_restart = 1L, maxeval = 40L))
mb <- admMoments(bad_fit, n_sim = 300L)
```

### Observed vs predicted spread

Aggregate-data fitting takes its variability information entirely from the
reported `V`, so a source off the identity line is one whose published spread
the model does not reproduce.

```{r spread, fig.height=4.4, fig.cap="Observed vs predicted SD of one observation across subjects. The low-concentration points fall below the line: proportional error puts too little spread there, where additive error had a fixed amount."}
lim <- range(c(mb$obs_sd, mb$pred_sd))
ggplot(mb, aes(pred_sd, obs_sd, colour = source, size = n)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey40") +
  geom_point(alpha = 0.85) +
  scale_size_area(max_size = 5, name = "n") +
  scale_x_log10() + scale_y_log10() +
  coord_fixed(xlim = lim, ylim = lim) +
  labs(x = "predicted SD (log scale)", y = "observed SD (log scale)",
       colour = NULL) +
  theme_bw()
```

### Residual against predicted mean

A fan or a curve here is a residual error model that does not fit; flat scatter
about zero is one that does. This is the plot that chooses between `add()`,
`prop()`, `pow()` and `t()`.

```{r zpred, fig.height=3.8, fig.cap="Standardised residual against predicted mean. The wedge -- large negative residuals at low concentrations, flat at high -- is the signature of proportional error fitted to additive data."}
ggplot(mb, aes(pred_mean, z, colour = source, size = n)) +
  geom_hline(yintercept = 0, colour = "grey40") +
  geom_hline(yintercept = c(-1.96, 1.96), linetype = "dashed",
             colour = "grey60") +
  geom_point(alpha = 0.85) +
  scale_size_area(max_size = 5, guide = "none") +
  labs(x = "predicted mean", y = "standardised residual", colour = NULL) +
  theme_bw()
```

### Residual against time

One line per source. A shape shared by **every** line is structural or an error
model; one line apart from the rest is that study.

```{r ztime, fig.height=3.8, fig.cap="Standardised residual against time, one line per source. All three bend the same way as concentrations fall, which is what makes it the model rather than any one paper."}
ggplot(mb, aes(time, z, colour = source, group = source)) +
  geom_hline(yintercept = 0, colour = "grey40") +
  geom_hline(yintercept = c(-1.96, 1.96), linetype = "dashed",
             colour = "grey60") +
  geom_line(alpha = 0.85) +
  geom_point(size = 1.9) +
  labs(x = "time (h)", y = "standardised residual", colour = NULL) +
  theme_bw()
```

For comparison, the same three on the correctly specified `renal_fit` are flat:
`max(abs(m$z))` is `r round(max(abs(m$z)), 2)`, against
`r round(max(abs(mb$z)), 1)` for the mis-specified one.

## NLL trace

The objective across optimizer iterations, coloured by restart:

```{r nll-panel, fig.height=4, fig.cap="NLL convergence trace. Each line is one optimizer restart."}
plots <- plot(fit, which = "nll")
```

Restarts landing on the same value support a unimodal landscape; a spread of
final values suggests local optima, so raise `n_restarts` and `restart_sd`.

## Parameter trace

One facet per parameter, over optimizer iterations:

```{r par-panel, fig.height=6, fig.cap="Parameter trace on the natural scale. Struct thetas back-transformed; sigma shown as SD; V(eta) = variance."}
plots <- plot(fit, which = "par")
```

Everything is on the natural scale: structural thetas back-transformed, sigma as
an SD, the Omega diagonal as a variance labelled `V(eta.x)`, its off-diagonal as
the raw Cholesky `L[i,j]`.

Traces that flatten well before `maxeval` mean the optimizer was not cut off. If
they still drift at the end, raise `maxeval`.

## Accessing individual panels

`plot()` returns its list invisibly; assign it to reach individual panels:

```{r access}
plots <- plot(fit, which = c("nll", "par"))
names(plots)
```

Per-study panels are named `<type>_<study>`, the traces `nll_trace` and
`par_trace`. With several observed outputs the label gains the output name —
`mean_lit.plasma`, `cov_lit.brain` — so each gets its own panel:

```{r access2, eval=FALSE}
plots$nll_trace
plots$par_trace
plots$mean_examplomycin
plots$cov_examplomycin
```

Every panel is an ordinary ggplot2 object, so `+ theme_minimal()` or
`+ labs()` works as usual.

## See also

- [Getting started](admixr2.html)
- [Multiple studies](multiple-studies.html) — meta-analysis across studies
- [Estimator comparison](estimator-comparison.html)

