---
title: "Diagnostic plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnostic plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 5
)
source(Filter(file.exists, c("_common.R", "../_common.R", "vignettes/_common.R"))[1])
```

```{r fit-hidden, include=FALSE}
library(admixr2)
library(rxode2)
library(nlmixr2)
library(ggplot2)

# The examplomycin fit from `vignette("admixr2")`, assembled by the shared
# helpers in this vignette's setup file.
s <- admVignetteStats()

fit <- nlmixr2(
  admVignetteModel, admData(), est = "admc",
  control = admControl(
    studies   = list(examplomycin = c(s, list(ev = rxode2::et(amt = 100)))),
    n_sim     = 5000L,
    cov_n_sim = 10000L,
    maxeval   = 300L,
    seed      = 1L
  )
)
```

`plot.admFit()` draws up to four panel types, all four by default. Each is
returned as a named ggplot2 object in a list.

```r
plots <- plot(fit, which = c("mean", "cov", "nll", "par"))
```

## Mean diagnostics

A 2×2 grid per study. `patchwork` composes it; without it you get four separate
plots.

```{r mean-panel, fig.height=6, fig.cap="Mean diagnostics for the examplomycin study."}
plots <- plot(fit, which = "mean")
```

**Top-left — Observed:** sample mean with ±1 SD ribbon, where
SD = √diag(**V**_obs).

**Top-right — Predicted:** predicted mean with ±1 SD ribbon, where
SD = √diag(**V**_pred), combining between-subject variability (Omega) and
residual error (sigma). Both top panels share a y-axis, so a difference in
magnitude is visible at once.

**Bottom-left — Raw residual:** `E_obs[t] − μ_pred[t]` as a lollipop. The grey
band is ±2 SE(mean), SE = √(**V**_pred[t,t] / n); points outside it are
systematic bias at that time.

**Bottom-right — Standardised residual:** `z[t] = residual[t] / SE[t]`, which a
well-specified model leaves approximately N(0, 1). Stars flag |z| > 1.96 (∗),
2.58 (∗∗), 3.29 (∗∗∗). Across 9 uncorrected time points, one flag is expected by
chance.

## Covariance diagnostics

Observed against predicted (co)variance, as heatmaps. Also composed with
`patchwork`.

```{r cov-panel, fig.height=6, fig.cap="Covariance diagnostics: observed and predicted covariance matrices with residuals."}
plots <- plot(fit, which = "cov")
```

**Top row — Observed | Predicted:** one shared blue-white-red scale. A good fit
matches in magnitude and sign, off-diagonal temporal structure included.

**Bottom-left — Residual (V_obs − V_pred):** diverging purple-white-teal.
Positive on the diagonal means the model under-predicts variance; negative,
over-predicts.

**Bottom-right — Standardised residual:** each entry over its asymptotic SE —
√(2 V[i,i]² / (n−1)) on the diagonal, √((V[i,i]·V[j,j] + V[i,j]²) / (n−1)) off
it. Stars as above.

## NLL trace

The objective across optimizer iterations, coloured by restart:

```{r nll-panel, fig.height=4, fig.cap="NLL convergence trace. Each line is one optimizer restart."}
plots <- plot(fit, which = "nll")
```

Restarts landing on the same value support a unimodal landscape; a spread of
final values suggests local optima, so raise `n_restarts` and `restart_sd`.

## Parameter trace

One facet per parameter, over optimizer iterations:

```{r par-panel, fig.height=6, fig.cap="Parameter trace on the natural scale. Struct thetas back-transformed; sigma shown as SD; V(eta) = variance."}
plots <- plot(fit, which = "par")
```

Everything is on the natural scale: structural thetas back-transformed, sigma as
an SD, the Omega diagonal as a variance labelled `V(eta.x)`, its off-diagonal as
the raw Cholesky `L[i,j]`.

Traces that flatten well before `maxeval` mean the optimizer was not cut off. If
they still drift at the end, raise `maxeval`.

## Accessing individual panels

`plot()` returns its list invisibly; assign it to reach individual panels:

```{r access}
plots <- plot(fit, which = c("nll", "par"))
names(plots)
```

Per-study panels are named `<type>_<study>`, the traces `nll_trace` and
`par_trace`. With several observed outputs the label gains the output name —
`mean_lit.plasma`, `cov_lit.brain` — so each gets its own panel:

```{r access2, eval=FALSE}
plots$nll_trace
plots$par_trace
plots$mean_examplomycin
plots$cov_examplomycin
```

## Customising with ggplot2

All returned objects are standard ggplot2 plots:

```{r custom, fig.height=4, fig.cap="NLL trace with a custom theme."}
plots$nll_trace +
  ggplot2::theme_minimal(base_size = 13) +
  ggplot2::labs(title = "Optimiser convergence", subtitle = NULL)
```

## IIV correlation heatmap

The estimated Omega reads as a correlation heatmap, which is worth a look
whenever the model has off-diagonal omega entries.

```{r omega-heatmap, fig.height=4.5, fig.cap="IIV correlation matrix derived from the estimated Omega."}
omega   <- fit$env$admExtra$omega
eta_nms <- fit$env$admExtra$eta_col_names
if (is.null(eta_nms)) eta_nms <- paste0("eta.", seq_len(nrow(omega)))

corr_mat <- cov2cor(omega)
rownames(corr_mat) <- colnames(corr_mat) <- eta_nms

df_corr <- expand.grid(
  eta_i = factor(eta_nms, levels = rev(eta_nms)),
  eta_j = factor(eta_nms, levels = eta_nms),
  stringsAsFactors = FALSE
)
df_corr$r <- as.vector(t(corr_mat))

ggplot2::ggplot(df_corr, ggplot2::aes(x = eta_j, y = eta_i, fill = r)) +
  ggplot2::geom_tile(colour = "white", linewidth = 0.5) +
  ggplot2::geom_text(ggplot2::aes(label = round(r, 2)), size = 3.2) +
  ggplot2::scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#D6604D",
    midpoint = 0, limits = c(-1, 1), name = "r"
  ) +
  ggplot2::labs(
    title    = "IIV correlation (estimated Omega)",
    subtitle = "Off-diagonal entries are zero for a diagonal Omega model",
    x = NULL, y = NULL
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
```

## See also

- [Getting started](admixr2.html)
- [Multiple studies](multiple-studies.html) — meta-analysis across studies
- [Estimator comparison](estimator-comparison.html)
