---
title: "Choosing a residual error model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing a residual error model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.2, dpi = 120
)
# nlmixr2's fit print method uses cli colour; when a build forces colour on (e.g.
# pkgdown/CI) the raw ANSI escapes leak into the HTML. Force plain-text output
# for this vignette only, saving the previous global state...
.adm_opts    <- options(cli.num_colors = 1, crayon.enabled = FALSE)
.adm_nocolor <- Sys.getenv("NO_COLOR", unset = NA)
Sys.setenv(NO_COLOR = "1")

# ...and restoring it once the document has finished knitting, so the setting
# does not leak into other articles when pkgdown builds them in one process.
knitr::knit_hooks$set(document = function(x) {
  options(.adm_opts)
  if (is.na(.adm_nocolor)) Sys.unsetenv("NO_COLOR") else Sys.setenv(NO_COLOR = .adm_nocolor)
  x
})

# The estimators keep rxode2's progress bar off during fitting (the control's
# nDisplayProgress argument), but a one-off bar still prints while each model is
# *compiled*. Strip those animation frames from the rendered output.
knitr::knit_hooks$set(output = local({
  .default <- knitr::knit_hooks$get("output")
  function(x, options) {
    x <- gsub("[^\n]*\\[[=| ]+\\][^\n]*\n?", "", x, perl = TRUE)
    .default(x, options)
  }
}))
```

## The question

Your model ends with `cp ~ prop(res.sd)`, as in an individual-level fit. The
input is not individual data, though. It is a mean and an error bar per time
point, digitised from a published figure or implied by a published model you
supplied through `datagen()` — and usually several studies of them at once.

Each error bar holds two things the fit has to tell apart: subjects genuinely
**differ** from one another, and each reported value carries variability the
structural model does not explain. Your residual model is the claim about how to
split them, and that split decides what goes to `omega` and what to the residual.

```{r libs}
library(admixr2)
library(rxode2)
library(nlmixr2)
library(ggplot2)
```

## How residual error lives in aggregate data

An error bar in a published figure is one number covering two sources of spread:
subjects differ from one another, and each reported value carries variability the
structural model does not explain. Only the second is your residual model's job.

Everything below uses one model and one study. `datagen()` stands in for a
digitised summary here so that the true values are known; a real analysis would
read `E`, `V` and `n` off the paper — see
[From a published figure to E, V and n](aggregate-data.html).

```{r study-model}
times <- c(0.5, 1, 2, 4, 8, 12, 24)

mod_prop <- function() {
  ini({
    tcl     <- log(5)   ; label("Log clearance (L/h)")
    tv      <- log(20)  ; label("Log volume (L)")
    res.sd  <- c(0, 0.2); label("Residual error SD")
    eta.cl  ~ 0.09
    eta.v   ~ 0.04
  })
  model({
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv  + eta.v)
    d/dt(central) <- -(cl / v) * central
    cp <- central / v
    cp ~ prop(res.sd)
  })
}

gen <- datagen(
  studies = list(s1 = list(times = times, ev = rxode2::et(amt = 100), n = 200L)),
  model = mod_prop, control = datagenControl(n_sim = 20000L, seed = 1L))
```

Simulating the individual data behind such a summary — 20 000 subjects, each
observed 25 times so that the two sources can be separated — shows how the two
parts sit inside the error bar:

```{r resid-setup, echo=FALSE}
b <- 0.2; N <- 20000L; R <- 25L

sim <- rxode2::rxode2({
  cl <- exp(tcl + eta.cl)
  v  <- exp(tv  + eta.v)
  d/dt(central) <- -(cl / v) * central
  cp <- central / v
})

set.seed(1)
s <- rxode2::rxSolve(
  sim,
  data.frame(tcl = log(5), tv = log(20),
             eta.cl = rnorm(N, 0, 0.3), eta.v = rnorm(N, 0, 0.2)),
  rxode2::et(amt = 100) |> rxode2::et(times),
  returnType = "data.frame", nDisplayProgress = .Machine$integer.max)

f      <- matrix(s$cp[s$time > 0], nrow = N, byrow = TRUE)
mean_f <- colMeans(f)
var_f  <- apply(f, 2, function(z) mean((z - mean(z))^2))

# Observing the same subject R times holds the between-subject part fixed, so the
# spread across replicates is residual error alone.
resid_var <- vapply(seq_along(times), function(j)
  mean(apply(f[, j] * (1 + b * matrix(rnorm(N * R), N, R)), 1, var)), 0)
```

```{r fig-split, echo=FALSE, fig.height=4, fig.cap="The light band is the whole error bar; the dark band is how much of it residual error alone would produce. The dark band is drawn inside the light one, not stacked on it: the total SD is sqrt(var_between + var_resid), not the sum of the two SDs."}
sd_tot <- sqrt(var_f + resid_var)
df <- data.frame(time = times, mean = mean_f,
                 lo_t = mean_f - sd_tot,   hi_t = mean_f + sd_tot,
                 lo_n = mean_f - sqrt(resid_var), hi_n = mean_f + sqrt(resid_var))

ggplot(df, aes(time, mean)) +
  geom_ribbon(aes(ymin = lo_t, ymax = hi_t, fill = "the whole error bar"), alpha = 0.9) +
  geom_ribbon(aes(ymin = lo_n, ymax = hi_n, fill = "residual error alone"), alpha = 0.9) +
  geom_line(linewidth = 0.9) + geom_point(size = 2) +
  scale_x_log10(breaks = times, labels = times) +
  scale_fill_manual(values = c("the whole error bar"   = "#CBD5E1",
                               "residual error alone"  = "#0072B2"),
                    breaks = c("the whole error bar", "residual error alone"),
                    name = NULL) +
  labs(x = "Time (h, log scale)", y = "Concentration (mg/L)",
       title = "How much of the error bar is residual error? (±1 SD)") +
  theme_bw() + theme(legend.position = "top")
```

The residual's share of the error bar is not constant: it dominates early and is
almost invisible by 24 h. That changing share is the shape information a fit uses
to tell the two apart, which is why the residual model you name matters and why a
wrong one has somewhere to hide.

## What a wrong residual model does to a fit

Fit the same study twice, changing only the error line:

```{r fit-both, results="hide"}
mod_add <- mod_prop |> model(cp ~ add(res.sd)) |> ini(res.sd = 0.5)

fit_prop <- nlmixr2(mod_prop, admData(), est = "adgh",
                    control = adghControl(studies = gen, print = 0L))
fit_add  <- nlmixr2(mod_add,  admData(), est = "adgh",
                    control = adghControl(studies = gen, print = 0L))
```

```{r fit-both-table}
pull <- function(fit, nm) unname(fit$parFixedDf[nm, "Estimate"])

knitr::kable(
  data.frame(
    parameter = c("CL (L/h)", "V (L)", "IIV on CL", "IIV on V",
                  "residual SD", "-2LL"),
    truth     = c(5, 20, 0.09, 0.04, 0.2, NA),
    prop      = c(exp(pull(fit_prop, "tcl")), exp(pull(fit_prop, "tv")),
                  fit_prop$omega[1, 1], fit_prop$omega[2, 2],
                  pull(fit_prop, "res.sd"), fit_prop$objective),
    add       = c(exp(pull(fit_add, "tcl")),  exp(pull(fit_add, "tv")),
                  fit_add$omega[1, 1],  fit_add$omega[2, 2],
                  pull(fit_add, "res.sd"),  fit_add$objective)),
  digits = 3,
  caption = paste("Same data, same structural model, two residual models.",
                  "The additive residual SD is in mg/L and is not comparable to",
                  "the proportional one."))
```

The proportional fit recovers all five generating values. The additive fit leaves
clearance and volume within a few percent, which is the dangerous part, since CL
and V are usually the first numbers anyone checks. Meanwhile the IIV on volume
comes back `r round(100 * (fit_add$omega[2, 2] / 0.04 - 1))`% too large: the
variance the residual failed to account for has to go somewhere, and a trial sized
from that fit would simulate the wrong spread.

The objective moves `r round(fit_add$objective - fit_prop$objective)` points, so
comparing the two fits catches it. The covariance panel shows the same thing
without needing a second fit to compare against:

```{r cov-panel, fig.height=5.5, fig.cap="Covariance diagnostics for the additive fit. In the lower-right panel, cells marked with an asterisk are where the predicted covariance differs from the observed by more than the sampling noise in V."}
plot(fit_add, which = "cov")
```

When the residual model fits, the standardised panel looks like noise. Here the
diagonal is starred early and late in opposite directions: one constant variance
spread over the whole profile, too small where concentrations are high and too
large where they have decayed. That opposite-sign pattern is the signature of an
additive model on proportional data, and the fix is to refit with `prop()` or
`add() + prop()`.

This panel is the aggregate-data replacement for a residual-vs-predicted plot; see
[Diagnostic plots](diagnostic-plots.html).

## What you can write

You write the error model in ordinary nlmixr2 syntax; admixr2 adds nothing to the
language. The estimator is the constraint: **`adirmc` supports the closed-form
continuous models only**, because it computes its mean by importance-reweighting
samples inside a single compiled routine that knows only those variance forms.
`admc`, `adfo` and `adgh` support all of it.

| Written as | Estimators | Notes |
|---|---|---|
| **Continuous — exact, closed form** | | |
| `cp ~ add(a)` | all four | |
| `cp ~ prop(b)` | all four | |
| `cp ~ pow(b, c)` | all four | exact at `c = 0.5` and `1`; otherwise the moments use a second-order expansion. `c` is an exponent, not an SD |
| `cp ~ add(a) + prop(b)` | all four | variance `a² + (b·f)²` |
| `cp ~ add(a) + prop(b) + combined1()` | all four | variance `(a + b·f)²` — a different model, not a reparameterisation of the row above |
| `cp ~ lnorm(a)` | all four | moment-matched log-normal |
| any of the above `+ t(nu)` | all four | **fix `nu`**: only `a²·nu/(nu−2)` is identified, so a free `nu` returns its starting value |
| **Transformed — integrated numerically** | | |
| `cp ~ boxCox(lambda) + add(a)` | not `adirmc` | `lambda` estimated or fixed |
| `cp ~ yeoJohnson(lambda) + add(a)` | not `adirmc` | `lambda` estimated or fixed |
| `cp ~ logitNorm(a, lo, hi)` | not `adirmc` | for an endpoint with a floor and a ceiling |
| `cp ~ probitNorm(a, lo, hi)` | not `adirmc` | as above |
| **Discrete — no residual parameters** | | |
| `y ~ pois(cp)` | not `adirmc` | the prediction is the *argument*, not the left-hand side |
| `y ~ binom(N, pp)` | not `adirmc` | `N` is the binomial denominator per observation, not the study's `n`, and must be constant |
| `y ~ nbinomMu(k, cp)` | not `adirmc` | |
| `cp ~ c(p1, p2, ...)` | not `adirmc` | ordinal; `p1, p2, ...` are the category probabilities, one observation block each |
| `y ~ beta(shape1, shape2)` | `admc`, `adgh` | the precision is implied by the two shapes, which `adfo` and `adirmc` cannot reach |
| **Correlated residuals** | | |
| `cp ~ add(a) + ar(rho)` | not `adirmc` | needs a **full** observed `V`, and pairs with `add()` only |

Each endpoint carries its own error model, so a multi-output model keeps them
separate — see [Several observed compartments](multi-compartment.html).

`nu` and `rho` are the two parameters aggregate data does not identify, and they
fail differently. `nu` enters only through the variance multiplier, so it is
aliased with the scale: admixr2 warns and returns your starting value, which is
why you should fix it. `rho` lives in the off-diagonal, so a diagonal `V` carries
no information about it at all, and that is a hard error rather than a warning.

Anything else is refused: `lnorm` mixed with `prop`/`pow`/`ar`/`t`, `ar` with a
prediction-dependent variance or a transform, `t()` with no scale term,
`cauchy()`, and the distributions with no admixr2 implementation. In each case the
combination has no single well-defined aggregate mean and variance, so admixr2
refuses rather than fitting something else:

```{r refusal, error=TRUE}
bad <- mod_prop |> model(cp ~ lnorm(res.sd) + prop(b)) |> ini(b = 0.2)

nlmixr2(bad, admData(), est = "adgh", control = adghControl(studies = gen))
```

## Setting `resid_nodes` on a transformed endpoint

Writing `cp ~ boxCox(lambda) + add(a)` says the residual is normal on the
**transformed** scale, not in mg/L. But the study you are fitting reports its mean
and SD in mg/L. So admixr2 has to answer a question that never comes up in an
individual-level fit: if the residual is normal after transformation, what are the
mean and variance back on the original scale?

For `add()` and `prop()` that back-transformation has a formula. For `boxCox`,
`yeoJohnson`, `logitNorm` and `probitNorm` it does not. Because the transform
bends the residual, the mean of the back-transformed values is not the
back-transform of the mean, and there is no closed form for either moment.

admixr2 integrates over the residual instead. This is the same machinery as the
Gaussian quadrature used to integrate over the **etas** in a mixed-effects
likelihood — nlmixr2's `agq`, or `nAGQ` in `lme4` — aimed at a different integral:
over the residual at one observation, rather than over a subject's random effects.

Concretely: rather than one prediction, the endpoint is evaluated at `resid_nodes`
different residual values, each weighted by how probable it is under the normal,
and the results averaged. Two things separate this from simulating residuals.

**The values are chosen, not drawn.** Gauss-Hermite places `n` points so that the
weighted average is exact whenever the quantity being averaged behaves like a
polynomial of degree `2n - 1` or less. A few well-placed points therefore do the
work of a great many random draws. They sit symmetrically about zero and reach out
to about `±sqrt(2n)` residual SDs, so a higher count both resolves the middle and
reaches further into the tails.

**The grid never moves.** The same points at every likelihood evaluation, so the
objective is smooth and its gradient usable. Drawing fresh residuals would add
Monte Carlo noise to every evaluation, which the optimiser would have to work
through.

How many nodes you need depends on how curved the back-transformation is over the
range the residual explores. A residual that is narrow on the transformed scale
stays where that curve is nearly straight, and a handful of nodes covers it.

The demanding case is a wide residual on a **bounded** endpoint — `logitNorm` or
`probitNorm` on a score with a floor and a ceiling — because the back-transform
flattens against both bounds. For a logit-normal endpoint with an SD of 2 on the
transformed scale, 7 nodes puts the 24 h variance 7% low and 15 nodes 0.6% low;
31 is converged. A more typical residual width converges by about 15. The default
of 81 covers the awkward cases with room to spare, at a cost invisible next to the
ODE solve. Lower it only if profiling shows the quadrature is your bottleneck, and
check that the objective does not move when you do.

`resid_nodes` is on all four estimator controls and on `datagenControl()`, so a
study you generate and the fit that consumes it integrate the residual the same
way unless you change one of them.

## How to choose

There is no residual plot to look at. The choice comes from the published error
bars, from what the source papers already report, and from the diagnostics above.

* **Read the error bars.** Widening roughly in proportion to the mean points to
  `prop()`; roughly constant width across the profile points to `add()`; wide at
  the top *and* not shrinking to nothing at the bottom points to `add() + prop()`.
* **Take the residual from the source where one is reported.** A study that
  published its own population model has already estimated a residual error model
  from the individual data you do not have — better information than an aggregate
  fit can recover. Supply that model to `datagen()` and it comes along with it, or
  carry it across by fixing yours to the same value:

  ```r
  ini({ res.sd <- fix(0.15) })   # residual error reported by the source study
  ```

* **Ask whether one residual across studies is credible.** Studies run in
  different laboratories, on different assays, years apart will not share a
  residual magnitude. A single pooled residual is an assumption, not a default. If
  one study's error bars are much wider relative to its means than the rest, give
  that study's endpoint its own residual rather than letting it set the value for
  every study.
* **Bounded endpoints belong on a transformed scale.** A score with a floor and a
  ceiling should use `logitNorm()` or `probitNorm()`, not `add()`, which will
  happily predict outside the range without warning.
* **Then check.** Fit the candidates, compare objectives, and read the covariance
  panel. A residual model that fits leaves no structure in the standardised
  residual.

**The residual and the IIV are separated only by their different time-course.**
This is the constraint behind everything above. Where the two have a similar shape
— a proportional residual against IIV on a parameter that scales the whole profile
— they trade off, and the split is weakly determined. The worked example is that
failure in miniature: the misspecified residual put its missing variance straight
into `omega.v`.

This is also an argument for the thing admixr2 is for. One study with few time
points barely identifies the split. Several studies at different doses and
sampling schedules constrain it from more directions at once, which is why a
meta-analysis across studies is better posed here than a reanalysis of any one of
them.

## How the residual enters the covariance

This is the one place where admixr2 changed behaviour in the current release, and
the change moves results for most models, so it is worth being precise about.

Both the old and the new rule start from the same structural covariance
`Cov(f)` — the spread caused by subjects differing from one another — and then
account for the residual on top of it. They differ in **what they add**.

**Up to version 0.3.0**, the residual variance was worked out once, at the
population mean prediction, and added to the diagonal:

$$V \;=\; \mathrm{Cov}(f) \;+\; \mathrm{diag}\big(\Sigma(\bar f)\big),
\qquad \text{proportional: } \Sigma(\bar f) \;=\; b^2\,\bar f^{\,2}$$

This is the "no eta-eps interaction" convention, familiar from individual-level
fitting, where each observation has its own prediction and the question does not
arise.

**From this release**, the residual variance is averaged over subjects — the law
of total variance:

$$V \;=\; \mathrm{Cov}(f) \;+\;
  \mathrm{diag}\big(\mathbb{E}_\eta[\mathrm{Var}(y \mid \eta)]\big),
\qquad \text{proportional: } b^2\,\mathbb{E}[f^2]
  \;=\; b^2\big(\bar f^{\,2} + \mathrm{Var}(f)\big)$$

The two differ by exactly `b^2 Var(f)`. A subject's residual scales with **that
subject's** prediction, not with the population mean, and the average of a square
exceeds the square of the average by the between-subject variance. The old rule
drops that term.

| error model | up to 0.3.0 added | now adds |
|---|---|---|
| `add(a)` | `a²` | `a²` — unchanged |
| `prop(b)` | `b²·mean²` | `b²·(mean² + Var(f))` |
| `pow(b, c)` | `b²·mean^(2c)` | `b²·E[f^(2c)]` |
| `lnorm(a)` | the diagonal only | the diagonal, **and** the off-diagonals scaled by `exp(s)` |

`lnorm()` is the one that is not only a diagonal correction: its conditional mean
is `f·exp(s/2)`, so the whole covariance is scaled, not just its diagonal.

Against the simulated subjects from earlier, where the residual's true
contribution is known:

```{r fig-rules, echo=FALSE, fig.height=4, fig.cap="Open circles are the residual variance measured from the simulated subjects. Averaging over subjects lands on them; evaluating at the population mean drifts low as subjects spread apart."}
avg_rule <- b^2 * (mean_f^2 + var_f)   # now: averaged over subjects
at_mean  <- b^2 * mean_f^2             # up to 0.3.0: at the population mean

lines_df <- rbind(
  data.frame(time = times, v = avg_rule, rule = "averaged over subjects (now)"),
  data.frame(time = times, v = at_mean,  rule = "at the population mean (<= 0.3.0)"))

ggplot(lines_df, aes(time, v, colour = rule, linetype = rule)) +
  geom_line(linewidth = 1) +
  geom_point(data = data.frame(time = times, v = resid_var),
             aes(time, v), inherit.aes = FALSE, size = 2.6, shape = 21,
             fill = "white", stroke = 1) +
  scale_x_log10(breaks = times, labels = times) +
  scale_y_log10() +
  scale_colour_manual(values = c("averaged over subjects (now)" = "#0072B2",
                                 "at the population mean (<= 0.3.0)" = "#D55E00"),
                      name = NULL) +
  scale_linetype_manual(values = c("averaged over subjects (now)" = "solid",
                                   "at the population mean (<= 0.3.0)" = "dashed"),
                        name = NULL) +
  labs(x = "Time (h, log scale)", y = "Residual variance (log scale)",
       title = "Two ways to compose the residual into V") +
  theme_bw() + theme(legend.position = "top")
```

The new rule agrees with the measurement to within
`r sprintf("%.1f%%", 100 * max(abs(avg_rule - resid_var) / resid_var))` at every
time. The old one is fine early, where subjects are alike, and misses
`r round(100 * (resid_var[length(times)] - at_mean[length(times)]) / resid_var[length(times)])`%
of the residual by 24 h, where they differ from one another by more than the mean
itself.

**What this means for results you already have.** Purely additive models are
unaffected, bit for bit — `a²` does not depend on the prediction, so there is
nothing to average. Every `prop()`, `pow()` and `lnorm()` fit changes. Checked
against individual-level simulation, the old rule carried a fixed 15-20% bias in
the predicted variance that did not shrink as the study grew; in a proportional
model with 30-50% IIV the reported residual SD and omega were each biased upward
by roughly 2-4%, and for `lnorm()` the effect is larger. A refit will not
reproduce the numbers a 0.3.0 fit gave you, and the new ones are the correct
comparison to a residual estimated from individual data.

## Two caveats

* **Point estimates are recovered; the reported uncertainty is approximate.** The
  covariance term of the objective treats your observed `V` as coming from
  normally distributed subjects. That holds for `add`/`prop`/`pow`/`lnorm`, and is
  an approximation for a heavy-tailed or bounded per-subject residual (small-`nu`
  `t`, counts, `beta`). State it when reporting SEs or intervals from those.
* **New since 0.3.0.** That release supported `add`, `prop`, `pow`, both combined
  forms and `lnorm`; everything else above was refused. This release adds `t(nu)`,
  the four transformed endpoints, the discrete endpoints and ordinal, `beta()`
  under `admc`/`adgh`, `ar(rho)` with `add()`, and `resid_nodes`. It also changes
  how the residual is composed into the covariance, so `prop()`, `pow()` and
  `lnorm()` results move — see the section above. If you have results from a
  version that accepted a now-refused model with only a warning — `pow()` above
  all — treat them as suspect: they were fitted as **additive** error.

## See also

- [From a published figure to E, V and n](aggregate-data.html) — where `E` and `V` come from
- [Comparing the estimators](estimator-comparison.html) — pairing an estimator with an endpoint
- [Diagnostic plots](diagnostic-plots.html) — reading the covariance panel
- [Simulating data & using published models](datagen.html) — `datagen()`, used throughout here
- [Getting started](admixr2.html)
