---
title: "Estimator comparison: adfo, admc and adirmc"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimator comparison: adfo, admc and adirmc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 5
)
```

## Three estimators, one interface

All three estimators accept the same model and study specification and return
the same `admFit` object. They share a common likelihood framework but differ
in how they approximate the population mean and covariance of the predicted
observations.

## The shared integration problem

For each study the observed data are a mean vector $\bar{y}$ and a covariance
matrix $S$ computed from $n$ subjects. Under a multivariate normal
approximation the aggregate log-likelihood is

$$
-2\ell = n \left[
  \log |V_\text{pred}| +
  \operatorname{tr}(V_\text{pred}^{-1} S) +
  r^\top V_\text{pred}^{-1} r
\right], \qquad r = \bar{y} - \mu_\text{pred}
$$

where $\mu_\text{pred}$ and $V_\text{pred}$ are the model-predicted population
mean and covariance of the observations. For a nonlinear model $f(\theta,\eta)$
these are integrals over the random-effects distribution
$\eta \sim \mathcal{N}(0, \Omega)$:

$$
\mu_\text{pred} = \mathbb{E}_\eta[f(\theta, \eta)], \qquad
V_\text{pred} = \operatorname{Var}_\eta[f(\theta, \eta)] + \Sigma
$$

These integrals have no closed form for nonlinear $f$. All three estimators
minimise the same objective but take different approaches to evaluating them.

## First-Order (FO): Taylor expansion at $\eta = 0$

`adfo` sidesteps the integral entirely by approximating $f(\theta, \eta)$ with
its first-order Taylor expansion around $\eta = 0$:

$$
f(\theta, \eta) \approx f(\theta, 0) + J\,\eta, \qquad
J_{t,j} = \left.\frac{\partial f_t}{\partial \eta_j}\right|_{\eta=0}
$$

Substituting into the population moments gives closed-form expressions:

$$
\mu_\text{pred} = f(\theta, 0), \qquad
V_\text{pred} = J\,\Omega\,J^\top + \Sigma
$$

The Jacobian $J$ is obtained in a single rxSolve call via sensitivity equations
that augment the ODE system with $\partial f / \partial \eta_j$ — no finite
differences, no extra solves per eta. This makes `adfo` the fastest estimator:
one rxSolve per NLL evaluation, fully deterministic.

**When the approximation holds:** the Taylor expansion is exact when $f$ is
linear in $\eta$. For nonlinear models or large IIV the true population
covariance exceeds $J\Omega J^\top$; FO underestimates $\Omega$ and bias grows
with the degree of nonlinearity.

**AIC note:** `adfo` evaluates a linearised likelihood. Its objective is not on
the same scale as `admc` or `adirmc` and must not be used for cross-estimator
AIC comparisons.

## Monte Carlo (MC): sample average over $\eta$

`admc` estimates the population integrals directly by drawing $N$ samples
$\eta_i \sim \mathcal{N}(0, \Omega)$ and computing sample moments:

$$
\mu_\text{pred} = \frac{1}{N} \sum_{i=1}^N f(\theta, \eta_i), \qquad
V_\text{pred} = \frac{1}{N} \sum_{i=1}^N
  \bigl(f(\theta,\eta_i) - \mu_\text{pred}\bigr)
  \bigl(f(\theta,\eta_i) - \mu_\text{pred}\bigr)^\top + \Sigma
$$

No approximation is made to $f$ itself — the estimator is asymptotically exact
as $N \to \infty$. Sobol quasi-random sequences (default) reduce MC variance
relative to plain normal draws, typically requiring 2–5× fewer samples for
equivalent precision.

The key cost is that **every NLL evaluation requires $N$ rxSolve calls** — one
per sample. The analytical gradient (sensitivity equations for mu-referenced
parameters; common random numbers FD otherwise) adds no further solves, but the
base cost per optimizer step remains $N \times$ (cost of one rxSolve).

**Advantages:**

- Asymptotically exact; AIC directly comparable across `admc` fits.
- Works well for standard 1–2 compartment PK models with moderate IIV.

**Limitations:**

- Each NLL evaluation requires $N$ rxSolve calls; slower than `adfo`.
- MC noise in the gradient can cause optimiser oscillation at low `n_sim`.

## Iterative Reweighting MC (IRMC): proposals fixed, inner loop free

`adirmc` addresses the main bottleneck of `admc` — the need for $N$ new
rxSolve calls at every optimizer step — by **decoupling proposal generation
from optimization**.

At each outer phase, $N$ proposals $\eta_i$ are drawn once and held fixed for
the entire inner optimization. Given fixed proposals, evaluating the NLL
requires only matrix operations — **no new rxSolve calls**. The proposals are
reweighted by their likelihood under the current $\Omega$ so that the inner
objective remains a valid approximation as parameters move:

$$
w_i \propto \frac{p(\eta_i \mid \Omega)}{q(\eta_i)}, \qquad
\mu_\text{pred} = \sum_i w_i\, f(\theta, \eta_i), \qquad
V_\text{pred} = \sum_i w_i\,
  \bigl(f(\theta,\eta_i) - \mu_\text{pred}\bigr)
  \bigl(\cdots\bigr)^\top + \Sigma
$$

The inner optimizer therefore runs to convergence at negligible cost; proposals
are refreshed only between phases. Box constraints on the parameters are
progressively tightened across phases to guide global convergence.

This decoupling makes `adirmc` particularly well-suited to **complex ODE
models** where each rxSolve is expensive: the total number of solves scales
with the number of phases rather than the number of optimizer steps. It is also
more robust to poor starting values because the inner optimisation is
deterministic and does not depend on re-sampling.

**Advantages:**

- Far fewer rxSolve calls per optimizer step than `admc` — especially
  beneficial for complex ODE systems with expensive solves.
- Robust to poor starting values; deterministic inner loop.
- AIC directly comparable to `admc`.

**Limitations:**

- Heavier per-phase overhead than `admc` for simple, well-initialised problems
  with cheap ODE solves.

## Common setup

```{r data}
library(admixr2)
library(rxode2)
library(nlmixr2)
library(ggplot2)

data("examplomycin")
obs   <- examplomycin[examplomycin$EVID == 0, ]
obs   <- obs[order(obs$ID, obs$TIME), ]
times <- sort(unique(obs$TIME))
ids   <- unique(obs$ID)
n     <- length(ids)

dv_mat <- matrix(NA_real_, nrow = n, ncol = length(times))
for (i in seq_along(ids)) {
  sub         <- obs[obs$ID == ids[i], ]
  dv_mat[i, ] <- sub$DV[order(sub$TIME)]
}
E <- colMeans(dv_mat)
V <- cov.wt(dv_mat, method = "ML")$cov

pk_model <- function() {
  ini({
    tcl     <- log(5)  ; label("Log clearance (L/hr)")
    tv1     <- log(10) ; label("Log central volume (L)")
    tv2     <- log(30) ; label("Log peripheral volume (L)")
    tq      <- log(10) ; label("Log inter-compartmental CL (L/hr)")
    tka     <- log(1)  ; label("Log absorption rate constant (1/hr)")
    prop.sd <- c(0, 0.2); label("Proportional residual error SD")
    eta.cl ~ 0.09
    eta.v1 ~ 0.09
    eta.v2 ~ 0.09
    eta.q  ~ 0.09
    eta.ka ~ 0.09
  })
  model({
    cl <- exp(tcl + eta.cl)
    v1 <- exp(tv1 + eta.v1)
    v2 <- exp(tv2 + eta.v2)
    q  <- exp(tq  + eta.q)
    ka <- exp(tka + eta.ka)
    d/dt(depot)      <- -ka * depot
    d/dt(central)    <- ka * depot - (cl/v1 + q/v1) * central + (q/v2) * peripheral
    d/dt(peripheral) <- (q/v1) * central - (q/v2) * peripheral
    cp <- central / v1
    cp ~ prop(prop.sd)
  })
}

study <- list(E = E, V = V, n = n, times = times, ev = rxode2::et(amt = 100))
```

The model uses mu-referenced parameterisation (`cl <- exp(tcl + eta.cl)`),
which enables analytical gradient computation via sensitivity equations in
all three estimators. See the
[Advanced usage](advanced.html#mu-referencing-and-sensitivity-equations)
vignette for details.

## Fitting with adfo

`adfo` is the fastest estimator and a natural starting point for model
screening or obtaining initial estimates for MC refinement.

```{r fit-adfo}
fit_fo <- nlmixr2(
  pk_model, admData(), est = "adfo",
  control = adfoControl(
    studies  = list(examplomycin = study),
    maxeval  = 500L,
    seed     = 1L
  )
)
```

## Fitting with admc

```{r fit-admc}
fit_mc <- nlmixr2(
  pk_model, admData(), est = "admc",
  control = admControl(
    studies   = list(examplomycin = study),
    n_sim     = 5000L,
    cov_n_sim = 10000L,
    maxeval   = 300L,
    seed      = 1L
  )
)
```

## Fitting with adirmc

`adirmc` is slower per evaluation but more robust to poor starting values and
high-dimensional $\Omega$. The settings below are tuned for a fast vignette
build; increase `n_sim` for production use.

```{r fit-irmc, eval=FALSE}
fit_irmc <- nlmixr2(
  pk_model, admData(), est = "adirmc",
  control = adirmcControl(
    studies         = list(examplomycin = study),
    n_sim           = 2000L,
    phases          = c(2, 1, 0.5, 0.1),
    cov_n_sim       = 10000L,
    omega_expansion = 1.5,
    seed            = 1L
  )
)
```

## Comparing parameter estimates

`adfo` and `admc` should recover values close to the true parameters
(CL = 5, V1 = 10, V2 = 30, Q = 10, ka = 1; IIV variance = 0.09 for all;
prop.sd = 0.2). For the examplomycin dataset IIV is moderate and the model is
near-linear in $\eta$, so FO bias is small.

```{r compare-table}
get_pars <- function(fit) {
  s  <- fit$env$admExtra$struct
  om <- diag(fit$env$admExtra$omega)
  sg <- sqrt(unlist(fit$env$admExtra$sigma_var))
  c(exp(s), om, sg)
}

tbl <- data.frame(
  Parameter = c(
    paste0("exp(", names(fit_fo$env$admExtra$struct), ")"),
    paste0("var(", fit_fo$env$admExtra$eta_col_names, ")"),
    "prop.sd"
  ),
  True = c(5, 10, 30, 10, 1, rep(0.09, 5), 0.2),
  adfo = round(get_pars(fit_fo), 4),
  admc = round(get_pars(fit_mc), 4)
)

knitr::kable(tbl, caption = "Parameter estimates vs true values")
```

## Comparing objectives

`admc` and `adirmc` evaluate the same likelihood and their -2LL values are
directly comparable. `adfo` evaluates a linearised likelihood — its objective
is on a different scale and **must not** be compared with MC objectives or used
for cross-estimator AIC:

```{r compare-obj}
cat(sprintf("adfo  -2LL = %.2f   AIC = %.2f\n", fit_fo$objective, AIC(fit_fo)))
cat(sprintf("admc  -2LL = %.2f   AIC = %.2f\n", fit_mc$objective, AIC(fit_mc)))
```

Use AIC only within the same estimator for model selection.

## NLL convergence traces

```{r nll-trace, fig.height=4, fig.cap="adfo and admc NLL convergence traces."}
plots_fo <- plot(fit_fo, which = "nll")
plots_mc <- plot(fit_mc, which = "nll")
```

For `adirmc`, the NLL trace reflects outer phase iterations rather than
individual LBFGS steps. Jumps between phases (as box constraints tighten) are
normal and expected.

## When to use each

| Situation | Recommendation |
|-----------|----------------|
| Rapid model screening, many candidate models | `adfo` — fastest per evaluation |
| Initial estimates before MC refinement | `adfo` then hand off to `admc` |
| Weak IIV (CV < 20 %) and near-linear model | `adfo` estimates reliable for inference |
| Standard 1–2 compartment PK, good starting values | `admc` with `grad = "sens"` |
| Initial exploration or poor starting values | `admc` with `n_restarts >= 3` |
| Complex ODE system with expensive solves | `adirmc` — inner loop needs no new rxSolve calls |
| High-dimensional Omega (≥ 5 etas) | `adirmc` — inner loop scales with phases not steps |
| Non-Gaussian or bounded IIV | `adirmc` with `omega_expansion > 1` |
| Need exact likelihood for AIC comparison across models | `admc` or `adirmc` (not `adfo`) |
| Maximum reproducibility | All three estimators accept `seed` |

## Control objects at a glance

```{r control-summary, eval=FALSE}
# adfo key arguments (fastest; linearised likelihood)
adfoControl(
  studies    = list(...),
  grad       = "none",      # "none" (BOBYQA), "analytical", "fd", "cfd"
  maxeval    = 500L,
  n_restarts = 1L,
  covMethod  = "r",         # SEs for struct+sigma only; omega SEs are not computed
  seed       = 1L
)

# admc key arguments
admControl(
  studies    = list(...),
  n_sim      = 5000L,       # MC sample count
  grad       = "sens",      # gradient mode: "sens", "fd", "cfd", "none"
  n_restarts = 1L,          # number of optimizer restarts
  workers    = 1L,          # parallel workers for restarts
  covMethod  = "r",         # SEs for struct+sigma only; omega SEs are not computed
  seed       = 1L
)

# adirmc key arguments
adirmcControl(
  studies         = list(...),
  n_sim           = 5000L,
  phases          = c(2, 1, 0.5, 0.1),   # box constraint half-widths per phase
  omega_expansion = 1.5,                   # inflate proposal Omega
  grad            = "analytical",          # "analytical", "none", "fd"
  kappa_method    = "exact",               # "exact" (default) or "linearized"
  seed            = 1L
)
```
