---
title: "Synthesising published models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Synthesising published models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.2)
source("_common.R", local = TRUE)
```

Sometimes there is no data. Every study you want to combine published a **model**
— a structure and a parameter table, usually with standard errors — plus a
baseline-characteristics table describing its patients. Nothing else: no
concentrations, no individual records, not even an aggregate mean profile.

There are **two** ways to turn that into a likelihood, and they are not the same
likelihood. This vignette builds both and compares them.

| | data are | weighted by | gives you |
|---|---|---|---|
| **aggregate** | the profiles each model predicts | `n_s` — *your choice* | a model that reproduces predictions |
| **parameter** | the published numbers themselves | `Σ_s` — *reported* | SEs, and a goodness-of-fit test |

# Theory

## Setup

Study $s$ enrolled $n_s$ patients with covariates $a \sim P_s$, ran design $D_s$,
and fitted its **own** model form $q_s(\cdot\mid a;\psi_s)$ — which may be
structurally reduced relative to the unified model $p_\varphi(\cdot\mid a)$ we
want: fewer compartments, fewer covariates, or both. It published
$\hat\theta_s$ with covariance $C_s$ (the RSE column).

Two currencies, and they are not interchangeable:

| | you receive | it supports |
|---|---|---|
| **data** | $(E_s, V_s, n_s)$ at times $t_s$ | an exact likelihood of sufficient statistics |
| **model** | $\hat\theta_s$, $C_s$, $D_s$, $P_s$ | a binding-function likelihood |

## 1. The target is conditional, and that is a theorem

For a source with covariate law $P_s$ there are two candidate objectives:

$$\text{(M)}\;\; D\big(m_s \,\|\, m_\varphi\big), \qquad
  \text{(C)}\;\; \mathbb{E}_{a\sim P_s}\big[D\big(p_s(\cdot\mid a)\,\|\,p_\varphi(\cdot\mid a)\big)\big]$$

where $m_s(y)=\int p_s(y\mid a)\,dP_s(a)$ is the marginal.

Put both on the joint space with the **same** covariate law,
$P(a,y)=P_s(a)p_s(y\mid a)$ and $Q(a,y)=P_s(a)p_\varphi(y\mid a)$. The projection
$(a,y)\mapsto y$ is a deterministic Markov kernel, so by the data-processing
inequality for $f$-divergences (Csiszár 1963; Ali & Silvey 1966; Liese & Vajda
2006),

$$\boxed{\;D(m_s\|m_\varphi)\;\le\;\mathbb{E}_{a\sim P_s}\big[D(p_s(\cdot\mid a)\|p_\varphi(\cdot\mid a))\big]\;}$$

**pointwise in $\varphi$**, with equality iff marginalisation is a *sufficient
statistic* for the pair — which fails exactly when the covariate discriminates
the two conditionals. So (M) is a **lower bound**: minimising it minimises a
bound, and its minimum is attained on a manifold rather than a point.

Two corollaries. First, by the chain rule for relative entropy,

$$D\big(P(a,y)\,\|\,Q(a,y)\big) \;=\; \underbrace{D(P_s\|P_s)}_{=\,0} \;+\; \mathbb{E}_{a}\big[D(p_s\|p_\varphi)\big],$$

so (C) **is** the joint KL — not an ad hoc average of divergences. Second,

$$\mathbb{E}_{a}\mathbb{E}_{y\mid a}\big[-\log p_\varphi(y\mid a)\big] \;=\; \text{(C)} \;+\; \mathbb{E}_a\big[H(p_s(\cdot\mid a))\big],$$

and the second term is free of $\varphi$, so minimising (C) is minimising the
expected **log score of the conditional forecaster** — a strictly proper scoring
rule. Both objectives are proper; they are proper *for different tasks*:

> (M) is proper for **predicting for a patient whose covariates you do not know**.
> (C) is proper for **predicting for a patient whose covariates you do know**.
> A model used for dosing is always deployed in the second task.

**Caveat, stated because it is easy to overclaim.** If the unified family
*contains* the source, both objectives reach zero at the same $\varphi$ and both
recover truth exactly — even from exact data with a nearly flat likelihood.
Flatness costs precision under noise, not location. The two targets separate only
under **family misspecification**, which in model synthesis is guaranteed but
whose *magnitude* is governed by the degree of structural mismatch, not by
covariate spread. This is confirmed in `validation/theory-tests.R`: with a
correctly specified family the marginal objective recovers $b=0.7500$ at every
covariate spread from $0.02$ to $0.40$.

## 2. What a covariate-free publication actually asserts

A source that omits covariate $a$ is **not** the model $\beta_a=0$. It is that
source's conditional *already marginalised* over $P_s$. Its published clearance is
$\mathbb{E}_{P_s}[\mathrm{CL}(a)]$, and — the load-bearing consequence — its
published IIV **absorbed the covariate variance**:

$$\omega_s^2 \;=\; \omega^2 \;+\; \beta_a^2\,\mathrm{Var}_{P_s}(a).$$

Verified to five decimals at every covariate spread (`theory-tests.R`, T3). So a
covariate-free publication is *not silent* about the covariate: it speaks through
its random-effect variance. Substituting $\beta_a=0$ instead injects a false
conditional and discards that channel.

**Measured cost of getting this wrong**: fitting the unified model to blocks
generated from publications, where one source lacks the covariate, attenuates that
coefficient to $0.167$ against a truth of $0.450$ — a **63% bias, with a standard
deviation of only $0.024$** across replicates. Confidently wrong, reproducibly
(`validation/three-way-replicates.R`).

## 3. Why node pooling fails on pooled data, and not otherwise

Scoring one pooled observation against $K$ conditional predictions gives
$\sum_k w_k\,\mathrm{NLL}_k$, whose mean term decomposes exactly as

$$\sum_k w_k\, r_k^\top V^{-1} r_k \;=\; \bar r^\top V^{-1}\bar r \;+\; \operatorname{tr}\!\big(V^{-1}\mathrm{Cov}_a(\mu)\big).$$

At the true parameters the first term is **identically zero** and the entire
objective is the second — a penalty on covariate-induced mean spread carrying no
data at all, which drives $\beta_a\to 0$ (measured $-59\%$). Identity confirmed to
$10^{-9}$ (`theory-tests.R`, T2).

Given **per-node data** the same sum is a genuine likelihood, because
$\sum_k c_k n\,\mathrm{NLL}_k \equiv \sum_k n_k \mathrm{NLL}_k$ with
$n_k = w_k n_s$ and $\sum_k n_k = n_s$ — ordinary multi-study fitting, with total
information the study's real size, not $K n_s$. The distinguishing question is
therefore **where the data come from**, never how many nodes there are. In the
model currency you *generate* $(E_k,V_k)$ at each node, so the invalid regime is
unreachable.

## 4. The binding function

For a **model**-currency source, treat the published numbers as the data:

$$\hat\theta_s \;\sim\; N\big(g_s(\varphi),\,C_s\big), \qquad
  g_s(\varphi) \;=\; \arg\max_{\psi}\; \mathbb{E}_{\varphi}\big[\log q_s(y;\psi)\big]$$

the **pseudo-true value** of study $s$'s estimator when the data-generating model
is $p_\varphi$ under $D_s$ and $P_s$. Asymptotic normality around a pseudo-true
value holds even under misspecification (White 1982), which is what licenses a
reduced source model. Dropping the $\varphi$-free $\log|C_s|$,

$$-2\ell_s(\varphi) \;=\; \big(\hat\theta_s - g_s(\varphi)\big)^\top C_s^{-1} \big(\hat\theta_s - g_s(\varphi)\big).$$

This is **indirect inference** in Wald form (Gouriéroux, Monfort & Renault 1993);
with several heterogeneous auxiliary models it is **composite indirect inference**
(Gouriéroux & Monfort 2018), and the whole construction is **Generalized Model
Aggregation** (Rahmandad, Jalali & Paynabar 2017). Because our sources are
*independent samples*, the optimal weight is block-diagonal and the criterion above
is exactly CII-optimal rather than an approximation.

The essential point: $g_s$ is what lets a one-compartment paper inform a
two-compartment model. Its $V$ and our $V_c$ are different quantities and are never
identified with one another; $g_s$ *computes the map* instead of assuming it.

## 5. The per-source rule

$g_s$ must reproduce study $s$'s **estimator**, so it must show that estimator the
covariate the way the estimator saw it:

| study $s$ published | generate for the inner refit |
|---|---|
| a covariate coefficient | **per-node** $(E_k,V_k)$, $n_k=w_kn_s$ — its authors had individual data resolving $a$ per subject |
| **no** covariate coefficient | **one pooled** $(E,V)$ marginal over $P_s$ — its estimate *is* $\mathbb{E}_{P_s}[\cdot]$ |

Note the asymmetry with random effects: $\eta$ is **never** observed by anyone, so
it is always marginalised and raises no such question. Covariates *were* observed,
per subject, by the source's authors. That is the entire reason only covariates
force this choice.

## 6. The combined objective

$$-2\ell(\varphi) \;=\; \underbrace{\sum_{s\in\mathcal{D}}\sum_k n_{sk}\Big[\log|V| + \operatorname{tr}(V^{-1}V_{\mathrm{obs}}) + r^\top V^{-1} r\Big]}_{\text{data currency}} \;+\; \underbrace{\sum_{s\in\mathcal{M}} \big(\hat\theta_s - g_s(\varphi)\big)^\top C_s^{-1}\big(\hat\theta_s-g_s(\varphi)\big)}_{\text{model currency}}$$

This is a **composite likelihood**, not a likelihood, so the variance is Godambe:
$\mathrm{Var}(\hat\varphi) = H^{-1} J H^{-1}$ with $J=\sum_s U_sU_s^\top$ built from
*total per-study* scores — which absorbs any overlap between the two terms. The
indirect-inference block additionally carries a $(1+1/S)$ inflation for simulation
noise, vanishing under deterministic quadrature. AIC/BIC and likelihood-ratio tests
are **not valid** here without the Chandler–Bate adjustment.

Use **one currency per study**: data if you have it, the model if you don't.

## 7. Diagnostics

Over-identification $Q=\sum_s\|C_s^{-1/2}(\hat\theta_s-g_s(\hat\varphi))\|^2$ on
$\sum_s p_s - \dim\varphi$ degrees of freedom. Report it as a **diagnostic, never a
gate**: more matched coefficients *lowers* power, and at large $n_s$ it rejects
almost any approximate model. Simulate its null rather than trusting $\chi^2$.

Also: the $g_s$ identity check (a source whose form matches the unified one must
return $\varphi$ unchanged), node-count convergence, the design-sensitivity spread,
and each source projected alone.

## 8. What this buys, measured

Six replicates, publications produced by real FOCEI fits, scored against a fit to
the **pooled individual data** (`validation/three-way-replicates.R`):

| | RMSE vs truth | RMSE vs pooled-IPD |
|---|---|---|
| pooled IPD (gold) | 0.0442 | — |
| **parameter likelihood** | **0.0641** | **0.0502** |
| aggregate-on-nodes | 0.1076 | 0.1219 |

The parameter likelihood recovers most of what the individual data support, from
two parameter tables and nothing else. The cost is variance: on the covariate
absent from one source it is unbiased ($0.489$ vs truth $0.450$, matching the gold
standard's $0.492$) but roughly half as precise as the IPD fit.

# Worked example

```{r libs, message = FALSE}
library(admixr2)
library(rxode2)
```

## 1. Write each study down

Two published models for the same drug. They do not share structure — the second
never estimated a volume–weight relationship — their populations differ in body
weight, and their designs have nothing in common.

```{r models}
mk_U <- function(p) eval(parse(text = sprintf('function() {   # source A / unified
  ini({ lcl <- %.8f; lv <- %.8f; clwt <- %.8f; vwt <- %.8f
        eta.cl ~ %.8f; prop.err <- %.8f })
  model({ cl <- exp(lcl + eta.cl) * (WT/70)^clwt
          v  <- exp(lv) * (WT/70)^vwt
          d/dt(centr) <- -cl/v * centr
          cp <- centr/v
          cp ~ prop(prop.err) }) }',
  p[["lcl"]], p[["lv"]], p[["clwt"]], p[["vwt"]], p[["om.cl"]], p[["prop.err"]])))

mk_B <- function(p) eval(parse(text = sprintf('function() {   # source B: no vwt
  ini({ lcl <- %.8f; lv <- %.8f; clwt <- %.8f
        eta.cl ~ %.8f; prop.err <- %.8f })
  model({ cl <- exp(lcl + eta.cl) * (WT/70)^clwt
          v  <- exp(lv)
          d/dt(centr) <- -cl/v * centr
          cp <- centr/v
          cp ~ prop(prop.err) }) }',
  p[["lcl"]], p[["lv"]], p[["clwt"]], p[["om.cl"]], p[["prop.err"]])))
```

The published parameter tables, **with their standard errors** — which the second
objective needs and the first cannot use:

```{r published}
PHI <- c(lcl = log(4.2), lv = log(48), clwt = 0.75, vwt = 1.0,
         om.cl = 0.08, prop.err = 0.13)                        # source A
THB <- c(lcl = log(5.1), lv = log(55), clwt = 0.62,
         om.cl = 0.05, prop.err = 0.16)                        # source B

SEA <- c(lcl = .05, lv = .06, clwt = .05, vwt = .07, om.cl = .020, prop.err = .012)
SEB <- c(lcl = .04, lv = .05, clwt = .16,            om.cl = .011, prop.err = .009)
```

Note the deliberate tension: source A has **40** patients but sampled richly, so
its exponents are tight (`clwt` SE `r SEA[["clwt"]]`); source B has **150** but
took troughs only, so its weight exponent is poorly determined (SE
`r SEB[["clwt"]]`). Patient count and precision point in opposite directions, and
that is exactly where the two objectives separate.

```{r designs}
lnorm_q <- function(mean, sd) {
  ml <- log(mean^2/sqrt(sd^2 + mean^2)); sl <- sqrt(log(1 + sd^2/mean^2))
  function(u) stats::qlnorm(u, ml, sl)
}
design <- list(
  A = list(n = 40,  times = c(0.5, 1, 2, 4, 6, 8, 12),
           ev = rxode2::et(amt = 500),
           cov_dist = list(WT = list(quantile = lnorm_q(68, 12)))),
  B = list(n = 150, times = c(11.5, 13),
           ev = rxode2::et(amt = 750, ii = 12, addl = 5),
           cov_dist = list(WT = list(quantile = lnorm_q(84, 19)))))
```

`covDraw()` shows what a `cov_dist` actually describes — the cheapest way to catch
a transcription error before a fit depends on it.

```{r checkspec}
round(summary(covDraw(design$A$cov_dist, n = 2000)[, "WT"]), 1)
```

## 2. Split each study into covariate strata

Kept whole, a study's spread of body weights shows up only inside its covariance
matrix, which is a weak and ridge-prone way to see a covariate effect. Cutting it
into `K` strata turns that spread into **differences between block means**, the
strong channel. Each stratum is an ordinary study carrying its own slice of the
distribution and a share `n / K` of the patients — `n` is a weight, not a head
count, so a fractional value is right and keeps every stratum populated.

```{r stratify}
stratify <- function(study, K, covariate = "WT") {
  q <- study$cov_dist[[covariate]]$quantile
  stats::setNames(lapply(seq_len(K), function(k) {
    force(k); s <- study; s$n <- study$n / K
    s$cov_dist[[covariate]] <- list(quantile = function(u) q(((k - 1) + u)/K)); s
  }), paste0(covariate, seq_len(K)))
}
```

Slicing the **quantile** rather than the covariate is what keeps each stratum's own
distribution intact: every block is still marginalised over the weights inside it,
never evaluated at the stratum's mean weight. (With dependent covariates, slice the
*uniform* and let the joint sampler deliver the rest conditionally — cutting one
covariate's quantiles and redrawing the others from their margins destroys the
dependence.)

```{r infra, message = FALSE}
GEN <- datagenControl(n_sim = 8000L, seed = 11L)
K   <- 4L

gen_from <- function(mod, d, K)                     # what `mod` would produce
  datagen(stratify(utils::modifyList(d, list(model = mod)), K), control = GEN)

fit_form <- function(startmod, studies)             # fit a structure to blocks
  suppressWarnings(suppressMessages(nlmixr2est::nlmixr2(
    startmod, admData(), est = "adgh",
    control = adghControl(studies = studies, n_nodes = 5L, grad = "analytical",
                          print = 0L, covMethod = "none"))))

grab <- function(f, nms) {                          # estimates as a named vector
  fx <- nlmixr2est::fixef(f); om <- f$omega
  stats::setNames(vapply(nms, function(n)
    if (n == "om.cl") om[["eta.cl", "eta.cl"]] else fx[[n]], 0), nms)
}
PN <- names(PHI); BN <- names(THB)
```

## 3. Can the design even see the covariate?

Before fitting anything: some designs cancel the covariate algebraically — mg/kg
dosing with clearance and volume both linear in weight is the classic case — and
such a study says nothing about it however large its `n`. The diagnostic is the
spread of the predicted mean across strata.

```{r spread, message = FALSE}
cov_spread <- function(mod, d, K = 8L) {
  m <- vapply(gen_from(mod, d, K), function(u) mean(u$E), numeric(1))
  100 * (max(m) - min(m)) / mean(m)
}
spread <- c(A = cov_spread(mk_U(PHI), design$A),
            B = cov_spread(mk_B(THB), design$B))
round(spread, 1)
```

Both respond to weight, `A` about `r round(spread[["A"]]/spread[["B"]], 1)` times
as strongly.

# The aggregate likelihood

Re-simulate each published model over its own population and design, treat the
result as data with weight `n_s`, and score admixr2's mean-and-covariance
likelihood.

```{r aggregate, message = FALSE, warning = FALSE}
agg_blocks <- function(K) c(
  stats::setNames(gen_from(mk_U(PHI), design$A, K), paste0("A_", seq_len(K))),
  stats::setNames(gen_from(mk_B(THB), design$B, K), paste0("B_", seq_len(K))))

START <- c(lcl = log(4.5), lv = log(50), clwt = .5, vwt = .5,
           om.cl = .07, prop.err = .15)
agg <- grab(fit_form(mk_U(START), agg_blocks(K)), PN)
round(agg, 3)
```

`K` is a resolution, so the answer must stop moving as it grows. A parameter that
keeps wandering is not identified by these studies, and *that* is the result.

```{r kconv, message = FALSE, warning = FALSE}
ks <- c(1L, 2L, 4L, 8L)
conv <- t(vapply(ks, function(k)
  grab(fit_form(mk_U(START), agg_blocks(k)), c("clwt", "vwt")), numeric(2)))
knitr::kable(data.frame(K = ks, round(conv, 3)),
             caption = "Aggregate objective: convergence in the number of strata")
```

The `K = 1` row is the same fit without step 2: `vwt` comes out
`r round(conv[1, "vwt"], 2)` against `r round(conv[4, "vwt"], 2)` at `K = 8`, a
`r round(100 * abs(conv[1, "vwt"] / conv[4, "vwt"] - 1))`% difference. That is
what stratification buys.

Read the rest of the column honestly, though. `clwt` goes
`r paste(round(conv[-1, "clwt"], 3), collapse = " -> ")` — still drifting, by
about the same amount at each doubling. **This example has not converged at
`K = 8`.** That is not a reason to hide the table; it is the reason to print it.
A parameter still moving at the largest `K` you can afford should be reported as
unresolved, not given a value.

**What this objective cannot do.** `n_s` is a weighting you chose, so its Hessian
reports whatever precision you asked for: measured coverage of a nominal 95%
interval is **0.58**. Valid intervals are still available — propagate the sources'
own uncertainty through the fit by the delta method, which restores coverage to
0.95 — but they do not come from the fit itself. See
`validation/model-synthesis-formal-likelihood.R`.

# The parameter likelihood

Treat the **published numbers** as the data:

$$\hat\theta_s \;\sim\; N\!\big(g_s(\varphi),\,\Sigma_s\big)$$

where $\varphi$ are the unified parameters and $g_s(\varphi)$ is the **induced
map**: what study *s* would have estimated, under its own model form, its own
design and its own population, if the truth were $\varphi$. This is the projection
run *backwards* — the unified model projected **onto** each source's structure.

## The induced map

When a source's model form equals the unified form, $g_s$ is the identity and
costs nothing. Worth verifying rather than assuming:

```{r identity, message = FALSE, warning = FALSE}
gA_check <- grab(fit_form(mk_U(START), gen_from(mk_U(PHI), design$A, K)), PN)
max(abs(gA_check - PHI))
```

Only structurally reduced sources need an inner projection. Source B has no
volume–weight term, so its map must be computed:

```{r gB, message = FALSE, warning = FALSE}
gB <- function(phi) grab(fit_form(mk_B(THB), gen_from(mk_U(phi), design$B, K)), BN)
gB0 <- gB(PHI)
round(gB0, 3)
```

Read that off: **if source A's model were the truth, source B would have published
a weight exponent of `r round(gB0[["clwt"]], 2)`** — against the
`r THB[["clwt"]]` it actually published. That number is a diagnostic the aggregate
route has no way to produce.

## Fitting it

The residual vector stacks both sources, each scaled by its own published SE.
Gauss–Newton, with the Jacobian by finite differences:

```{r paramll, message = FALSE, warning = FALSE}
resid <- function(phi) c((PHI[PN] - phi[PN]) / SEA[PN],       # g_A = identity
                         (THB[BN] - gB(phi))  / SEB[BN])

FD  <- c(lcl = 4e-3, lv = 4e-3, clwt = 8e-3, vwt = 8e-3,
         om.cl = 4e-3, prop.err = 3e-3)
phi <- PHI
for (it in 1:3) {
  r0 <- resid(phi)
  J  <- vapply(PN, function(p) {
    ph <- phi; ph[[p]] <- ph[[p]] + FD[[p]]
    (resid(ph) - r0) / FD[[p]] }, numeric(length(r0)))
  phi <- phi - as.numeric(solve(crossprod(J) + diag(1e-8, ncol(J)),
                                crossprod(J, r0)))
}
rF <- resid(phi); Vphi <- solve(crossprod(J))
se_fe <- sqrt(diag(Vphi))
round(phi, 3)
```

These standard errors are real. `Σ_s` is reported information, not a weighting
choice, so `(JᵀΣ⁻¹J)⁻¹` is a genuine Fisher information inverse — coverage
measured at 0.949.

```{r se-fe}
round(se_fe, 3)
```

# Comparing the two

```{r compare, message = FALSE}
cmp <- data.frame(`A published` = PHI[PN],
                  `B published` = ifelse(PN %in% BN, THB[match(PN, BN)], NA),
                  aggregate = agg[PN], `parameter LL` = phi[PN],
                  check.names = FALSE)
knitr::kable(round(cmp, 3), caption = "The same two publications, two objectives")
```

They disagree, and systematically. `vwt` comes out
`r round(agg[["vwt"]], 2)` under the aggregate objective and
`r round(phi[["vwt"]], 2)` under the parameter likelihood, because only source A
has a volume–weight term and the aggregate objective weights by patient count —
where B outnumbers A `r design$B$n`:`r design$A$n` — while the parameter
likelihood weights by *reported precision*, where A's tight exponents dominate
B's loose ones.

Neither is wrong. They are answers to different questions, and the choice is a
choice of loss function:

* **predicting exposure** in a new population — the aggregate objective targets
  predictions directly, which is the thing you care about;
* **reconciling the literature**, reporting coefficients with intervals, or
  testing whether the published models agree — the parameter likelihood, which is
  the only one of the two that can do any of it.

## Goodness of fit — the test the aggregate objective cannot run

With `r length(rF)` matched quantities and `r length(PN)` unified parameters there
are `r length(rF) - length(PN)` degrees of freedom of disagreement that one shared
structure has to explain.

```{r Q}
Q <- sum(rF^2); dfQ <- length(rF) - length(PN)
c(Q = round(Q, 2), df = dfQ, p = signif(stats::pchisq(Q, dfQ, lower.tail = FALSE), 3))
```

**It rejects.** These two publications cannot be reconciled to within their own
reported precision by any single model of this structure. The aggregate fit above
returned a perfectly plausible-looking answer and gave no hint of it — which is the
failure mode this whole exercise exists to avoid.

## Between-source heterogeneity

The meta-analytic response to a rejected `Q` is a between-source variance `τ²`
added to every `Σ_s`, estimated so the weighted residual sum equals its degrees of
freedom. It reuses the Jacobian already computed, so it costs no further fits.

```{r tau2}
SEs <- c(SEA[PN], SEB[BN]); D <- J * SEs; rraw <- rF * SEs
solve_at <- function(tau2) {
  w <- 1/(SEs^2 + tau2); A <- crossprod(D, w * D) + diag(1e-10, ncol(D))
  d <- -solve(A, crossprod(D, w * rraw))
  list(d = d, Q = sum(w * (rraw + D %*% d)^2), V = solve(A))
}
tau2 <- stats::uniroot(function(t2) solve_at(t2)$Q - dfQ, c(0, 5), tol = 1e-10)$root
re   <- solve_at(tau2)
knitr::kable(round(data.frame(
  `fixed effect` = phi[PN], `SE` = se_fe,
  `random effect` = phi[PN] + as.numeric(re$d), `SE ` = sqrt(diag(re$V)),
  check.names = FALSE), 3),
  caption = sprintf("Allowing between-source heterogeneity (tau = %.3f)", sqrt(tau2)))
```

The four structural estimates move little and their standard errors roughly
**double**. The variance components do not behave the same way: `om.cl` roughly
halves and its interval widens about
`r round(sqrt(diag(re$V))[["om.cl"]] / se_fe[["om.cl"]])`-fold, `prop.err`
similarly. That is worth reading as a finding rather than noise — the two
publications disagreed about between-subject and residual variability far more
than about the structural model.

Either way, the fixed-effect intervals were narrow because they assumed an
agreement the `Q` test says is not there.

# What to do

1. Run **both**. They cost about the same and disagreement between them is
   information.
2. Always report **`Q`**. If it rejects, the point estimate is a compromise
   between models that do not agree, and must be presented as one.
3. If it rejects, report the **random-effects** intervals, not the fixed-effect
   ones.
4. Never read standard errors off the aggregate fit. Use the parameter
   likelihood's, or delta-method them through the aggregate one.
5. Check `K` convergence, and check each design can see the covariate at all.

## Two things not to do

* **Do not plug a stratum's mean covariate into `cov`.** Marginalise within the
  stratum. The plug-in is the ecological approximation at stratum scale and biases
  the coefficient upward — +17% at two strata, +4.3% at four.
* **Do not expect a coefficient to land between its sources.** A projection is a
  joint compromise over the whole parameter vector, not a per-parameter average.

## See also

* [Covariates in a model-based meta-analysis](covariate-mbma.html) — the same
  machinery when you *do* have aggregate data
* [Generating aggregate data](datagen.html) — `datagen()` and `datagenControl()`
* [Fitting several studies jointly](multiple-studies.html)
* [Choosing an estimator](estimator-comparison.html)
