---
title: "Several observed compartments (plasma and brain)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Several observed compartments (plasma and brain)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.2, dpi = 120
)
source(Filter(file.exists, c("_common.R", "../_common.R", "vignettes/_common.R"))[1])
```

## The problem

You are developing a CNS drug and you want one number: **how much of it reaches
the brain?** There is no patient-level data, only a published paper with two
figures — a **plasma** and a **brain** concentration–time curve, each a mean with
error bars, which you digitise into a mean and an SD per sampling time.

That summary is what admixr2 is built for: it fits a population PK model directly
to means and covariances, so a digitised figure is a fittable dataset. Two steps:

1. **Plasma only** — the ordinary single-output workflow, to set the scene.
2. **Plasma + brain** — add the brain as a second observed output and read the
   brain-penetration ratio straight off the fit.

```{r libs}
library(admixr2)
library(rxode2)
library(nlmixr2)
library(ggplot2)

ev <- rxode2::et(amt = 100, cmt = "central")   # single 100-unit dose, shared throughout
```

## Step 1 — Plasma only

Start where every PK analysis starts: the plasma curve. A **study** is the
digitised summary bundled with its design — the mean vector `E`, its variance `V`
(here `SD^2`, read as a diagonal covariance, which is all error bars give you),
the sample size `n`, the sampling `times` and the dosing `ev`.

```{r plasma-data}
plasma_times <- c(0.5, 1, 2, 4, 8, 12)
plasma_mean  <- c(8.793, 7.812, 6.850, 5.597, 3.985, 2.910)
plasma_sd    <- c(1.151, 0.911, 0.765, 0.649, 0.770, 0.828)

plasma_study <- list(
  E = plasma_mean, V = plasma_sd^2, n = 60L,
  times = plasma_times, ev = ev
)
```

An ordinary two-compartment model with one observed output, `cp`, fitted in one
call to `nlmixr2()` with admixr2's Gauss–Hermite estimator:

```{r plasma-fit, comment = NA}
pk_plasma <- function() {
  ini({
    tcl <- log(1);  tv1 <- log(10);  tq <- log(3);  tv2 <- log(8)
    prop.cp <- 0.05
    eta.cl ~ 0.09
    eta.v1 ~ 0.04
  })
  model({
    cl <- exp(tcl + eta.cl); v1 <- exp(tv1 + eta.v1)
    q  <- exp(tq);           v2 <- exp(tv2)
    d/dt(central) <- -(cl/v1)*central - (q/v1)*central + (q/v2)*periph
    d/dt(periph)  <-  (q/v1)*central - (q/v2)*periph
    cp <- central / v1
    cp ~ prop(prop.cp)
  })
}

fit_plasma <- nlmixr2(pk_plasma, admData(), est = "adgh",
                      control = adghControl(studies = list(trial = plasma_study)))
fit_plasma
```

A perfectly good plasma model — and look at what it cannot answer. `periph` is a
mathematical distribution store: never measured, and connected to the brain by
nothing. Brain exposure needs brain data **and** a real brain compartment.

## Step 2 — Add the brain

Here are the brain concentrations digitised from the same paper:

```{r brain-data}
brain_times <- c(1, 2, 4, 8, 12)
brain_mean  <- c(3.004, 3.394, 3.018, 2.157, 1.551)
brain_sd    <- c(0.353, 0.349, 0.309, 0.369, 0.405)
```

Swap the anonymous peripheral compartment for a **mechanistic brain
compartment**: drug moves plasma → brain with influx clearance `qin` and back
with efflux clearance `qout`. The steady-state brain:plasma ratio is what we are
after:

$$K_{p,uu} = \frac{q_{in}}{q_{out}}$$

With **two** observed outputs — plasma `cp` and brain `cb` — the model carries a
residual-error term for each. (`vb`, the brain volume, is a fixed physiological
constant, not an estimated parameter.)

```{r brain-model}
pk_cns <- function() {
  ini({
    tcl  <- log(1);  tv1  <- log(10)
    tqin <- log(3);  tqout <- log(6)
    prop.cp <- 0.05      # plasma residual (proportional)
    add.cb  <- 0.02      # brain residual (additive)
    eta.cl ~ 0.09
    eta.v1 ~ 0.04
  })
  model({
    cl <- exp(tcl + eta.cl); v1 <- exp(tv1 + eta.v1)
    qin <- exp(tqin);        qout <- exp(tqout)
    vb  <- 5
    d/dt(central) <- -(cl/v1)*central - (qin/v1)*central + (qout/vb)*brain
    d/dt(brain)   <-  (qin/v1)*central - (qout/vb)*brain
    cp <- central / v1        # plasma concentration
    cb <- brain   / vb        # brain concentration
    cp ~ prop(prop.cp)
    cb ~ add(add.cb)
  })
}
```

Two outputs means two summaries. In place of a single `E`/`V`, give the study an
**`observations` list** — one named entry per observed compartment, pairing a
model output with its own `times`, `E` and `V`:

```{r brain-study}
cns_study <- list(
  n = 60L, ev = ev,
  observations = list(
    plasma = list(output = "cp", times = plasma_times, E = plasma_mean, V = plasma_sd^2),
    brain  = list(output = "cb", times = brain_times,  E = brain_mean,  V = brain_sd^2)
  )
)
```

The only other change is telling `admData()` which outputs to expect; the fit call
is otherwise identical to Step 1:

```{r brain-fit, comment = NA}
fit_cns <- nlmixr2(pk_cns, admData(c("cp", "cb")), est = "adgh",
                   control = adghControl(studies = list(lit = cns_study)))
fit_cns
```

## The same study in long format

If you have used nlmixr2's [multiple
endpoints](https://nlmixr2.org/articles/multiple-endpoints.html), the
`observations` list may feel like a detour: nlmixr2 stacks observations in **one**
data frame and labels each row with its endpoint (`DVID` / `CMT`).

admixr2 takes a study written that way too — a `data` frame with one row per
observed *endpoint × time*: an endpoint column (`DVID`, `CMT` or `output`), a
time column (`TIME`), the mean (`E`) and its variance (`V`, or an `SD` column):

```{r long-study}
cns_long <- list(
  n = 60L, ev = ev,
  data = data.frame(
    DVID = c(rep("cp", length(plasma_times)), rep("cb", length(brain_times))),
    TIME = c(plasma_times, brain_times),
    E    = c(plasma_mean,  brain_mean),
    V    = c(plasma_sd,    brain_sd)^2
  )
)
```

The same study, written differently: admixr2 normalises it into the same
likelihood blocks, so the fits agree to the last digit.

```{r long-fit, comment = NA}
fit_long <- nlmixr2(pk_cns, admData(c("cp", "cb")), est = "adgh",
                    control = adghControl(studies = list(lit = cns_long)))

c(observations = fit_cns$objective, long = fit_long$objective)
```

Which to use is taste. `observations` keeps each compartment's design visibly
together, the natural way to write a study *by hand*. Long format is the natural
way to write one *from data* — the shape summaries already have in a spreadsheet
or out of a `dplyr` pipeline, and the shape nlmixr2 itself uses.

### Same-subject data: one stacked covariance

Long format earns its keep when plasma and brain were measured in the **same
subjects**. The two curves are then correlated, and that correlation is
information the fit should use — but it lives *between* the compartments, where
per-compartment `V` matrices have nowhere to put it.

Long format does: every observation being a row, the study takes **one**
covariance matrix spanning all of them, aligned with the rows of `data` — exactly
what `cov.wt(dv_mat, method = "ML")$cov` returns from the subject-level matrix
(one row per subject, one column per observation, plasma columns then brain).

```{r long-joint, eval = FALSE}
cns_joint <- list(
  n = 60L, ev = ev,
  data = data.frame(
    DVID = c(rep("cp", length(plasma_times)), rep("cb", length(brain_times))),
    TIME = c(plasma_times, brain_times),
    E    = c(plasma_mean,  brain_mean)
  ),
  V = V_joint   # 11 x 11: plasma block, brain block, and the cross-covariance
)
```

admixr2 then scores all 11 observations with a **single** multivariate normal,
simulating both compartments from shared random effects. A study-level `V` is
what marks the study same-subject; without one each endpoint stays an independent
block. (`observations` can express this too, via a `cross` list of per-pair
blocks — see `?admControl` — but assembling those by hand is the chore long
format removes.)

The choice is a **modelling** decision, not a formatting one: a joint fit with
zero cross-covariances is *not* two independent blocks. The model predicts plasma
and brain to co-vary, sharing `eta.cl` and `eta.v1`, so telling it you observed
no covariance is a real statement about the data and the likelihood will hold you
to it. Joint when the compartments came from the same subjects, independent when
they did not — here, two separate figures.

### Model against data

One fit now describes both compartments. Below, the observed summaries (points,
mean ± SD) sit under the fitted population prediction — the mean curve and the
±SD band implied by the estimated between-subject variability, over 1000
simulated subjects with residual error excluded.

```{r brain-plot, fig.height = 4.6}
# 1. Pull the fitted estimates from the fit (standard nlmixr2 accessors).
theta <- fit_cns$theta   # fixed effects (tcl, tv1, tqin, tqout, residuals)
omega <- fit_cns$omega   # between-subject covariance (rows/cols already named)

# 2. Simulate the fitted population. We solve a plain, residual-free copy of the
#    model so the band shows between-subject variability alone -- and because a
#    multi-endpoint fit can only be re-solved with per-endpoint DVID/CMT tags,
#    whereas this gives cp and cb directly.
sim_model <- rxode2::rxode2({
  cl <- exp(tcl + eta.cl); v1 <- exp(tv1 + eta.v1)
  qin <- exp(tqin); qout <- exp(tqout); vb <- 5
  d/dt(central) <- -(cl/v1)*central - (qin/v1)*central + (qout/vb)*brain
  d/dt(brain)   <-  (qin/v1)*central - (qout/vb)*brain
  cp <- central / v1
  cb <- brain / vb
})
grid <- seq(0.25, 13, by = 0.25)
sim  <- rxode2::rxSolve(sim_model,
  params = theta[c("tcl", "tv1", "tqin", "tqout")],
  omega  = omega, nSub = 1000L,
  events = ev |> rxode2::et(grid), returnType = "data.frame")

# 3. Summarise model and data as mean +/- SD per time, per compartment.
band <- function(value)
  data.frame(time = sort(unique(sim$time)),
             mean = tapply(value, sim$time, mean),
             sd   = tapply(value, sim$time, sd))

pred <- rbind(cbind(compartment = "Plasma", band(sim$cp)),
              cbind(compartment = "Brain",  band(sim$cb)))
obs  <- rbind(
  data.frame(compartment = "Plasma", time = plasma_times, mean = plasma_mean, sd = plasma_sd),
  data.frame(compartment = "Brain",  time = brain_times,  mean = brain_mean,  sd = brain_sd))

# 4. Model (line + band) over data (points + error bars), coloured by compartment.
pal <- c(Plasma = "#0072B2", Brain = "#D55E00")   # Okabe-Ito, colour-blind safe

ggplot() +
  geom_ribbon(data = pred,
              aes(time, ymin = mean - sd, ymax = mean + sd, fill = compartment),
              alpha = 0.15) +
  geom_line(data = pred,
            aes(time, mean, colour = compartment), linewidth = 1) +
  geom_errorbar(data = obs,
                aes(time, ymin = mean - sd, ymax = mean + sd, colour = compartment),
                width = 0.4, linewidth = 0.6, show.legend = FALSE) +
  geom_point(data = obs,
             aes(time, mean, fill = compartment),
             shape = 21, colour = "white", size = 3, stroke = 0.7) +
  scale_colour_manual(values = pal, breaks = c("Plasma", "Brain")) +
  scale_fill_manual(values = pal, guide = "none") +
  scale_x_continuous(breaks = seq(0, 12, 2),
                     expand = expansion(mult = c(0.01, 0.03))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  coord_cartesian(ylim = c(0, NA), clip = "off") +
  guides(colour = guide_legend(override.aes = list(linewidth = 1.4))) +
  labs(x = "Time (h)", y = "Concentration (mg/L)",
       title = "Fitted population prediction vs. observed data",
       subtitle = "Line + band: model (mean ± SD)   ·   Points + bars: data (mean ± SD)",
       colour = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    legend.justification = "left",
    legend.margin = margin(b = 2),
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(colour = "grey40", size = 9, margin = margin(b = 9)),
    axis.title = element_text(colour = "grey25"),
    axis.title.x = element_text(margin = margin(t = 6)),
    axis.title.y = element_text(margin = margin(r = 6)),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(colour = "grey90", linewidth = 0.4),
    plot.margin = margin(12, 16, 10, 12)
  )
```

The prediction tracks both compartments. Now the payoff, read straight off the
estimates:

```{r kp}
Kp_uu <- exp(theta[["tqin"]]) / exp(theta[["tqout"]])
round(Kp_uu, 2)
```

`Kp,uu` ≈ `r round(Kp_uu, 1)`: at steady state the brain sees about half the
plasma concentration. This is why the brain data was needed at all — with plasma
alone `qin` and `qout` are not separately identifiable, and `Kp,uu` cannot be
estimated.

### Built-in diagnostics

That overlay was assembled by hand for a custom figure. You don't have to:
`plot()` draws observed-vs-predicted panels directly, one per observed output.

```{r plot-fit, fig.width = 7, fig.height = 3.2}
plot(fit_cns, which = "mean")
```

## Notes

* **Estimators.** `adgh` (used here), `adfo` and `admc` support several observed
  outputs, each with its own analytical / sensitivity gradient. `adirmc` does
  not. See the [estimator comparison](estimator-comparison.html) to choose.
* **Structural vs. observed compartments.** How many compartments the ODEs hold
  is irrelevant; what counts is how many outputs you *observe* and fit — one in
  Step 1, two in Step 2.
* **Hard-coded constants** such as `vb <- 5` keep their value. Not every
  physiological constant has to be estimated.
* **Two ways to write a study.** `observations` and the long-format `data` frame
  are interchangeable: same normalisation, same likelihood, same numbers.
  Independent experiments can also carry a per-endpoint `n` column and a
  per-endpoint `ev`, a list of event tables keyed by endpoint.
* **Same-subject data.** Plasma and brain came from separate figures here, so
  they are independent likelihood blocks. Measured in the *same* subjects, supply
  the cross-covariance instead — one stacked `V` in long format, or a
  per-output-pair `cross` list with `observations`; see `?admControl`.

## See also

- [PD and PK/PD data](pkpd.html) — add a pharmacodynamic endpoint
- [Multiple studies](multiple-studies.html) — meta-analysis across studies
- [Diagnostic plots](diagnostic-plots.html)
