---
title: "Several observed compartments (plasma and brain)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Several observed compartments (plasma and brain)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.2, dpi = 120
)
source(Filter(file.exists, c("_common.R", "../_common.R", "vignettes/_common.R"))[1])
```

## The problem

You are developing a CNS drug and you want one number: **how much of it reaches
the brain?** There is no patient-level data, only a published paper with two
figures — a **plasma** and a **brain** concentration–time curve, each a mean with
error bars, which you digitise into a mean and an SD per sampling time.

That summary is what admixr2 is built for: it fits a population PK model directly
to means and covariances, so a digitised figure is a fittable dataset. Two steps:

1. **Plasma only** — the ordinary single-output workflow, to set the scene.
2. **Plasma + brain** — add the brain as a second observed output and read the
   brain-penetration ratio straight off the fit.

```{r libs}
library(admixr2)
library(rxode2)
library(nlmixr2)
library(ggplot2)

ev <- rxode2::et(amt = 100, cmt = "central")   # single 100-unit dose, shared throughout
```

## Step 1 — Plasma only

Start where every PK analysis starts: the plasma curve. A **study** is the
digitised summary bundled with its design — the mean vector `E`, its variance `V`
(here `SD^2`, read as a diagonal covariance, which is all error bars give you),
the sample size `n`, the sampling `times` and the dosing `ev`.

```{r plasma-data}
plasma_times <- c(0.5, 1, 2, 4, 8, 12)
plasma_mean  <- c(8.793, 7.812, 6.850, 5.597, 3.985, 2.910)
plasma_sd    <- c(1.151, 0.911, 0.765, 0.649, 0.770, 0.828)

plasma_study <- list(
  E = plasma_mean, V = plasma_sd^2, n = 60L,
  times = plasma_times, ev = ev
)
```

An ordinary two-compartment model with one observed output, `cp`, fitted in one
call to `nlmixr2()` with admixr2's Gauss–Hermite estimator:

```{r plasma-fit, comment = NA}
pk_plasma <- function() {
  ini({
    tcl <- log(1);  tv1 <- log(10);  tq <- log(3);  tv2 <- log(8)
    prop.cp <- 0.05
    eta.cl ~ 0.09
    eta.v1 ~ 0.04
  })
  model({
    cl <- exp(tcl + eta.cl); v1 <- exp(tv1 + eta.v1)
    q  <- exp(tq);           v2 <- exp(tv2)
    d/dt(central) <- -(cl/v1)*central - (q/v1)*central + (q/v2)*periph
    d/dt(periph)  <-  (q/v1)*central - (q/v2)*periph
    cp <- central / v1
    cp ~ prop(prop.cp)
  })
}

fit_plasma <- nlmixr2(pk_plasma, admData(), est = "adgh",
                      control = adghControl(studies = list(trial = plasma_study)))
fit_plasma
```

A perfectly good plasma model — and look at what it cannot answer. `periph` is a
mathematical distribution store: never measured, and connected to the brain by
nothing. Brain exposure needs brain data **and** a real brain compartment.

## Step 2 — Add the brain

Here are the brain concentrations digitised from the same paper:

```{r brain-data}
brain_times <- c(1, 2, 4, 8, 12)
brain_mean  <- c(3.004, 3.394, 3.018, 2.157, 1.551)
brain_sd    <- c(0.353, 0.349, 0.309, 0.369, 0.405)
```

Swap the anonymous peripheral compartment for a **mechanistic brain
compartment**: drug moves plasma → brain with influx clearance `qin` and back
with efflux clearance `qout`. The steady-state brain:plasma ratio is what we are
after:

$$K_{p,uu} = \frac{q_{in}}{q_{out}}$$

```{r cmt-diagram, echo=FALSE, fig.height=2.9, fig.width=7, fig.cap="The two structures. Step 1's peripheral compartment is a distribution store: drug goes in and comes back, but nothing was ever measured there, so no amount of plasma data says what concentration it holds. Step 2 gives that compartment a volume, a sampled concentration and a name."}
# One panel per structure, same geometry and different labels: the point is that
# Step 2 swaps an unmeasured distribution store for a compartment that was
# actually sampled.
.pan <- c("Step 1 - plasma only", "Step 2 - plasma + brain")
box <- rbind(
  data.frame(panel = .pan[1], x = 0,   lab = "central\nV1",   out = "cp"),
  data.frame(panel = .pan[1], x = 3.2, lab = "periph\nV2",    out = NA),
  data.frame(panel = .pan[2], x = 0,   lab = "central\nV1",   out = "cp"),
  data.frame(panel = .pan[2], x = 3.2, lab = "brain\nVb = 5", out = "cb"))
box$panel <- factor(box$panel, .pan)
box$seen  <- !is.na(box$out)
box$tag   <- ifelse(box$seen, paste0("observed as ", box$out),
                    "never measured")

arr  <- arrow(length = unit(0.16, "cm"), type = "closed")
flow <- rbind(
  data.frame(panel = .pan, x = 0.95, xend = 2.25, y =  0.22, yend =  0.22,
             lab = c("Q", "qin"),  vj = -0.6),
  data.frame(panel = .pan, x = 2.25, xend = 0.95, y = -0.22, yend = -0.22,
             lab = c("Q", "qout"), vj =  1.5))
flow$panel <- factor(flow$panel, .pan)
dose <- data.frame(panel = factor(.pan, .pan), x = -1.9, xend = -0.95,
                   y = 0, yend = 0)
elim <- data.frame(panel = factor(.pan, .pan), x = 0, xend = 0,
                   y = -0.5, yend = -1.35)

ggplot() +
  geom_segment(data = dose, aes(x, y, xend = xend, yend = yend),
               arrow = arr, linewidth = 0.5, colour = "grey25") +
  geom_text(data = dose, aes(x - 0.05, y + 0.28, label = "dose"),
            size = 3.1, hjust = 0, colour = "grey25") +
  geom_segment(data = elim, aes(x, y, xend = xend, yend = yend),
               arrow = arr, linewidth = 0.5, colour = "grey25") +
  geom_text(data = elim, aes(x + 0.14, yend + 0.2, label = "CL"),
            size = 3.1, hjust = 0, colour = "grey25") +
  geom_segment(data = flow, aes(x, y, xend = xend, yend = yend),
               arrow = arr, linewidth = 0.5, colour = "grey25") +
  geom_text(data = flow, aes(x = 1.6, y = y, label = lab, vjust = vj),
            size = 3.1, colour = "grey25") +
  geom_tile(data = box, aes(x, 0, fill = seen, colour = seen),
            width = 1.9, height = 1, linewidth = 0.7) +
  geom_text(data = box, aes(x, 0, label = lab), size = 3.3,
            lineheight = 0.95) +
  geom_text(data = box, aes(x, -0.78, label = tag, colour = seen),
            size = 2.9, fontface = "italic", show.legend = FALSE) +
  scale_fill_manual(values = c(`TRUE` = "#CFE3F2", `FALSE` = "grey93"),
                    guide = "none") +
  scale_colour_manual(values = c(`TRUE` = "#0072B2", `FALSE` = "grey55"),
                      guide = "none") +
  facet_wrap(~panel) +
  coord_cartesian(xlim = c(-2.1, 4.3), ylim = c(-1.6, 0.95)) +
  theme_void(base_size = 12) +
  theme(strip.text = element_text(face = "bold", size = 10,
                                  margin = margin(b = 6)),
        panel.spacing = unit(1.2, "lines"))
```

With **two** observed outputs — plasma `cp` and brain `cb` — the model carries a
residual-error term for each. (`vb`, the brain volume, is a fixed physiological
constant, not an estimated parameter.)

```{r brain-model}
pk_cns <- function() {
  ini({
    tcl  <- log(1);  tv1  <- log(10)
    tqin <- log(3);  tqout <- log(6)
    prop.cp <- 0.05      # plasma residual (proportional)
    add.cb  <- 0.02      # brain residual (additive)
    eta.cl ~ 0.09
    eta.v1 ~ 0.04
  })
  model({
    cl <- exp(tcl + eta.cl); v1 <- exp(tv1 + eta.v1)
    qin <- exp(tqin);        qout <- exp(tqout)
    vb  <- 5
    d/dt(central) <- -(cl/v1)*central - (qin/v1)*central + (qout/vb)*brain
    d/dt(brain)   <-  (qin/v1)*central - (qout/vb)*brain
    cp <- central / v1        # plasma concentration
    cb <- brain   / vb        # brain concentration
    cp ~ prop(prop.cp)
    cb ~ add(add.cb)
  })
}
```

Two outputs means two summaries. In place of a single `E`/`V`, give the study an
**`observations` list** — one named entry per observed compartment, pairing a
model output with its own `times`, `E` and `V`:

```{r brain-study}
cns_study <- list(
  n = 60L, ev = ev,
  observations = list(
    plasma = list(output = "cp", times = plasma_times, E = plasma_mean, V = plasma_sd^2),
    brain  = list(output = "cb", times = brain_times,  E = brain_mean,  V = brain_sd^2)
  )
)
```

The only other change is telling `admData()` which outputs to expect; the fit call
is otherwise identical to Step 1:

```{r brain-fit, comment = NA}
fit_cns <- nlmixr2(pk_cns, admData(c("cp", "cb")), est = "adgh",
                   control = adghControl(studies = list(lit = cns_study)))
fit_cns
```

## The same study in long format

If you have used nlmixr2's [multiple
endpoints](https://nlmixr2.org/articles/multiple-endpoints.html), the
`observations` list may feel like a detour: nlmixr2 stacks observations in **one**
data frame and labels each row with its endpoint (`DVID` / `CMT`).

admixr2 takes a study written that way too — a `data` frame with one row per
observed *endpoint × time*: an endpoint column (`DVID`, `CMT` or `output`), a
time column (`TIME`), the mean (`E`) and its variance (`V`, or an `SD` column):

```{r long-study}
cns_long <- list(
  n = 60L, ev = ev,
  data = data.frame(
    DVID = c(rep("cp", length(plasma_times)), rep("cb", length(brain_times))),
    TIME = c(plasma_times, brain_times),
    E    = c(plasma_mean,  brain_mean),
    V    = c(plasma_sd,    brain_sd)^2
  )
)
```

The same study, written differently: admixr2 normalises it into the same
likelihood blocks, so the fits agree to the last digit.

```{r long-fit, comment = NA}
fit_long <- nlmixr2(pk_cns, admData(c("cp", "cb")), est = "adgh",
                    control = adghControl(studies = list(lit = cns_long)))

c(observations = fit_cns$objective, long = fit_long$objective)
```

Which to use is taste. `observations` keeps each compartment's design visibly
together, the natural way to write a study *by hand*. Long format is the natural
way to write one *from data* — the shape summaries already have in a spreadsheet
or out of a `dplyr` pipeline, and the shape nlmixr2 itself uses.

### Same-subject data: one stacked covariance

Long format earns its keep when plasma and brain were measured in the **same
subjects**. The two curves are then correlated, and that correlation is
information the fit should use — but it lives *between* the compartments, where
per-compartment `V` matrices have nowhere to put it.

Long format does: every observation being a row, the study takes **one**
covariance matrix spanning all of them, aligned with the rows of `data` — exactly
what `cov.wt(dv_mat, method = "ML")$cov` returns from the subject-level matrix
(one row per subject, one column per observation, plasma columns then brain).

```{r long-joint, eval = FALSE}
cns_joint <- list(
  n = 60L, ev = ev,
  data = data.frame(
    DVID = c(rep("cp", length(plasma_times)), rep("cb", length(brain_times))),
    TIME = c(plasma_times, brain_times),
    E    = c(plasma_mean,  brain_mean)
  ),
  V = V_joint   # 11 x 11: plasma block, brain block, and the cross-covariance
)
```

admixr2 then scores all 11 observations with a **single** multivariate normal,
simulating both compartments from shared random effects. A study-level `V` is
what marks the study same-subject; without one each endpoint stays an independent
block. (`observations` can express this too, via a `cross` list of per-pair
blocks — see `?admControl` — but assembling those by hand is the chore long
format removes.)

The choice is a **modelling** decision, not a formatting one: a joint fit with
zero cross-covariances is *not* two independent blocks. The model predicts plasma
and brain to co-vary, sharing `eta.cl` and `eta.v1`, so telling it you observed
no covariance is a real statement about the data and the likelihood will hold you
to it. Joint when the compartments came from the same subjects, independent when
they did not — here, two separate figures.

### Model against data

`plot()` draws observed against predicted for each observed output, one panel
per compartment:

```{r plot-fit, fig.width = 7, fig.height = 3.2}
plot(fit_cns, which = "mean")
```

For a figure of your own, re-solve a residual-free copy of the model over the
estimated `Omega` -- the band is then between-subject variability alone, and a
plain rxode2 model gives `cp` and `cb` directly, where a multi-endpoint fit can
only be re-solved with per-endpoint `DVID`/`CMT` tags:

```{r brain-sim, eval = FALSE}
sim_model <- rxode2::rxode2({
  cl <- exp(tcl + eta.cl); v1 <- exp(tv1 + eta.v1)
  qin <- exp(tqin); qout <- exp(tqout); vb <- 5
  d/dt(central) <- -(cl/v1)*central - (qin/v1)*central + (qout/vb)*brain
  d/dt(brain)   <-  (qin/v1)*central - (qout/vb)*brain
  cp <- central / v1
  cb <- brain / vb
})
rxode2::rxSolve(sim_model, params = fit_cns$theta[c("tcl", "tv1", "tqin", "tqout")],
                omega = fit_cns$omega, nSub = 1000L,
                events = ev |> rxode2::et(seq(0.25, 13, by = 0.25)))
```

Now the payoff, read straight off the estimates:

```{r kp}
theta <- fit_cns$theta
Kp_uu <- exp(theta[["tqin"]]) / exp(theta[["tqout"]])
round(Kp_uu, 2)
```

`Kp,uu` ≈ `r round(Kp_uu, 1)`: at steady state the brain sees about half the
plasma concentration. This is why the brain data was needed at all — with plasma
alone `qin` and `qout` are not separately identifiable, and `Kp,uu` cannot be
estimated.

## Notes

* **Estimators.** `adgh` (used here), `adfo` and `admc` support several observed
  outputs, each with its own analytical / sensitivity gradient. `adirmc` does
  not. See the [estimator comparison](estimator-comparison.html) to choose.
* **Structural vs. observed compartments.** How many compartments the ODEs hold
  is irrelevant; what counts is how many outputs you *observe* and fit — one in
  Step 1, two in Step 2.
* **Hard-coded constants** such as `vb <- 5` keep their value. Not every
  physiological constant has to be estimated.
* **Two ways to write a study.** `observations` and the long-format `data` frame
  are interchangeable: same normalisation, same likelihood, same numbers.
  Independent experiments can also carry a per-endpoint `n` column and a
  per-endpoint `ev`, a list of event tables keyed by endpoint.
* **Same-subject data.** Plasma and brain came from separate figures here, so
  they are independent likelihood blocks. Measured in the *same* subjects, supply
  the cross-covariance instead — one stacked `V` in long format, or a
  per-output-pair `cross` list with `observations`; see `?admControl`.

## See also

- [PD and PK/PD data](pkpd.html) — add a pharmacodynamic endpoint
- [Multiple studies](multiple-studies.html) — meta-analysis across studies
- [Diagnostic plots](diagnostic-plots.html)
