---
title: "Multiple studies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multiple studies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 5
)
source(Filter(file.exists, c("_common.R", "../_common.R", "vignettes/_common.R"))[1])
```

## Why multiple studies?

Passing several studies to `admControl()` fits them simultaneously, minimising
the sum of per-study NLLs under one shared set of population parameters. This is
**meta-analysis**, the core use case: summary statistics from several trials that
may differ in dose, size or schedule, and one population model consistent with
all of them. Each study's `E`, `V` and `n` can come from a digitised figure
(`vignette("aggregate-data")`) or from its own published model
(`vignette("datagen")`).

## Two trials that differ

The point of a meta-analysis is that the studies are *not* interchangeable, so
this one combines a trial we hold individual records for with a trial we have
only summary statistics from --- at a different dose, a sparser schedule and a
different size.

```{r data}
library(admixr2)
library(rxode2)
library(nlmixr2)
library(ggplot2)

# Trial A: 500 subjects at 100 mg, richly sampled. Individual records, reduced
# to the E, V and n a publication would print.
dv_mat  <- admVignetteDvMatrix()      # 500 subjects x 9 times
trial_a <- admVignetteStats(dv_mat)
times_a <- trial_a$times

# Trial B: published as summary statistics only -- 120 subjects at 200 mg, four
# sampling times. datagen() turns its model into the aggregate data it implies,
# which is what makes a published study a direct input; see vignette("datagen").
times_b <- c(0.5, 2, 6, 12)
trial_b <- datagen(
  studies = list(b = list(times = times_b,
                          ev    = rxode2::et(amt = 200),
                          n     = 120L)),
  model   = admVignetteModel,
  control = datagenControl(n_sim = 10000L, seed = 1L)
)$b
```

`admVignetteDvMatrix()` and `admVignetteStats()` come from this vignette's setup
file: the first reshapes `examplomycin` into one row per subject, the second
takes `E`, `V` and `n` off it.

Trial B is simulated here so the vignette has a second study to combine; in
practice its `E`, `V` and `n` would be digitised from the paper
(`vignette("aggregate-data")`) or derived from the model the paper published.

## Comparing the two observed profiles

The two are on different doses, so they should *not* lie on top of each other ---
that separation is the information a joint fit uses.

```{r obs-compare, fig.cap="Observed mean ± 1 SD for each trial on a log time axis. Trial B is at twice the dose and four sampling times."}
df_obs <- rbind(
  data.frame(trial = "Trial A (100 mg, n = 500)", time = times_a,
             mean = trial_a$E,
             lo   = trial_a$E - sqrt(diag(trial_a$V)),
             hi   = trial_a$E + sqrt(diag(trial_a$V))),
  data.frame(trial = "Trial B (200 mg, n = 120)", time = times_b,
             mean = trial_b$E,
             lo   = trial_b$E - sqrt(diag(trial_b$V)),
             hi   = trial_b$E + sqrt(diag(trial_b$V)))
)

ggplot(df_obs, aes(x = time, y = mean, colour = trial, fill = trial)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_x_log10(breaks = sort(unique(c(times_a, times_b)))) +
  scale_colour_manual(values = c("Trial A (100 mg, n = 500)" = "#0072B2",
                                 "Trial B (200 mg, n = 120)" = "#D55E00")) +
  scale_fill_manual(  values = c("Trial A (100 mg, n = 500)" = "#0072B2",
                                 "Trial B (200 mg, n = 120)" = "#D55E00")) +
  labs(title  = "Observed mean ± 1 SD by trial",
       x      = "Time (h, log scale)",
       y      = "Concentration",
       colour = NULL, fill = NULL) +
  theme_bw()
```

## Model definition

The two-compartment model from [Getting started](admixr2.html), supplied by this
vignette's setup file and fitted to both trials at once:

```{r model}
pk_model <- admVignetteModel
```

## Fitting with two studies

Pass both trials as a named list. Each entry carries its own `times`, `ev`, `V`,
`n` and `method`, which is what lets them differ:

```{r fit, comment = NA}
fit_multi <- nlmixr2(
  pk_model, admData(), est = "admc",
  control = admControl(
    studies = list(
      trial_a = list(E = trial_a$E, V = trial_a$V, n = trial_a$n,
                     times = times_a, ev = rxode2::et(amt = 100)),
      trial_b = list(E = trial_b$E, V = trial_b$V, n = trial_b$n,
                     times = times_b, ev = rxode2::et(amt = 200))
    ),
    n_sim     = 5000L,
    cov_n_sim = 10000L,
    maxeval   = 300L,
    seed      = 1L
  )
)

print(fit_multi)
```

Nothing above assumes the two share a dose or a schedule. The objective is the
sum of per-study negative log-likelihoods under one set of population
parameters, and each study is predicted at its own dosing and its own times.

## Per-study diagnostic plots

`plot()` produces separate panels per study, named `mean_<study>` and
`cov_<study>`:

```{r diag, fig.height=6, fig.cap="Mean diagnostics for both trials (one panel per study)."}
plots <- plot(fit_multi, which = "mean")
names(plots)
```

Access individual panels to compare studies side by side:

```{r panels, eval=FALSE}
plots$mean_trial_a
plots$mean_trial_b

# Combine with patchwork if installed
if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork::wrap_plots(plots, ncol = 1)
}
```

## Scaling to a programme

A third and fourth study are more entries in the same list, so a development
programme is written the same way the two above are --- one entry per trial,
each with its own dose, schedule and size.

A study with a diagonal `V` (or a plain vector of variances) gets
`method = "var"`, skipping the O(n_t³) Cholesky solve there is no off-diagonal
structure to justify.

## See also

- [From a published figure to E, V and n](aggregate-data.html) — prepare each study's `E`, `V` and `n`
- [Several observed compartments](multi-compartment.html) — several outputs per study
- [Estimator comparison](estimator-comparison.html)
