---
title: "Several observed compartments (plasma and brain)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Several observed compartments (plasma and brain)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.2, dpi = 120
)
# nlmixr2's fit print method uses cli colour; when a build forces colour on (e.g.
# pkgdown/CI) the raw ANSI escapes leak into the HTML. Force plain-text output
# for this vignette only, saving the previous global state...
.adm_opts    <- options(cli.num_colors = 1, crayon.enabled = FALSE)
.adm_nocolor <- Sys.getenv("NO_COLOR", unset = NA)
Sys.setenv(NO_COLOR = "1")

# ...and restoring it once the document has finished knitting, so the setting
# does not leak into other articles when pkgdown builds them in one process.
knitr::knit_hooks$set(document = function(x) {
  options(.adm_opts)
  if (is.na(.adm_nocolor)) Sys.unsetenv("NO_COLOR") else Sys.setenv(NO_COLOR = .adm_nocolor)
  x
})

# The estimators keep rxode2's progress bar off during fitting (the control's
# nDisplayProgress argument), but a one-off bar still prints while each model is
# *compiled*. Strip those animation frames from the rendered output.
knitr::knit_hooks$set(output = local({
  .default <- knitr::knit_hooks$get("output")
  function(x, options) {
    x <- gsub("[^\n]*\\[[=| ]+\\][^\n]*\n?", "", x, perl = TRUE)
    .default(x, options)
  }
}))
```

## The problem

You are developing a CNS drug and you want one number: **how much of it
reaches the brain?** You don't have patient-level data. What you *do* have is a
published paper with two figures — a **plasma** and a **brain**
concentration–time curve, each drawn as a mean with error bars. You digitise
them into a mean and an SD at every sampling time.

That aggregate summary is exactly what admixr2 is built for. It fits a
population PK model *directly* to means-and-covariances, so a digitised figure
becomes a fittable dataset — no individual records required. This vignette
builds the analysis in two steps:

1. **Plasma only** — the ordinary single-output workflow, to set the scene.
2. **Plasma + brain** — add the brain as a second observed output and read the
   brain-penetration ratio straight off the fit.

```{r libs}
library(admixr2)
library(rxode2)
library(nlmixr2)
library(ggplot2)

ev <- rxode2::et(amt = 100, cmt = "central")   # single 100-unit dose, shared throughout
```

## Step 1 — Plasma only

Start where every PK analysis starts: the plasma curve. In admixr2 a **study**
is the digitised summary bundled with its design — the mean vector `E`, its
variance `V` (here `SD^2`, read as a diagonal covariance, exactly what error
bars give you), the sample size `n`, the sampling `times`, and the dosing
`ev`.

```{r plasma-data}
plasma_times <- c(0.5, 1, 2, 4, 8, 12)
plasma_mean  <- c(8.793, 7.812, 6.850, 5.597, 3.985, 2.910)
plasma_sd    <- c(1.151, 0.911, 0.765, 0.649, 0.770, 0.828)

plasma_study <- list(
  E = plasma_mean, V = plasma_sd^2, n = 60L,
  times = plasma_times, ev = ev
)
```

The model is an ordinary two-compartment PK model with one observed output,
`cp`. Fitting is a single call to `nlmixr2()` with `est = "adgh"`, admixr2's
Gauss–Hermite estimator:

```{r plasma-fit, comment = NA}
pk_plasma <- function() {
  ini({
    tcl <- log(1);  tv1 <- log(10);  tq <- log(3);  tv2 <- log(8)
    prop.cp <- 0.05
    eta.cl ~ 0.09
    eta.v1 ~ 0.04
  })
  model({
    cl <- exp(tcl + eta.cl); v1 <- exp(tv1 + eta.v1)
    q  <- exp(tq);           v2 <- exp(tv2)
    d/dt(central) <- -(cl/v1)*central - (q/v1)*central + (q/v2)*periph
    d/dt(periph)  <-  (q/v1)*central - (q/v2)*periph
    cp <- central / v1
    cp ~ prop(prop.cp)
  })
}

fit_plasma <- nlmixr2(pk_plasma, admData(), est = "adgh",
                      control = adghControl(studies = list(trial = plasma_study)))
fit_plasma
```

This is a perfectly good plasma model — but look at what it *cannot* answer. Its
`periph` compartment is a mathematical distribution store: we never measured it,
and nothing connects it to the brain. To quantify brain exposure we need brain
data **and** a model with a real brain compartment.

## Step 2 — Add the brain

Here are the brain concentrations digitised from the same paper:

```{r brain-data}
brain_times <- c(1, 2, 4, 8, 12)
brain_mean  <- c(3.004, 3.394, 3.018, 2.157, 1.551)
brain_sd    <- c(0.353, 0.349, 0.309, 0.369, 0.405)
```

Now swap the anonymous peripheral compartment for a **mechanistic brain
compartment**. Drug moves plasma → brain with influx clearance `qin` and back
brain → plasma with efflux clearance `qout`. The steady-state brain:plasma
ratio is the quantity we want:

$$K_{p,uu} = \frac{q_{in}}{q_{out}}$$

The model now has **two** observed outputs — plasma `cp` and brain `cb` — so it
carries a residual-error term for each. (`vb`, the brain volume, is a fixed
physiological constant, not an estimated parameter.)

```{r brain-model}
pk_cns <- function() {
  ini({
    tcl  <- log(1);  tv1  <- log(10)
    tqin <- log(3);  tqout <- log(6)
    prop.cp <- 0.05      # plasma residual (proportional)
    add.cb  <- 0.02      # brain residual (additive)
    eta.cl ~ 0.09
    eta.v1 ~ 0.04
  })
  model({
    cl <- exp(tcl + eta.cl); v1 <- exp(tv1 + eta.v1)
    qin <- exp(tqin);        qout <- exp(tqout)
    vb  <- 5
    d/dt(central) <- -(cl/v1)*central - (qin/v1)*central + (qout/vb)*brain
    d/dt(brain)   <-  (qin/v1)*central - (qout/vb)*brain
    cp <- central / v1        # plasma concentration
    cb <- brain   / vb        # brain concentration
    cp ~ prop(prop.cp)
    cb ~ add(add.cb)
  })
}
```

Two observed outputs means the study needs two summaries. Instead of a single
`E`/`V`, give it an **`observations` list** — one named entry per observed
compartment, each pairing a model output with its own `times`, `E` and `V`:

```{r brain-study}
cns_study <- list(
  n = 60L, ev = ev,
  observations = list(
    plasma = list(output = "cp", times = plasma_times, E = plasma_mean, V = plasma_sd^2),
    brain  = list(output = "cb", times = brain_times,  E = brain_mean,  V = brain_sd^2)
  )
)
```

The only other change is telling `admData()` which outputs to expect. Then the
fit call is identical to Step 1:

```{r brain-fit, comment = NA}
fit_cns <- nlmixr2(pk_cns, admData(c("cp", "cb")), est = "adgh",
                   control = adghControl(studies = list(lit = cns_study)))
fit_cns
```

## The same study in long format

If you have used nlmixr2 with [multiple
endpoints](https://nlmixr2.org/articles/multiple-endpoints.html), the
`observations` list above may feel like a detour: nlmixr2 does not group
observations into per-endpoint objects — it stacks them in **one** data frame and
labels each row with the endpoint it belongs to (`DVID` / `CMT`).

admixr2 accepts a study written that way too. Give the study a `data` frame with
one row per observed *endpoint × time* — an endpoint column (`DVID`, `CMT` or
`output`), a time column (`TIME`), the mean (`E`) and its variance (`V`, or an
`SD` column):

```{r long-study}
cns_long <- list(
  n = 60L, ev = ev,
  data = data.frame(
    DVID = c(rep("cp", length(plasma_times)), rep("cb", length(brain_times))),
    TIME = c(plasma_times, brain_times),
    E    = c(plasma_mean,  brain_mean),
    V    = c(plasma_sd,    brain_sd)^2
  )
)
```

This is the same study, written differently — admixr2 normalises it into exactly
the same likelihood blocks, so the fits agree to the last digit:

```{r long-fit, comment = NA}
fit_long <- nlmixr2(pk_cns, admData(c("cp", "cb")), est = "adgh",
                    control = adghControl(studies = list(lit = cns_long)))

c(observations = fit_cns$objective, long = fit_long$objective)
```

Which form to use is a matter of taste. The `observations` list keeps each
compartment's design visibly together and is the more natural way to write a
study *by hand*. The long format is the more natural way to write one *from
data*: it is the shape you already have if your summaries live in a spreadsheet
or come out of a `dplyr` pipeline, and it is the shape nlmixr2 itself uses.

### Same-subject data: one stacked covariance

The long format earns its keep when plasma and brain were measured in the **same
subjects**. Then the two curves are correlated, and that correlation is
information the fit should use — but it lives *between* the compartments, so
there is nowhere to put it in per-compartment `V` matrices.

In long format there is: because every observation is just a row, the study takes
**one** covariance matrix spanning all of them — rows and columns aligned with
the rows of `data`, exactly what `cov.wt(dv_mat, method = "ML")$cov` hands you
when you still have the subject-level matrix (`dv_mat`: one row per subject, one
column per observation, plasma columns then brain columns).

```{r long-joint, eval = FALSE}
cns_joint <- list(
  n = 60L, ev = ev,
  data = data.frame(
    DVID = c(rep("cp", length(plasma_times)), rep("cb", length(brain_times))),
    TIME = c(plasma_times, brain_times),
    E    = c(plasma_mean,  brain_mean)
  ),
  V = V_joint   # 11 x 11: plasma block, brain block, and the cross-covariance
)
```

admixr2 then scores all 11 observations with a **single** multivariate normal,
simulating both compartments from shared random effects. Supplying a study-level
`V` is what marks the study as same-subject; without one, each endpoint stays an
independent likelihood block. (The `observations` form can express this too, via
a `cross` list of per-pair blocks — see `?admControl` — but assembling those by
hand is precisely the chore the long format removes.)

Note that the choice is a **modelling** decision, not a formatting one: a joint
fit with zero cross-covariances is *not* the same as two independent blocks. The
model predicts plasma and brain to co-vary (they share `eta.cl` and `eta.v1`);
telling it you observed no covariance is a real statement about the data, and the
likelihood will hold you to it. Use the joint form when the compartments came
from the same subjects, the independent form when they came from different ones —
here, from two separate figures.

### Model against data

A single fit now describes both compartments. The figure overlays the observed
summaries (points, mean ± SD) with the fitted population prediction — the mean
curve and the ±SD band implied by the estimated between-subject variability
(1000 simulated subjects, residual error excluded). Plasma and brain are
distinguished by colour.

```{r brain-plot, fig.height = 4.6}
# 1. Pull the fitted estimates from the fit (standard nlmixr2 accessors).
theta <- fit_cns$theta   # fixed effects (tcl, tv1, tqin, tqout, residuals)
omega <- fit_cns$omega   # between-subject covariance (rows/cols already named)

# 2. Simulate the fitted population. We solve a plain, residual-free copy of the
#    model so the band shows between-subject variability alone -- and because a
#    multi-endpoint fit can only be re-solved with per-endpoint DVID/CMT tags,
#    whereas this gives cp and cb directly.
sim_model <- rxode2::rxode2({
  cl <- exp(tcl + eta.cl); v1 <- exp(tv1 + eta.v1)
  qin <- exp(tqin); qout <- exp(tqout); vb <- 5
  d/dt(central) <- -(cl/v1)*central - (qin/v1)*central + (qout/vb)*brain
  d/dt(brain)   <-  (qin/v1)*central - (qout/vb)*brain
  cp <- central / v1
  cb <- brain / vb
})
grid <- seq(0.25, 13, by = 0.25)
sim  <- rxode2::rxSolve(sim_model,
  params = theta[c("tcl", "tv1", "tqin", "tqout")],
  omega  = omega, nSub = 1000L,
  events = ev |> rxode2::et(grid), returnType = "data.frame")

# 3. Summarise model and data as mean +/- SD per time, per compartment.
band <- function(value)
  data.frame(time = sort(unique(sim$time)),
             mean = tapply(value, sim$time, mean),
             sd   = tapply(value, sim$time, sd))

pred <- rbind(cbind(compartment = "Plasma", band(sim$cp)),
              cbind(compartment = "Brain",  band(sim$cb)))
obs  <- rbind(
  data.frame(compartment = "Plasma", time = plasma_times, mean = plasma_mean, sd = plasma_sd),
  data.frame(compartment = "Brain",  time = brain_times,  mean = brain_mean,  sd = brain_sd))

# 4. Model (line + band) over data (points + error bars), coloured by compartment.
pal <- c(Plasma = "#0072B2", Brain = "#D55E00")   # Okabe-Ito, colour-blind safe

ggplot() +
  geom_ribbon(data = pred,
              aes(time, ymin = mean - sd, ymax = mean + sd, fill = compartment),
              alpha = 0.15) +
  geom_line(data = pred,
            aes(time, mean, colour = compartment), linewidth = 1) +
  geom_errorbar(data = obs,
                aes(time, ymin = mean - sd, ymax = mean + sd, colour = compartment),
                width = 0.4, linewidth = 0.6, show.legend = FALSE) +
  geom_point(data = obs,
             aes(time, mean, fill = compartment),
             shape = 21, colour = "white", size = 3, stroke = 0.7) +
  scale_colour_manual(values = pal, breaks = c("Plasma", "Brain")) +
  scale_fill_manual(values = pal, guide = "none") +
  scale_x_continuous(breaks = seq(0, 12, 2),
                     expand = expansion(mult = c(0.01, 0.03))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  coord_cartesian(ylim = c(0, NA), clip = "off") +
  guides(colour = guide_legend(override.aes = list(linewidth = 1.4))) +
  labs(x = "Time (h)", y = "Concentration (mg/L)",
       title = "Fitted population prediction vs. observed data",
       subtitle = "Line + band: model (mean ± SD)   ·   Points + bars: data (mean ± SD)",
       colour = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    legend.justification = "left",
    legend.margin = margin(b = 2),
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(colour = "grey40", size = 9, margin = margin(b = 9)),
    axis.title = element_text(colour = "grey25"),
    axis.title.x = element_text(margin = margin(t = 6)),
    axis.title.y = element_text(margin = margin(r = 6)),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(colour = "grey90", linewidth = 0.4),
    plot.margin = margin(12, 16, 10, 12)
  )
```

The prediction tracks both compartments. Now the payoff — read the brain
penetration directly off the estimates:

```{r kp}
Kp_uu <- exp(theta[["tqin"]]) / exp(theta[["tqout"]])
round(Kp_uu, 2)
```

`Kp,uu ≈ 0.5`: at steady state the brain sees about half the plasma
concentration. This is the whole reason the brain data was needed — with plasma
alone, `qin` and `qout` are not separately identifiable and `Kp,uu` cannot be
estimated. The brain measurements resolve it.

### Built-in diagnostics

The overlay above was assembled by hand for a custom figure, but you don't have
to: calling `plot()` on the fit draws observed-vs-predicted panels directly —
one per observed output.

```{r plot-fit, fig.width = 7, fig.height = 3.2}
plot(fit_cns, which = "mean")
```

## Notes

* **Estimators.** `adgh` (used here), `adfo` and `admc` all support several
  observed outputs, each with its own analytical / sensitivity gradient.
  `adirmc` does not — use one of the other three. See the
  [estimator comparison](estimator-comparison.html) vignette to choose.
* **Structural vs. observed compartments.** How many compartments the ODEs
  contain is irrelevant to admixr2; what matters is how many outputs you
  *observe* and fit — one in Step 1, two in Step 2.
* **Hard-coded constants** such as `vb <- 5` keep their value; not every
  physiological constant has to be an estimated parameter.
* **Two ways to write a study.** The `observations` list and the long-format
  `data` frame are interchangeable — same normalisation, same likelihood, same
  numbers. Independent experiments can also carry a per-endpoint `n` column and a
  per-endpoint `ev` (a list of event tables keyed by endpoint).
* **Same-subject data.** Here plasma and brain came from separate figures, so
  they are treated as independent likelihood blocks. If they had instead been
  measured in the *same* subjects, supply the plasma–brain cross-covariance for a
  joint fit — one stacked `V` in long format, or a per-output-pair `cross` list
  with `observations`; see `?admControl`.
```