---
title: "PD and PK/PD data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PD and PK/PD data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.2, dpi = 120
)
source(Filter(file.exists, c("_common.R", "../_common.R", "vignettes/_common.R"))[1])
```

## The problem

A blood-pressure trial reports two doses, 50 mg and 400 mg. You are choosing the
dose for the next study and you want **150 mg**, which nobody measured. This
vignette fits a PK/PD model to the two published arms and predicts the arm that
was never run.

```{r libs}
library(admixr2)
library(rxode2)
library(nlmixr2)
library(ggplot2)
```

## The data

Two arms, 60 subjects each, reporting a plasma concentration–time curve and a
diastolic blood pressure (DBP) curve. Concentrations at seven times, DBP at four:
PD is usually measured more sparsely than PK, and the two need not share a grid.

```{r data}
pk_times <- c(0.5, 1, 2, 4, 8, 12, 24)
pd_times <- c(1, 4, 12, 24)

# 50 mg arm
conc50_mean <- c(0.969, 0.920, 0.831, 0.679, 0.460, 0.317, 0.111)  # mg/L
conc50_sd   <- c(0.207, 0.188, 0.159, 0.125, 0.106, 0.097, 0.062)
dbp50_mean  <- c(91.5, 92.4, 94.1, 95.3)                           # mmHg
dbp50_sd    <- c( 8.9,  8.9,  8.9,  8.9)

# 400 mg arm
conc400_mean <- c(7.709, 7.291, 6.528, 5.256, 3.462, 2.326, 0.780)
conc400_sd   <- c(1.557, 1.440, 1.270, 1.106, 1.027, 0.951, 0.599)
dbp400_mean  <- c(83.1, 84.0, 86.8, 90.7)
dbp400_sd    <- c( 8.5,  8.5,  8.7,  8.8)
```

Both arms are simulated from a known model — CL = 5 L/h, V = 50 L, baseline
DBP = 95 mmHg, Emax = 15 mmHg, EC50 = 2 mg/L — so the 150 mg prediction can be
checked against a truth at the end. Being a sample rather than the population,
the numbers carry the scatter any 60-subject trial would. A digitised figure
gives you exactly these fields.

By 24 h the 50 mg arm is back at baseline with the drug almost gone (0.1 mg/L),
while the 400 mg arm is still about 4 mmHg below it. The DBP standard deviations
barely move across time, the spread being dominated by between-subject
differences in *baseline* blood pressure — the same people at every visit.

These are standard deviations. Published figures often plot a standard *error* or
a model-based least-squares-mean SE, which must be converted first; see
[From a published figure to E, V and n](aggregate-data.html).

## The model

One compartment with a direct-effect `Emax` term that *lowers* DBP. Two observed
outputs make it a multiple-endpoint model, carrying a residual error term for
each, as in [Several observed compartments](multi-compartment.html):

```{r model}
pkpd_model <- function() {
  ini({
    tcl     <- log(4)  ; label("Log clearance (L/h)")
    tv      <- log(40) ; label("Log volume (L)")
    te0     <- log(90) ; label("Log baseline DBP (mmHg)")
    temax   <- log(10) ; label("Log maximum DBP reduction (mmHg)")
    tec50   <- log(1)  ; label("Log EC50 (mg/L)")
    prop.cp <- 0.1     ; label("Proportional residual error, concentration")
    add.dbp <- 3       ; label("Additive residual error, DBP (mmHg)")
    eta.cl ~ 0.09
    eta.v  ~ 0.04
    eta.e0 ~ 0.007
  })
  model({
    cl   <- exp(tcl + eta.cl)
    v    <- exp(tv  + eta.v)
    e0   <- exp(te0 + eta.e0)   # baseline DBP
    emax <- exp(temax)
    ec50 <- exp(tec50)
    d/dt(central) <- -(cl/v) * central
    cp  <- central / v                          # output 1: concentration
    dbp <- e0 - emax * cp / (ec50 + cp)         # output 2: DBP (drug lowers it)
    cp  ~ prop(prop.cp)
    dbp ~ add(add.dbp)
  })
}
```

A few points to note:

* Each output carries its own residual error term (`prop.cp`, `add.dbp`) —
  ordinary nlmixr2 multiple-endpoint syntax.
* `e0` is the baseline. With no pre-dose DBP observation it is identified by
  extrapolating the Emax curve to zero concentration, and the 50 mg arm's 24 h
  point — where the drug contributes about 5% of `emax` — is what keeps that
  extrapolation short. A pre-dose observation or a placebo arm is the robust way
  to pin a baseline; without one, `e0` and `emax` trade off.
* The minus sign is because the drug lowers DBP, so `emax` is the maximum
  *reduction*, in mmHg.
* `emax` and `ec50` carry no `eta`: DBP SDs nearly flat across time and dose say
  little about PD-parameter IIV, and `eta.e0` already reproduces the spread.

## Assembling the study specification

One `observations` entry per output, naming the model variable it corresponds to
and its own `times`, `E` and `V`:

```{r studies}
study50 <- list(
  n = 60L, ev = rxode2::et(amt = 50, cmt = "central"),
  observations = list(
    conc = list(output = "cp",  times = pk_times, E = conc50_mean, V = conc50_sd^2),
    bp   = list(output = "dbp", times = pd_times, E = dbp50_mean,  V = dbp50_sd^2)
  ))

study400 <- list(
  n = 60L, ev = rxode2::et(amt = 400, cmt = "central"),
  observations = list(
    conc = list(output = "cp",  times = pk_times, E = conc400_mean, V = conc400_sd^2),
    bp   = list(output = "dbp", times = pd_times, E = dbp400_mean,  V = dbp400_sd^2)
  ))
```

## Fitting

`admData()` builds the placeholder data frame nlmixr2's interface expects, and
takes the names of the observed outputs — the observations themselves live in the
control, not the data argument. Otherwise an ordinary admixr2 fit:

```{r fit, comment = NA}
fit <- nlmixr2(pkpd_model, admData(c("cp", "dbp")), est = "adgh",
               control = adghControl(studies = list(mg50  = study50,
                                                    mg400 = study400)))
fit
```

`plot()` returns one observed-vs-predicted panel per observed output:

```{r plot, fig.height = 3.2}
plot(fit, which = "mean")
```

## Predicting an unstudied dose

The fit gives both things the question needs: how a dose becomes a concentration
over time (`cl`, `v`), and how concentration becomes an effect (`emax`, `ec50`).
Here both are closed form — a bolus decays exponentially and the effect follows
instantly:

```{r predict}
theta <- fit$theta
cl <- exp(theta[["tcl"]]); v    <- exp(theta[["tv"]])
e0 <- exp(theta[["te0"]]); emax <- exp(theta[["temax"]])
ec50 <- exp(theta[["tec50"]])

conc <- function(dose, t) (dose / v) * exp(-(cl / v) * t)
drop <- function(dose, t) { cc <- conc(dose, t); emax * cc / (ec50 + cc) }

# DBP reduction at 1 h, the first sampled PD time
round(c(mg50 = drop(50, 1), mg150 = drop(150, 1), mg400 = drop(400, 1)), 1)
```

At the first sampled time the model puts 150 mg at
**`r round(drop(150, 1), 1)` mmHg** below baseline, against
`r round(drop(50, 1), 1)` mmHg for 50 mg and `r round(drop(400, 1), 1)` mmHg for
400 mg. The two studied arms are the check: the model's DBP at 1 h is
`r round(e0 - drop(50, 1), 1)` and `r round(e0 - drop(400, 1), 1)` mmHg against
the observed `r dbp50_mean[1]` and `r dbp400_mean[1]`, so the prediction sits on
data rather than beside it.

Naive linear interpolation in dose would put 150 mg at
`r round(drop(50,1) + (150-50)/(400-50) * (drop(400,1) - drop(50,1)), 1)` mmHg.
The curve gives more, the response already flattening: 150 mg buys much of what
400 mg does.

The whole predicted time course follows, laid over the arms that were measured:

```{r dr-plot, fig.height = 4.4}
tt  <- seq(0, 24, length.out = 200)
pred <- rbind(
  data.frame(t = tt, dbp = e0 - drop( 50, tt), arm = "50 mg (studied)"),
  data.frame(t = tt, dbp = e0 - drop(150, tt), arm = "150 mg (predicted)"),
  data.frame(t = tt, dbp = e0 - drop(400, tt), arm = "400 mg (studied)"))
obs <- rbind(
  data.frame(t = pd_times, dbp = dbp50_mean,  sd = dbp50_sd,  arm = "50 mg (studied)"),
  data.frame(t = pd_times, dbp = dbp400_mean, sd = dbp400_sd, arm = "400 mg (studied)"))

pal <- c("50 mg (studied)"    = "#0072B2",
         "150 mg (predicted)" = "#D55E00",
         "400 mg (studied)"   = "#009E73")

ggplot(pred, aes(t, dbp, colour = arm)) +
  geom_hline(yintercept = e0, linetype = "dashed", colour = "grey55") +
  geom_line(aes(linetype = arm), linewidth = 1) +
  geom_point(data = obs, aes(fill = arm), shape = 21, colour = "white",
             size = 2.6, stroke = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = pal) +
  scale_fill_manual(values = pal) +
  scale_linetype_manual(values = c("50 mg (studied)"    = "solid",
                                   "150 mg (predicted)" = "22",
                                   "400 mg (studied)"   = "solid")) +
  scale_x_continuous(breaks = seq(0, 24, 4),
                     expand = expansion(mult = c(0.01, 0.02))) +
  labs(x = "Time (h)", y = "DBP (mmHg)", colour = NULL, linetype = NULL,
       title = "150 mg predicted from the 50 mg and 400 mg arms",
       subtitle = "Points: observed means. Dashed line: baseline.") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", legend.justification = "left",
        plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(colour = "grey40", size = 9,
                                     margin = margin(b = 9)),
        panel.grid.minor = element_blank())
```

Because the data were simulated, the answer is known. The truth was `Emax` = 15
mmHg and `EC50` = 2 mg/L; the fit gives `r round(emax, 1)` and
`r round(ec50, 2)`. At 1 h the true 150 mg drop is
`r round(15 * ((150/50)*exp(-(5/50)*1)) / (2 + (150/50)*exp(-(5/50)*1)), 1)` mmHg
against the predicted `r round(drop(150, 1), 1)` — an overshoot of about
`r round(100 * (drop(150,1) / (15 * ((150/50)*exp(-(5/50)*1)) / (2 + (150/50)*exp(-(5/50)*1))) - 1))`%.
That is what sixty subjects and two dose levels buy: the right shape and a
usable dose, not a precise `Emax`.

A few points to note:

* **This is a typical subject, not a population mean.** `drop()` uses
  `exp(theta)`, i.e. η = 0, whereas the estimator matched `E` to a population
  *mean* — and the mean of a nonlinear function is not that function at the mean.
  The two agree to about 0.01 mmHg here, the median-to-mean shift and the
  curvature nearly cancelling at `eta.v ~ 0.04`. The gap grows with IIV and would
  matter if `emax` or `ec50` carried an `eta`. For a population mean, simulate
  over the estimated `Omega`.
* **Each arm sweeps a range of concentrations** as the drug clears, which is why
  one arm carries more than a static dose–response intuition suggests: 400 mg
  alone spans 0.4–4 × `EC50` and largely identifies `emax` and `ec50`.
* **The 50 mg arm's job is the baseline.** With 400 mg alone the lowest observed
  concentration is still ~29% of `emax`, so the drug-free state is never
  approached and `e0` and `emax` trade off. The 50 mg arm's 24 h point is the only
  near-drug-free observation there is.
* **Check before predicting.** `ec50` comes back at
  `r round(fit$parFixedDf["tec50", "%RSE"], 0)`% RSE — identified but not
  precise, and a dose prediction inherits that. A `temax`/`tec50` correlation
  near ±1 would mean the two are trading off and the plateau is not identified at
  all:

```{r corr}
cv <- fit$cov
if (is.null(cv)) {
  message("Covariance not computed; refit with a larger cov_h_outer.")
} else {
  round(cv["temax", "tec50"] / sqrt(cv["temax", "temax"] * cv["tec50", "tec50"]), 3)
}
```

## Notes

* **Getting E and V from a paper.** Error bars are not always SDs; see
  [From a published figure to E, V and n](aggregate-data.html).
* **Which estimators.** `adgh` (used here), `adfo` and `admc` support several
  observed outputs; `adirmc` errors on multi-output models. See the
  [estimator comparison](estimator-comparison.html).
* **Same-subject PK and PD.** Concentration and DBP are independent likelihood
  blocks here, as they would be from two figures. Measured in the same subjects —
  usually the case in a PK/PD study — they are correlated, and a joint fit with
  zero cross-covariance is not two independent blocks. Supply the
  cross-covariance; see `?admControl` and
  [Several observed compartments](multi-compartment.html).
* **Delayed effects.** `dbp` responds to `cp` instantly, which is what makes the
  prediction a formula. If the effect lags, use an effect compartment or a
  turnover model and simulate instead.
* **Placebo arms.** With only active arms, drug effect and the disease's natural
  time course are confounded. A placebo arm is just another study with a
  zero-amount `ev`.
* **Bounded endpoints.** An additive residual can predict outside a bounded
  score's range. Transform instead — `logitNorm()`, `probitNorm()`, `boxCox()`
  and `yeoJohnson()`, lambda estimated or fixed. See
  [Choosing a residual error model](error-models.html).

## See also

- [Several observed compartments](multi-compartment.html) — multiple outputs and joint fits
- [Multiple studies](multiple-studies.html) — meta-analysis across studies
- [From a published figure to E, V and n](aggregate-data.html)
