---
title: "Valid inference for aggregate data modelling"
subtitle: "Standard errors, likelihood-ratio tests and model selection in admixr2, for digitised summaries, published models, and both together"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, comment = "#>")
```

# The short version

The aggregate objective is a **quasi-likelihood**. Point estimates are
consistent; standard errors, the likelihood-ratio reference distribution and the
AIC penalty are all wrong, and all three are wrong for the same reason. One
matrix product, $H^{-1}J$, repairs all three.

In practice that means five things.

| | |
|---|---|
| **1** | Fit with `covMethod = "r,s"`. Always. |
| **2** | Compare models with `anova(full, reduced)`, or on **TIC**. |
| **3** | Never use **AIC**, and never compare raw $\Delta$OFV. |
| **4** | Nest by *fixing* a coefficient, not by deleting the term. |
| **5** | `model_cov` must span every parameter the source estimated. |

Everything below is why.

# What the objective is

A study reports a mean vector and a covariance matrix over $m$ times from $n$
subjects. Those are the sufficient statistics, and under independent
multivariate-normal subjects Cochran's theorem splits them into two independent
pieces,
$$
\bar y \sim N\!\left(\mu,\ \tilde V / n\right),
\qquad
n V_{\mathrm{obs}} \sim \mathrm{Wishart}(n-1,\ \tilde V),
$$
whose combined $-2\log L$ is what admixr2 minimises:
$$
-2\log L \;=\; n\Bigl(\log|\tilde V| \;+\; \operatorname{tr}(\tilde V^{-1} V_{\mathrm{obs}}) \;+\; r'\tilde V^{-1} r\Bigr),
\qquad r = E_{\mathrm{obs}} - \mu_{\mathrm{pred}} .
$$

The model supplies $\mu$ and $\tilde V$ through the **law of total variance** —
not $V_{\text{struct}} + \Sigma(\mu)$, which is the NONMEM "no $\eta$–$\epsilon$
interaction" convention and understates the diagonal:
$$
\mu_{\mathrm{pred}} = m_s \circ E_\eta[f],
\qquad
\tilde V = (m_s \otimes m_s)\circ \operatorname{Cov}_\eta(f) \;+\; \operatorname{diag}\bigl(E_\eta[\operatorname{Var}(y\mid\eta)]\bigr).
$$

Covariates enter by integrating the prediction over the population the study
enrolled, declared as `cov_dist`. Where the covariates reach the model through
fewer scalars than there are covariates, that integral is reduced to its true
dimension by a change of variables rather than approximated.

**The same formula is used for every study, whatever its provenance.** This is
the single most important structural fact in this note, and section 6 returns to
it.

# Why it is a quasi-likelihood, and what that costs

Equation (1) assumes the individuals behind a summary are multivariate normal.
They are not. The marginal over the random effects is a mixture, normal only
where $f$ is linear in $\eta$ — so marginalising over the random effect alone
already breaks it, before any covariate is involved.

For a study generated from a **published model** there is a second and much
larger reason: its $(E, V)$ is a *deterministic function of somebody else's
published estimate* $\hat\theta_{\mathrm{src}}$. It is not a sample of $n$
patients in any sense.

The consequence is precise, and narrower than it first appears.

> The score has zero expectation at $\Psi$ under **any** weight
> (Gouriéroux–Monfort–Trognon). **Estimates remain consistent.** What fails is
> the information equality $H = J$ — and therefore every quantity derived from
> it: standard errors, the LRT reference distribution, and AIC's penalty.

Write $F = -2\log L$, $H$ for its expected Hessian and $J$ for the variance of
its score. Under correct specification $J = 2H$. Nothing in an aggregate fit
guarantees that, and with model sources it is not remotely true.

# Standard errors

`covMethod = "r,s"` returns the sandwich
$$
\operatorname{Avar}(\hat\Psi) \;=\; H^{-1} J H^{-1},
\qquad
J \;=\; \sum_s G_s\, \Omega_s\, G_s' .
$$

The meat is assembled **per study, according to what that study actually is**:

| study kind | contribution to $J$ |
|---|---|
| digitised summary | $G_s \Omega_s G_s'$ with $\Omega_s = \operatorname{Cov}(t_s)$ computed **from the model** — conditional third and fourth moments, not normal theory |
| published model | its data block is *removed*, and replaced by $G_s\, C_{\mathrm{src}}\, G_s'$ |

Two properties follow, and both matter:

* Under correct specification the sandwich reduces **exactly** to $2H^{-1}$,
  which is what `covMethod = "r"` returns. It is never wrong to ask for it.
* Point estimates are **bit-identical** — only the reported uncertainty changes.

## What it fixes, measured

Digitised data, 800 replicates, $\omega = 0.5$:

| quantity | naive `"r"` | sandwich `"r,s"` |
|---|---|---|
| SE/SD, covariate carried by $V$ | 0.529 | 1.059 |
| SE/SD, $\omega$ | 0.745 | 0.974 |
| SE/SD, clearance | 0.676 | 0.983 |
| coverage | 0.825 | 0.946 |

A published model source, lone, reproducible:

| declared $C_{\mathrm{src}}$ | `"r,s"`, $n = 100$ | `"r,s"`, $n = 6400$ | `"r"`, $n=100$ | `"r"`, $n = 6400$ |
|---|---|---|---|---|
| 0.08000 | 0.08000 | 0.08000 | 0.01793 | 0.00224 |
| 0.06000 | 0.06000 | 0.06000 | 0.00160 | 0.00020 |
| 0.00400 | 0.00400 | 0.00400 | 0.00244 | 0.00031 |

The sandwich returns the source's own uncertainty exactly, and **independently
of $n$**. The naive route falls as $1/\sqrt n$ — a precision the analyst chooses
by typing a number. That is the single clearest argument for `"r,s"`.

# Model comparison

## Likelihood-ratio test

Under a quasi-likelihood $\Delta$OFV does not converge to $\chi^2_q$. It
converges to a weighted sum,
$$
\Delta\mathrm{OFV} \;\longrightarrow\; \sum_{i=1}^{q} \lambda_i\, \chi^2_1,
\qquad
\lambda_i = \operatorname{eig}\bigl(A^{-1}B\bigr),
$$
with $A = (2H^{-1})_{[\gamma,\gamma]}$ and $B = (H^{-1}JH^{-1})_{[\gamma,\gamma]}$.
Both blocks are taken **after inverting**, so each is a Schur complement and the
nuisance parameters are profiled out rather than conditioned on. (Doing it the
other way round inflated $\sum\lambda_i$ twelve-fold in testing, in the
conservative direction, so no covariate would ever have been selected.)

At $q = 1$ — covariate selection, the main use case — there is exactly one
weight, so $\Delta\mathrm{OFV}/\lambda \sim \chi^2_1$ and the corrected test is
asymptotically **exact**, not an approximation. `anova()` handles $q>1$ with a
Ruben series, and degrades to a trace approximation when the tested block is
ill-conditioned.

It **refuses** rather than approximating in three cases: a boundary restriction
(fixing a variance to zero puts the null on the edge of the parameter space,
where the limit is a $\bar\chi^2$ mixture), a non-nested pair (that needs Vuong),
and two fits built at different stratum resolutions.

## AIC, TIC

AIC's $2p$ is **derived, not conventional**. The optimism of a maximised
objective — how much better it fits its own data than it would fit new data — is
$\operatorname{tr}(H^{-1}J)/2$. The information equality makes that equal $p$,
and that is the *only* reason AIC has the form it does. Where $H \neq J$ the
penalty is simply the wrong size.

Takeuchi's criterion replaces it with the honest quantity:
$$
\mathrm{TIC} = F(\hat\Psi) + \operatorname{tr}(H^{-1}J),
\qquad
p_{\mathrm{eff}} = \tfrac12\operatorname{tr}(H^{-1}J).
$$
Under correct specification $J = 2H$, so $p_{\mathrm{eff}} = p$ and
$\mathrm{TIC} = \mathrm{AIC}$ exactly. Any implementation failing that check is
wrong.

## BIC

Not corrected for misspecification, and not tested in this setting. admixr2
keeps nlmixr2's `nobs` convention so BIC stays comparable with other packages,
and reports `BIC_h` beside it as the principled alternative. **Do not select on
BIC for these fits.**

# Reading `Weight` and `p_eff`

These two columns are diagnostics, not bookkeeping.

$\lambda$ (the `Weight` column) is **exactly linear in the declared source
uncertainty** and proportional to the weight the source carries:
$$
\lambda \;\propto\; n \times C_{\mathrm{src}} .
$$
Measured by scaling every `model_cov` over a 64-fold range on one fitted design:

| $C_{\mathrm{src}}$ scale | $\lambda$ | $\lambda$/scale | estimate | SE |
|---|---|---|---|---|
| 0.25 | 0.850 | 3.3986 | 0.553290 | 0.0216 |
| 1 | 3.399 | 3.3986 | 0.553290 | 0.0432 |
| 4 | 13.594 | 3.3986 | 0.553290 | 0.0864 |
| 16 | 54.378 | 3.3986 | 0.553290 | 0.1727 |

The ratio is constant to four decimals and the estimate is bit-identical
throughout — because the objective never reads `model_cov`; only the sandwich
does. So $\lambda$ has a plain reading: **the share of the fit's apparent
information that is really the source analysts' uncertainty rather than
evidence.** Near 1, the sources are precise enough to behave like data. Below 1
— which does happen — the uncorrected test was *conservative*, not liberal.

$p_{\mathrm{eff}}$ scales exactly with $n$ and is **not** an effective parameter
count. Only its *difference* between nested models is meaningful, and that
difference is all TIC uses. `anova()` prints a note when
$p_{\mathrm{eff}}/N_{\mathrm{par}}$ is large enough that AIC should not be read.

# Why this is valid for each kind of input

## Digitised aggregate data

The objective is the exact $-2\log L$ of the sufficient statistics under
iid-normal subjects. The departure from normality is real but is repaired
entirely in $J$: estimates are untouched, and SE, LRT and TIC are each corrected
by the same matrix.

## Published models supplied directly

`datagen()` maps a published model plus the population it was fitted in onto the
$(E, V, n)$ a digitised figure would have given, so it enters the *same*
objective. The only random object in the chain is $\hat\theta_{\mathrm{src}}$,
and it is carried into $J$ as $G C_{\mathrm{src}} G'$. This is why the reported
interval reproduces the source's own and does not shrink with $n$.

`n` still matters, but only as **relative weight** between sources. Set it to the
sample size each source was developed on.

## Both together

The objective sums study terms regardless of provenance. $J$ sums per-study
meats, each built for its own provenance. There is no third mechanism, and no
interaction between the two — which is precisely why they combine.

# How to run one

```{r recipe}
fit <- nlmixr2(meta_model, admData(), est = "adgh",
               control = adghControl(studies = studies,
                                     covMethod = "r,s"))   # <- always

# nest by FIXING the coefficient: `cov_dist` declares CRCL, and admixr2
# refuses a distribution naming a covariate the model never reads
null_model <- function() {
  ini({
    tcl   <- log(4)
    bcrcl <- fix(0)          # the restriction under test
    ...
  })
  model({ ... })
}
fit_null <- nlmixr2(null_model, admData(), est = "adgh",
                    control = adghControl(studies = studies,
                                          covMethod = "r,s"))

anova(fit, fit_null)         # read `p`, never `p.naive`, never AIC
```

Two rules that are easy to get wrong and quiet when you do:

* **`model_cov` must cover every parameter the source estimated.** A focei
  `$cov` carries the structural thetas only; the residual SD and IIV rows have to
  be added from the paper's %RSE. A missing row reads as a parameter the source
  knew *exactly*, and admixr2 refuses the sandwich — which means
  `covMethod = "r,s"` silently falls back to `"r"` and you get an interval that
  is too narrow. It warns, but a chunk with `warning = FALSE` will eat it.
* **Stratify the covariates each source's own model estimated**, and marginalise
  the rest. This drives the point estimate, not just the interval: splitting a
  source on a covariate its model never fitted manufactures a contrast, and
  marginalising one it *did* fit throws the contrast away.

# Evidence

Two independent campaigns. The first established the machinery for digitised
data (probe suite, ~40 scripts); the second checked it where the studies are
published models, which is the case in which $J$ is a completely different
object.

## Digitised data

| | uncorrected | corrected |
|---|---|---|
| LRT size, nominal 0.05 | 0.152 – 0.179 | **0.0540 [0.0484, 0.0600]** (6000 reps) |
| AIC keeps a spurious covariate | 0.292 | 0.147 (nominal 0.157) |

Robust across $\omega = 0.2/0.5/0.8$ — where the uncorrected size swings
0.075–0.330 — proportional residual error, two random effects, and triple $N$.

## Published model sources

400 replicates, three sources, null covariate:

| | corrected | naive |
|---|---|---|
| size at $\alpha = 0.05$ | **0.0600 [0.0367, 0.0833]** | 0.7950 [0.7554, 0.8346] |
| size at $\alpha = 0.01$ | **0.0200 [0.0063, 0.0337]** | 0.7100 [0.6655, 0.7545] |
| KS vs Uniform(0,1) | $D=0.031$, $p = 0.851$ | $D = 0.745$, $p < 10^{-15}$ |

The corrected $p$-values are **uniform under the null**, which is a stronger
statement than correct size at one $\alpha$.

Selection, 200 replicates, same design:

| criterion | keeps the null covariate | should be |
|---|---|---|
| AIC | 0.8400 [0.789, 0.891] | ~0.157 |
| raw $\Delta$OFV | 0.7550 [0.695, 0.815] | 0.05 |
| **TIC** | **0.1450 [0.096, 0.194]** | ~0.157 |
| **corrected $p$** | **0.0750 [0.039, 0.112]** | 0.05 |

with $\Delta p_{\mathrm{eff}}/\lambda = 1.019$ — TIC's per-parameter charge and
the LRT's rescaling are the same quantity, as the theory requires. AIC charges
2 where the honest charge was 92.

Standard errors on the same replicates:

| parameter | SE/SD | coverage |
|---|---|---|
| null covariate | 0.969 | 0.910 [0.882, 0.938] |
| sex effect | 0.974 | 0.948 [0.926, 0.969] |
| clearance | 0.975 | 0.908 [0.879, 0.936] |

# Limitations

Stated plainly, because a methods note that only lists successes is not useful.

* **A mixed fit** — digitised studies and model sources together — is
  implemented and correct by construction, but has **not** been validated by
  simulation. This is the first gap I would close.
* **Coverage is 0.91** against nominal 0.95 for two of three parameters, while
  SE/SD is 0.97. The scale is right, so this reads as mild tail behaviour rather
  than a mis-sized interval. Unexplained.
* **BIC** is uncorrected and untested here.
* **`adfo` understates** a model source's standard error: it scores FO-linearised
  moments, so its map from the source's parameters is not the identity. Prefer
  `adgh` or `admc`, which reproduce the source.
* **Structural misfit** — a source whose model is simpler than the one being
  fitted — was deliberately excluded from the validation design. Estimates
  behaved well in a worked example, but LRT and TIC were not checked there.
* **Non-nested comparison** is not provided; that needs a Vuong test.
* **Boundary restrictions** are refused rather than approximated. To compare a
  model with and without a random effect, use TIC or simulate under the null.

# Provenance

Derivations and raw results: `DERIVATION-DOFV.md`, `DERIVATION-SANDWICH.md`,
`HANDOFF-INFERENCE.md`, and `FINDING-mbma-inference.md`. Every figure in this
note is a measured value from those campaigns; none is asymptotic theory quoted
without a check.
